---
title: "DEgenes_heatmap"
author: "Shuo Shan"
date: "01/05/2023"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Load packages  
```{r package, warning=FALSE, include=FALSE}
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)

Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
setwd(Dir)
#save.image(file=paste0(Dir,"/myEnvironment_version2.RData"))
load(paste0(Dir,"/myEnvironment_version2.RData"))

Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_FC5_avgCPM10.RData"))
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_FC10_avgCPM10.RData"))
```
# temp, delete after 09/04/2024
```{r temp-plot}
make_paired_plot_temp <- function(this.title,this.df) {
  df = data.frame(CPM=as.numeric(this.df[1,]), sample=colnames(this.df)) %>%
    separate( . , col=sample, into=c("donor","treatment","celltype")) 
  df$treatment <- gsub("IFN","IFNG",df$treatment)
  df$treatment <- factor(df$treatment, levels = c("PBS","IFNG"))
  df$celltype <- df$celltype %>%
    gsub("FRB","Fibroblast", . ) %>%
    gsub("MEL","Melanocyte", . ) %>%
    gsub("KRT","Keratinocyte", . )
  
  position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
  line.color = "grey"; linetype = "solid"; palette="bright"
  
  p <- ggplot(df, aes(x = treatment, y = CPM, color = treatment)) +
    geom_jitter(width=0.1, alpha=0.8) +
    scale_color_manual(values=c("#999999", "#E69F00")) +
        ggtitle(this.title) +
    theme(legend.position="none") +
    theme(plot.title = element_text(size = 10, face = "bold")) +
    facet_wrap(~ celltype)
  p
  return(p)
}

g="IFNGR1"
pdf(paste0("~/Desktop/temp_plot_",g,".pdf"),width=5,height=5)
make_paired_plot_temp(g,CPM_rna[g,])
dev.off()




make_paired_plot_no3donorcolor <- function(this.title,this.df) {
  df = data.frame(CPM=as.numeric(this.df[1,]), sample=colnames(this.df)) %>%
    separate( . , col=sample, into=c("donor","treatment","celltype")) 
  df$treatment <- gsub("IFN","IFNG",df$treatment)
  df$treatment <- factor(df$treatment, levels = c("PBS","IFNG"))
  df$celltype <- df$celltype %>%
    gsub("FRB","Fibroblast", . ) %>%
    gsub("MEL","Melanocyte", . ) %>%
    gsub("KRT","Keratinocyte", . )
  
  position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
  line.color = "grey"; linetype = "solid"; palette="bright"

  
  p <- ggplot(df, create_aes(list(x = "treatment", y = "CPM"))) +
    geom_exec(geom_line, data = df, group = "donor",
              color = line.color, size = line.size, linetype = linetype,
              position = position) +
    geom_exec(geom_point, data = df, color = "treatment", size = point.size,
              position = position) +
    scale_color_manual(values=c("#999999", "#E69F00")) +
    ggtitle(this.title) +
    theme(legend.position="none") +
    theme(plot.title = element_text(size = 10, face = "bold")) +
    facet_wrap(~ celltype)
  p
  return(p)
}

for (i in 1:14) {
  this_gene_list <- data.table::fread(paste0(Dir,"/heatmap_RNA_round2/reordered_inducedGenes_log2FC1.5_cluster",i,".txt"),header=F)$V1
  
  this_plot_list <- list()
  for (j in 1:length(this_gene_list)) {
    g=this_gene_list[j]
    this_plot_list[[j]] <- make_paired_plot_no3donorcolor(g,CPM_rna[g,])
  }
  
  pdf(paste0("~/Desktop/cluster",i,".pdf"), width=5, height=5)
  for (k in 1:length(this_plot_list)) {
    print(this_plot_list[[k]])
    }
  dev.off()
  i
}
```
# Load file
## loading ATACseq data
```{r cov-atac, eval=FALSE, echo=FALSE, include=FALSE}
atac_coverage_dir="~/Downloads/nl/human/skin/eQTLs/chromatin/RegulatoryElements/method3"
# load raw coverage for enhancer and promoters
count_cRE <- data.table::fread(paste0(atac_coverage_dir,"/merged_coverage_ATACseq_allcts_regulatory_region.bed")) %>%
  dplyr::select(-c(chr,start,end)) %>% column_to_rownames("name")
# CPM
col.sum=apply(count_cRE,2,sum); 
CPM_atac=sweep(count_cRE, 2, col.sum, `/`)*1000000
CPM_atac <- CPM_atac[,c("F25_KRT_PBS","F49_KRT_PBS","F55_KRT_PBS",
                        "F25_MEL_PBS","F49_MEL_PBS","F55_MEL_PBS",
                        "F25_FRB_PBS","F49_FRB_PBS","F55_FRB_PBS",
                        "F25_KRT_IFN","F49_KRT_IFN","F55_KRT_IFN",
                        "F25_MEL_IFN","F49_MEL_IFN","F55_MEL_IFN",
                        "F25_FRB_IFN","F49_FRB_IFN","F55_FRB_IFN")]
rm(col.sum,count_cRE,atac_coverage_dir)
```
## loading H3K27ac ChIPseq data
```{r cov-chip, eval=FALSE, echo=FALSE, include=FALSE}
chip_coverage_dir="~/Downloads/nl/human/skin/eQTLs/chromatin/RegulatoryElements/method3"
# load raw coverage for enhancer and promoters
count_cRE <- data.table::fread(paste0(chip_coverage_dir,"/merged_coverage_ChIPseq_allcts_regulatory_region.bed")) %>%
  dplyr::select(-c(chr,start,end)) %>% column_to_rownames("name")
# CPM
col.sum=apply(count_cRE,2,sum); 
CPM_chip=sweep(count_cRE, 2, col.sum, `/`)*1000000
CPM_chip <- CPM_chip[,c("F25_KRT_PBS","F49_KRT_PBS","F55_KRT_PBS",
                        "F25_MEL_PBS","F49_MEL_PBS","F55_MEL_PBS",
                        "F25_FRB_PBS","F49_FRB_PBS","F55_FRB_PBS",
                        "F25_KRT_IFN","F49_KRT_IFN","F55_KRT_IFN",
                        "F25_MEL_IFN","F49_MEL_IFN","F55_MEL_IFN",
                        "F25_FRB_IFN","F49_FRB_IFN","F55_FRB_IFN")]
rm(col.sum,count_cRE,chip_coverage_dir)
```
## loading RNAseq data and DESeq2 results
```{r load-rna, eval=FALSE, echo=FALSE, include=FALSE}
rna_dir="~/Downloads/nl/human/skin/eQTLs/RNA-Seq"
CPM_rna=data.table::fread(paste0(rna_dir,"/CPM_3cts.txt"))
CPM_rna = CPM_rna %>% column_to_rownames(. , var="V1")
new_colnames=c(colnames(CPM_rna)[grepl("PBS_KRT",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("PBS_MEL",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("PBS_FRB",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("IFN_KRT",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("IFN_MEL",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("IFN_FRB",colnames(CPM_rna))])
CPM_rna <- CPM_rna[,new_colnames]

dir.krt <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/keratinocytes/pipeline_11192022/analysis"
dir.mel <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/melanocytes/pipeline_12022022/analysis"
dir.frb <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis"
rna.res.krt <- read.table(paste0(dir.krt,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()
rna.res.mel <- read.table(paste0(dir.mel,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()
rna.res.frb <- read.table(paste0(dir.frb,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()
```
## Link of gene-promoter-enhancer
```{r link, eval=FALSE, echo=FALSE, include=FALSE}
# link between gene-promoter and gene-enhancer
cRE_dict_dir="~/Downloads/nl/human/skin/eQTLs/chromatin/RegulatoryElements/method3"
link_gene_promoter=data.table::fread(paste0(cRE_dict_dir,"/dictionary_gene_promoter_links_allcts.txt")) %>%
  set_colnames(c("gene","promoter","distance","celltype"))
link_gene_enhancer=data.table::fread(paste0(cRE_dict_dir,"/dictionary_closest_gene_enhancer_links_allcts.txt")) %>% 
  set_colnames(c("gene","enhancer","distance","celltype")) 
```

# plot heatmaps Round One
## load rnaseq CPM for all DE genes, and corresponding ATACseq and ChIPseq CPM for their promoter and enhancers
```{r DE_CPM, eval=FALSE, echo=FALSE, include=FALSE}
promoter_df <- link_gene_promoter[which(gene %in% DEG),]
enhancer_df <- link_gene_enhancer[which(gene %in% DEG),]


CPM_atac_DEG_promoter <- CPM_atac[unique(promoter_df$promoter),]
CPM_atac_DEG_enhancer <- CPM_atac[unique(enhancer_df$enhancer),]
CPM_chip_DEG_promoter <- CPM_chip[unique(promoter_df$promoter),]
CPM_chip_DEG_enhancer <- CPM_chip[unique(enhancer_df$enhancer),]
```

## find the matching enhancers for each gene and plot
```{r plots, eval=FALSE, echo=FALSE, include=FALSE}
plot_list = list()
for (g in DEG.shared$gene) {
  p <- create_compiled_heatmap(g)
  plot_list[[i]] = p
}
pdf(paste0(outDir,"/compiled_heatmaps_gene_CRE_shared_DEG.pdf"),width=15,height=8)
for (i in 1:length(plot_list)) {
  print(plot_list[[i]])
}
dev.off()


```
## useful plotting functions
```{r nicekmeans, echo=FALSE, include=FALSE}
# useful function
niceKmeans <- function (df, k, kmeans.algorithm = "Lloyd", iter.max = 1000, nstart=100) {
  
  result = data.frame();
  source = t(scale(t(df)));
  kmeans = kmeans(source, centers = k, iter.max = iter.max, algorithm=kmeans.algorithm, nstart=nstart);
  clustered = data.frame();
  breaks = c();
  for (i in 1:k) {
    print(sprintf("Doing cluster %i",i))
    cluster = source[kmeans$cluster==i,];
    rows = row.names(cluster);
    clust = heatmap.2(as.matrix(cluster), distfun=function(x) as.dist((1-cor(t(x)))/2))
    cluster.ordered = cluster[clust$rowInd,];
    cluster.ordered.genes = rows[clust$rowInd];
    row.names(cluster.ordered) = cluster.ordered.genes;
    class = data.frame(row.names=cluster.ordered.genes);
    class[,"class"] = i 
    cluster.ordered = cbind(cluster.ordered, class);
    clustered = rbind(clustered, cluster.ordered);
    if(i > 1 & i < k) {
      breaks[i] = as.numeric(breaks[i-1]) + length(rows);
    } else if(i==1) {
      breaks[i] = length(rows);
    }
  }
  
  result <- list();
  result$clustered <- clustered;
  result$breaks <- breaks;
  return(result);
}

```
```{r gene-enhancer-heatmap, eval=FALSE, echo=FALSE, include=FALSE}
create_compiled_heatmap <- function(g) {
#g = "IRF7"
this.dict.p <- link_gene_promoter[which(link_gene_promoter$gene == g),c("promoter","distance")] %>% distinct() %>% arrange(gene,distance)
this.dict.e <- link_gene_enhancer[which(link_gene_enhancer$gene == g),c("enhancer","distance")] %>% distinct() %>% arrange(gene,distance)

header <- c("F25_PBS_KRT","F49_PBS_KRT","F55_PBS_KRT",
            "F25_PBS_MEL","F49_PBS_MEL","F55_PBS_MEL",
            "F25_PBS_FRB","F49_PBS_FRB","F55_PBS_FRB",
            "F25_IFN_KRT","F49_IFN_KRT","F55_IFN_KRT",
            "F25_IFN_MEL","F49_IFN_MEL","F55_IFN_MEL",
            "F25_IFN_FRB","F49_IFN_FRB","F55_IFN_FRB")
this.CPM.gene <- log2(CPM_rna[g,header] + 1)
this.CPM.atac.p <- log2(CPM_atac_DEG_promoter[unique(this.dict.p$promoter),] + 1)
this.CPM.atac.e <- log2(CPM_atac_DEG_enhancer[unique(this.dict.e$enhancer),] + 1)
this.CPM.chip.p <- log2(CPM_chip_DEG_promoter[unique(this.dict.p$promoter),] + 1)
this.CPM.chip.e <- log2(CPM_chip_DEG_enhancer[unique(this.dict.e$enhancer),] + 1)

CPM.gene.scaled <- t(scale(t(as.matrix(this.CPM.gene)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.atac.p.scaled <- t(scale(t(as.matrix(this.CPM.atac.p)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.atac.e.scaled <- t(scale(t(as.matrix(this.CPM.atac.e)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.chip.p.scaled <- t(scale(t(as.matrix(this.CPM.chip.p)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.chip.e.scaled <- t(scale(t(as.matrix(this.CPM.chip.e)), center = TRUE, scale = TRUE)) %>% as.data.frame()
#CPM.atac.p.scaled <- CPM.p.scaled %>% rownames_to_column("promoter") %>% left_join( . , unique(this.dict.p[,c("name","dist")]), by="name") %>% mutate(tag=paste0(name,"_",dist)) %>% column_to_rownames("tag") %>% dplyr::select(-c("name", "dist"))
#CPM.atac.e.scaled <- CPM.e.scaled %>% rownames_to_column("name") %>% left_join( . , unique(this.dict.e[,c("name","dist")]), by="name") %>% mutate(tag=paste0(name,"_",dist)) %>% column_to_rownames("tag") %>% dplyr::select(-c("name", "dist"))

# heatmaps for gene expression, promoter and enhancer ATACseq signal.
column_ha = HeatmapAnnotation(
  condition = colnames(CPM.gene.scaled) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  celltype = colnames(CPM.gene.scaled) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  col = list(celltype = c("FRB" = "#97de8c", "MEL" = "#80391e", "KRT" = "#1a8bb5"),
             condition = c("PBS" = "grey", "IFN" = "#Ec5003")),
  show_legend = c(TRUE,TRUE))


heatmap.gene <- Heatmap(CPM.gene.scaled, 
        name = "gene z-log2CPM",
        top_annotation = column_ha,
        column_title = g,
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmap.atac.promoter <- Heatmap(CPM.atac.p.scaled, 
        name = "z-log2CPM",
        column_title = "ATACseq signal promoter ",
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmap.atac.enhancer <- Heatmap(CPM.atac.e.scaled, 
        name = "enhancer z-log2CPM",
        column_title = "ATACseq signal enhancer",
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmap.chip.promoter <- Heatmap(CPM.chip.p.scaled, 
        name = "z-log2CPM",
        column_title = "H3K27ac ChIPseq signal promoter ",
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmap.chip.enhancer <- Heatmap(CPM.chip.e.scaled, 
        name = "enhancer z-log2CPM",
        column_title = "H3K27ac ChIPseq signal enhancer",
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmaps <- heatmap.gene %v% heatmap.atac.promoter %v% heatmap.chip.promoter %v% heatmap.atac.enhancer %v% heatmap.chip.enhancer # vertically concatenating heatmaps
return(heatmaps)
}

```
```{r paired_plot_function, eval=FALSE, echo=FALSE, include=FALSE}
make_paired_plot <- function(this.title,this.df) {
  df = data.frame(CPM=as.numeric(this.df[1,]), sample=colnames(this.df)) %>%
    separate( . , col=sample, into=c("donor","treatment","celltype")) 
  df$treatment <- factor(df$treatment, levels = c("PBS","IFN"))
  
  position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
  line.color = "grey"; linetype = "solid"; palette="bright"
  
  highlightdata1 <- df %>% filter(donor=="F25")
  highlightdata2 <- df %>% filter(donor=="F49")
  highlightdata3 <- df %>% filter(donor=="F55")
  
  p <- ggplot(df, create_aes(list(x = "treatment", y = "CPM"))) +
    geom_exec(geom_line, data = df, group = "donor",
              color = line.color, size = line.size, linetype = linetype,
              position = position) +
    geom_exec(geom_point, data = df, color = "treatment", size = point.size,
              position = position) +
    scale_color_manual(values=c("#999999", "#E69F00")) +
    geom_exec(geom_line, data = highlightdata1, group = "donor",
              color = "blue", size = 0.7, linetype = linetype,
              position = position) +
    geom_exec(geom_line, data = highlightdata2, group = "donor",
              color = "brown", size = 0.7, linetype = linetype,
              position = position) +
    geom_exec(geom_line, data = highlightdata3, group = "donor",
              color = "black", size = 0.7, linetype = linetype,
              position = position) +
    ggtitle(this.title) +
    theme(legend.position="none") +
    theme(plot.title = element_text(size = 10, face = "bold")) +
    facet_wrap(~ celltype)
  p
  return(p)
}
```
```{r gene-enhancer-paired-plot, eval=FALSE, echo=FALSE, include=FALSE}
create_compiled_paired_plots <- function(g) {
this.dict.p <- link_gene_promoter[which(link_gene_promoter$gene == g),] %>% arrange(gene,dist)
this.dict.e <- link_gene_enhancer[which(link_gene_enhancer$gene == g),] %>% arrange(gene,dist)

header <- c("F25_FRB_PBS","F25_FRB_IFN","F49_FRB_PBS","F49_FRB_IFN","F55_FRB_PBS","F55_FRB_IFN",
            "F25_MEL_PBS","F25_MEL_IFN","F49_MEL_PBS","F49_MEL_IFN","F55_MEL_PBS","F55_MEL_IFN",
            "F25_KRT_PBS","F25_KRT_IFN","F49_KRT_PBS","F49_KRT_IFN","F55_KRT_PBS","F55_KRT_IFN")

this.CPM.gene <- CPM_rna[g,header] 
this.CPM.p <- CPM_promoter[unique(this.dict.p$name),header] %>% rownames_to_column("name") %>% left_join( . , unique(this.dict.p[,c("name","dist")]), by="name") %>% mutate(tag=paste0(name,"_",dist)) %>% column_to_rownames("tag") %>% dplyr::select(-c("name", "dist"))
this.CPM.e <- CPM_enhancer[unique(this.dict.e$name),header] %>% rownames_to_column("name") %>% left_join( . , unique(this.dict.e[,c("name","dist")]), by="name") %>% mutate(tag=paste0(name,"_",dist)) %>% column_to_rownames("tag") %>% dplyr::select(-c("name", "dist"))

this.logCPM.gene <- log2(CPM_rna[g,header] + 1)
this.logCPM.p <- log2(CPM_promoter[unique(this.dict.p$name),header] + 1)
this.logCPM.e <- log2(CPM_enhancer[unique(this.dict.e$name),header] + 1)

CPM.gene.scaled <- t(scale(t(as.matrix(this.logCPM.gene)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.p.scaled <- t(scale(t(as.matrix(this.logCPM.p)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.e.scaled <- t(scale(t(as.matrix(this.logCPM.e)), center = TRUE, scale = TRUE)) %>% as.data.frame()

# make paired plots
this.CPM.all <- rbind(this.CPM.gene, this.CPM.p, this.CPM.e)

paired_plot_gene <- make_paired_plot(g,this.CPM.gene)

plot_list_promoter = list()
for (i in 1:nrow(this.CPM.p)) {
  p <- make_paired_plot(rownames(this.CPM.p)[i],this.CPM.p[i,])
  plot_list_promoter[[i]] = p
}
paired_plot_promoter <- do.call("grid.arrange", c(plot_list_promoter, ncol=1, nrow=nrow(this.CPM.p)))

plot_list_enhancer = list()
for (i in 1:nrow(this.CPM.p)) {
  p <- make_paired_plot(rownames(this.CPM.e)[i],this.CPM.e[i,])
  plot_list_enhancer[[i]] = p
}
paired_plot_enhancer <- do.call("grid.arrange", c(plot_list_enhancer, ncol=1, nrow=nrow(this.CPM.p)))

p <- plot_grid(paired_plot_gene, paired_plot_promoter, paired_plot_enhancer, ncol=3, nrow=1)
return(p)
}
```
## Heatmap of RNAseq
```{r RNA-heatmap-DEgenes-CPM-rowClustered-kmm-hclust-log2FC1.5-heatmap, eval=FALSE, include=FALSE}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)


pbs_cols <- colnames(CPM_rna)[which(grepl("PBS",colnames(CPM_rna)))]
ifn_cols <- colnames(CPM_rna)[which(grepl("IFN",colnames(CPM_rna)))]
rows_above_CPM_cutoff_PBS <- rownames(CPM_rna[rowMeans(CPM_rna[,pbs_cols]) >= 10,]) %>% unlist()
rows_above_CPM_cutoff_IFN <- rownames(CPM_rna[rowMeans(CPM_rna[,ifn_cols]) >= 10,]) %>% unlist()
selected_rows <- unique(c(rows_above_CPM_cutoff_PBS,rows_above_CPM_cutoff_IFN))
CPM_rna_new <- CPM_rna %>% dplyr::filter(rownames(.) %in% selected_rows)

DEG <- unique(c(rna.res.krt[which(rna.res.krt$padj<0.05 & (rna.res.krt$log2FoldChange)>1.5),"gene"],
                rna.res.mel[which(rna.res.mel$padj<0.05 & (rna.res.mel$log2FoldChange)>1.5),"gene"],
                rna.res.frb[which(rna.res.frb$padj<0.05 & (rna.res.frb$log2FoldChange)>1.5),"gene"]))
CPM_rna_DEG <- CPM_rna_new %>% dplyr::filter(rownames(.) %in% DEG)

# 1. center and scale data
logCPM <- log2(CPM_rna_DEG + 1)

# 2. hierarchically cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 7)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,km_res$breaks[5]-km_res$breaks[4]),
                        rep(6,km_res$breaks[6]-km_res$breaks[5]),
                        rep(7,nrow(km_res$clustered)-km_res$breaks[6]))

km_res.rna <- km_res
split_instructions.rna <- split_instructions

# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1

zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_rna_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  celltype = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))
pdf(paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, control vs IGNg stimulated, known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(CPM_rna_DEG)[grepl("PBS_KRT",colnames(CPM_rna_DEG))])), 
                                             rep(2,length(colnames(CPM_rna_DEG)[grepl("PBS_MEL",colnames(CPM_rna_DEG))])), 
                                             rep(3,length(colnames(CPM_rna_DEG)[grepl("PBS_FRB",colnames(CPM_rna_DEG))])), 
                                             rep(4,length(colnames(CPM_rna_DEG)[grepl("IFN_KRT",colnames(CPM_rna_DEG))])), 
                                             rep(5,length(colnames(CPM_rna_DEG)[grepl("IFN_MEL",colnames(CPM_rna_DEG))])),
                                             rep(6,length(colnames(CPM_rna_DEG)[grepl("IFN_FRB",colnames(CPM_rna_DEG))]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

cluster1_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==1)]
cluster2_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==2)]
cluster3_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==3)]
cluster4_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==4)]
cluster5_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==5)]
cluster6_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==6)]
cluster7_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==7)]


write.table(cluster1_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10_cluster1_genes_cts_induced_genes_KRT.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster2_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10_cluster2_genes_paired_induced_genes_KRT_MEL.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster3_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10_cluster3_genes_cts_induced_genes_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster4_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10_cluster4_commonly_induced_genes.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster5_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10_cluster5_genes_paired_induced_genes_MEL_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster6_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10_cluster6_genes_paired_induced_genes_KRT_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster7_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10_cluster7_genes_cts_induced_genes_MEL.txt"),quote=F,row.names = F, col.names=F)

save.image(file=paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
```
```{r RNA-heatmap-DEgenes-CPM-rowClustered-kmm-hclust-log2FC1.5-heatmap-FC5cutoff, eval=FALSE, include=FALSE}
# first load log2FC1.5 heatmap .rdata
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)

# filter for genes with fold change > 5 AND average CPM >= 10.
cutoff=2.321928
DEG_FC5 = unique(c(rna.res.krt[which(rna.res.krt$padj<0.05 & (rna.res.krt$log2FoldChange) > cutoff),"gene"],
                rna.res.mel[which(rna.res.mel$padj<0.05 & (rna.res.mel$log2FoldChange) > cutoff),"gene"],
                rna.res.frb[which(rna.res.frb$padj<0.05 & (rna.res.frb$log2FoldChange) > cutoff),"gene"]))
DEG_FC5_CPM10 = intersect(DEG_FC5, selected_rows)

# pick from clustered matrix only genes that satisfy the cutoffs
km_res_FC5_CPM10 = data.frame()
split_instructions_FC5_CPM10 = c()

for (i in 1:7) {
  if (i == 1) {
    this.cluster <- km_res.rna$clustered[1 : km_res.rna$breaks[1],]
  } 
  else if (i == 7){
    this.cluster <- km_res.rna$clustered[km_res.rna$breaks[6] : nrow(km_res.rna$clustered),]
  }
  else{
    this.cluster <- km_res.rna$clustered[km_res.rna$breaks[i] : km_res.rna$breaks[i-1],]
  }
  
  this.cluster.filtered <- this.cluster %>% dplyr::filter(rownames(.) %in% DEG_FC5_CPM10) %>% dplyr::select(-class)
  km_res_FC5_CPM10 <- rbind(km_res_FC5_CPM10, this.cluster.filtered)
  split_instructions_FC5_CPM10 = c(split_instructions_FC5_CPM10, rep( i , nrow(this.cluster.filtered)))
}

# pick row indices for known IFNg responding genes
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res_FC5_CPM10
zlogCPM.heatmap_with_gene <- km_res_FC5_CPM10 %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  celltype = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_03072023_padj0.05_FC5_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, control vs IGNg stimulated, known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM.heatmap)[grepl("PBS_KRT",colnames(zlogCPM.heatmap))])), 
                                             rep(2,length(colnames(zlogCPM.heatmap)[grepl("PBS_MEL",colnames(zlogCPM.heatmap))])), 
                                             rep(3,length(colnames(zlogCPM.heatmap)[grepl("PBS_FRB",colnames(zlogCPM.heatmap))])), 
                                             rep(4,length(colnames(zlogCPM.heatmap)[grepl("IFN_KRT",colnames(zlogCPM.heatmap))])), 
                                             rep(5,length(colnames(zlogCPM.heatmap)[grepl("IFN_MEL",colnames(zlogCPM.heatmap))])),
                                             rep(6,length(colnames(zlogCPM.heatmap)[grepl("IFN_FRB",colnames(zlogCPM.heatmap))]))),
                            row_split = split_instructions_FC5_CPM10, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

cluster1_genes = rownames(km_res_FC5_CPM10)[which(split_instructions_FC5_CPM10==1)]
cluster2_genes = rownames(km_res_FC5_CPM10)[which(split_instructions_FC5_CPM10==2)]
cluster3_genes = rownames(km_res_FC5_CPM10)[which(split_instructions_FC5_CPM10==3)]
cluster4_genes = rownames(km_res_FC5_CPM10)[which(split_instructions_FC5_CPM10==4)]
cluster5_genes = rownames(km_res_FC5_CPM10)[which(split_instructions_FC5_CPM10==5)]
cluster6_genes = rownames(km_res_FC5_CPM10)[which(split_instructions_FC5_CPM10==6)]
cluster7_genes = rownames(km_res_FC5_CPM10)[which(split_instructions_FC5_CPM10==7)]

write.table(cluster1_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC5_avgCPM10_cluster1_genes_cts_induced_genes_KRT.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster2_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC5_avgCPM10_cluster2_genes_paired_induced_genes_KRT_MEL.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster3_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC5_avgCPM10_cluster3_genes_cts_induced_genes_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster4_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC5_avgCPM10_cluster4_commonly_induced_genes.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster5_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC5_avgCPM10_cluster5_genes_paired_induced_genes_MEL_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster6_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC5_avgCPM10_cluster6_genes_paired_induced_genes_KRT_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster7_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC5_avgCPM10_cluster7_genes_cts_induced_genes_MEL.txt"),quote=F,row.names = F, col.names=F)

save.image(file=paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_FC5_avgCPM10.RData"))

```
```{r RNA-heatmap-DEgenes-CPM-rowClustered-kmm-hclust-log2FC1.5-heatmap-FC10cutoff, eval=FALSE, include=FALSE}
# first load log2FC1.5 heatmap .rdata
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_FC10_avgCPM10.RData"))
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)
# filter for genes with fold change > 10 AND average CPM >= 10.
cutoff=3.321928
DEG_FC10 = unique(c(rna.res.krt[which(rna.res.krt$padj<0.05 & (rna.res.krt$log2FoldChange) > cutoff),"gene"],
                rna.res.mel[which(rna.res.mel$padj<0.05 & (rna.res.mel$log2FoldChange) > cutoff),"gene"],
                rna.res.frb[which(rna.res.frb$padj<0.05 & (rna.res.frb$log2FoldChange) > cutoff),"gene"]))
DEG_FC10_CPM10 = intersect(DEG_FC10, selected_rows)

# pick from clustered matrix only genes that satisfy the cutoffs
km_res_FC10_CPM10 = data.frame()
split_instructions_FC10_CPM10 = c()

for (i in 1:7) {
  if (i == 1) {
    this.cluster <- km_res.rna$clustered[1 : km_res.rna$breaks[1],]
  } 
  else if (i == 7){
    this.cluster <- km_res.rna$clustered[km_res.rna$breaks[6] : nrow(km_res.rna$clustered),]
  }
  else{
    this.cluster <- km_res.rna$clustered[km_res.rna$breaks[i] : km_res.rna$breaks[i-1],]
  }
  
  this.cluster.filtered <- this.cluster %>% dplyr::filter(rownames(.) %in% DEG_FC10_CPM10) %>% dplyr::select(-class)
  km_res_FC10_CPM10 <- rbind(km_res_FC10_CPM10, this.cluster.filtered)
  split_instructions_FC10_CPM10 = c(split_instructions_FC10_CPM10, rep( i , nrow(this.cluster.filtered)))
}

# pick row indices for known IFNg responding genes
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res_FC10_CPM10
zlogCPM.heatmap_with_gene <- km_res_FC10_CPM10 %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  celltype = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_03072023_padj0.05_FC10_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, control vs IGNg stimulated, known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM.heatmap)[grepl("PBS_KRT",colnames(zlogCPM.heatmap))])), 
                                             rep(2,length(colnames(zlogCPM.heatmap)[grepl("PBS_MEL",colnames(zlogCPM.heatmap))])), 
                                             rep(3,length(colnames(zlogCPM.heatmap)[grepl("PBS_FRB",colnames(zlogCPM.heatmap))])), 
                                             rep(4,length(colnames(zlogCPM.heatmap)[grepl("IFN_KRT",colnames(zlogCPM.heatmap))])), 
                                             rep(5,length(colnames(zlogCPM.heatmap)[grepl("IFN_MEL",colnames(zlogCPM.heatmap))])),
                                             rep(6,length(colnames(zlogCPM.heatmap)[grepl("IFN_FRB",colnames(zlogCPM.heatmap))]))),
                            row_split = split_instructions_FC10_CPM10, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

cluster1_genes = rownames(km_res_FC10_CPM10)[which(split_instructions_FC10_CPM10==1)]
cluster2_genes = rownames(km_res_FC10_CPM10)[which(split_instructions_FC10_CPM10==2)]
cluster3_genes = rownames(km_res_FC10_CPM10)[which(split_instructions_FC10_CPM10==3)]
cluster4_genes = rownames(km_res_FC10_CPM10)[which(split_instructions_FC10_CPM10==4)]
cluster5_genes = rownames(km_res_FC10_CPM10)[which(split_instructions_FC10_CPM10==5)]
cluster6_genes = rownames(km_res_FC10_CPM10)[which(split_instructions_FC10_CPM10==6)]
cluster7_genes = rownames(km_res_FC10_CPM10)[which(split_instructions_FC10_CPM10==7)]


write.table(cluster1_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC10_avgCPM10_cluster1_genes_cts_induced_genes_KRT.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster2_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC10_avgCPM10_cluster2_genes_paired_induced_genes_KRT_MEL.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster3_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC10_avgCPM10_cluster3_genes_cts_induced_genes_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster4_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC10_avgCPM10_cluster4_commonly_induced_genes.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster5_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC10_avgCPM10_cluster5_genes_paired_induced_genes_MEL_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster6_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC10_avgCPM10_cluster6_genes_paired_induced_genes_KRT_FRB.txt"),quote=F,row.names = F, col.names=F)
write.table(cluster7_genes,paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_FC10_avgCPM10_cluster7_genes_cts_induced_genes_MEL.txt"),quote=F,row.names = F, col.names=F)

save.image(file=paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_FC10_avgCPM10.RData"))

```
## Correlation heatmap of RNAseq: 1) DEG and TFs; 2) DEG and DEG
```{r correlation-heatmap-log2FC1.5DEG-TF}
# first load log2FC1.5 heatmap .rdata
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)

# pick DE genes and TFs
TF_list=c("STAT1","STAT2","STAT3","STAT4","STAT5A","STAT5B","STAT6","NFKB1","NFKB2","RELA","RELB","REL","RUNX1","RUNX2","GATA2","GATA3","ATF1","ATF2","ATF3","ATF4","ATF5","ATF6","ATF6B","ATF7","BATF","BATF2","BATF3","IRF1","IRF2","IRF3","IRF4","IRF5","IRF6","IRF7","IRF8","IRF9","JUN","JUNB","JUND","FOS","FOSB","FOSL1","FOSL2","ETS1","ETS2","ERG","ETV3","ETV1","ETV4","KLF2","KLF3","KLF4","KLF5","KLF6","KLF8","KLF9","KLF10","KLF11","KLF12","KLF13","KLF15","KLF16","MAF","MAFA","MAFB","MAFF","MAFK","TEAD1","TEAD2","TEAD3","TEAD4","PRDM1","SOX1","SOX4","PRDM2","ELF1","ELF2","CBFB","FOXA1","AR")
gene_list=c(cluster1_genes, cluster2_genes, cluster3_genes, 
            cluster4_genes, cluster5_genes, cluster6_genes, cluster7_genes)
CPM_picked_genes <- CPM_rna[gene_list,]
CPM_TF <- CPM_rna[TF_list,]

# generate correlation matrix
mat.correlation <- matrix(nrow=length(gene_list), ncol=length(TF_list))
for (i in 1:length(gene_list)) {
  for (j in 1:length(TF_list)) {
    mat.correlation[i,j] <- cor(as.numeric(CPM_picked_genes[i,]), as.numeric(CPM_TF[j,]) , method="pearson")
  }
}
rownames(mat.correlation) <- gene_list
colnames(mat.correlation) <- TF_list

# plot heatmap of correlation matrix
pdf(paste0(Dir,"/heatmap_RNA/correlation_heatmap_RNAseq_TF_vs_DEgenes_7kmm-hc_03212023_padj0.05_log2FC1.5_avgCPM10_pearsonColumnClustered.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(mat.correlation, 
                            name = "PCC",
                            column_title = "PCC correlation between DEG and TFs",
                            cluster_columns = TRUE,
                            clustering_distance_columns = "pearson",
                            cluster_rows = FALSE,
                            row_split = split_instructions.rna, 
                            show_row_names = TRUE,
                            show_column_names = TRUE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            row_names_gp = gpar(fontsize=3),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r correlation-heatmap-FC5DEG-TF}
# first load log2FC1.5 heatmap .rdata
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_FC5_avgCPM10.RData"))
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)

# pick DE genes and TFs
TF_list=c("STAT1","STAT2","STAT3","STAT4","STAT5A","STAT5B","STAT6","NFKB1","NFKB2","RELA","RELB","REL","RUNX1","RUNX2","GATA2","GATA3","ATF1","ATF2","ATF3","ATF4","ATF5","ATF6","ATF6B","ATF7","BATF","BATF2","BATF3","IRF1","IRF2","IRF3","IRF4","IRF5","IRF6","IRF7","IRF8","IRF9","JUN","JUNB","JUND","FOS","FOSB","FOSL1","FOSL2","ETS1","ETS2","ERG","ETV3","ETV1","ETV4","KLF2","KLF3","KLF4","KLF5","KLF6","KLF8","KLF9","KLF10","KLF11","KLF12","KLF13","KLF15","KLF16","MAF","MAFA","MAFB","MAFF","MAFK","TEAD1","TEAD2","TEAD3","TEAD4","PRDM1","SOX1","SOX4","PRDM2","ELF1","ELF2","CBFB","FOXA1","AR")
gene_list=c(cluster1_genes, cluster2_genes, cluster3_genes, 
            cluster4_genes, cluster5_genes, cluster6_genes, cluster7_genes)
CPM_picked_genes <- CPM_rna[gene_list,]
CPM_TF <- CPM_rna[TF_list,]

# generate correlation matrix
mat.correlation <- matrix(nrow=length(gene_list), ncol=length(TF_list))
for (i in 1:length(gene_list)) {
  for (j in 1:length(TF_list)) {
    mat.correlation[i,j] <- cor(as.numeric(CPM_picked_genes[i,]), as.numeric(CPM_TF[j,]) , method="pearson")
  }
}
rownames(mat.correlation) <- gene_list
colnames(mat.correlation) <- TF_list

# plot heatmap of correlation matrix
pdf(paste0(Dir,"/heatmap_RNA/correlation_heatmap_RNAseq_TF_vs_DEgenes_7kmm-hc_03212023_padj0.05_FC5_avgCPM10_pearsonClustered.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(mat.correlation, 
                            name = "PCC",
                            column_title = "PCC correlation between DEG and TFs",
                            cluster_columns = TRUE,
                            clustering_distance_columns = "pearson",
                            cluster_rows = FALSE,
                            row_split = split_instructions_FC5_CPM10, 
                            show_row_names = TRUE,
                            show_column_names = TRUE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            row_names_gp = gpar(fontsize=3),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r correlation-heatmap-FC10DEG-TF}
# first load log2FC1.5 heatmap .rdata
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_FC10_avgCPM10.RData"))
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)

# pick DE genes and TFs
TF_list=c("STAT1","STAT2","STAT3","STAT4","STAT5A","STAT5B","STAT6","NFKB1","NFKB2","RELA","RELB","REL","RUNX1","RUNX2","GATA2","GATA3","ATF1","ATF2","ATF3","ATF4","ATF5","ATF6","ATF6B","ATF7","BATF","BATF2","BATF3","IRF1","IRF2","IRF3","IRF4","IRF5","IRF6","IRF7","IRF8","IRF9","JUN","JUNB","JUND","FOS","FOSB","FOSL1","FOSL2","ETS1","ETS2","ERG","ETV3","ETV1","ETV4","KLF2","KLF3","KLF4","KLF5","KLF6","KLF8","KLF9","KLF10","KLF11","KLF12","KLF13","KLF15","KLF16","MAF","MAFA","MAFB","MAFF","MAFK","TEAD1","TEAD2","TEAD3","TEAD4","PRDM1","SOX1","SOX4","PRDM2","ELF1","ELF2","CBFB","FOXA1","AR")
gene_list=c(cluster1_genes, cluster2_genes, cluster3_genes, 
            cluster4_genes, cluster5_genes, cluster6_genes, cluster7_genes)
CPM_picked_genes <- CPM_rna[gene_list,]
CPM_TF <- CPM_rna[TF_list,]

# generate correlation matrix
mat.correlation <- matrix(nrow=length(gene_list), ncol=length(TF_list))
for (i in 1:length(gene_list)) {
  for (j in 1:length(TF_list)) {
    mat.correlation[i,j] <- cor(as.numeric(CPM_picked_genes[i,]), as.numeric(CPM_TF[j,]) , method="pearson")
  }
}
rownames(mat.correlation) <- gene_list
colnames(mat.correlation) <- TF_list

# plot heatmap of correlation matrix
pdf(paste0(Dir,"/heatmap_RNA/correlation_heatmap_RNAseq_TF_vs_DEgenes_7kmm-hc_03212023_padj0.05_FC10_avgCPM10_pearsonClustered.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(mat.correlation, 
                            name = "PCC",
                            column_title = "PCC correlation between DEG and TFs",
                            cluster_columns = TRUE,
                            clustering_distance_columns = "pearson",
                            cluster_rows = FALSE,
                            row_split = split_instructions_FC10_CPM10, 
                            show_row_names = TRUE,
                            show_column_names = TRUE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            row_names_gp = gpar(fontsize=3),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r correlation-of-PCC-acrosslog2FCinducedgenes-heatmap-TF-TF}
# first load log2FC1.5 heatmap .rdata
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
library(ComplexHeatmap)
library(InteractiveComplexHeatmap)

# pick DE genes and TFs
TF_list=c("STAT1","STAT2","STAT3","STAT4","STAT5A","STAT5B","STAT6","NFKB1","NFKB2","RELA","RELB","REL","RUNX1","RUNX2","GATA2","GATA3","ATF1","ATF2","ATF3","ATF4","ATF5","ATF6","ATF6B","ATF7","BATF","BATF2","BATF3","IRF1","IRF2","IRF3","IRF4","IRF5","IRF6","IRF7","IRF8","IRF9","JUN","JUNB","JUND","FOS","FOSB","FOSL1","FOSL2","ETS1","ETS2","ERG","ETV3","ETV1","ETV4","KLF2","KLF3","KLF4","KLF5","KLF6","KLF8","KLF9","KLF10","KLF11","KLF12","KLF13","KLF15","KLF16","MAF","MAFA","MAFB","MAFF","MAFK","TEAD1","TEAD2","TEAD3","TEAD4","PRDM1","SOX1","SOX4","PRDM2","ELF1","ELF2","CBFB","FOXA1","AR")
gene_list=c(cluster1_genes, cluster2_genes, cluster3_genes, 
            cluster4_genes, cluster5_genes, cluster6_genes, cluster7_genes)
CPM_picked_genes <- CPM_rna[gene_list,]
CPM_TF <- CPM_rna[TF_list,]

# generate correlation matrix
mat.correlation <- matrix(nrow=length(gene_list), ncol=length(TF_list))
for (i in 1:length(gene_list)) {
  for (j in 1:length(TF_list)) {
    mat.correlation[i,j] <- cor(as.numeric(CPM_picked_genes[i,]), as.numeric(CPM_TF[j,]) , method="pearson")
  }
}
rownames(mat.correlation) <- gene_list
colnames(mat.correlation) <- TF_list

# generate correlation of correlation matrix
mat.correlation.of.pcc <- matrix(nrow=length(TF_list), ncol=length(TF_list))
for (i in 1:length(TF_list)) {
  for (j in 1:length(TF_list)) {
    mat.correlation.of.pcc[i,j] <- cor(as.numeric(mat.correlation[,i]), as.numeric(mat.correlation[,j]))
  }
}
rownames(mat.correlation.of.pcc) <- TF_list
colnames(mat.correlation.of.pcc) <- TF_list

# plot heatmap of correlation matrix
pdf(paste0(Dir,"/heatmap_RNA/correlation_heatmap_RNAseq_TF_vs_TF_across_pcc_of_DEgenes_7kmm-hc_03212023_padj0.05_log2FC1.5_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(mat.correlation.of.pcc, 
                            name = "PCC",
                            column_title = "PCC correlation between TFs across the PCC correlation with induced genes",
                            cluster_columns = TRUE,
                            cluster_rows = TRUE,
                            show_row_names = TRUE,
                            clustering_distance_rows = 'pearson',
                            clustering_distance_columns = 'pearson',
                            show_column_names = TRUE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            row_names_gp = gpar(fontsize=10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
#htShiny(heatmap.rna)
dev.off()

save.image(file=paste0(Dir,"/heatmap_RNA/correlation_heatmap_RNAseq_TF_vs_DEgenes_7kmm-hc_03132023_padj0.05_log2FC1.5_avgCPM10.RData"))
```
```{r TF-paired-plots}
plot_list <- list()
for (i in 1:length(TF_list)) {
  g=TF_list[i]
  plot_list[[i]] <- make_paired_plot(g,CPM_rna[g,])
}
pdf(paste0(Dir,"/heatmap_RNA/","paired_plot_TF",".pdf"),width=15,height=8)
for (i in 1:length(TF_list)) {
  print(plot_list[[i]])
}
dev.off()
```
## Keeping tally of human TFs
```{r TF-in-clusters}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)
tf <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/Lambert_2018_human_TFs.txt",header=F) %>%
  dplyr::filter(V1 %in% rownames(CPM_rna)) %>%
  pull(V1)

zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_rna_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% tf)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  celltype = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))
pdf(paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_7kmm-hc_03222023_padj0.05_log2FC1.5_avgCPM10_TF-labeled.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, control vs IGNg stimulated, TF labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(CPM_rna_DEG)[grepl("PBS_KRT",colnames(CPM_rna_DEG))])), 
                                             rep(2,length(colnames(CPM_rna_DEG)[grepl("PBS_MEL",colnames(CPM_rna_DEG))])), 
                                             rep(3,length(colnames(CPM_rna_DEG)[grepl("PBS_FRB",colnames(CPM_rna_DEG))])), 
                                             rep(4,length(colnames(CPM_rna_DEG)[grepl("IFN_KRT",colnames(CPM_rna_DEG))])), 
                                             rep(5,length(colnames(CPM_rna_DEG)[grepl("IFN_MEL",colnames(CPM_rna_DEG))])),
                                             rep(6,length(colnames(CPM_rna_DEG)[grepl("IFN_FRB",colnames(CPM_rna_DEG))]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r TF-tally}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)
tf <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/Lambert_2018_human_TFs.txt",header=F) %>%
  dplyr::filter(V1 %in% rownames(CPM_rna)) %>%
  pull(V1) %>%
  unique()

CPM_tf <- CPM_rna[tf,]

# how many of them have expression in only one cell type
tf_means <- c()
for (i in 1:nrow(CPM_tf)) {
  mean.pbs.mel <- CPM_tf[i,grepl("PBS_MEL",colnames(CPM_tf))] %>% as.numeric() %>% mean()
  mean.pbs.frb <- CPM_tf[i,grepl("PBS_FRB",colnames(CPM_tf))] %>% as.numeric() %>% mean()
  mean.pbs.krt <- CPM_tf[i,grepl("PBS_KRT",colnames(CPM_tf))] %>% as.numeric() %>% mean()
  
  this.mean <- c(mean.pbs.frb, mean.pbs.mel, mean.pbs.krt)
  tf_means <- rbind(tf_means, this.mean)
}
colnames(tf_means) <- c("PBS_FRB", "PBS_MEL", "PBS_KRT")
rownames(tf_means) <- rownames(CPM_tf)
tf_means <- as.data.frame(tf_means)
tf_means$count_cpm10 <- unlist(rowSums(tf_means > 10))
tf_means$celltypes <- apply(tf_means[,1:3],1,function(x) c("FRB","MEL","KRT")[which(x > 10)]) %>% paste( . )

table(tf_means$count_cpm10)
table(tf_means$celltypes)
library("ggalluvial")
df.alluv <- data.frame(all=rep("human_TF",nrow(CPM_tf)),
                       expressed=c(rep("not expressed",762),
                                   rep("1 celltype",179),
                                   rep("2 celltypes",136),
                                   rep("3 celltypes",553)),
                       celltype=c(rep("NA",762),
                                  rep("1.MEL",67),
                                  rep("1.FRB",72),
                                  rep("1.KRT",40),
                                  rep("2.MEL_FRB",70),
                                  rep("2.MEL_KRT",25),
                                  rep("2.FRB_KRT",41),
                                  rep("3.all celltypes",553)))
ggplot(data = df.alluv,
       aes(axis1 = all, axis2 = expressed, axis3=celltype)) +
  scale_x_discrete(limits = c("all", "expressed","celltype"), expand = c(.2, .05)) +
  xlab("TF") +
  geom_alluvium(aes(fill = celltype)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal() +
  ggtitle("human TF expression in skin cells")

# among the expressed factors, how many of them are also induced?
dir.krt <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/keratinocytes/pipeline_11192022/analysis"
dir.mel <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/melanocytes/pipeline_12022022/analysis"
dir.frb <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis"
rna.res.krt <- read.table(paste0(dir.krt,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()
rna.res.mel <- read.table(paste0(dir.mel,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()
rna.res.frb <- read.table(paste0(dir.frb,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()

induced.krt <- rna.res.krt %>% dplyr::filter(gene %in% tf) %>% dplyr::filter(pvalue<0.05) %>% dplyr::filter(log2FoldChange > 1.5) %>% pull(gene)
induced.frb <- rna.res.frb %>% dplyr::filter(gene %in% tf) %>% dplyr::filter(pvalue<0.05) %>% dplyr::filter(log2FoldChange > 1.5) %>% pull(gene)
induced.mel <- rna.res.mel %>% dplyr::filter(gene %in% tf) %>% dplyr::filter(pvalue<0.05) %>% dplyr::filter(log2FoldChange > 1.5) %>% pull(gene)

df.krt <- data.frame(gene=induced.krt,celltype=rep("KRT",length(induced.krt)))
df.mel <- data.frame(gene=induced.mel,celltype=rep("MEL",length(induced.mel)))
df.frb <- data.frame(gene=induced.frb,celltype=rep("FRB",length(induced.frb)))

temp <- rbind(df.krt, df.mel, df.frb) %>%
  group_by(gene) %>% 
  summarize(celltype=paste(celltype,collapse=','))

df.expressed <- tf_means %>% rownames_to_column("gene") %>% dplyr::select(c(gene,count_cpm10,celltypes)) %>%
  set_colnames(c("gene","expressed_celltypes_count","expressed_celltypes"))

df.induced <- rbind(df.krt, df.mel, df.frb) %>%
  group_by(gene) %>% 
  mutate(count=n()) %>%
  distinct_at( . , "gene", .keep_all=TRUE) %>%
  dplyr::select(-celltype) %>%
  left_join( . , temp, by="gene") 

df.expressed_induced <- left_join(df.expressed, df.induced, by="gene")

```
## Heatmaps of ATACseq-promoter
```{r ATAC-promoter-all-genes-general-dynamics}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch promoters of all genes and scale
this.cpm <- CPM_atac[which(rownames(CPM_atac) %in% link_gene_promoter$promoter),]
this.logCPM <- log2(this.cpm + 1)

# pick 5 clusters
df <- this.logCPM
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
rm(df,wss)

# niceKmeans
km_res_atac_promoter <- niceKmeans(this.logCPM, 5)
split_instructions_atac_promoter <- c(rep(1,km_res_atac_promoter$breaks[1]), 
                        rep(2,km_res_atac_promoter$breaks[2]-km_res_atac_promoter$breaks[1]),
                        rep(3,km_res_atac_promoter$breaks[3]-km_res_atac_promoter$breaks[2]),
                        rep(4,km_res_atac_promoter$breaks[4]-km_res_atac_promoter$breaks[3]),
                        rep(5,nrow(km_res_atac_promoter$clustered)-km_res_atac_promoter$breaks[4]))
zlogCPM_atac_promoter <- km_res_atac_promoter$clustered[,1:ncol(this.logCPM)]

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ATACseq_promoters_all-genes.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM_atac_promoter) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM_atac_promoter) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))
heatmap.rna <- draw(Heatmap(zlogCPM_atac_promoter, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "ATACseq signal of all promoters in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_atac_promoter)[grepl("KRT_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(2,length(colnames(zlogCPM_atac_promoter)[grepl("MEL_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(3,length(colnames(zlogCPM_atac_promoter)[grepl("FRB_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(4,length(colnames(zlogCPM_atac_promoter)[grepl("KRT_IFN",colnames(zlogCPM_atac_promoter))])), 
                                             rep(5,length(colnames(zlogCPM_atac_promoter)[grepl("MEL_IFN",colnames(zlogCPM_atac_promoter))])),
                                             rep(6,length(colnames(zlogCPM_atac_promoter)[grepl("FRB_IFN",colnames(zlogCPM_atac_promoter))]))),
                            row_split = split_instructions_atac_promoter, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r ATAC-promoter-induced-genes-general-dynamics}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch promoters of all genes and scale
induced.genes <- rownames(km_res.rna$clustered)
induced.genes.promoters <- link_gene_promoter[which(link_gene_promoter$gene %in% induced.genes),"promoter"]
this.cpm <- CPM_atac[which(rownames(CPM_atac) %in% induced.genes.promoters$promoter),]
this.logCPM <- log2(this.cpm + 1)

# pick 5 clusters
df <- this.logCPM
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
rm(df,wss)

# niceKmeans
km_res_atac_promoter <- niceKmeans(this.logCPM, 5)
split_instructions_atac_promoter <- c(rep(1,km_res_atac_promoter$breaks[1]), 
                        rep(2,km_res_atac_promoter$breaks[2]-km_res_atac_promoter$breaks[1]),
                        rep(3,km_res_atac_promoter$breaks[3]-km_res_atac_promoter$breaks[2]),
                        rep(4,km_res_atac_promoter$breaks[4]-km_res_atac_promoter$breaks[3]),
                        rep(5,nrow(km_res_atac_promoter$clustered)-km_res_atac_promoter$breaks[4]))
zlogCPM_atac_promoter <- km_res_atac_promoter$clustered[,1:ncol(this.logCPM)]

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ATACseq_promoters_induced-genes_5kmm.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM_atac_promoter) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM_atac_promoter) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))
heatmap.rna <- draw(Heatmap(zlogCPM_atac_promoter, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "ATACseq signal of all promoters in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_atac_promoter)[grepl("KRT_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(2,length(colnames(zlogCPM_atac_promoter)[grepl("MEL_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(3,length(colnames(zlogCPM_atac_promoter)[grepl("FRB_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(4,length(colnames(zlogCPM_atac_promoter)[grepl("KRT_IFN",colnames(zlogCPM_atac_promoter))])), 
                                             rep(5,length(colnames(zlogCPM_atac_promoter)[grepl("MEL_IFN",colnames(zlogCPM_atac_promoter))])),
                                             rep(6,length(colnames(zlogCPM_atac_promoter)[grepl("FRB_IFN",colnames(zlogCPM_atac_promoter))]))),
                            row_split = split_instructions_atac_promoter, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r ATAC-promoter-induced-genes-cluster-matched}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch promoters of all genes and scale
induced.genes <- rownames(km_res.rna$clustered)
induced.genes.cluster <- data.frame(gene=induced.genes,
                                    cluster=split_instructions.rna) %>%
  dplyr::filter(gene %in% link_gene_promoter$gene)
induced.genes.and.promoter <- data.frame()
for (i in 1:nrow(induced.genes.cluster)) {
  this.promoter <- unique(link_gene_promoter[which(link_gene_promoter$gene==induced.genes.cluster$gene[i]),"promoter"])
  this.promoter.cluster <- induced.genes.cluster[which(induced.genes.cluster$gene==induced.genes.cluster$gene[i]),"cluster"]
    this.promoter.df <- data.frame(gene=rep(induced.genes.cluster$gene[i],length(this.promoter)),
                                   promoter=this.promoter,
                                   cluster=rep(this.promoter.cluster,length(this.promoter)))
    induced.genes.and.promoter <- rbind(induced.genes.and.promoter, this.promoter.df)
}

this.cpm <- CPM_atac[induced.genes.and.promoter$promoter,]
this.logCPM <- log2(this.cpm + 1)
this.zlogCPM <- t(scale(t(this.logCPM)))

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ATACseq_promoters_induced-genes_gene-cluster-matched.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(this.zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(this.zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

heatmap.atac <- draw(Heatmap(this.zlogCPM, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "ATACseq signal of all promoters in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_atac_promoter)[grepl("KRT_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(2,length(colnames(zlogCPM_atac_promoter)[grepl("MEL_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(3,length(colnames(zlogCPM_atac_promoter)[grepl("FRB_PBS",colnames(zlogCPM_atac_promoter))])), 
                                             rep(4,length(colnames(zlogCPM_atac_promoter)[grepl("KRT_IFN",colnames(zlogCPM_atac_promoter))])), 
                                             rep(5,length(colnames(zlogCPM_atac_promoter)[grepl("MEL_IFN",colnames(zlogCPM_atac_promoter))])),
                                             rep(6,length(colnames(zlogCPM_atac_promoter)[grepl("FRB_IFN",colnames(zlogCPM_atac_promoter))]))),
                            row_split = induced.genes.and.promoter$cluster, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(1, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
## Heatmaps of ATACseq-enhancer
```{r ATAC-enhancer-all-genes-general-dynamics}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch enhancers of all genes and scale
this.cpm <- CPM_atac[which(rownames(CPM_atac) %in% link_gene_enhancer$enhancer),]
this.logCPM <- log2(this.cpm + 1)

# pick 5 clusters
df <- this.logCPM
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
rm(df,wss)

# niceKmeans
km_res_atac_enhancer <- niceKmeans(this.logCPM, 5)
split_instructions_atac_enhancer <- c(rep(1,km_res_atac_enhancer$breaks[1]), 
                        rep(2,km_res_atac_enhancer$breaks[2]-km_res_atac_enhancer$breaks[1]),
                        rep(3,km_res_atac_enhancer$breaks[3]-km_res_atac_enhancer$breaks[2]),
                        rep(4,km_res_atac_enhancer$breaks[4]-km_res_atac_enhancer$breaks[3]),
                        rep(5,nrow(km_res_atac_enhancer$clustered)-km_res_atac_enhancer$breaks[4]))
zlogCPM_atac_enhancer <- km_res_atac_enhancer$clustered[,1:ncol(this.logCPM)]

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ATACseq_enhancers_all-genes.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM_atac_enhancer) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM_atac_enhancer) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))
heatmap.rna <- draw(Heatmap(zlogCPM_atac_enhancer, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "ATACseq signal of all enhancers in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_atac_enhancer)[grepl("KRT_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(2,length(colnames(zlogCPM_atac_enhancer)[grepl("MEL_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(3,length(colnames(zlogCPM_atac_enhancer)[grepl("FRB_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(4,length(colnames(zlogCPM_atac_enhancer)[grepl("KRT_IFN",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(5,length(colnames(zlogCPM_atac_enhancer)[grepl("MEL_IFN",colnames(zlogCPM_atac_enhancer))])),
                                             rep(6,length(colnames(zlogCPM_atac_enhancer)[grepl("FRB_IFN",colnames(zlogCPM_atac_enhancer))]))),
                            row_split = split_instructions_atac_enhancer, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r ATAC-enhancer-induced-genes-general-dynamics}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch enhancers of all genes and scale
induced.genes <- rownames(km_res.rna$clustered)
induced.genes.enhancers <- link_gene_enhancer[which(link_gene_enhancer$gene %in% induced.genes),"enhancer"]
this.cpm <- CPM_atac[which(rownames(CPM_atac) %in% induced.genes.enhancers$enhancer),]
this.logCPM <- log2(this.cpm + 1)

# pick 5 clusters
df <- this.logCPM
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
rm(df,wss)

# niceKmeans
km_res_atac_enhancer <- niceKmeans(this.logCPM, 5)
split_instructions_atac_enhancer <- c(rep(1,km_res_atac_enhancer$breaks[1]), 
                        rep(2,km_res_atac_enhancer$breaks[2]-km_res_atac_enhancer$breaks[1]),
                        rep(3,km_res_atac_enhancer$breaks[3]-km_res_atac_enhancer$breaks[2]),
                        rep(4,km_res_atac_enhancer$breaks[4]-km_res_atac_enhancer$breaks[3]),
                        rep(5,nrow(km_res_atac_enhancer$clustered)-km_res_atac_enhancer$breaks[4]))
zlogCPM_atac_enhancer <- km_res_atac_enhancer$clustered[,1:ncol(this.logCPM)]

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ATACseq_enhancers_induced-genes_5kmm.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM_atac_enhancer) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM_atac_enhancer) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))
heatmap.rna <- draw(Heatmap(zlogCPM_atac_enhancer, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "ATACseq signal of all enhancers in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_atac_enhancer)[grepl("KRT_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(2,length(colnames(zlogCPM_atac_enhancer)[grepl("MEL_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(3,length(colnames(zlogCPM_atac_enhancer)[grepl("FRB_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(4,length(colnames(zlogCPM_atac_enhancer)[grepl("KRT_IFN",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(5,length(colnames(zlogCPM_atac_enhancer)[grepl("MEL_IFN",colnames(zlogCPM_atac_enhancer))])),
                                             rep(6,length(colnames(zlogCPM_atac_enhancer)[grepl("FRB_IFN",colnames(zlogCPM_atac_enhancer))]))),
                            row_split = split_instructions_atac_enhancer, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r ATAC-enhancer-induced-genes-cluster-matched}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch enhancers of all genes and scale
induced.genes <- rownames(km_res.rna$clustered)
induced.genes.cluster <- data.frame(gene=induced.genes,
                                    cluster=split_instructions.rna) %>%
  dplyr::filter(gene %in% link_gene_enhancer$gene)
induced.genes.and.enhancer <- data.frame()
for (i in 1:nrow(induced.genes.cluster)) {
  this.enhancer <- unique(link_gene_enhancer[which(link_gene_enhancer$gene==induced.genes.cluster$gene[i]),"enhancer"])
  this.enhancer.cluster <- induced.genes.cluster[which(induced.genes.cluster$gene==induced.genes.cluster$gene[i]),"cluster"]
    this.enhancer.df <- data.frame(gene=rep(induced.genes.cluster$gene[i],length(this.enhancer)),
                                   enhancer=this.enhancer,
                                   cluster=rep(this.enhancer.cluster,length(this.enhancer)))
    induced.genes.and.enhancer <- rbind(induced.genes.and.enhancer, this.enhancer.df)
}

this.cpm <- CPM_atac[induced.genes.and.enhancer$enhancer,]
this.logCPM <- log2(this.cpm + 1)
this.zlogCPM <- t(scale(t(this.logCPM)))

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ATACseq_enhancers_induced-genes_gene-cluster-matched.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(this.zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(this.zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

heatmap.atac <- draw(Heatmap(this.zlogCPM, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "ATACseq signal of all enhancers in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_atac_enhancer)[grepl("KRT_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(2,length(colnames(zlogCPM_atac_enhancer)[grepl("MEL_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(3,length(colnames(zlogCPM_atac_enhancer)[grepl("FRB_PBS",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(4,length(colnames(zlogCPM_atac_enhancer)[grepl("KRT_IFN",colnames(zlogCPM_atac_enhancer))])), 
                                             rep(5,length(colnames(zlogCPM_atac_enhancer)[grepl("MEL_IFN",colnames(zlogCPM_atac_enhancer))])),
                                             rep(6,length(colnames(zlogCPM_atac_enhancer)[grepl("FRB_IFN",colnames(zlogCPM_atac_enhancer))]))),
                            row_split = induced.genes.and.enhancer$cluster, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(1, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
## Other Heatmaps of ATACseq
```{r ATAC-promoter-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))
# 1. center and scale data
logCPM <- log2(CPM_atac_DEG_promoter + 1)

# 2. hierarchically cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#lot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.atac.promoter <- km_res
split_instructions.atac.promoter <- split_instructions
```
```{r ATAC-promoter-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2}
draw_atac_promoter_heatmap <- function(g) {# 4. load peaks of interest
  peaks <- link_gene_promoter[which(link_gene_promoter$gene %in% g),] %>% dplyr::pull(promoter) %>% unique
  zlogCPM.heatmap <- km_res.atac.promoter$clustered[,1:ncol(CPM_atac_DEG_promoter)]
  zlogCPM.heatmap_with_peaks_kegg <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% peaks)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_peaks_kegg))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.atac.promoter <- Heatmap(zlogCPM.heatmap, 
                                   name = "z-score log2CPM",
                                   top_annotation = column_ha,
                                   right_annotation = row_ha,
                                   column_title = "ATACseq heatmap of promoters",
                                   cluster_columns = FALSE,
                                   cluster_rows = FALSE,
                                   column_split = c(rep(1,9), rep(2,9)),
                                   row_split = split_instructions.atac.promoter, 
                                   show_row_names = FALSE,
                                   row_title_rot = 0,
                                   column_names_rot = 45,
                                   row_gap = unit(2, "mm"),
                                   border = FALSE,
                                   column_names_gp = gpar(fontsize = 5),
                                   column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                   row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.atac.promoter)
  
  # clean-up
  rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_peaks_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
}

```
```{r ATAC-enhancer-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
# 1. center and scale data
zlogCPM <- log2(CPM_atac_DEG_enhancer + 1)

# 2. hierarchically cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(zlogCPM, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.atac.enhancer <- km_res
split_instructions.atac.enhancer <- split_instructions
```
```{r ATAC-enhancer-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2}
draw_atac_enhancer_heatmap <- function(glst){
  # 4. load peaks of interest
  peaks <- link_gene_enhancer[which(link_gene_enhancer$gene %in% glst),] %>% pull(enhancer) %>% unique
  zlogCPM.heatmap <- km_res.atac.enhancer$clustered[,1:ncol(CPM_atac_DEG_enhancer)]
  zlogCPM.heatmap_with_peaks_kegg <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% peaks)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_peaks_kegg))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.atac.enhancer <- Heatmap(zlogCPM.heatmap, 
                                   name = "z-score log2CPM",
                                   top_annotation = column_ha,
                                   right_annotation = row_ha,
                                   column_title = "ATACseq heatmap of enhancers",
                                   cluster_columns = FALSE,
                                   cluster_rows = FALSE,
                                   column_split = c(rep(1,9), rep(2,9)),
                                   row_split = split_instructions.atac.enhancer, 
                                   show_row_names = FALSE,
                                   row_title_rot = 0,
                                   column_names_rot = 45,
                                   row_gap = unit(2, "mm"),
                                   border = FALSE,
                                   column_names_gp = gpar(fontsize = 5),
                                   column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                   row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.atac.enhancer)
  
  # clean-up
  rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_peaks_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
}
```
## Heatmaps of ChIPseq-promoter
```{r ChIP-promoter-all-genes-general-dynamics}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch promoters of all genes and scale
this.cpm <- CPM_chip[which(rownames(CPM_chip) %in% link_gene_promoter$promoter),]
this.logCPM <- log2(this.cpm + 1)

# pick 5 clusters
df <- this.logCPM
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
rm(df,wss)

# niceKmeans
km_res_chip_promoter <- niceKmeans(this.logCPM, 5)
split_instructions_chip_promoter <- c(rep(1,km_res_chip_promoter$breaks[1]), 
                        rep(2,km_res_chip_promoter$breaks[2]-km_res_chip_promoter$breaks[1]),
                        rep(3,km_res_chip_promoter$breaks[3]-km_res_chip_promoter$breaks[2]),
                        rep(4,km_res_chip_promoter$breaks[4]-km_res_chip_promoter$breaks[3]),
                        rep(5,nrow(km_res_chip_promoter$clustered)-km_res_chip_promoter$breaks[4]))
zlogCPM_chip_promoter <- km_res_chip_promoter$clustered[,1:ncol(this.logCPM)]

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ChIPseq_promoters_all-genes.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM_chip_promoter) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM_chip_promoter) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))
heatmap.rna <- draw(Heatmap(zlogCPM_chip_promoter, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "chipseq signal of all promoters in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_chip_promoter)[grepl("KRT_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(2,length(colnames(zlogCPM_chip_promoter)[grepl("MEL_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(3,length(colnames(zlogCPM_chip_promoter)[grepl("FRB_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(4,length(colnames(zlogCPM_chip_promoter)[grepl("KRT_IFN",colnames(zlogCPM_chip_promoter))])), 
                                             rep(5,length(colnames(zlogCPM_chip_promoter)[grepl("MEL_IFN",colnames(zlogCPM_chip_promoter))])),
                                             rep(6,length(colnames(zlogCPM_chip_promoter)[grepl("FRB_IFN",colnames(zlogCPM_chip_promoter))]))),
                            row_split = split_instructions_chip_promoter, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r ChIPseq-promoter-induced-genes-general-dynamics}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch promoters of all genes and scale
induced.genes <- rownames(km_res.rna$clustered)
induced.genes.promoters <- link_gene_promoter[which(link_gene_promoter$gene %in% induced.genes),"promoter"]
this.cpm <- CPM_chip[which(rownames(CPM_chip) %in% induced.genes.promoters$promoter),]
this.logCPM <- log2(this.cpm + 1)

# pick 5 clusters
df <- this.logCPM
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
rm(df,wss)

# niceKmeans
km_res_chip_promoter <- niceKmeans(this.logCPM, 5)
split_instructions_chip_promoter <- c(rep(1,km_res_chip_promoter$breaks[1]), 
                        rep(2,km_res_chip_promoter$breaks[2]-km_res_chip_promoter$breaks[1]),
                        rep(3,km_res_chip_promoter$breaks[3]-km_res_chip_promoter$breaks[2]),
                        rep(4,km_res_chip_promoter$breaks[4]-km_res_chip_promoter$breaks[3]),
                        rep(5,nrow(km_res_chip_promoter$clustered)-km_res_chip_promoter$breaks[4]))
zlogCPM_chip_promoter <- km_res_chip_promoter$clustered[,1:ncol(this.logCPM)]

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_chipseq_promoters_induced-genes_5kmm.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM_chip_promoter) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM_chip_promoter) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))
heatmap.rna <- draw(Heatmap(zlogCPM_chip_promoter, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "chipseq signal of all promoters in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_chip_promoter)[grepl("KRT_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(2,length(colnames(zlogCPM_chip_promoter)[grepl("MEL_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(3,length(colnames(zlogCPM_chip_promoter)[grepl("FRB_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(4,length(colnames(zlogCPM_chip_promoter)[grepl("KRT_IFN",colnames(zlogCPM_chip_promoter))])), 
                                             rep(5,length(colnames(zlogCPM_chip_promoter)[grepl("MEL_IFN",colnames(zlogCPM_chip_promoter))])),
                                             rep(6,length(colnames(zlogCPM_chip_promoter)[grepl("FRB_IFN",colnames(zlogCPM_chip_promoter))]))),
                            row_split = split_instructions_chip_promoter, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r ChIPseq-promoter-induced-genes-cluster-matched}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch promoters of all genes and scale
induced.genes <- rownames(km_res.rna$clustered)
induced.genes.cluster <- data.frame(gene=induced.genes,
                                    cluster=split_instructions.rna) %>%
  dplyr::filter(gene %in% link_gene_promoter$gene)
induced.genes.and.promoter <- data.frame()
for (i in 1:nrow(induced.genes.cluster)) {
  this.promoter <- unique(link_gene_promoter[which(link_gene_promoter$gene==induced.genes.cluster$gene[i]),"promoter"])
  this.promoter.cluster <- induced.genes.cluster[which(induced.genes.cluster$gene==induced.genes.cluster$gene[i]),"cluster"]
    this.promoter.df <- data.frame(gene=rep(induced.genes.cluster$gene[i],length(this.promoter)),
                                   promoter=this.promoter,
                                   cluster=rep(this.promoter.cluster,length(this.promoter)))
    induced.genes.and.promoter <- rbind(induced.genes.and.promoter, this.promoter.df)
}

this.cpm <- CPM_chip[induced.genes.and.promoter$promoter,]
this.logCPM <- log2(this.cpm + 1)
this.zlogCPM <- t(scale(t(this.logCPM)))

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ChIPseq_promoters_induced-genes_gene-cluster-matched.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(this.zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(this.zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

heatmap.chip <- draw(Heatmap(this.zlogCPM, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "chipseq signal of all promoters in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_chip_promoter)[grepl("KRT_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(2,length(colnames(zlogCPM_chip_promoter)[grepl("MEL_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(3,length(colnames(zlogCPM_chip_promoter)[grepl("FRB_PBS",colnames(zlogCPM_chip_promoter))])), 
                                             rep(4,length(colnames(zlogCPM_chip_promoter)[grepl("KRT_IFN",colnames(zlogCPM_chip_promoter))])), 
                                             rep(5,length(colnames(zlogCPM_chip_promoter)[grepl("MEL_IFN",colnames(zlogCPM_chip_promoter))])),
                                             rep(6,length(colnames(zlogCPM_chip_promoter)[grepl("FRB_IFN",colnames(zlogCPM_chip_promoter))]))),
                            row_split = induced.genes.and.promoter$cluster, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(1, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
## Heatmaps of ChIPseq-enhancer
```{r ChIPseq-enhancer-all-genes-general-dynamics}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch enhancers of all genes and scale
this.cpm <- CPM_chip[which(rownames(CPM_chip) %in% link_gene_enhancer$enhancer),]
this.logCPM <- log2(this.cpm + 1)

# pick 5 clusters
df <- this.logCPM
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
rm(df,wss)

# niceKmeans
km_res_chip_enhancer <- niceKmeans(this.logCPM, 5)
split_instructions_chip_enhancer <- c(rep(1,km_res_chip_enhancer$breaks[1]), 
                        rep(2,km_res_chip_enhancer$breaks[2]-km_res_chip_enhancer$breaks[1]),
                        rep(3,km_res_chip_enhancer$breaks[3]-km_res_chip_enhancer$breaks[2]),
                        rep(4,km_res_chip_enhancer$breaks[4]-km_res_chip_enhancer$breaks[3]),
                        rep(5,nrow(km_res_chip_enhancer$clustered)-km_res_chip_enhancer$breaks[4]))
zlogCPM_chip_enhancer <- km_res_chip_enhancer$clustered[,1:ncol(this.logCPM)]

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ChIPseq_enhancers_all-genes.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM_chip_enhancer) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM_chip_enhancer) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))
heatmap.rna <- draw(Heatmap(zlogCPM_chip_enhancer, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "H3K27ac ChIPseq signal of all enhancers in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_chip_enhancer)[grepl("KRT_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(2,length(colnames(zlogCPM_chip_enhancer)[grepl("MEL_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(3,length(colnames(zlogCPM_chip_enhancer)[grepl("FRB_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(4,length(colnames(zlogCPM_chip_enhancer)[grepl("KRT_IFN",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(5,length(colnames(zlogCPM_chip_enhancer)[grepl("MEL_IFN",colnames(zlogCPM_chip_enhancer))])),
                                             rep(6,length(colnames(zlogCPM_chip_enhancer)[grepl("FRB_IFN",colnames(zlogCPM_chip_enhancer))]))),
                            row_split = split_instructions_chip_enhancer, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r ChIPseq-enhancer-induced-genes-general-dynamics}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch enhancers of all genes and scale
induced.genes <- rownames(km_res.rna$clustered)
induced.genes.enhancers <- link_gene_enhancer[which(link_gene_enhancer$gene %in% induced.genes),"enhancer"]
this.cpm <- CPM_chip[which(rownames(CPM_chip) %in% induced.genes.enhancers$enhancer),]
this.logCPM <- log2(this.cpm + 1)

# pick 5 clusters
df <- this.logCPM
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
rm(df,wss)

# niceKmeans
km_res_chip_enhancer <- niceKmeans(this.logCPM, 5)
split_instructions_chip_enhancer <- c(rep(1,km_res_chip_enhancer$breaks[1]), 
                        rep(2,km_res_chip_enhancer$breaks[2]-km_res_chip_enhancer$breaks[1]),
                        rep(3,km_res_chip_enhancer$breaks[3]-km_res_chip_enhancer$breaks[2]),
                        rep(4,km_res_chip_enhancer$breaks[4]-km_res_chip_enhancer$breaks[3]),
                        rep(5,nrow(km_res_chip_enhancer$clustered)-km_res_chip_enhancer$breaks[4]))
zlogCPM_chip_enhancer <- km_res_chip_enhancer$clustered[,1:ncol(this.logCPM)]

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ChIPseq_enhancers_induced-genes_5kmm.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM_chip_enhancer) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM_chip_enhancer) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))
heatmap.rna <- draw(Heatmap(zlogCPM_chip_enhancer, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "chipseq signal of all enhancers in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_chip_enhancer)[grepl("KRT_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(2,length(colnames(zlogCPM_chip_enhancer)[grepl("MEL_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(3,length(colnames(zlogCPM_chip_enhancer)[grepl("FRB_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(4,length(colnames(zlogCPM_chip_enhancer)[grepl("KRT_IFN",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(5,length(colnames(zlogCPM_chip_enhancer)[grepl("MEL_IFN",colnames(zlogCPM_chip_enhancer))])),
                                             rep(6,length(colnames(zlogCPM_chip_enhancer)[grepl("FRB_IFN",colnames(zlogCPM_chip_enhancer))]))),
                            row_split = split_instructions_chip_enhancer, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
```{r ChIPseq-enhancer-induced-genes-cluster-matched}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
load(paste0(Dir,"/heatmap_RNA/myEnvironment_7kmm-hc_padj0.05_log2FC1.5_avgCPM10.RData"))

# fetch enhancers of all genes and scale
induced.genes <- rownames(km_res.rna$clustered)
induced.genes.cluster <- data.frame(gene=induced.genes,
                                    cluster=split_instructions.rna) %>%
  dplyr::filter(gene %in% link_gene_enhancer$gene)
induced.genes.and.enhancer <- data.frame()
for (i in 1:nrow(induced.genes.cluster)) {
  this.enhancer <- unique(link_gene_enhancer[which(link_gene_enhancer$gene==induced.genes.cluster$gene[i]),"enhancer"])
  this.enhancer.cluster <- induced.genes.cluster[which(induced.genes.cluster$gene==induced.genes.cluster$gene[i]),"cluster"]
    this.enhancer.df <- data.frame(gene=rep(induced.genes.cluster$gene[i],length(this.enhancer)),
                                   enhancer=this.enhancer,
                                   cluster=rep(this.enhancer.cluster,length(this.enhancer)))
    induced.genes.and.enhancer <- rbind(induced.genes.and.enhancer, this.enhancer.df)
}

this.cpm <- CPM_chip[induced.genes.and.enhancer$enhancer,]
this.logCPM <- log2(this.cpm + 1)
this.zlogCPM <- t(scale(t(this.logCPM)))

# heatmap
pdf(paste0(Dir,"/heatmap_cRE/heatmap_ChIPseq_enhancers_induced-genes_gene-cluster-matched.pdf"),width=24,height=30)
column_ha = HeatmapAnnotation(
  condition = colnames(this.zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(this.zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

heatmap.chip <- draw(Heatmap(this.zlogCPM, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            column_title = "chipseq signal of all enhancers in 3 celltypes, control vs IGNg stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(zlogCPM_chip_enhancer)[grepl("KRT_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(2,length(colnames(zlogCPM_chip_enhancer)[grepl("MEL_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(3,length(colnames(zlogCPM_chip_enhancer)[grepl("FRB_PBS",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(4,length(colnames(zlogCPM_chip_enhancer)[grepl("KRT_IFN",colnames(zlogCPM_chip_enhancer))])), 
                                             rep(5,length(colnames(zlogCPM_chip_enhancer)[grepl("MEL_IFN",colnames(zlogCPM_chip_enhancer))])),
                                             rep(6,length(colnames(zlogCPM_chip_enhancer)[grepl("FRB_IFN",colnames(zlogCPM_chip_enhancer))]))),
                            row_split = induced.genes.and.enhancer$cluster, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(1, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

```
## Other Heatmaps of ChIPseq
```{r ChIP-promoter-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
# 1. center and scale data
logCPM <- log2(CPM_chip_DEG_promoter + 1)

# 2. hierarchially cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.chip.promoter <- km_res
split_instructions.chip.promoter <- split_instructions
```
```{r ChIP-promoter-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2}
draw_chip_promoter_heatmap <- function(g){
  # 4. load peaks of interest
  peaks <- link_gene_promoter[which(link_gene_promoter$gene %in% g),] %>% pull(promoter) %>% unique
  zlogCPM.heatmap <- km_res.chip.promoter$clustered[,1:ncol(CPM_chip_DEG_promoter)]
  zlogCPM.heatmap_with_peaks_kegg <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% peaks)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_peaks_kegg))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.chip.promoter <- Heatmap(zlogCPM.heatmap, 
                                   name = "z-score log2CPM",
                                   top_annotation = column_ha,
                                   right_annotation = row_ha,
                                   column_title = "H3K27ac ChIPseq heatmap of promoters",
                                   cluster_columns = FALSE,
                                   cluster_rows = FALSE,
                                   column_split = c(rep(1,9), rep(2,9)),
                                   row_split = split_instructions.chip.promoter, 
                                   show_row_names = FALSE,
                                   row_title_rot = 0,
                                   column_names_rot = 45,
                                   row_gap = unit(2, "mm"),
                                   border = FALSE,
                                   column_names_gp = gpar(fontsize = 5),
                                   column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                   row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.chip.promoter)
  
  # clean-up
  rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_peaks_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
}
```
```{r chip-enhancer-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
# 1. center and scale data
logCPM <- log2(CPM_chip_DEG_enhancer + 1)

# 2. hierarchially cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.chip.enhancer <- km_res
split_instructions.chip.enhancer <- split_instructions
```
```{r chip-enhancer-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2}
draw_chip_enhancer_heatmap <- function(g){
  # 4. load peaks of interest
  peaks <- link_gene_enhancer[which(link_gene_enhancer$gene %in% g),] %>% pull(enhancer) %>% unique
  zlogCPM.heatmap <- km_res.chip.enhancer$clustered[,1:ncol(CPM_chip_DEG_enhancer)]
  zlogCPM.heatmap_with_peaks_kegg <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% peaks)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_peaks_kegg))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.chip.enhancer <- Heatmap(zlogCPM.heatmap, 
                                   name = "z-score log2CPM",
                                   top_annotation = column_ha,
                                   right_annotation = row_ha,
                                   column_title = "chipseq heatmap of enhancers",
                                   cluster_columns = FALSE,
                                   cluster_rows = FALSE,
                                   column_split = c(rep(1,9), rep(2,9)),
                                   row_split = split_instructions.chip.enhancer, 
                                   show_row_names = FALSE,
                                   row_title_rot = 0,
                                   column_names_rot = 45,
                                   row_gap = unit(2, "mm"),
                                   border = FALSE,
                                   column_names_gp = gpar(fontsize = 5),
                                   column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                   row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.chip.enhancer)
  
  # clean-up
  rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_peaks_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
}
```

# plot heatmaps Round Two
## Refined clustering of round one RNAseq heatmap
```{r rna-heatmap-round2-avgCPM10-log2FC1.5}
Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
setwd(Dir)
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)

pbs_cols <- colnames(CPM_rna)[which(grepl("PBS",colnames(CPM_rna)))]
ifn_cols <- colnames(CPM_rna)[which(grepl("IFN",colnames(CPM_rna)))]
rows_above_CPM_cutoff_PBS <- rownames(CPM_rna[rowMeans(CPM_rna[,pbs_cols]) >= 10,]) %>% unlist()
rows_above_CPM_cutoff_IFN <- rownames(CPM_rna[rowMeans(CPM_rna[,ifn_cols]) >= 10,]) %>% unlist()
selected_rows <- unique(c(rows_above_CPM_cutoff_PBS,rows_above_CPM_cutoff_IFN))
CPM_rna_new <- CPM_rna %>% dplyr::filter(rownames(.) %in% selected_rows)

DEG <- unique(c(rna.res.krt[which(rna.res.krt$padj<0.05 & (rna.res.krt$log2FoldChange)>1.5),"gene"],
                rna.res.mel[which(rna.res.mel$padj<0.05 & (rna.res.mel$log2FoldChange)>1.5),"gene"],
                rna.res.frb[which(rna.res.frb$padj<0.05 & (rna.res.frb$log2FoldChange)>1.5),"gene"]))
CPM_rna_DEG <- CPM_rna_new %>% dplyr::filter(rownames(.) %in% DEG)
logCPM <- log2(CPM_rna_DEG + 1)

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 14)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,km_res$breaks[5]-km_res$breaks[4]),
                        rep(6,km_res$breaks[6]-km_res$breaks[5]),
                        rep(7,km_res$breaks[7]-km_res$breaks[6]),
                        rep(8,km_res$breaks[8]-km_res$breaks[7]),
                        rep(9,km_res$breaks[9]-km_res$breaks[8]),
                        rep(10,km_res$breaks[10]-km_res$breaks[9]),
                        rep(11,km_res$breaks[11]-km_res$breaks[10]),
                        rep(12,km_res$breaks[12]-km_res$breaks[11]),
                        rep(13,km_res$breaks[13]-km_res$breaks[12]),
                        rep(14,nrow(km_res$clustered)-km_res$breaks[13]))
km_res.rna <- km_res
split_instructions.rna <- split_instructions

# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_rna_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  celltype = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNA/heatmap_RNAseq_DEgenes_14kmm-hc_03282023_padj0.05_log2FC1.5_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, control vs IGNg stimulated, known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(CPM_rna_DEG)[grepl("PBS_KRT",colnames(CPM_rna_DEG))])), 
                                             rep(2,length(colnames(CPM_rna_DEG)[grepl("PBS_MEL",colnames(CPM_rna_DEG))])), 
                                             rep(3,length(colnames(CPM_rna_DEG)[grepl("PBS_FRB",colnames(CPM_rna_DEG))])), 
                                             rep(4,length(colnames(CPM_rna_DEG)[grepl("IFN_KRT",colnames(CPM_rna_DEG))])), 
                                             rep(5,length(colnames(CPM_rna_DEG)[grepl("IFN_MEL",colnames(CPM_rna_DEG))])),
                                             rep(6,length(colnames(CPM_rna_DEG)[grepl("IFN_FRB",colnames(CPM_rna_DEG))]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()
cluster1_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==1)]

save.image(file=paste0(Dir,"/heatmap_RNA/myEnvironment_heatmap_RNAseq_DEgenes_7kmm-hc_02292023_padj0.05_log2FC1.5_avgCPM10.RData"))
```
## PLOT
```{r plot}
gene <- rownames(km_res.rna$clustered)[which(split_instructions.rna==3)]
p1 <- draw_rna_heatmap(gene)
p2 <- draw_atac_promoter_heatmap(gene)
p3 <- draw_atac_enhancer_heatmap(gene)
p4 <- draw_chip_promoter_heatmap(gene)
p5 <- draw_chip_enhancer_heatmap(gene)

p.promoter <- p2 %v% p4
p.enhancer <- p3 %v% p5

grid.newpage()
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
grid.rect(gp = gpar(fill = "#FF000020"))
draw(p1, newpage = FALSE)
popViewport()

pushViewport(viewport(x = 0.5, y = 1, width = 0.5, height = 0.5, just = c("left", "top")))
grid.rect(gp = gpar(fill = "#00FF0020"))
draw(p.promoter, newpage = FALSE)
popViewport()

pushViewport(viewport(x = 0.5, y = 0.5, width = 0.5, height = 0.5, just = c("left", "top")))
grid.rect(gp = gpar(fill = "#0000FF20"))
draw(p.enhancer, newpage = FALSE)
popViewport()
```
# For genes that are up/down-regulated in all 3cts, are their enhancers shared across celltypes?
## For commonly regulated genes, check if their enhancers are shared across celltypes in ATACseq
```{r gene-enhancer-sharing-atac}
# to view which genes belong to clsuter of (commonly up-regulated genes)
commonly_induced_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4)]
gene <- commonly_induced_genes[2]
make_paired_plot(gene,CPM_rna_DEG[gene,])

# get the enhancers of these genes
enhancer_of_gene_group1 <- link_gene_enhancer %>% dplyr::filter(gene %in% commonly_induced_genes) %>% arrange(gene) %>% pull(enhancer) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
logCPM <- log2(CPM_atac_DEG_enhancer[enhancer_of_gene_group1,] + 1)
zlogCPM <- t(scale(t(log2(CPM_atac_DEG_enhancer[enhancer_of_gene_group1,] + 1))))

# # # cluster the enhancers
# km_res <- niceKmeans(logCPM, 7)
# split_instructions <- c(rep(1,km_res$breaks[1]),
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,km_res$breaks[5]-km_res$breaks[4]),
#                         rep(6,km_res$breaks[6]-km_res$breaks[5]),
#                         rep(7,nrow(km_res$clustered)-km_res$breaks[6]))
# km_res.atac.enhancer <- km_res
# split_instructions.atac.enhancer <- split_instructions

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.atac.enhancer <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ATACseq heatmap of enhancers of commonly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.atac.enhancer); enhancer_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# pick the commonly open enhancers of commonly induced genes
enhancers_shared <- rownames(zlogCPM)[sort(row_order(ht)$`5`)]
enhancers_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`2`,row_order(ht)$`4`,row_order(ht)$`7`))]
enhancers_paired <- rownames(zlogCPM)[sort(c(row_order(ht)$`1`,row_order(ht)$`3`,row_order(ht)$`6`))]

df1 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_shared) %>% mutate(pattern="shared")
df2 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_cts) %>% mutate(pattern="cts")
df3 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_paired) %>% mutate(pattern="paired")
df <- rbind(df1,df2,df3)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)
```
```{r gene-promoter-sharing-atac}
# to view which genes belong to clsuter of (commonly up-regulated genes)
commonly_induced_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==6)]

# get the enhancers of these genes
promoter_of_gene_group1 <- link_gene_promoter %>% dplyr::filter(gene %in% commonly_induced_genes) %>% arrange(gene) %>% pull(promoter) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_atac_DEG_promoter[promoter_of_gene_group1,] + 1))))

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.atac.promoter <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ATACseq heatmap of promoters of commonly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.atac.promoter); promoter_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# plot the distance of common promoters and cts promoters
# pick the commonly open enhancers of commonly induced genes
promoters_shared <- rownames(zlogCPM)[sort(c(row_order(ht)$`3`))]
promoters_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`4`,row_order(ht)$`6`))]

df1 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_shared) %>% mutate(pattern="shared")
df2 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_cts) %>% mutate(pattern="cts")
df <- rbind(df1,df2)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)

# Could the cell-type specific promoters come from genes with multiple promoters?
genes_of_df1 <- link_gene_promoter %>% dplyr::filter(gene %in% df1$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="common")
genes_of_df2 <- link_gene_promoter %>% dplyr::filter(gene %in% df2$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="cts")

df <- rbind(genes_of_df1, genes_of_df2)

get_box_stats <- function(y, upper_limit = max(df$count) * 1.15) {
  return(data.frame(
    y = 0.95 * upper_limit,
    label = paste(
      "Count =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n",
      "Median =", round(median(y), 2), "\n"
    )
  ))
}
ggplot(df, aes(x=pattern, y=count)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  stat_summary(fun.data = get_box_stats, geom="text", hjust=0.5, vjust=0.8)
```
## For commonly regulated genes, check if their enhancers are shared across celltypes in ChIPseq
```{r gene-enhancer-sharing-chip}
# to view which genes belong to clsuter of (commonly up-regulated genes)
commonly_induced_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==6)]

# get the enhancers of these genes
enhancer_of_gene_group1 <- link_gene_enhancer %>% dplyr::filter(gene %in% commonly_induced_genes) %>% arrange(gene) %>% pull(enhancer) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_chip_DEG_enhancer[enhancer_of_gene_group1,] + 1))))

# # cluster the enhancers
# km_res <- niceKmeans(zlogCPM, 7)
# split_instructions <- c(rep(1,km_res$breaks[1]), 
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,km_res$breaks[5]-km_res$breaks[4]),
#                         rep(6,km_res$breaks[6]-km_res$breaks[5]),
#                         rep(7,nrow(km_res$clustered)-km_res$breaks[6]))
# km_res.chip.enhancer <- km_res
# split_instructions.chip.enhancer <- split_instructions

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.chip.enhancer <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ChIPseq heatmap of enhancers of commonly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.chip.enhancer); enhancer_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# pick the commonly open enhancers of commonly induced genes
enhancers_shared <- rownames(zlogCPM)[sort(row_order(ht)$`5`)]
enhancers_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`2`,row_order(ht)$`4`,row_order(ht)$`7`))]
enhancers_paired <- rownames(zlogCPM)[sort(c(row_order(ht)$`1`,row_order(ht)$`3`,row_order(ht)$`6`))]

df1 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_shared) %>% mutate(pattern="shared")
df2 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_cts) %>% mutate(pattern="cts")
df3 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_paired) %>% mutate(pattern="paired")
df <- rbind(df1,df2,df3)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)
```
```{r gene-promoter-sharing-chip}
# to view which genes belong to clsuter of (commonly up-regulated genes)
commonly_induced_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==6)]

# get the enhancers of these genes
promoter_of_gene_group1 <- link_gene_promoter %>% dplyr::filter(gene %in% commonly_induced_genes) %>% arrange(gene) %>% pull(promoter) %>% unique

# prepare data to plot the ChIPseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_chip_DEG_promoter[promoter_of_gene_group1,] + 1))))

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.chip.promoter <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ChIPseq heatmap of promoters of commonly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.chip.promoter); promoter_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# plot the distance of common promoters and cts promoters
# pick the commonly open enhancers of commonly induced genes
promoters_shared <- rownames(zlogCPM)[sort(c(row_order(ht)$`3`))]
promoters_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`4`,row_order(ht)$`6`))]

df1 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_shared) %>% mutate(pattern="shared")
df2 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_cts) %>% mutate(pattern="cts")
df <- rbind(df1,df2)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)

# Could the cell-type specific promoters come from genes with multiple promoters?
genes_of_df1 <- link_gene_promoter %>% dplyr::filter(gene %in% df1$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="common")
genes_of_df2 <- link_gene_promoter %>% dplyr::filter(gene %in% df2$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="cts")

df <- rbind(genes_of_df1, genes_of_df2)

get_box_stats <- function(y, upper_limit = max(df$count) * 1.15) {
  return(data.frame(
    y = 0.95 * upper_limit,
    label = paste(
      "Count =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n",
      "Median =", round(median(y), 2), "\n"
    )
  ))
}
ggplot(df, aes(x=pattern, y=count)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  stat_summary(fun.data = get_box_stats, geom="text", hjust=0.5, vjust=0.8)
```
## For cell-type-specificly expressed genes, check if their enhancers are unique across celltypes in ATACseq
```{r cts-gene-enhancer-sharing-atac}
# to view which genes belong to clsuter of (cts-regulated genes)
these_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4 | split_instructions.rna==6)]
gene <- these_genes[200]
make_paired_plot(gene,CPM_rna_DEG[gene,])

# get the enhancers of these genes
enhancer_of_gene_group1 <- link_gene_enhancer %>% dplyr::filter(gene %in% these_genes) %>% arrange(gene) %>% pull(enhancer) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_atac_DEG_enhancer[enhancer_of_gene_group1,] + 1))))

# # cluster the enhancers
# km_res <- niceKmeans(zlogCPM, 7)
# split_instructions <- c(rep(1,km_res$breaks[1]), 
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,km_res$breaks[5]-km_res$breaks[4]),
#                         rep(6,km_res$breaks[6]-km_res$breaks[5]),
#                         rep(7,nrow(km_res$clustered)-km_res$breaks[6]))
# km_res.atac.enhancer <- km_res
# split_instructions.atac.enhancer <- split_instructions

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.atac.enhancer <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ATACseq heatmap of enhancers of cell-type-spcificly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.atac.enhancer); enhancer_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.
```
```{r gene-promoter-sharing-atac}
# to view which genes belong to clsuter of (commonly up-regulated genes)
these_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4 | split_instructions.rna==6)]

# get the enhancers of these genes
promoter_of_gene_group1 <- link_gene_promoter %>% dplyr::filter(gene %in% these_genes) %>% arrange(gene) %>% pull(promoter) %>% unique

# prepare data to plot the atacseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_atac_DEG_promoter[promoter_of_gene_group1,] + 1))))

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.atac.promoter <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "atacseq heatmap of promoters of cell-type-specificly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.atac.promoter); promoter_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# plot the distance of common promoters and cts promoters
# pick the commonly open enhancers of commonly induced genes
promoters_shared <- rownames(zlogCPM)[sort(c(row_order(ht)$`3`))]
promoters_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`4`,row_order(ht)$`6`))]

df1 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_shared) %>% mutate(pattern="shared")
df2 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_cts) %>% mutate(pattern="cts")
df <- rbind(df1,df2)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)

# Could the cell-type specific promoters come from genes with multiple promoters?
genes_of_df1 <- link_gene_promoter %>% dplyr::filter(gene %in% df1$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="common")
genes_of_df2 <- link_gene_promoter %>% dplyr::filter(gene %in% df2$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="cts")

df <- rbind(genes_of_df1, genes_of_df2)

get_box_stats <- function(y, upper_limit = max(df$count) * 1.15) {
  return(data.frame(
    y = 0.95 * upper_limit,
    label = paste(
      "Count =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n",
      "Median =", round(median(y), 2), "\n"
    )
  ))
}
ggplot(df, aes(x=pattern, y=count)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  stat_summary(fun.data = get_box_stats, geom="text", hjust=0.5, vjust=0.8)
```
## For cell-type-specificly expressed genes, check if their enhancers are unique across celltypes in ChIPseq
```{r cts-gene-enhancer-sharing-chip}
# to view which genes belong to clsuter of (cts-regulated genes)
these_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4 | split_instructions.rna==6)]

# get the enhancers of these genes
enhancer_of_gene_group1 <- link_gene_enhancer %>% dplyr::filter(gene %in% these_genes) %>% arrange(gene) %>% pull(enhancer) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_chip_DEG_enhancer[enhancer_of_gene_group1,] + 1))))

# # cluster the enhancers
# km_res <- niceKmeans(zlogCPM, 7)
# split_instructions <- c(rep(1,km_res$breaks[1]), 
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,km_res$breaks[5]-km_res$breaks[4]),
#                         rep(6,km_res$breaks[6]-km_res$breaks[5]),
#                         rep(7,nrow(km_res$clustered)-km_res$breaks[6]))
# km_res.chip.enhancer <- km_res
# split_instructions.chip.enhancer <- split_instructions

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.chip.enhancer <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ChIPseq heatmap of enhancers of cell-type-spcificly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.chip.enhancer); enhancer_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.
```
```{r gene-promoter-sharing-chip}
# to view which genes belong to clsuter of (commonly up-regulated genes)
these_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4 | split_instructions.rna==6)]

# get the enhancers of these genes
promoter_of_gene_group1 <- link_gene_promoter %>% dplyr::filter(gene %in% these_genes) %>% arrange(gene) %>% pull(promoter) %>% unique

# prepare data to plot the ChIPseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_chip_DEG_promoter[promoter_of_gene_group1,] + 1))))

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.chip.promoter <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ChIPseq heatmap of promoters of cell-type-specificly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.chip.promoter); promoter_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# plot the distance of common promoters and cts promoters
# pick the commonly open enhancers of commonly induced genes
promoters_shared <- rownames(zlogCPM)[sort(c(row_order(ht)$`3`))]
promoters_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`4`,row_order(ht)$`6`))]

df1 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_shared) %>% mutate(pattern="shared")
df2 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_cts) %>% mutate(pattern="cts")
df <- rbind(df1,df2)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)

# Could the cell-type specific promoters come from genes with multiple promoters?
genes_of_df1 <- link_gene_promoter %>% dplyr::filter(gene %in% df1$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="common")
genes_of_df2 <- link_gene_promoter %>% dplyr::filter(gene %in% df2$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="cts")

df <- rbind(genes_of_df1, genes_of_df2)

get_box_stats <- function(y, upper_limit = max(df$count) * 1.15) {
  return(data.frame(
    y = 0.95 * upper_limit,
    label = paste(
      "Count =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n",
      "Median =", round(median(y), 2), "\n"
    )
  ))
}
ggplot(df, aes(x=pattern, y=count)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  stat_summary(fun.data = get_box_stats, geom="text", hjust=0.5, vjust=0.8)
```