---
title: "DEgenes_heatmap"
author: "Shuo Shan"
date: "01/05/2023"
output: html_document
params:
  this.gene:
    value: "MAFB"
    label: "gene name"
    input: text
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Load packages  
```{r package, warning=FALSE, include=FALSE}
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)

Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
setwd(Dir)
#save.image(file=paste0(Dir,"/myEnvironment.RData"))
load(paste0(Dir,"/myEnvironment.RData"))
```
# Load file
## loading ATACseq data
```{r cov-atac, eval=FALSE, echo=FALSE, include=FALSE}
atac_coverage_dir="~/Downloads/nl/human/skin/eQTLs/chromatin/RegulatoryElements/method3"
# load raw coverage for enhancer and promoters
count_cRE <- data.table::fread(paste0(atac_coverage_dir,"/merged_coverage_ATACseq_allcts_regulatory_region.bed")) %>%
  dplyr::select(-c(chr,start,end)) %>% column_to_rownames("name")
# CPM
col.sum=apply(count_cRE,2,sum); 
CPM_atac=sweep(count_cRE, 2, col.sum, `/`)*1000000
CPM_atac <- CPM_atac[,c("F25_KRT_PBS","F49_KRT_PBS","F55_KRT_PBS",
                        "F25_MEL_PBS","F49_MEL_PBS","F55_MEL_PBS",
                        "F25_FRB_PBS","F49_FRB_PBS","F55_FRB_PBS",
                        "F25_KRT_IFN","F49_KRT_IFN","F55_KRT_IFN",
                        "F25_MEL_IFN","F49_MEL_IFN","F55_MEL_IFN",
                        "F25_FRB_IFN","F49_FRB_IFN","F55_FRB_IFN")]
rm(col.sum,count_cRE,atac_coverage_dir)
```
## loading H3K27ac ChIPseq data
```{r cov-chip, eval=FALSE, echo=FALSE, include=FALSE}
chip_coverage_dir="~/Downloads/nl/human/skin/eQTLs/chromatin/RegulatoryElements/method3"
# load raw coverage for enhancer and promoters
count_cRE <- data.table::fread(paste0(chip_coverage_dir,"/merged_coverage_ChIPseq_allcts_regulatory_region.bed")) %>%
  dplyr::select(-c(chr,start,end)) %>% column_to_rownames("name")
# CPM
col.sum=apply(count_cRE,2,sum); 
CPM_chip=sweep(count_cRE, 2, col.sum, `/`)*1000000
CPM_chip <- CPM_chip[,c("F25_KRT_PBS","F49_KRT_PBS","F55_KRT_PBS",
                        "F25_MEL_PBS","F49_MEL_PBS","F55_MEL_PBS",
                        "F25_FRB_PBS","F49_FRB_PBS","F55_FRB_PBS",
                        "F25_KRT_IFN","F49_KRT_IFN","F55_KRT_IFN",
                        "F25_MEL_IFN","F49_MEL_IFN","F55_MEL_IFN",
                        "F25_FRB_IFN","F49_FRB_IFN","F55_FRB_IFN")]
rm(col.sum,count_cRE,chip_coverage_dir)
```
## loading RNAseq data and DESeq2 results
```{r load-rna, eval=FALSE, echo=FALSE, include=FALSE}
rna_dir="~/Downloads/nl/human/skin/eQTLs/RNA-Seq"
CPM_rna=data.table::fread(paste0(rna_dir,"/CPM_3cts.txt"))
CPM_rna = CPM_rna %>% column_to_rownames(. , var="V1")
new_colnames=c(colnames(CPM_rna)[grepl("PBS_KRT",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("PBS_MEL",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("PBS_FRB",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("IFN_KRT",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("IFN_MEL",colnames(CPM_rna))],
               colnames(CPM_rna)[grepl("IFN_FRB",colnames(CPM_rna))])
CPM_rna <- CPM_rna[,new_colnames]

dir.krt <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/keratinocytes/pipeline_11192022/analysis"
dir.mel <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/melanocytes/pipeline_12022022/analysis"
dir.frb <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis"
rna.res.krt <- read.table(paste0(dir.krt,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()
rna.res.mel <- read.table(paste0(dir.mel,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()
rna.res.frb <- read.table(paste0(dir.frb,"/DESeq2_results_IFNvsPBS.txt"),header=T,sep=" ") %>% na.omit()
DEG <- unique(c(rna.res.krt[which(rna.res.krt$padj<0.05 & abs(rna.res.krt$log2FoldChange)>3),"gene"],
                rna.res.mel[which(rna.res.mel$padj<0.05 & abs(rna.res.mel$log2FoldChange)>3),"gene"],
                rna.res.frb[which(rna.res.frb$padj<0.05 & abs(rna.res.frb$log2FoldChange)>3),"gene"]))

```
## Link of gene-promoter-enhancer
```{r link, eval=FALSE, echo=FALSE, include=FALSE}
# link between gene-promoter and gene-enhancer
cRE_dict_dir="~/Downloads/nl/human/skin/eQTLs/chromatin/RegulatoryElements/method3"
link_gene_promoter=data.table::fread(paste0(cRE_dict_dir,"/dictionary_gene_promoter_links_allcts.txt")) %>%
  set_colnames(c("gene","promoter","distance","celltype"))
link_gene_enhancer=data.table::fread(paste0(cRE_dict_dir,"/dictionary_gene_enhancer_links_allcts.txt")) %>% 
  set_colnames(c("gene","enhancer","distance","celltype")) 
```

# plot heatmaps
## load rnaseq CPM for all DE genes, and corresponding ATACseq and ChIPseq CPM for their promoter and enhancers
```{r DE_CPM, eval=FALSE, echo=FALSE, include=FALSE}
promoter_df <- link_gene_promoter[which(gene %in% DEG),]
enhancer_df <- link_gene_enhancer[which(gene %in% DEG),]

CPM_rna_DEG <- CPM_rna[DEG,]
CPM_atac_DEG_promoter <- CPM_atac[unique(promoter_df$promoter),]
CPM_atac_DEG_enhancer <- CPM_atac[unique(enhancer_df$enhancer),]
CPM_chip_DEG_promoter <- CPM_chip[unique(promoter_df$promoter),]
CPM_chip_DEG_enhancer <- CPM_chip[unique(enhancer_df$enhancer),]
```

## find the matching enhancers for each gene and plot
```{r plots, eval=FALSE, echo=FALSE, include=FALSE}
plot_list = list()
for (g in DEG.shared$gene) {
  p <- create_compiled_heatmap(g)
  plot_list[[i]] = p
}
pdf(paste0(outDir,"/compiled_heatmaps_gene_CRE_shared_DEG.pdf"),width=15,height=8)
for (i in 1:length(plot_list)) {
  print(plot_list[[i]])
}
dev.off()
```
## useful plotting functions
```{r nicekmeans, echo=FALSE, include=FALSE}
# useful function
niceKmeans <-function (df, k, kmeans.algorithm = "Lloyd", iter.max = 1000, nstart=100) {
  
  result = data.frame();
  source = t(scale(t(df)));
  kmeans = kmeans(source, centers = k, iter.max = iter.max, algorithm=kmeans.algorithm, nstart=nstart);
  clustered = data.frame();
  breaks = c();
  for (i in 1:k) {
    print(sprintf("Doing cluster %i",i))
    cluster = source[kmeans$cluster==i,];
    rows = row.names(cluster);
    clust = heatmap.2(as.matrix(cluster), distfun=function(x) as.dist((1-cor(t(x)))/2))
    cluster.ordered = cluster[clust$rowInd,];
    cluster.ordered.genes = rows[clust$rowInd];
    row.names(cluster.ordered) = cluster.ordered.genes;
    class = data.frame(row.names=cluster.ordered.genes);
    class[,"class"] = i 
    cluster.ordered = cbind(cluster.ordered, class);
    clustered = rbind(clustered, cluster.ordered);
    if(i > 1 & i < k) {
      breaks[i] = as.numeric(breaks[i-1]) + length(rows);
    } else if(i==1) {
      breaks[i] = length(rows);
    }
  }
  
  result <- list();
  result$clustered <- clustered;
  result$breaks <- breaks;
  return(result);
}

```
```{r gene-enhancer-heatmap, eval=FALSE, echo=FALSE, include=FALSE}
create_compiled_heatmap <- function(g) {
g = "IRF7"
this.dict.p <- link_gene_promoter[which(link_gene_promoter$gene == g),c("promoter","distance")] %>% distinct() %>% arrange(gene,distance)
this.dict.e <- link_gene_enhancer[which(link_gene_enhancer$gene == g),c("enhancer","distance")] %>% distinct() %>% arrange(gene,distance)

header <- c("F25_PBS_KRT","F49_PBS_KRT","F55_PBS_KRT",
            "F25_PBS_MEL","F49_PBS_MEL","F55_PBS_MEL",
            "F25_PBS_FRB","F49_PBS_FRB","F55_PBS_FRB",
            "F25_IFN_KRT","F49_IFN_KRT","F55_IFN_KRT",
            "F25_IFN_MEL","F49_IFN_MEL","F55_IFN_MEL",
            "F25_IFN_FRB","F49_IFN_FRB","F55_IFN_FRB")
this.CPM.gene <- log2(CPM_rna[g,header] + 1)
this.CPM.atac.p <- log2(CPM_atac_DEG_promoter[unique(this.dict.p$promoter),] + 1)
this.CPM.atac.e <- log2(CPM_atac_DEG_enhancer[unique(this.dict.e$enhancer),] + 1)
this.CPM.chip.p <- log2(CPM_chip_DEG_promoter[unique(this.dict.p$promoter),] + 1)
this.CPM.chip.e <- log2(CPM_chip_DEG_enhancer[unique(this.dict.e$enhancer),] + 1)

CPM.gene.scaled <- t(scale(t(as.matrix(this.CPM.gene)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.atac.p.scaled <- t(scale(t(as.matrix(this.CPM.atac.p)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.atac.e.scaled <- t(scale(t(as.matrix(this.CPM.atac.e)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.chip.p.scaled <- t(scale(t(as.matrix(this.CPM.chip.p)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.chip.e.scaled <- t(scale(t(as.matrix(this.CPM.chip.e)), center = TRUE, scale = TRUE)) %>% as.data.frame()
#CPM.atac.p.scaled <- CPM.p.scaled %>% rownames_to_column("promoter") %>% left_join( . , unique(this.dict.p[,c("name","dist")]), by="name") %>% mutate(tag=paste0(name,"_",dist)) %>% column_to_rownames("tag") %>% dplyr::select(-c("name", "dist"))
#CPM.atac.e.scaled <- CPM.e.scaled %>% rownames_to_column("name") %>% left_join( . , unique(this.dict.e[,c("name","dist")]), by="name") %>% mutate(tag=paste0(name,"_",dist)) %>% column_to_rownames("tag") %>% dplyr::select(-c("name", "dist"))

# heatmaps for gene expression, promoter and enhancer ATACseq signal.
column_ha = HeatmapAnnotation(
  condition = colnames(CPM.gene.scaled) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  celltype = colnames(CPM.gene.scaled) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  col = list(celltype = c("FRB" = "#97de8c", "MEL" = "#80391e", "KRT" = "#1a8bb5"),
             condition = c("PBS" = "grey", "IFN" = "#Ec5003")),
  show_legend = c(TRUE,TRUE))


heatmap.gene <- Heatmap(CPM.gene.scaled, 
        name = "gene z-log2CPM",
        top_annotation = column_ha,
        column_title = g,
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmap.atac.promoter <- Heatmap(CPM.atac.p.scaled, 
        name = "z-log2CPM",
        column_title = "ATACseq signal promoter ",
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmap.atac.enhancer <- Heatmap(CPM.atac.e.scaled, 
        name = "enhancer z-log2CPM",
        column_title = "ATACseq signal enhancer",
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmap.chip.promoter <- Heatmap(CPM.chip.p.scaled, 
        name = "z-log2CPM",
        column_title = "H3K27ac ChIPseq signal promoter ",
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmap.chip.enhancer <- Heatmap(CPM.chip.e.scaled, 
        name = "enhancer z-log2CPM",
        column_title = "H3K27ac ChIPseq signal enhancer",
        column_split = rep(1:3, each = 6),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(2, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_names_gp = gpar(fontsize = 5),
        row_title_gp = gpar(fontsize = 5, fontface = "bold"))

heatmaps <- heatmap.gene %v% heatmap.atac.promoter %v% heatmap.chip.promoter %v% heatmap.atac.enhancer %v% heatmap.chip.enhancer # vertically concatenating heatmaps
return(heatmaps)
}

```
```{r paired_plot_function, eval=FALSE, echo=FALSE, include=FALSE}
make_paired_plot <- function(this.title,this.df) {
  df = data.frame(CPM=as.numeric(this.df[1,]), sample=colnames(this.df)) %>%
    separate( . , col=sample, into=c("donor","treatment","celltype")) 
  df$treatment <- factor(df$treatment, levels = c("PBS","IFN"))
  
  position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
  line.color = "grey"; linetype = "solid"; palette="bright"
  
  highlightdata1 <- df %>% filter(donor=="F25")
  highlightdata2 <- df %>% filter(donor=="F49")
  highlightdata3 <- df %>% filter(donor=="F55")
  
  p <- ggplot(df, create_aes(list(x = "treatment", y = "CPM"))) +
    geom_exec(geom_line, data = df, group = "donor",
              color = line.color, size = line.size, linetype = linetype,
              position = position) +
    geom_exec(geom_point, data = df, color = "treatment", size = point.size,
              position = position) +
    scale_color_manual(values=c("#999999", "#E69F00")) +
    geom_exec(geom_line, data = highlightdata1, group = "donor",
              color = "blue", size = 0.7, linetype = linetype,
              position = position) +
    geom_exec(geom_line, data = highlightdata2, group = "donor",
              color = "brown", size = 0.7, linetype = linetype,
              position = position) +
    geom_exec(geom_line, data = highlightdata3, group = "donor",
              color = "black", size = 0.7, linetype = linetype,
              position = position) +
    ggtitle(this.title) +
    theme(legend.position="none") +
    theme(plot.title = element_text(size = 5, face = "bold")) +
    facet_wrap(~ celltype)
  p
  return(p)
}
```
```{r gene-enhancer-paired-plot, eval=FALSE, echo=FALSE, include=FALSE}
create_compiled_paired_plots <- function(g) {
this.dict.p <- link_gene_promoter[which(link_gene_promoter$gene == g),] %>% arrange(gene,dist)
this.dict.e <- link_gene_enhancer[which(link_gene_enhancer$gene == g),] %>% arrange(gene,dist)

header <- c("F25_FRB_PBS","F25_FRB_IFN","F49_FRB_PBS","F49_FRB_IFN","F55_FRB_PBS","F55_FRB_IFN",
            "F25_MEL_PBS","F25_MEL_IFN","F49_MEL_PBS","F49_MEL_IFN","F55_MEL_PBS","F55_MEL_IFN",
            "F25_KRT_PBS","F25_KRT_IFN","F49_KRT_PBS","F49_KRT_IFN","F55_KRT_PBS","F55_KRT_IFN")

this.CPM.gene <- CPM_rna[g,header] 
this.CPM.p <- CPM_promoter[unique(this.dict.p$name),header] %>% rownames_to_column("name") %>% left_join( . , unique(this.dict.p[,c("name","dist")]), by="name") %>% mutate(tag=paste0(name,"_",dist)) %>% column_to_rownames("tag") %>% dplyr::select(-c("name", "dist"))
this.CPM.e <- CPM_enhancer[unique(this.dict.e$name),header] %>% rownames_to_column("name") %>% left_join( . , unique(this.dict.e[,c("name","dist")]), by="name") %>% mutate(tag=paste0(name,"_",dist)) %>% column_to_rownames("tag") %>% dplyr::select(-c("name", "dist"))

this.logCPM.gene <- log2(CPM_rna[g,header] + 1)
this.logCPM.p <- log2(CPM_promoter[unique(this.dict.p$name),header] + 1)
this.logCPM.e <- log2(CPM_enhancer[unique(this.dict.e$name),header] + 1)

CPM.gene.scaled <- t(scale(t(as.matrix(this.logCPM.gene)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.p.scaled <- t(scale(t(as.matrix(this.logCPM.p)), center = TRUE, scale = TRUE)) %>% as.data.frame()
CPM.e.scaled <- t(scale(t(as.matrix(this.logCPM.e)), center = TRUE, scale = TRUE)) %>% as.data.frame()

# make paired plots
this.CPM.all <- rbind(this.CPM.gene, this.CPM.p, this.CPM.e)

paired_plot_gene <- make_paired_plot(g,this.CPM.gene)

plot_list_promoter = list()
for (i in 1:nrow(this.CPM.p)) {
  p <- make_paired_plot(rownames(this.CPM.p)[i],this.CPM.p[i,])
  plot_list_promoter[[i]] = p
}
paired_plot_promoter <- do.call("grid.arrange", c(plot_list_promoter, ncol=1, nrow=nrow(this.CPM.p)))

plot_list_enhancer = list()
for (i in 1:nrow(this.CPM.p)) {
  p <- make_paired_plot(rownames(this.CPM.e)[i],this.CPM.e[i,])
  plot_list_enhancer[[i]] = p
}
paired_plot_enhancer <- do.call("grid.arrange", c(plot_list_enhancer, ncol=1, nrow=nrow(this.CPM.p)))

p <- plot_grid(paired_plot_gene, paired_plot_promoter, paired_plot_enhancer, ncol=3, nrow=1)
return(p)
}
```

## Compile a list of DE genes
```{r DEG, eval=FALSE, echo=FALSE, include=FALSE}
DEG <- unique(c(res.krt[which(res.krt$padj<0.05 & abs(res.krt$log2FoldChange)>1.5),"gene"],
                res.mel[which(res.mel$padj<0.05 & abs(res.mel$log2FoldChange)>1.5),"gene"],
                res.frb[which(res.frb$padj<0.05 & abs(res.frb$log2FoldChange)>1.5),"gene"]))
```
## select gene name
```{r gene-select}
g = params$this.gene
```
## Heatmap of RNAseq
```{r RNA-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
# 1. center and scale data
logCPM <- log2(CPM_rna_DEG + 1)

# 2. hierarchically cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 7)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,km_res$breaks[5]-km_res$breaks[4]),
                        rep(6,km_res$breaks[6]-km_res$breaks[5]),
                        rep(7,nrow(km_res$clustered)-km_res$breaks[6]))

km_res.rna <- km_res
split_instructions.rna <- split_instructions
```
```{r RNA-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2, include=FALSE}
# 4. load genes of interest
gene <- rownames(km_res.rna$clustered)[which(split_instructions.rna==7)]
draw_rna_heatmap <- function(g) {
  zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_rna_DEG)]
  zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% gene)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
    show_legend = c(TRUE,TRUE))
  
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.rna <- Heatmap(zlogCPM.heatmap, 
                         name = "z-score log2CPM",
                         top_annotation = column_ha,
                         right_annotation = row_ha,
                         column_title = "DE genes in 3 celltypes, control vs IGNg stimulated, KEGG genes labeled",
                         cluster_columns = FALSE,
                         cluster_rows = FALSE,
                         column_split = c(rep(1,length(colnames(CPM_rna_DEG)[grepl("PBS_KRT",colnames(CPM_rna_DEG))])), 
                                          rep(2,length(colnames(CPM_rna_DEG)[grepl("PBS_MEL",colnames(CPM_rna_DEG))])), 
                                          rep(3,length(colnames(CPM_rna_DEG)[grepl("PBS_FRB",colnames(CPM_rna_DEG))])), 
                                          rep(4,length(colnames(CPM_rna_DEG)[grepl("IFN_KRT",colnames(CPM_rna_DEG))])), 
                                          rep(5,length(colnames(CPM_rna_DEG)[grepl("IFN_MEL",colnames(CPM_rna_DEG))])),
                                          rep(6,length(colnames(CPM_rna_DEG)[grepl("IFN_FRB",colnames(CPM_rna_DEG))]))),
                         row_split = split_instructions.rna, 
                         show_row_names = FALSE,
                         row_title_rot = 0,
                         column_names_rot = 45,
                         row_gap = unit(2, "mm"),
                         border = FALSE,
                         column_names_gp = gpar(fontsize = 5),
                         column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                         row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.rna)
}
# clean-up
rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_gene_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
```

## Heatmaps of ATACseq
```{r ATAC-promoter-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
# 1. center and scale data
logCPM <- log2(CPM_atac_DEG_promoter + 1)

# 2. hierarchically cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#lot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.atac.promoter <- km_res
split_instructions.atac.promoter <- split_instructions
```
```{r ATAC-promoter-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2}
draw_atac_promoter_heatmap <- function(g) {# 4. load peaks of interest
  peaks <- link_gene_promoter[which(link_gene_promoter$gene==g),] %>% dplyr::pull(promoter) %>% unique
  zlogCPM.heatmap <- km_res.atac.promoter$clustered[,1:ncol(CPM_atac_DEG_promoter)]
  zlogCPM.heatmap_with_peaks_kegg <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% peaks)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_peaks_kegg))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.atac.promoter <- Heatmap(zlogCPM.heatmap, 
                                   name = "z-score log2CPM",
                                   top_annotation = column_ha,
                                   right_annotation = row_ha,
                                   column_title = "ATACseq heatmap of promoters",
                                   cluster_columns = FALSE,
                                   cluster_rows = FALSE,
                                   column_split = c(rep(1,9), rep(2,9)),
                                   row_split = split_instructions.atac.promoter, 
                                   show_row_names = FALSE,
                                   row_title_rot = 0,
                                   column_names_rot = 45,
                                   row_gap = unit(2, "mm"),
                                   border = FALSE,
                                   column_names_gp = gpar(fontsize = 5),
                                   column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                   row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.atac.promoter)
  
  # clean-up
  rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_peaks_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
}

```
```{r ATAC-enhancer-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
# 1. center and scale data
zlogCPM <- log2(CPM_atac_DEG_enhancer + 1)

# 2. hierarchically cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.atac.enhancer <- km_res
split_instructions.atac.enhancer <- split_instructions
```
```{r ATAC-enhancer-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2}
draw_atac_enhancer_heatmap <- function(g){
  # 4. load peaks of interest
  peaks <- link_gene_enhancer[which(link_gene_enhancer$gene==g),] %>% pull(enhancer) %>% unique
  zlogCPM.heatmap <- km_res.atac.enhancer$clustered[,1:ncol(CPM_atac_DEG_enhancer)]
  zlogCPM.heatmap_with_peaks_kegg <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% peaks)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_peaks_kegg))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.atac.enhancer <- Heatmap(zlogCPM.heatmap, 
                                   name = "z-score log2CPM",
                                   top_annotation = column_ha,
                                   right_annotation = row_ha,
                                   column_title = "ATACseq heatmap of enhancers",
                                   cluster_columns = FALSE,
                                   cluster_rows = FALSE,
                                   column_split = c(rep(1,9), rep(2,9)),
                                   row_split = split_instructions.atac.enhancer, 
                                   show_row_names = FALSE,
                                   row_title_rot = 0,
                                   column_names_rot = 45,
                                   row_gap = unit(2, "mm"),
                                   border = FALSE,
                                   column_names_gp = gpar(fontsize = 5),
                                   column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                   row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.atac.enhancer)
  
  # clean-up
  rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_peaks_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
}
```

## Heatmaps of ChIPseq
```{r ChIP-promoter-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
# 1. center and scale data
logCPM <- log2(CPM_chip_DEG_promoter + 1)

# 2. hierarchially cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.chip.promoter <- km_res
split_instructions.chip.promoter <- split_instructions
```
```{r ChIP-promoter-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2}
draw_chip_promoter_heatmap <- function(g){
  # 4. load peaks of interest
  peaks <- link_gene_promoter[which(link_gene_promoter$gene==g),] %>% pull(promoter) %>% unique
  zlogCPM.heatmap <- km_res.chip.promoter$clustered[,1:ncol(CPM_chip_DEG_promoter)]
  zlogCPM.heatmap_with_peaks_kegg <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% peaks)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_peaks_kegg))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.chip.promoter <- Heatmap(zlogCPM.heatmap, 
                                   name = "z-score log2CPM",
                                   top_annotation = column_ha,
                                   right_annotation = row_ha,
                                   column_title = "H3K27ac ChIPseq heatmap of promoters",
                                   cluster_columns = FALSE,
                                   cluster_rows = FALSE,
                                   column_split = c(rep(1,9), rep(2,9)),
                                   row_split = split_instructions.chip.promoter, 
                                   show_row_names = FALSE,
                                   row_title_rot = 0,
                                   column_names_rot = 45,
                                   row_gap = unit(2, "mm"),
                                   border = FALSE,
                                   column_names_gp = gpar(fontsize = 5),
                                   column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                   row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.chip.promoter)
  
  # clean-up
  rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_peaks_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
}
```
```{r chip-enhancer-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-1, eval=FALSE, include=FALSE}
# 1. center and scale data
logCPM <- log2(CPM_chip_DEG_enhancer + 1)

# 2. hierarchially cluster K-means clusters
# perform k-means-clustering to determine k_row.
#df <- logCPM
#wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
#for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
#plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# split rows by determined number of clusters
km_res <- niceKmeans(logCPM, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.chip.enhancer <- km_res
split_instructions.chip.enhancer <- split_instructions
```
```{r chip-enhancer-heatmap-DEgenes-CPM-kegg-rowClustered-kmm-hclust-2}
draw_chip_enhancer_heatmap <- function(g){
  # 4. load peaks of interest
  peaks <- link_gene_enhancer[which(link_gene_enhancer$gene==g),] %>% pull(enhancer) %>% unique
  zlogCPM.heatmap <- km_res.chip.enhancer$clustered[,1:ncol(CPM_chip_DEG_enhancer)]
  zlogCPM.heatmap_with_peaks_kegg <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% peaks)
  row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_peaks_kegg))
  row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]
  
  # 5. plot heatmap
  column_ha = HeatmapAnnotation(
    condition = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
    celltype = colnames(zlogCPM.heatmap) %>% 
      lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
    col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
               celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))
  row_ha = rowAnnotation(
    foo = anno_mark(at = row_anno_index, 
                    labels = row_anno_label,
                    which="row",
                    labels_gp = gpar(fontsize=3),
                    padding=unit(0.5,"mm")))
  #pdf(paste0(Dir,"/heatmap_RNAseq_DEgenes_7kmm-hc.pdf"),width=24,height=30)
  heatmap.chip.enhancer <- Heatmap(zlogCPM.heatmap, 
                                   name = "z-score log2CPM",
                                   top_annotation = column_ha,
                                   right_annotation = row_ha,
                                   column_title = "chipseq heatmap of enhancers",
                                   cluster_columns = FALSE,
                                   cluster_rows = FALSE,
                                   column_split = c(rep(1,9), rep(2,9)),
                                   row_split = split_instructions.chip.enhancer, 
                                   show_row_names = FALSE,
                                   row_title_rot = 0,
                                   column_names_rot = 45,
                                   row_gap = unit(2, "mm"),
                                   border = FALSE,
                                   column_names_gp = gpar(fontsize = 5),
                                   column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                   row_title_gp = gpar(fontsize = 5, fontface = "bold"))
  #dev.off()
  return(heatmap.chip.enhancer)
  
  # clean-up
  rm(logCPM, df, wss, km_res, split_instructions, zlogCPM.heatmap, zlogCPM.heatmap_with_peaks_kegg, row_anno_index, row_anno_label, column_ha, row_ha)
}
```

## PLOT
```{r plot}
p1 <- draw_rna_heatmap(gene)
p2 <- draw_atac_promoter_heatmap(gene)
p3 <- draw_atac_enhancer_heatmap(gene)
p4 <- draw_chip_promoter_heatmap(gene)
p5 <- draw_chip_enhancer_heatmap(gene)

p.promoter <- p2 %v% p4
p.enhancer <- p3 %v% p5

grid.newpage()
pushViewport(viewport(x = 0, width = 0.5, just = "left"))
grid.rect(gp = gpar(fill = "#FF000020"))
draw(p1, newpage = FALSE)
popViewport()

pushViewport(viewport(x = 0.5, y = 1, width = 0.5, height = 0.5, just = c("left", "top")))
grid.rect(gp = gpar(fill = "#00FF0020"))
draw(p.promoter, newpage = FALSE)
popViewport()

pushViewport(viewport(x = 0.5, y = 0.5, width = 0.5, height = 0.5, just = c("left", "top")))
grid.rect(gp = gpar(fill = "#0000FF20"))
draw(p.enhancer, newpage = FALSE)
popViewport()
```
# For genes that are up/down-regulated in all 3cts, are their enhancers shared across celltypes?
## For commonly regulated genes, check if their enhancers are shared across celltypes in ATACseq
```{r gene-enhancer-sharing-atac}
# to view which genes belong to clsuter of (commonly up-regulated genes)
commonly_induced_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==6)]
gene <- commonly_induced_genes[2]
make_paired_plot(gene,CPM_rna_DEG[gene,])

# get the enhancers of these genes
enhancer_of_gene_group1 <- link_gene_enhancer %>% dplyr::filter(gene %in% commonly_induced_genes) %>% arrange(gene) %>% pull(enhancer) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_atac_DEG_enhancer[enhancer_of_gene_group1,] + 1))))

# # cluster the enhancers
# km_res <- niceKmeans(zlogCPM, 7)
# split_instructions <- c(rep(1,km_res$breaks[1]), 
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,km_res$breaks[5]-km_res$breaks[4]),
#                         rep(6,km_res$breaks[6]-km_res$breaks[5]),
#                         rep(7,nrow(km_res$clustered)-km_res$breaks[6]))
# km_res.atac.enhancer <- km_res
# split_instructions.atac.enhancer <- split_instructions

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.atac.enhancer <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ATACseq heatmap of enhancers of commonly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.atac.enhancer); enhancer_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# pick the commonly open enhancers of commonly induced genes
enhancers_shared <- rownames(zlogCPM)[sort(row_order(ht)$`5`)]
enhancers_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`2`,row_order(ht)$`4`,row_order(ht)$`7`))]
enhancers_paired <- rownames(zlogCPM)[sort(c(row_order(ht)$`1`,row_order(ht)$`3`,row_order(ht)$`6`))]

df1 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_shared) %>% mutate(pattern="shared")
df2 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_cts) %>% mutate(pattern="cts")
df3 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_paired) %>% mutate(pattern="paired")
df <- rbind(df1,df2,df3)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)
```
```{r gene-promoter-sharing-atac}
# to view which genes belong to clsuter of (commonly up-regulated genes)
commonly_induced_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==6)]

# get the enhancers of these genes
promoter_of_gene_group1 <- link_gene_promoter %>% dplyr::filter(gene %in% commonly_induced_genes) %>% arrange(gene) %>% pull(promoter) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_atac_DEG_promoter[promoter_of_gene_group1,] + 1))))

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.atac.promoter <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ATACseq heatmap of promoters of commonly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.atac.promoter); promoter_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# plot the distance of common promoters and cts promoters
# pick the commonly open enhancers of commonly induced genes
promoters_shared <- rownames(zlogCPM)[sort(c(row_order(ht)$`3`))]
promoters_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`4`,row_order(ht)$`6`))]

df1 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_shared) %>% mutate(pattern="shared")
df2 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_cts) %>% mutate(pattern="cts")
df <- rbind(df1,df2)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)

# Could the cell-type specific promoters come from genes with multiple promoters?
genes_of_df1 <- link_gene_promoter %>% dplyr::filter(gene %in% df1$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="common")
genes_of_df2 <- link_gene_promoter %>% dplyr::filter(gene %in% df2$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="cts")

df <- rbind(genes_of_df1, genes_of_df2)

get_box_stats <- function(y, upper_limit = max(df$count) * 1.15) {
  return(data.frame(
    y = 0.95 * upper_limit,
    label = paste(
      "Count =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n",
      "Median =", round(median(y), 2), "\n"
    )
  ))
}
ggplot(df, aes(x=pattern, y=count)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  stat_summary(fun.data = get_box_stats, geom="text", hjust=0.5, vjust=0.8)
```
## For commonly regulated genes, check if their enhancers are shared across celltypes in ChIPseq
```{r gene-enhancer-sharing-chip}
# to view which genes belong to clsuter of (commonly up-regulated genes)
commonly_induced_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==6)]

# get the enhancers of these genes
enhancer_of_gene_group1 <- link_gene_enhancer %>% dplyr::filter(gene %in% commonly_induced_genes) %>% arrange(gene) %>% pull(enhancer) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_chip_DEG_enhancer[enhancer_of_gene_group1,] + 1))))

# # cluster the enhancers
# km_res <- niceKmeans(zlogCPM, 7)
# split_instructions <- c(rep(1,km_res$breaks[1]), 
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,km_res$breaks[5]-km_res$breaks[4]),
#                         rep(6,km_res$breaks[6]-km_res$breaks[5]),
#                         rep(7,nrow(km_res$clustered)-km_res$breaks[6]))
# km_res.chip.enhancer <- km_res
# split_instructions.chip.enhancer <- split_instructions

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.chip.enhancer <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ChIPseq heatmap of enhancers of commonly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.chip.enhancer); enhancer_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# pick the commonly open enhancers of commonly induced genes
enhancers_shared <- rownames(zlogCPM)[sort(row_order(ht)$`5`)]
enhancers_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`2`,row_order(ht)$`4`,row_order(ht)$`7`))]
enhancers_paired <- rownames(zlogCPM)[sort(c(row_order(ht)$`1`,row_order(ht)$`3`,row_order(ht)$`6`))]

df1 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_shared) %>% mutate(pattern="shared")
df2 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_cts) %>% mutate(pattern="cts")
df3 <- link_gene_enhancer %>% dplyr::filter(enhancer %in% enhancers_paired) %>% mutate(pattern="paired")
df <- rbind(df1,df2,df3)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)
```
```{r gene-promoter-sharing-chip}
# to view which genes belong to clsuter of (commonly up-regulated genes)
commonly_induced_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==6)]

# get the enhancers of these genes
promoter_of_gene_group1 <- link_gene_promoter %>% dplyr::filter(gene %in% commonly_induced_genes) %>% arrange(gene) %>% pull(promoter) %>% unique

# prepare data to plot the ChIPseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_chip_DEG_promoter[promoter_of_gene_group1,] + 1))))

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.chip.promoter <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ChIPseq heatmap of promoters of commonly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.chip.promoter); promoter_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# plot the distance of common promoters and cts promoters
# pick the commonly open enhancers of commonly induced genes
promoters_shared <- rownames(zlogCPM)[sort(c(row_order(ht)$`3`))]
promoters_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`4`,row_order(ht)$`6`))]

df1 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_shared) %>% mutate(pattern="shared")
df2 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_cts) %>% mutate(pattern="cts")
df <- rbind(df1,df2)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)

# Could the cell-type specific promoters come from genes with multiple promoters?
genes_of_df1 <- link_gene_promoter %>% dplyr::filter(gene %in% df1$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="common")
genes_of_df2 <- link_gene_promoter %>% dplyr::filter(gene %in% df2$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="cts")

df <- rbind(genes_of_df1, genes_of_df2)

get_box_stats <- function(y, upper_limit = max(df$count) * 1.15) {
  return(data.frame(
    y = 0.95 * upper_limit,
    label = paste(
      "Count =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n",
      "Median =", round(median(y), 2), "\n"
    )
  ))
}
ggplot(df, aes(x=pattern, y=count)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  stat_summary(fun.data = get_box_stats, geom="text", hjust=0.5, vjust=0.8)
```
## For cell-type-specificly expressed genes, check if their enhancers are unique across celltypes in ATACseq
```{r cts-gene-enhancer-sharing-atac}
# to view which genes belong to clsuter of (cts-regulated genes)
these_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4 | split_instructions.rna==6)]
gene <- these_genes[200]
make_paired_plot(gene,CPM_rna_DEG[gene,])

# get the enhancers of these genes
enhancer_of_gene_group1 <- link_gene_enhancer %>% dplyr::filter(gene %in% these_genes) %>% arrange(gene) %>% pull(enhancer) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_atac_DEG_enhancer[enhancer_of_gene_group1,] + 1))))

# # cluster the enhancers
# km_res <- niceKmeans(zlogCPM, 7)
# split_instructions <- c(rep(1,km_res$breaks[1]), 
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,km_res$breaks[5]-km_res$breaks[4]),
#                         rep(6,km_res$breaks[6]-km_res$breaks[5]),
#                         rep(7,nrow(km_res$clustered)-km_res$breaks[6]))
# km_res.atac.enhancer <- km_res
# split_instructions.atac.enhancer <- split_instructions

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.atac.enhancer <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ATACseq heatmap of enhancers of cell-type-spcificly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.atac.enhancer); enhancer_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.
```
```{r gene-promoter-sharing-atac}
# to view which genes belong to clsuter of (commonly up-regulated genes)
these_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4 | split_instructions.rna==6)]

# get the enhancers of these genes
promoter_of_gene_group1 <- link_gene_promoter %>% dplyr::filter(gene %in% these_genes) %>% arrange(gene) %>% pull(promoter) %>% unique

# prepare data to plot the atacseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_atac_DEG_promoter[promoter_of_gene_group1,] + 1))))

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.atac.promoter <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "atacseq heatmap of promoters of cell-type-specificly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.atac.promoter); promoter_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# plot the distance of common promoters and cts promoters
# pick the commonly open enhancers of commonly induced genes
promoters_shared <- rownames(zlogCPM)[sort(c(row_order(ht)$`3`))]
promoters_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`4`,row_order(ht)$`6`))]

df1 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_shared) %>% mutate(pattern="shared")
df2 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_cts) %>% mutate(pattern="cts")
df <- rbind(df1,df2)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)

# Could the cell-type specific promoters come from genes with multiple promoters?
genes_of_df1 <- link_gene_promoter %>% dplyr::filter(gene %in% df1$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="common")
genes_of_df2 <- link_gene_promoter %>% dplyr::filter(gene %in% df2$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="cts")

df <- rbind(genes_of_df1, genes_of_df2)

get_box_stats <- function(y, upper_limit = max(df$count) * 1.15) {
  return(data.frame(
    y = 0.95 * upper_limit,
    label = paste(
      "Count =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n",
      "Median =", round(median(y), 2), "\n"
    )
  ))
}
ggplot(df, aes(x=pattern, y=count)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  stat_summary(fun.data = get_box_stats, geom="text", hjust=0.5, vjust=0.8)
```
## For cell-type-specificly expressed genes, check if their enhancers are unique across celltypes in ChIPseq
```{r cts-gene-enhancer-sharing-chip}
# to view which genes belong to clsuter of (cts-regulated genes)
these_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4 | split_instructions.rna==6)]

# get the enhancers of these genes
enhancer_of_gene_group1 <- link_gene_enhancer %>% dplyr::filter(gene %in% these_genes) %>% arrange(gene) %>% pull(enhancer) %>% unique

# prepare data to plot the ATACseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_chip_DEG_enhancer[enhancer_of_gene_group1,] + 1))))

# # cluster the enhancers
# km_res <- niceKmeans(zlogCPM, 7)
# split_instructions <- c(rep(1,km_res$breaks[1]), 
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,km_res$breaks[5]-km_res$breaks[4]),
#                         rep(6,km_res$breaks[6]-km_res$breaks[5]),
#                         rep(7,nrow(km_res$clustered)-km_res$breaks[6]))
# km_res.chip.enhancer <- km_res
# split_instructions.chip.enhancer <- split_instructions

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.chip.enhancer <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ChIPseq heatmap of enhancers of cell-type-spcificly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.chip.enhancer); enhancer_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.
```
```{r gene-promoter-sharing-chip}
# to view which genes belong to clsuter of (commonly up-regulated genes)
these_genes <- rownames(km_res.rna$clustered)[which(split_instructions.rna==4 | split_instructions.rna==6)]

# get the enhancers of these genes
promoter_of_gene_group1 <- link_gene_promoter %>% dplyr::filter(gene %in% these_genes) %>% arrange(gene) %>% pull(promoter) %>% unique

# prepare data to plot the ChIPseq heatmap of the enhancers
zlogCPM <- t(scale(t(log2(CPM_chip_DEG_promoter[promoter_of_gene_group1,] + 1))))

# plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = colnames(zlogCPM) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")))

heatmap.chip.promoter <- Heatmap(zlogCPM, 
                                 name = "z-score log2CPM",
                                 row_km = 7,
                                 top_annotation = column_ha,
                                 column_title = "ChIPseq heatmap of promoters of cell-type-specificly induced genes",
                                 cluster_columns = FALSE,
                                 cluster_rows = TRUE,
                                 column_split = c(rep(1,9), rep(2,9)),
                                 show_row_names = FALSE,
                                 row_title_rot = 0,
                                 column_names_rot = 45,
                                 row_gap = unit(2, "mm"),
                                 border = FALSE,
                                 column_names_gp = gpar(fontsize = 5),
                                 column_title_gp = gpar(fontsize = 5, fontface = "bold"),
                                 row_title_gp = gpar(fontsize = 5, fontface = "bold"))
ht = draw(heatmap.chip.promoter); promoter_clusters = row_order(ht) # this ensures the cluster info doesn't change every time I execute this line of code.

# plot the distance of common promoters and cts promoters
# pick the commonly open enhancers of commonly induced genes
promoters_shared <- rownames(zlogCPM)[sort(c(row_order(ht)$`3`))]
promoters_cts <- rownames(zlogCPM)[sort(c(row_order(ht)$`4`,row_order(ht)$`6`))]

df1 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_shared) %>% mutate(pattern="shared")
df2 <- link_gene_promoter %>% dplyr::filter(promoter %in% promoters_cts) %>% mutate(pattern="cts")
df <- rbind(df1,df2)

ggplot(df , aes(x=pattern, y=distance)) +
  geom_violin() +
  geom_boxplot(width=0.1)

# Could the cell-type specific promoters come from genes with multiple promoters?
genes_of_df1 <- link_gene_promoter %>% dplyr::filter(gene %in% df1$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="common")
genes_of_df2 <- link_gene_promoter %>% dplyr::filter(gene %in% df2$gene) %>% arrange(gene,promoter) %>% 
  dplyr::filter(celltype=="KRT") %>% dplyr::select(c("gene","promoter")) %>% 
  group_by(gene) %>% mutate(count = n()) %>% dplyr::select(c("gene","count")) %>%
  distinct(gene,.keep_all=TRUE) %>% mutate(pattern="cts")

df <- rbind(genes_of_df1, genes_of_df2)

get_box_stats <- function(y, upper_limit = max(df$count) * 1.15) {
  return(data.frame(
    y = 0.95 * upper_limit,
    label = paste(
      "Count =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n",
      "Median =", round(median(y), 2), "\n"
    )
  ))
}
ggplot(df, aes(x=pattern, y=count)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  stat_summary(fun.data = get_box_stats, geom="text", hjust=0.5, vjust=0.8)
```