---
title: "commonly-induced-vs-cellTypeSpecicallyInduced-promoters"
author: "Crystal Shan"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages  
```{r package, warning=FALSE, include=FALSE}
library(dplyr)
library(magrittr)
library(tidyverse)
library(gplots)
library(ggfortify)
library(gridExtra)
library(grid)
library(cowplot)
library(ggpubr)
library(DESeq2)
library(ComplexHeatmap)

Dir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis"
setwd(Dir)

outDir="~/Downloads/nl/human/skin/eQTLs/integrative_analysis/XSTREME/commonly_induced_genes_promoter"

load(paste0(Dir,"/heatmap_RNA_round2/myEnvironment_heatmap_RNAseq_DEgenes_14kmm-hc_04182023_padj0.05_log2FC1.5_avgCPM10.RData"))
res.atac.dds.frb = readRDS("~/Downloads/nl/human/skin/eQTLs/ATACseq_DolphinNext/DESeq2/DESeq2_results_ATACseq_peaks_1kb_FRB.Rdata")
res.atac.dds.krt = readRDS("~/Downloads/nl/human/skin/eQTLs/ATACseq_DolphinNext/DESeq2/DESeq2_results_ATACseq_peaks_1kb_KRT.Rdata")
res.atac.dds.mel = readRDS("~/Downloads/nl/human/skin/eQTLs/ATACseq_DolphinNext/DESeq2/DESeq2_results_ATACseq_peaks_1kb_MEL.Rdata")
res.chip.dds.frb = readRDS("~/Downloads/nl/human/skin/eQTLs/ChIP-seq/DESeq2/DESeq2_results_H3K27acChIPseq_peaks_1kb_FRB.Rdata")
res.chip.dds.krt = readRDS("~/Downloads/nl/human/skin/eQTLs/ChIP-seq/DESeq2/DESeq2_results_H3K27acChIPseq_peaks_1kb_KRT.Rdata")
res.chip.dds.mel = readRDS("~/Downloads/nl/human/skin/eQTLs/ChIP-seq/DESeq2/DESeq2_results_H3K27acChIPseq_peaks_1kb_MEL.Rdata")
link_gene_promoter = readRDS(paste0(Dir,"/heatmap_RNA_round2/link_gene_promoter_within_300kbp_of_TSS_allcts.Rdata"))
link_gene_enhancer = readRDS(paste0(Dir,"/heatmap_RNA_round2/link_gene_enhancer_within_300kbp_of_TSS_allcts.Rdata"))
FRB_cRE_dynamics = readRDS(paste0(Dir,"/heatmap_RNA_round2/FRB_gene_and_cRE_RNA_ATAC_and_H3K27ac_dynamics.Rdata"))
MEL_cRE_dynamics = readRDS(paste0(Dir,"/heatmap_RNA_round2/MEL_gene_and_cRE_RNA_ATAC_and_H3K27ac_dynamics.Rdata"))
KRT_cRE_dynamics = readRDS(paste0(Dir,"/heatmap_RNA_round2/KRT_gene_and_cRE_RNA_ATAC_and_H3K27ac_dynamics.Rdata"))
FRB_cRE_dynamics = readRDS(paste0(Dir,"/heatmap_RNA_round2/FRB_gene_and_cRE_RNA_ATAC_and_H3K27ac_dynamics.Rdata"))
bed <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/chromatin/RegulatoryElements/method3/regulatory_regions.bed")
colnames(bed) <- c("chr","start","end","region")
df3 = readRDS(paste0(Dir,"/heatmap_RNA_round2/enhancer_count_by_gene.rds"))
```

## plot heatmap to check which clusters are commonly induced promoters
```{r plot}
# km_res.rna.new variable is loaded in line28 from script heatmap_RNA_round2/3cts_interactive_heatmap_RNAseq_ATACseq_ChIPseq_version2.Rmd
split_instructions.new <- km_res.rna.new$newclass

# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap.new <- km_res.rna.new[,1:ncol(CPM_rna_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap.new %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap.new) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap.new)[row_anno_index]

# 5. plot heatmap
column_ha = HeatmapAnnotation(
  condition = colnames(zlogCPM.heatmap.new) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  celltype = colnames(zlogCPM.heatmap.new) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  col = list(condition = c("PBS" = "black", "IFN" = "#FA8072"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(outDir,"/heatmap_reordered_RNAseq_DEgenes_14kmm-hc_padj0.05_log2FC1.5_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap.new, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, control vs IGNg stimulated, known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(colnames(CPM_rna_DEG)[grepl("PBS_KRT",colnames(CPM_rna_DEG))])), 
                                             rep(2,length(colnames(CPM_rna_DEG)[grepl("PBS_MEL",colnames(CPM_rna_DEG))])), 
                                             rep(3,length(colnames(CPM_rna_DEG)[grepl("PBS_FRB",colnames(CPM_rna_DEG))])), 
                                             rep(4,length(colnames(CPM_rna_DEG)[grepl("IFN_KRT",colnames(CPM_rna_DEG))])), 
                                             rep(5,length(colnames(CPM_rna_DEG)[grepl("IFN_MEL",colnames(CPM_rna_DEG))])),
                                             rep(6,length(colnames(CPM_rna_DEG)[grepl("IFN_FRB",colnames(CPM_rna_DEG))]))),
                            row_split = split_instructions.new, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()


```

# Cluster 1-3 are genes with low expression in PBS, but are highly induced upon IFNg in all three cell types.
## save promoters of commonly induced promoters
```{r fetch}
# get gene list
gene_list <- c()
for (i in c(1,2,3)){
 this_cluster_genes = rownames(km_res.rna.new)[which(km_res.rna.new$newclass==i)]
 gene_list <- c(gene_list, this_cluster_genes)
}

# get promoter list
promoter_df <- link_gene_promoter %>% dplyr::filter(gene %in% gene_list)

# only pick promoters that exist in all three cell types
promoter_list <- table( promoter_df[,c("region")]) %>% as.data.frame() %>% dplyr::filter(Freq==3) %>% pull(region) %>% as.character()
# 187 out of 263 genes have at least 1 promoter that's active in all three celltypes
# link_gene_promoter %>% dplyr::filter(gene %in% gene_list & region %in% promoter_list) %>% pull(gene) %>% unique() %>% length()

# save promoter list
write(promoter_list, file=paste0(outDir,"/commonly_induced_promoters.txt"))

# fetch bedfile
promoter_bed <- bed %>% dplyr::filter(region %in% promoter_list)
data.table::fwrite(promoter_bed, file=paste0(outDir,"/commonly_induced_promoters.bed"),quote=F, sep="\t",col.names=F, row.names=F)
```
## analysis5: targets being the commonly induced genes’ promoters that gain H3K27ac signal
```{r target_analysis5}
# get gene list
gene_list <- c()
for (i in c(1,2,3)){
 this_cluster_genes = rownames(km_res.rna.new)[which(km_res.rna.new$newclass==i)]
 gene_list <- c(gene_list, this_cluster_genes)
}

# get promoter list
promoter_df <- link_gene_promoter %>% dplyr::filter(gene %in% gene_list)

# only pick promoters that gain H3K27ac signal in at least 1 celltype. 517 regions.
MEL_cRE_dynamics[which(MEL_cRE_dynamics$region %in% promoter_df$region),]$DK27ac

temp1 <- MEL_cRE_dynamics %>% dplyr::filter(region %in% promoter_df$region & DK27ac=="gainH3K27ac") %>% pull(region)
temp2 <- KRT_cRE_dynamics %>% dplyr::filter(region %in% promoter_df$region & DK27ac=="gainH3K27ac") %>% pull(region)
temp3 <- FRB_cRE_dynamics %>% dplyr::filter(region %in% promoter_df$region & DK27ac=="gainH3K27ac") %>% pull(region)

promoter_df <- link_gene_promoter %>% 
  dplyr::filter(gene %in% gene_list) %>%
  dplyr::filter(region %in% unique(c(temp1, temp2, temp3)))

# save promoter list
write(promoter_list, file=paste0(outDir,"/commonly_induced_promoters_that_gained_H3K27ac.txt"))

# fetch bedfile
promoter_bed <- bed %>% dplyr::filter(region %in% promoter_list)
data.table::fwrite(promoter_bed, file=paste0(outDir,"/commonly_induced_promoters_that_gained_H3K27ac.bed"),quote=F, sep="\t",col.names=F, row.names=F)

```

# negative controls
## negCtrl 1: random shuffle
## negCtrl 2: all lowly expressed genes in PBS in 3 celltypes and no induction in IFN
## negCtrl 3: all heatmap genes except for cluster1-3 genes
```{r negCtrl3}
# get negative control genes
negCtrl_genes <- c()
for (i in c(4:14)){
 this_cluster_genes = rownames(km_res.rna.new)[which(km_res.rna.new$newclass==i)]
 negCtrl_genes <- c(negCtrl_genes, this_cluster_genes)
}

# get promoter list: 1078 promoters. don't care about which celltype express it
promoter_df <- link_gene_promoter %>% dplyr::filter(gene %in% negCtrl_genes)
promoter_list <- promoter_df  %>% pull(region) %>% as.character() %>% unique()

# save promoter list
write(promoter_list, file=paste0(outDir,"/commonly_induced_promoters_negCtrl_analysis3.txt"))

# fetch bedfile
promoter_bed <- bed %>% dplyr::filter(region %in% promoter_list)
data.table::fwrite(promoter_bed, file=paste0(outDir,"/commonly_induced_promoters_negCtrl_analysis3.bed"),quote=F, sep="\t",col.names=F, row.names=F)

```
## negCtrl 4: all non-induced genes whose average PBS levels across 3 celltypes match that of the target genes.
```{r negCtrl4}
# get target gene mean PBS expression for each celltype
# get gene list
gene_list <- c()
for (i in c(1,2,3)){
 this_cluster_genes = rownames(km_res.rna.new)[which(km_res.rna.new$newclass==i)]
 gene_list <- c(gene_list, this_cluster_genes)
}
targetExprs_PBS_MEL <- CPM_rna[gene_list,] %>% dplyr::select(ends_with("_PBS_MEL")) %>% apply( . , 1, mean)
targetExprs_PBS_KRT <- CPM_rna[gene_list,] %>% dplyr::select(ends_with("_PBS_KRT")) %>% apply( . , 1, mean)
targetExprs_PBS_FRB <- CPM_rna[gene_list,] %>% dplyr::select(ends_with("_PBS_FRB")) %>% apply( . , 1, mean)

# get all non-induced genes
all_genes <- rownames(CPM_rna)
nonInduced_genes <- all_genes[which(!all_genes %in% rownames(km_res.rna.new))]
CPM_rna_nonInduced <- CPM_rna[nonInduced_genes,]
CPM_rna_nonInduced$meanPBSexprs.MEL <- CPM_rna_nonInduced %>% dplyr::select(contains("_PBS_MEL")) %>% rowMeans()
CPM_rna_nonInduced$meanPBSexprs.KRT <- CPM_rna_nonInduced %>% dplyr::select(contains("_PBS_KRT")) %>% rowMeans()
CPM_rna_nonInduced$meanPBSexprs.FRB <- CPM_rna_nonInduced %>% dplyr::select(contains("_PBS_FRB")) %>% rowMeans()

# get negative control genes (i.e. non-induced genes with similar PBS expression)
# at 10% threshold, 120 genes out of 263 don't have a negative control gene
# at 20% threshold, 42 genes out of 263 don't have a negative control gene
# at 30% threshold, 8 genes out of 263 don't have a negative control gene
# at 35% threshold, 4 genes out of 263 don't have a negative control gene
negCtrl_genes <- c()
for (target_gene in gene_list) {
  threshold <- 0.30
  # MEL
  target_PBSmean <- CPM_rna[target_gene,] %>% dplyr::select(contains("_PBS_MEL")) %>% rowMeans()
  this_negCtrl_genes_MEL <- CPM_rna_nonInduced %>% 
    dplyr::filter(meanPBSexprs.MEL >= (target_PBSmean - (target_PBSmean * threshold))  & 
                    meanPBSexprs.MEL <= (target_PBSmean + (target_PBSmean * threshold))) %>%
    rownames_to_column("gene") %>% pull(gene)
  # KRT
    target_PBSmean <- CPM_rna[target_gene,] %>% dplyr::select(contains("_PBS_KRT")) %>% rowMeans()
  this_negCtrl_genes_KRT <- CPM_rna_nonInduced %>% 
    dplyr::filter(meanPBSexprs.KRT >= (target_PBSmean - (target_PBSmean * threshold))  &  
                    meanPBSexprs.KRT <= (target_PBSmean + (target_PBSmean * threshold))) %>%
    rownames_to_column("gene") %>% pull(gene)
    # FRB
    target_PBSmean <- CPM_rna[target_gene,] %>% dplyr::select(contains("_PBS_FRB")) %>% rowMeans()
  this_negCtrl_genes_FRB <- CPM_rna_nonInduced %>% 
    dplyr::filter(meanPBSexprs.FRB >= (target_PBSmean - (target_PBSmean * threshold))  &  
                    meanPBSexprs.FRB <= (target_PBSmean + (target_PBSmean * threshold))) %>%
    rownames_to_column("gene") %>% pull(gene)
  
  # build negative control genes and pick one
  this_negCtrl_genes <- intersect(this_negCtrl_genes_FRB, intersect(this_negCtrl_genes_KRT, this_negCtrl_genes_MEL))
  if (length(this_negCtrl_genes) > 0) {
    this_sample <- sample(this_negCtrl_genes,1)
    # if the neg control set ALL exist in the main negCtrl set, then randomly pick one and add to the main list
    if ( all(this_negCtrl_genes %in% negCtrl_genes)) {
      negCtrl_genes <- c(negCtrl_genes, this_sample)
    } else {
      # otherwise, pick a random gene from this target's negCtrl set that hasn't been picked, and add to main list
      while (this_sample %in% negCtrl_genes) {
          this_sample <- sample(this_negCtrl_genes,1)
      }
      negCtrl_genes <- c(negCtrl_genes, this_sample)
    }
  } else {
    negCtrl_genes <- c(negCtrl_genes, NA)
  }
}

negCtrl_genes <- negCtrl_genes %>% na.omit()

# get promoter list: 1078 promoters. don't care about which celltype express it
promoter_df <- link_gene_promoter %>% dplyr::filter(gene %in% negCtrl_genes)
promoter_list <- promoter_df  %>% pull(region) %>% as.character() %>% unique()

# save promoter list
write(promoter_list, file=paste0(outDir,"/commonly_induced_promoters_negCtrl_analysis4.txt"))

# fetch bedfile
promoter_bed <- bed %>% dplyr::filter(region %in% promoter_list)
data.table::fwrite(promoter_bed, file=paste0(outDir,"/commonly_induced_promoters_negCtrl_analysis4.bed"),quote=F, sep="\t",col.names=F, row.names=F)

```
## negCtrl 5: all commonly induced genes' promoters that DO NOT gain H3K27ac signal in any celltype
```{r negCtrl5}
# get gene list
gene_list <- c()
for (i in c(1,2,3)){
 this_cluster_genes = rownames(km_res.rna.new)[which(km_res.rna.new$newclass==i)]
 gene_list <- c(gene_list, this_cluster_genes)
}

# get promoter list
promoter_df <- link_gene_promoter %>% dplyr::filter(gene %in% gene_list)

# only pick promoters that DON"T gain H3K27ac signal in any celltype. 364 regions.
temp1 <- MEL_cRE_dynamics %>% dplyr::filter(region %in% promoter_df$region & DK27ac=="gainH3K27ac") %>% pull(region)
temp2 <- KRT_cRE_dynamics %>% dplyr::filter(region %in% promoter_df$region & DK27ac=="gainH3K27ac") %>% pull(region)
temp3 <- FRB_cRE_dynamics %>% dplyr::filter(region %in% promoter_df$region & DK27ac=="gainH3K27ac") %>% pull(region)

promoter_df <- link_gene_promoter %>% 
  dplyr::filter(gene %in% gene_list) %>%
  dplyr::filter(! region %in% unique(c(temp1, temp2, temp3)))

# save promoter list
write(promoter_list, file=paste0(outDir,"/commonly_induced_promoters_negCtrl_analysis5.txt"))

# fetch bedfile
promoter_bed <- bed %>% dplyr::filter(region %in% promoter_list)
data.table::fwrite(promoter_bed, file=paste0(outDir,"/commonly_induced_promoters_negCtrl_analysis5.bed"),quote=F, sep="\t",col.names=F, row.names=F)
```