---
title: "celseq2_MEL_seqDepth_analysis"
author: "Shuo Shan"
date: "11/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### sequencing depth analysis

Examine how sequencing depth of Celseq2 libraries affect variability in data.
Do one before batch and non-batch corrected (donor or other variates)
Group samples by library total reads to low (<5M), mid (5-10M), high (>10M)
1.	Correlation between PBS samples sequenced at 10M+ reads, 5-10M reads, and 1-5M reads.
    a.	Range of cycle number. 
    b.	Concatenate sample name 
2.	Cycle number vs. variability. 
### Load packages.    
```{r package, warning=FALSE}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(DEXSeq)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
library(edgeR)
library(gridExtra)
library(reshape2)

Dir="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/seqDepthAnalysis"
setwd(Dir)
#save.image(file=paste0(Dir,'/myEnvironment.seqDepthAnalysis.RData'))
#load(paste0(Dir,'/myEnvironment.seqDepthAnalysis.RData'))
```
### Load ESAT raw count datasets

Celseq2 data for melanotytes (MEL) from 28 donors X 2 culture conditions. Libraries were pooled and prepared by celseq2 protocol, pair-end sequenced by BGI DNBseq or Illumina NextSeq550/NextSeq500 in house. Barcode in read 1 name was used to separate out pooled samples within a library, and read 2 was used for downstream processing and treated as single-end sequencing. Read 2 was aligned by star to hg38, and gene expression count matrix generated by ESAT.
  
The rest of donors that are not included are:  
- not sequenced (F22,F35,F36)
- genotyping failed (F47)
- sequencing failed (F24M, F28M)
```{r, echo=FALSE, message=FALSE}
file <- paste0("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/ESAT/output/MEL/gene/","all.gene.txt")
esat <- read.table(file,sep="\t",header=TRUE,row.names=1) %>% dplyr::select(-starts_with(c("F47","F24","F28")))
count <- esat[,-c(1,2)]
colnames(count) <- colnames(count) %>% gsub("_MEL_GTCTAG","M_IFN",.) %>% gsub("_MEL_AGACTC","M_PBS",.)
samples <- colnames(count) %>% gsub("_IFN","", . ) %>% gsub("_PBS","", .) %>% unique(.)

# library size distribution
df=data.frame(sample=samples,total_count=colSums(count))
df=df[order(-df$total_count),]
plot_ly(x = df$sample,y = df$total_count,type = "bar")

# histogram for gene counts, all samples
h <- hist(log10(rowSums(count)), breaks = as.numeric(100), plot = FALSE)
plot_ly(x = h$mids, y = h$counts, width = 500, height=300) %>% layout(title = "Histogram") %>% add_bars()
``` 

### Load metadata
```{r metadata}
lib_info <- read.table(file="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/ESAT/output/MEL/metadata.txt",header=T,sep="\t") %>% dplyr::filter(.$Name %in% colnames(count))
discard <- c("X","X100.","Total_Reads_Lib","Multimapped.Reads.Aligned..RSEM.","Unique.Aligned.Reads..RSEM.","Percent.Unique.Aligned.Reads.RSEM","Percent.Aligned.Reads.RSEM")
lib_info <- lib_info %>% dplyr::select(-discard)
metadata <- lib_info
low <- metadata %>% dplyr::filter(Total_Reads_Sample < 5000000)
med <- metadata %>% dplyr::filter(Total_Reads_Sample < 10000000) %>% dplyr::filter(Total_Reads_Sample >= 5000000)
high <- metadata %>% dplyr::filter(Total_Reads_Sample >= 10000000)
df <- data.frame(Name=c(low$Name,med$Name,high$Name),
                 Lib_Size_Tier=c(rep("low",nrow(low)),
                        rep("med",nrow(med)),
                        rep("high",nrow(high))))
metadata <- inner_join(df,metadata,by="Name")
metadata$NewName <- paste(metadata$Name,"_",as.character(metadata$PCR.cycle.number)) %>% gsub(" ","",.)
rm(lib_info,discard,df,low,med,high)
```
### Handy functions
```{r functions}
# handy function
get_lower_tri <- function(cormat){
  cormat[upper.tri(cormat)]<- NA
  return(cormat)
}

# handy function2
get_melted_cormat <- function(mat){
  # Get upper triangle of the correlation matrix
  cormat <- round(cor(mat),2)
  lower_tri <- get_lower_tri(cormat)
  melted_cormat <- melt(lower_tri, na.rm = TRUE) %>% magrittr::set_colnames(c("Name1","Name2","PCC"))
  return(melted_cormat)
}

# handy function3
plot_cor_heatmap <- function(melted_cormat, caption, subcaption){
  p <- ggplot(data = melted_cormat, aes(Name1, Name2, fill = PCC))+
    geom_tile(color = "white")+
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                         midpoint = 0.75, limit = c(0.5,1), space = "Lab", 
                         name="Pearson\nCorrelation") +
    theme_minimal()+ 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1))+
    coord_fixed() +
    ggtitle(caption) +
    labs(subtitle=subcaption)
  return(p)
}
```
### Calculate CPM matrix
```{r cpm}
lib.size=data.frame(Sample=colnames(count),sum=colSums(count))
mat.pca.num <- t(count) %>% as.data.frame() %>% rownames_to_column("Sample")
mat.pca.num <- inner_join(lib.size,mat.pca.num,by="Sample") %>% column_to_rownames("Sample")
mat.pca.cpm <- mat.pca.num[,-1]/mat.pca.num$sum*1000000 
mat.cpm <- t(mat.pca.cpm) %>% as.data.frame() %>% magrittr::set_colnames(colnames(count))
rm(mat.pca.num, mat.pca.cpm, lib.size)
```
### Calculate filterd CPM matrix (gene count > 10)
```{r cpm-filtered}
mat.cpm.filt <- mat.cpm[rowSums(mat.cpm)>=10,] %>% as.data.frame()
```
### Batch-correct filtered-CPM matrix
```{r batch-correction}
df <- count[rownames(mat.cpm.filt),metadata$Name]
dds <- DESeqDataSetFromMatrix(countData = df,
                              colData = metadata,
                              design = ~ Donor + Condition)
dds <- DESeq(dds)
vsd <- vst(dds, blind=FALSE)
mat <- assay(vsd)
mat <- limma::removeBatchEffect(mat, vsd$Donor)
assay(vsd) <- mat
counts_batch_corrected <- assay(vsd)
```




### Correlation of CPM matrix of PBS samples before batch correction.
```{r cor_cpm_low}
dat="CPM"
condition="PBS"
tier="low"
anno="reads < 5M"
title=paste0(dat,"_",condition,"_",anno)

subset <- metadata %>% 
  dplyr::filter(Condition==condition) %>%
  dplyr::filter(Lib_Size_Tier==tier)
cpm.this <- mat.cpm[,subset$Name] %>% magrittr::set_colnames(subset$NewName)
melted_cormat <- get_melted_cormat(cpm.this)
subtitle <- paste0("min PCC: ",as.character(min(melted_cormat$PCC)),
                   "; max PCC: ",as.character(max(melted_cormat$PCC)),
                   "; median PCC: ",as.character(median(melted_cormat$PCC)))

p.low <- plot_cor_heatmap(melted_cormat,title,subtitle)
p.low

png(file=paste0(Dir,"/",dat,"_",condition,"_",tier,".png"))
p.low
dev.off()

rm(cpm.this,melted_cormat)
```
```{r cor_cpm_med}
dat="CPM"
condition="PBS"
tier="med"
anno="reads: 5M-10M"
title=paste0(dat,"_",condition,"_",anno)

subset <- metadata %>% 
  dplyr::filter(Condition==condition) %>%
  dplyr::filter(Lib_Size_Tier==tier)
cpm.this <- mat.cpm[,subset$Name] %>% magrittr::set_colnames(subset$NewName)
melted_cormat <- get_melted_cormat(cpm.this)
subtitle <- paste0("min PCC: ",as.character(min(melted_cormat$PCC)),
                   "; max PCC: ",as.character(max(melted_cormat$PCC)),
                   "; median PCC: ",as.character(median(melted_cormat$PCC)))

p.med <- plot_cor_heatmap(melted_cormat,title,subtitle)
p.med

png(file=paste0(Dir,"/",dat,"_",condition,"_",tier,".png"))
p.med
dev.off()

rm(cpm.this,melted_cormat)
```
```{r cor_cpm_high}
dat="CPM"
condition="PBS"
tier="high"
anno="reads: >10M"
title=paste0(dat,"_",condition,"_",anno)

subset <- metadata %>% 
  dplyr::filter(Condition==condition) %>%
  dplyr::filter(Lib_Size_Tier==tier)
cpm.this <- mat.cpm[,subset$Name]  %>% magrittr::set_colnames(subset$NewName)
melted_cormat <- get_melted_cormat(cpm.this)
subtitle <- paste0("min PCC: ",as.character(min(melted_cormat$PCC)),
                   "; max PCC: ",as.character(max(melted_cormat$PCC)),
                   "; median PCC: ",as.character(median(melted_cormat$PCC)))

p.high <- plot_cor_heatmap(melted_cormat,title,subtitle)
p.high

png(file=paste0(Dir,"/",dat,"_",condition,"_",tier,".png"))
p.high
dev.off()

rm(cpm.this,melted_cormat)
```


### Correlation of filtered CPM matrix of PBS samples before batch correction.
Surprisingly, filtering out genes with no to low reads didn't affect the correlation for most samples, and made the correlation slightly worse.
```{r cor_cpm-filt_low}
dat="CPM-filtered"
condition="PBS"
tier="low"
anno="reads < 5M"
title=paste0(dat,"_",condition,"_",anno)

subset <- metadata %>% 
  dplyr::filter(Condition==condition) %>%
  dplyr::filter(Lib_Size_Tier==tier)
cpm.this <- mat.cpm.filt[,subset$Name] %>% magrittr::set_colnames(subset$NewName)
melted_cormat <- get_melted_cormat(cpm.this)
subtitle <- paste0("min PCC: ",as.character(min(melted_cormat$PCC)),
                   "; max PCC: ",as.character(max(melted_cormat$PCC)),
                   "; median PCC: ",as.character(median(melted_cormat$PCC)))

p.low <- plot_cor_heatmap(melted_cormat,title,subtitle)
p.low

png(file=paste0(Dir,"/",dat,"_",condition,"_",tier,".png"))
p.low
dev.off()

rm(cpm.this,melted_cormat)
```
```{r cor_cpm-filt_med}
dat="CPM-filtered"
condition="PBS"
tier="med"
anno="reads: 5M-10M"
title=paste0(dat,"_",condition,"_",anno)

subset <- metadata %>% 
  dplyr::filter(Condition==condition) %>%
  dplyr::filter(Lib_Size_Tier==tier)
cpm.this <-  mat.cpm.filt[,subset$Name]  %>% magrittr::set_colnames(subset$NewName)
melted_cormat <- get_melted_cormat(cpm.this)
subtitle <- paste0("min PCC: ",as.character(min(melted_cormat$PCC)),
                   "; max PCC: ",as.character(max(melted_cormat$PCC)),
                   "; median PCC: ",as.character(median(melted_cormat$PCC)))

p.med <- plot_cor_heatmap(melted_cormat,title,subtitle)
p.med

png(file=paste0(Dir,"/",dat,"_",condition,"_",tier,".png"))
p.med
dev.off()

rm(cpm.this,melted_cormat)
```
```{r cor_cpm-filt_high}
dat="CPM-filt"
condition="PBS"
tier="high"
anno="reads: >10M"
title=paste0(dat,"_",condition,"_",anno)

subset <- metadata %>% 
  dplyr::filter(Condition==condition) %>%
  dplyr::filter(Lib_Size_Tier==tier)
cpm.this <- mat.cpm.filt[,subset$Name]  %>% magrittr::set_colnames(subset$NewName)
melted_cormat <- get_melted_cormat(cpm.this)
subtitle <- paste0("min PCC: ",as.character(min(melted_cormat$PCC)),
                   "; max PCC: ",as.character(max(melted_cormat$PCC)),
                   "; median PCC: ",as.character(median(melted_cormat$PCC)))

p.high <- plot_cor_heatmap(melted_cormat,title,subtitle)
p.high

png(file=paste0(Dir,"/",dat,"_",condition,"_",tier,".png"))
p.high
dev.off()

rm(cpm.this,melted_cormat)
```





