---
title: "celseq2_analysis_MEL"
author: "Shuo Shan"
date: "6/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## RNAseq analysis
  
Celseq2 data for melanotytes (MEL) from 16 donors X 2 culture conditions. Libraries were pooled and prepared by celseq2 protocol, pair-end sequenced by BGI, trimmed to (read 1 3bp, read 2 36bp), barcode in read 1 name was used to separate out pooled samples within a library, and read 2 was used for downstream processing and treated as single-end sequencing. Read 2 was aligned by star to hg38, and gene expression count matrix generated by ESAT.
  
### Load packages.    
```{r package, warning=FALSE}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(DEXSeq)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
library(edgeR)
library(gridExtra)

Dir="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq"
setwd(Dir)
```
  
### load metadata for libraries.  
```{r load_metadata}
#############################
ancestry <- data.frame(donor=c("F27","F30","F31","F34","F41","F45","F46","F49",
                              "F50","F51","F52","F55","F56","F25","F62","F63"),
                       ancestry=c("NorthernEuropean","NorthernEuropean","Mixed","Mediterranean",
                                  "Mixed","Mediterranean","Mixed","African",
                                  "NorthernEuropean","Mixed","NorthernEuropean","African",
                                  "Mediterranean","NorthernEuropean","NorthernEuropean","African"))
lib_info <- read.table(file=paste0(Dir,"/16_donors/celseq2_metadata.txt"),header=T,sep="\t")
lib_info$donor <- lib_info$donor %>% gsub("M","",.)
lib_info <- inner_join(lib_info,ancestry,by="donor")
lib_info$name <- lib_info$name %>% gsub("IFNg", "IFN", .)
lib_info$Lib.Prep.Date <- ordered(lib_info$Lib.Prep.Date,levels=c("11/17/20","8/26/20","12/16/20","1/3/21"))
lib_info$Lib.Size <- apply(mat.count,2,sum)

# explore ChIP-seq data:
tmp <- read.table(file="/nl/umw_manuel_garber/human/skin/eQTLs/Nextseq550/20210608_ATAC_ChIPseq/DolphinNext_ChIP/report4819/chip/merged.peak.sizes.txt",header=F,sep="\t")
tmp$bin <- cut(tmp$V1, 20)
table(tmp$bin)
###############################
#lib_info.log2fc <- lib_info %>% filter(condition=="PBS") %>% dplyr::select(c("donor","Lib.Prep.Date","PCR.cycle.number","ancestry"))
```

### Load gene expression matrix.  
```{r load_gene_expression_matrix}
mat <- read.table(file=paste0(Dir,"/16_donors/gene_count_matrix/16_donors.gene.txt"),header=T,row.names=1)
# count matrix and cpm matrix. (cpm=counts per million)
mat <- mat %>% dplyr::select(-contains("Symbol")) %>% 
  dplyr::select(-contains(".1")) %>%
  dplyr::select(-contains("chr.")) %>%
  dplyr::select(-contains("strand.")) %>%
  dplyr::select(-contains("F47"))
colnames(mat) <- colnames(mat) %>%
  gsub("F63M_PBS.ACCATG", "F63M_PBS", .) %>%
  gsub("F63M_IFN.ACTCGA", "F63M_IFN", .) %>%
  gsub("F62M_PBS.GTCTAG", "F62M_PBS", .) %>%
  gsub("F62M_IFN.GTTGCA", "F62M_IFN", .) %>%
  gsub("IFNg"           , "IFN", .)
mat.count <- mat %>% dplyr::select(-c(1,2))
mat.count <- mat.count[,order(colnames(mat.count))]
mat <- cbind(mat[,c(1:2)],mat.count)
mat.count.filtered <- mat.count[rowMedians(as.matrix(mat.count)) > 0,]

samples.PBS <- sort(colnames(mat.count)[grep("PBS",colnames(mat.count))])
samples.IFN <- sort(colnames(mat.count)[grep("IFN",colnames(mat.count))])

raw_counts_wn <- mat.count[rowMedians(as.matrix(mat.count)) > 0,]
pseudo_counts <- log2(raw_counts_wn + 1)
```

```{r mean_vs_cv}
df <- data.frame(mean = apply(raw_counts_wn[ ,lib_info$condition == "IFNg"], 1, mean),
                 var = apply(raw_counts_wn[ ,lib_info$condition == "IFNg"], 1, var))
df <- df[df$mean <= 5000, ]
p <- ggplot(data=df, aes(x = mean, y = var))
p <- p + geom_point(colour = "orange",alpha=0.4)
p <- p + theme_bw()
p <- p + geom_abline(aes(intercept=0, slope=1))
p <- p + ggtitle("Variance versus mean in counts") + ylab("variance")
print(p)

this.highvar <- df[which(df$var>1500000),]
this.lowvar <- df[which(df$mean>3000 & df$var<1000000),]


# df <- mat.cpm.filtered
# ### generate stats (mean, std, cv) of log2-transformed data
# df.mean <- apply(df,MARGIN=1,FUN=mean)
# df.variance <- (apply(df,MARGIN=1,FUN=sd)) ^ 2
# #df.poissonVariance <- rpois(length(df.mean),df.mean)
# # plot: mean vs. CV
# df.plot <- data.frame(mean=df.mean,
#                  cv=df.lncv)
# df.plot <- data.frame(mean=log10(df.mean),
#                       variance=log10(df.variance),
#                       poissonVariance=log10(df.poissonVariance))
# # plot
# ggplot(df.plot,aes(x=mean,y=variance)) + 
#   geom_point(aes(y=variance),color="black",alpha=0.2) +
#   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#         panel.background = element_blank(), axis.line = element_line(colour = "black")) +
#   xlab("log10(mean)") + ylab("log10(variance)")
# # poisson isn't a good fit
# ggplot(df.plot,aes(x=mean,y=variance)) + 
#   geom_point(aes(y=variance,color="a"),alpha=0.2) +
#   geom_point(aes(y=poissonVariance,color="b"),alpha=0.2) +
#   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#         panel.background = element_blank(), axis.line = element_line(colour = "black")) +
#   xlab("log(mean)") + ylab("log(variance)") +
#   scale_color_manual(values=c("a"="black","b"="blue"),
#                      labels=c("observed","poisson"),
#                      name="")
```
### generate counts-per-million matrix.  
```{r cpm}
### counts per million
cSum <- apply(mat.count,2,sum)
mat.cpm <- 1000000*sweep(mat.count,MARGIN=2,FUN="/",STATS=cSum)
# another way of filtering
mat.cpm.filtered <- mat.cpm[rowMedians(as.matrix(mat.count)) > 0,]
mat.count.filtered <- mat.count[rowMedians(as.matrix(mat.count)) > 0,]
# # filter out empty/low-expressed genes with < 1 max cpm across all samples
# mat.cpm$max = apply(mat.cpm, 1, function (x) max(x))
# mat.cpm.filtered <- mat.cpm %>% rowwise() %>% mutate(max=max(F25M_IFN:F63M_PBS)) %>% 
#   as.data.frame() %>% magrittr::set_rownames(rownames(mat.cpm)) %>%
#   filter(max>=1) %>% dplyr::select(-max)
# mat.count.filtered <- mat.count[rownames(mat.cpm.filtered),]
# no threshold, 45,384
# so 30,375 genes have 0 raw count in PBS and IFN.
mat.cpm.PBS <- mat.count.filtered[,samples.PBS]
mat.cpm.IFN <- mat.count.filtered[,samples.IFN]
mat.cpm.log2FC <- log2(mat.cpm.IFN+1) - log2(mat.cpm.PBS+1)
colnames(mat.cpm.log2FC) <- colnames(mat.cpm.log2FC) %>% gsub("M","",.)
```
```{r count_distribution}

df <- data.frame(rcounts = raw_counts_wn[ ,1], prcounts = pseudo_counts[ ,1])

p <- ggplot(data=df, aes(x = rcounts, y = ..density..))
p <- p + geom_histogram(fill = "lightblue")
p <- p + theme_bw()
p <- p + ggtitle(lib_info$name[1])
p <- p + xlab("counts")

p2 <- ggplot(data=df, aes(x = prcounts, y = ..density..))
p2 <- p2 + geom_histogram(fill = "lightblue")
p2 <- p2 + theme_bw()
p2 <- p2 + ggtitle(lib_info$name[1])
p2 <- p2 + xlab(expression(log[2](counts + 1)))

grid.arrange(p, p2, ncol = 2)

```
```{r normalization_comparison}
df <- raw_counts_wn
### generate stats (mean, std, cv)
df.mean <- apply(df,MARGIN=1,FUN=mean)
df.variance <- (apply(df,MARGIN=1,FUN=sd)) ^ 2
# order by variance
df <- df[order(df.variance,decreasing=T),]
df = df[1:2500,]
pseudo_counts <- log2(df + 1)

# raw counts
df_raw <- melt(as.matrix(pseudo_counts), id = rownames(pseudo_counts))
names(df_raw)[1:2]<- c("id", "sample")
df_raw$method <- rep("Raw counts", nrow(df_raw))

# cpm
df_cpm <- log2(mat.cpm[rownames(df),]+1)
df_cpm <- melt(as.matrix(df_cpm), id = rownames(df_cpm))
names(df_cpm)[1:2]<- c("id", "sample")
df_cpm$method <- rep("CPM", nrow(df_cpm))

# DESeq2 normalization
# lib_info$Lib.Prep.Date <- as.character(lib_info$Lib.Prep.Date)
dge <- DESeqDataSetFromMatrix(countData = df,
                              colData = lib_info,
                              design = ~ condition)
dge <- estimateSizeFactors(dge)
deseq_normcount <- counts(dge, normalized = TRUE)
pseudo_deseq <- log2(deseq_normcount + 1)
df_deseq <- melt(pseudo_deseq, id = rownames(df))
names(df_deseq)[1:2]<- c("id", "sample")
df_deseq$method <- rep("DESeq (RLE)", nrow(df_deseq))  
#test_normcount <- sweep(raw_counts_wn, 2, sizeFactors(dge), "/")
#sum(test_normcount != deseq_normcount)

# edgeR normalization
dge2 <- DGEList(df)
#dge2$samples
# edgeR: normalize by TMM
dge2 <- calcNormFactors(dge2, method = "TMM")
pseudo_TMM <- log2(cpm(dge2) + 1)
df_TMM <- reshape2::melt(pseudo_TMM, id = rownames(df))
names(df_TMM)[1:2] <- c ("id", "sample")
df_TMM$method <- rep("TMM", nrow(df_TMM))
# edgeR: normalize by total count
pseudo_TC <- log2(cpm(dge2) + 1)
df_TC <- melt(pseudo_TC, id = rownames(df))
names(df_TC)[1:2] <- c ("id", "sample")
df_TC$method <- rep("TC", nrow(df_TC))
# edgeR: normalize by upper quartile
dge2 <- calcNormFactors(dge2, method = "upperquartile")
pseudo_UQ <- log2(cpm(dge2) + 1)
df_UQ <- melt(pseudo_UQ, id = rownames(df))
names(df_UQ)[1:2] <- c ("id", "sample")
df_UQ$method <- rep("UQ", nrow(df_UQ))
#test_normcount <- sweep(dge2$counts, 2, dge2$samples$lib.size*dge2$samples$norm.factors / 10^6,"/")
#range(as.vector(test_normcount - cpm(dge2)))

# compare
df_allnorm <- rbind(df_raw, df_cpm, df_deseq, df_TMM)
df_allnorm$method <- factor(df_allnorm$method,
                            levels = c("Raw counts", "CPM","DESeq (RLE)","TMM"))

p <- ggplot(data=df_allnorm, aes(x=sample, y=value, fill=method))
p <- p + geom_boxplot(outlier.size = 0.1)  
p <- p + theme_bw()
p <- p + ggtitle("Boxplots of normalized pseudo counts\n
for all samples by normalization methods")
p <- p + facet_grid(. ~ method) 
p <- p + ylab(expression(log[2] ~ (normalized ~ count + 1))) + xlab("")
p <- p + theme(title = element_text(size=10), axis.text.x = element_blank(), 
               axis.ticks.x = element_blank())

# compare #2
p <- ggplot(data=df_allnorm, aes(x=value, colour=sample))
p <- p + geom_density()  
p <- p + theme_bw()
p <- p + ggtitle("Density of normalized pseudo counts\n
for all samples by normalization methods")
p <- p + facet_grid(. ~ method) 
p <- p + ylab(expression(log[2] ~ (normalized ~ count + 1))) + xlab("")
p <- p + theme(title = element_text(size=10), axis.text.x = element_blank(), 
               axis.ticks.x = element_blank())
print(p)

# conclusion: visually, DESeq or EdgeR TMM normalization methods both give similar results
# we'll stick with DESeq2 normalization
```
```{r pca-post-DESeq2normalization}
df <- raw_counts_wn
# order by variance
df.mean <- apply(df,MARGIN=1,FUN=mean)
df.variance <- (apply(df,MARGIN=1,FUN=sd)) ^ 2
df <- df[order(df.variance,decreasing=T),]
df = df[1:2500,]
dge <- DESeqDataSetFromMatrix(countData = df,
                              colData = lib_info,
                              design = ~ condition)
dge <- estimateSizeFactors(dge)
deseq_normcount <- counts(dge, normalized = TRUE)
pseudo_deseq <- log2(deseq_normcount + 1)

mat.pca.num <- t(pseudo_deseq) %>% as.data.frame()
pca = prcomp(as.matrix(mat.pca.num),centered= T,scale = T )
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
df.pca <- data.frame(PCs = colnames(pca$x), pctVar = pca$sdev^2/sum(pca$sdev^2))
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)
ggplot(df.pca[1:6,],aes(x=PCs)) + 
  geom_bar(aes(y=pctVar),stat="identity",fill="blue",alpha=0.6) +
  xlab("Principle Components (PCs)") + 
  ylab("percent variance captured by each PC") +
  geom_label(aes(y=-0.03,label=round(pctVar,3))) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# top PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], PC4=pca$x[,4], PC5=pca$x[,5],PC6=pca$x[,6],name=rownames(pcs))
d=inner_join(d,lib_info,by="name") %>% magrittr::set_rownames(d$name)
colnames(d)

# PCA plots
# PC1 vs. PC2, by condition
ggplot(d, aes(x=PC1,y=PC2,col=condition,label=name)) + 
  geom_point() + 
  geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'condition') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC1 vs. PC2, by batch
d %>%
  mutate(q=factor(Lib.Prep.Date)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=Lib.Prep.Date,label=name)) + 
  geom_point(size=3) + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'Lib.Prep.Date') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC1 vs. PC2, by condition
ggplot(d, aes(x=PC1,y=PC2,col=as.factor(ancestry),label=name)) + 
  geom_point() + 
  geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'barcode') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC2 vs. PC3, by Sequencing Platform
d %>%
  mutate(platform=as.factor(Sequencing.Platform)) %>%
  ggplot(., aes(y=PC2,x=PC3,col=platform,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'platform') +
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) + 
  xlab(sprintf("PC3 (%2.1f%% variance)",100*pca.pctVar[3])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# check for correlation
cor(d$PC1,as.numeric(as.factor(d$condition)),method="pearson")
cor(d$PC2,as.numeric(d$Lib.Prep.Date),method="pearson")
cor(d$PC3,as.numeric(as.factor(d$donor)),method="pearson")
cor(d$PC3,d$donor,method="pearson")
```

## PCA OF CPM
### get top 2500 genes with most variable cpm across all samples
```{r top2500, include=F}
### generate stats (mean, std, cv) of log2-transformed data
mat.cpm.filtered.log2 <- log2(mat.cpm.filtered+1)
mat.cpm.filtered.mean <- log2(apply(mat.cpm.filtered,MARGIN=1,FUN=mean)+1)
mat.cpm.filtered.lnsd <- apply(mat.cpm.filtered.log2,MARGIN=1,FUN=sd)
mat.cpm.filtered.lncv <- sqrt( (exp(mat.cpm.filtered.lnsd) )^2 -1) #coefficient of variance

# order by cv
mat.cpm.filtered.log2.ordered <- mat.cpm.filtered.log2[order(mat.cpm.filtered.lncv, decreasing=T),]
mat.cpm.filtered.ordered <- mat.cpm.filtered[order(mat.cpm.filtered.lncv, decreasing=T),]
mat.count.filtered.ordered <- mat.count.filtered[order(mat.cpm.filtered.lncv, decreasing=T),]

# pick top 2500 genes, largest cv first
mat.cpm.filtered.log2.top = mat.cpm.filtered.log2.ordered[1:2500,]
mat.cpm.filtered.top = mat.cpm.filtered.ordered[rownames(mat.cpm.filtered.log2.top),]

# plot: mean vs. CV
df <- data.frame(mean=mat.cpm.filtered.mean,
                 cv=mat.cpm.filtered.lncv)
ggplot(df,aes(x=mean,y=cv)) + geom_point()

# plot: density distribution
n=1
sample=mat.cpm.filtered[,colnames(mat.cpm.filtered)[n]] %>% as.data.frame() %>% magrittr::set_colnames("cpm")
ggplot(sample, aes(x=cpm)) + geom_density()
```

### PCA
```{r pca, echo=F}
#df <- raw_counts_wn
df <- mat.cpm.filtered
### generate stats (mean, std, cv) of log2-transformed data
df.mean <- apply(df,MARGIN=1,FUN=mean)
df.variance <- (apply(df,MARGIN=1,FUN=sd)) ^ 2

# order by variance
df.ordered <- df[order(df.variance,decreasing=T),]

# pick top genes, largest variance first
df.top = df.ordered[1:5000,]
df.top = df.ordered[rownames(df.top),]
df.top = log2(df.top+1)

mat.pca.num <- t(df.top) %>% as.data.frame()
pca = prcomp(as.matrix(mat.pca.num),centered= T,scale = T )

# loadings in each pc
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x 
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# top 3 PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], PC4=pca$x[,4], PC5=pca$x[,5],PC6=pca$x[,6],name=rownames(pcs))
d=inner_join(d,lib_info,by="name") %>% magrittr::set_rownames(d$name)
colnames(d)
```
```{r pca-plots}
# PC1 vs. PC2, by condition
ggplot(d, aes(y=PC1,x=PC2,col=condition,label=name)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'condition') +
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC1 vs. PC2, by batch
d %>%
  mutate(q=factor(Lib.Prep.Date)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=Lib.Prep.Date,shape=condition,label=name)) + 
  geom_point(size=3) + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'Lib.Prep.Date') +
  scale_shape_manual(values=c(16,17)) +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC1 vs. PC2, by batch
d %>%
  mutate(q=factor(Lib.Prep.Date)) %>%
  ggplot(., aes(x=PC3,y=PC2,col=Lib.Prep.Date,label=name)) + 
  geom_point(size=3) + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'Lib.Prep.Date') +
  #scale_shape_manual(values=c(16,17)) +
  xlab(sprintf("PC3 (%2.1f%% variance)",100*pca.pctVar[3])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC1 vs. PC2, by Uniquely Mapped Reads
d$group_tags <- cut(d$Unique.Reads.Aligned..STAR., breaks=6)
d %>%
  mutate(platform=d$group_tags) %>%
  ggplot(., aes(y=PC1,x=PC2,col=platform,label=name)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'Uniquely Aligned Reads') +
  scale_shape_manual(values=c(16,17)) +
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# check for correlation
cor(d$PC1,as.numeric(as.factor(d$condition)),method="pearson")
cor(d$PC4,as.numeric(as.factor(d[,5]),method="pearson"))

cor(d$PC2,as.numeric(d$Lib.Prep.Date),method="pearson")
cor(d$PC3,as.numeric(as.factor(d$Sequencing.Platform)),method="pearson")
```
```{r batch-correction}
library(sva) # for loading ComBat-seq
count_matrix <- as.matrix(mat.count)
write.table(count_matrix,file=paste0(Dir,"/16_donors/combat_ready_count_matrix.txt"),quote=F,sep="\t",row.names=T,col.names=T)
write.table(d,file=paste0(Dir,"/16_donors/combat_ready_metadata.txt"),quote=F,sep="\t",row.names=T,col.names=T)
batch <- d[names(mat.count),]$Lib.Prep.Date %>% as.numeric()
group <- d[names(mat.count),]$condition %>% ordered(.,levels=c("PBS","IFNg")) %>% as.numeric()
# this sentence couldn't be ran yet since sva is outdated and doesn't contain ConBat_seq yet.
adjusted_counts <- sva::ComBat_seq(count_matrix, batch=batch, group=group, full_mod=TRUE)
adjusted_counts <- read.table(file=paste0(Dir,"/16_donors/combat_adjusted_count_matrix.txt"),header=T,sep="\t")
```
```{r pca-after-combat}
df <- adjusted_counts
### generate stats (mean, std, cv) of log2-transformed data
df.mean <- apply(df,MARGIN=1,FUN=mean)
df.variance <- (apply(df,MARGIN=1,FUN=sd)) ^ 2
# order by variance
df.ordered <- df[order(df.variance,decreasing=T),]
# pick top 2500 genes, largest variance first
df.top = df.ordered[1:2500,]
df.top = df.ordered[rownames(df.top),]
df.top = log2(df.top+1)
# pca
mat.pca.num <- t(df.top) %>% as.data.frame()
pca = prcomp(as.matrix(mat.pca.num),centered= T,scale = T )
# loadings in each pc
loadings=as.data.frame(pca$rotation)
# percent variance in each pc
plot(pca)
df.pca <- data.frame(PCs = colnames(pca$x), pctVar = pca$sdev^2/sum(pca$sdev^2))
ggplot(df.pca[1:6,],aes(x=PCs)) + 
  geom_bar(aes(y=pctVar),stat="identity",fill="blue",alpha=0.6) +
  xlab("Principle Components (PCs)") + 
  ylab("percent variance captured by each PC") +
  geom_label(aes(y=-0.03,label=round(pctVar,3))) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# top PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], PC4=pca$x[,4], PC5=pca$x[,5],PC6=pca$x[,6],name=rownames(pcs))
d=inner_join(d,lib_info,by="name") %>% magrittr::set_rownames(d$name)
colnames(d)

# PCA plots
# PC1 vs. PC2, by condition
ggplot(d, aes(x=PC1,y=PC2,col=condition,label=name)) + 
  geom_point() + 
  geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'condition') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC1 vs. PC2, by batch
d %>%
  mutate(q=factor(Lib.Prep.Date)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=Lib.Prep.Date,label=name)) + 
  geom_point(size=3) + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'Lib.Prep.Date') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC1 vs. PC2, by condition
ggplot(d, aes(x=PC1,y=PC2,col=as.factor(ancestry),label=name)) + 
  geom_point() + 
  geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'barcode') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC2 vs. PC3, by Sequencing Platform
d %>%
  mutate(platform=as.factor(Sequencing.Platform)) %>%
  ggplot(., aes(y=PC2,x=PC3,col=platform,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'platform') +
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) + 
  xlab(sprintf("PC3 (%2.1f%% variance)",100*pca.pctVar[3])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# PC3 vs. PC4, by donor
d %>%
  mutate(platform=as.factor(donor)) %>%
  ggplot(., aes(y=PC3,x=PC4,col=platform,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'donor') +
  ylab(sprintf("PC3 (%2.1f%% variance)",100*pca.pctVar[3])) + 
  xlab(sprintf("PC4 (%2.1f%% variance)",100*pca.pctVar[4])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

# check for correlation
cor(d$PC1,as.numeric(as.factor(d$condition)),method="pearson")
cor(d$PC2,as.numeric(d$Lib.Prep.Date),method="pearson")
cor(d$PC3,as.numeric(as.factor(d$Sequencing.Platform)),method="pearson")

```
### DESeq analysis
```{r DESeq analysis}
lib_info$condition <- lib_info$condition %>%
  gsub("IFNg","stimulated", .) %>%
  gsub("PBS", "ctrl", .)
lib_info$Lib.Prep.Date <- as.character(lib_info$Lib.Prep.Date)
# df <- raw_counts_wn
# # order by variance
# df.mean <- apply(df,MARGIN=1,FUN=mean)
# df.variance <- (apply(df,MARGIN=1,FUN=sd)) ^ 2
# df <- df[order(df.variance,decreasing=T),]
# df = df[1:2500,]
# dge <- DESeqDataSetFromMatrix(countData = df,
#                               colData = lib_info,
#                               design = ~ condition)
# dge <- estimateSizeFactors(dge)
# deseq_normcount <- counts(dge, normalized = TRUE)
# pseudo_deseq <- log2(deseq_normcount + 1)
# dds <- DESeq(dge)
df <- raw_counts_wn
# order by variance
df.mean <- apply(df,MARGIN=1,FUN=mean)
df.variance <- (apply(df,MARGIN=1,FUN=sd)) ^ 2
df <- df[order(df.variance,decreasing=T),]
df = df[1:2500,]
mat.count.filtered2 <- df %>% dplyr::select(starts_with(c("F25","F51","F49","F27","F55","F52")))
lib_info2 <- lib_info %>% filter(donor %in% c("F25","F51","F49","F27","F55","F52"))
dds <- DESeqDataSetFromMatrix(countData = mat.count.filtered2,
                              colData = lib_info2,
                              design = ~ condition)
dds <- DESeq(dds) # turn off fold change shrinkage 
res <- results(dds)
summary(res)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
#ntd <- normTransform(dds)
#resLFC <- lfcShrink(dds, coef="condition_stimulated_vs_ctrl", type="apeglm")
resOrdered <- res[order(res$pvalue),] %>% as.data.frame() 
resOrdered <- resOrdered %>% filter(padj < 0.05)
sum(res$padj < 0.05, na.rm=TRUE)

res.up <- resOrdered %>% filter(log2FoldChange < -1.5) %>% filter(padj < 0.05)
res.down <- resOrdered %>% filter(log2FoldChange > 1.5) %>% filter(padj < 0.05)
res.nochange <- resOrdered %>% filter(abs(log2FoldChange) <= 1.5)

##############################
# heatmap of log2FC
df <- rbind(res.up,res.nochange[1:30,],res.down[1:17,]) 
df <- df %>% rownames_to_column("id")
df$tag <- c(rep("up",nrow(res.up)),rep("nochange",nrow(res.nochange[1:30,])),rep("down",nrow(res.down)))
df <- df %>% dplyr::select("id","tag")

samples.PBS <- sort(colnames(normalized_counts)[grep("PBS",colnames(normalized_counts))])
samples.IFN <- sort(colnames(normalized_counts)[grep("IFN",colnames(normalized_counts))])
normalized_counts.PBS <- normalized_counts[,samples.PBS]
normalized_counts.IFN <- normalized_counts[,samples.IFN]
normalized_counts.log2FC <- log2(normalized_counts.IFN+1) - log2(normalized_counts.PBS+1)
colnames(normalized_counts.log2FC) <- colnames(normalized_counts.log2FC) %>% gsub("_IFN","",.)

df2 <- normalized_counts.log2FC %>% rownames_to_column("id") %>% inner_join(., df, by="id")

# scale data
df.scaled <- df2
df.scaled[, c(2:7)] <- scale(df.scaled[, 2:7])

# Run clustering
df.matrix <- as.matrix(df.scaled[, c(2:7)])
rownames(df.matrix) <- df.scaled$id
df.dendro <- as.dendrogram(hclust(d = dist(x = df.matrix)))

# Create dendro
library(ggdendro)
dendro.plot <- ggdendrogram(data = df.dendro, rotate = TRUE) + theme(axis.text.y = element_text(size = 6))
# Preview the plot
print(dendro.plot)

# melt data to long format
df.long <- reshape2::melt(df.scaled, id.vars = c("id", "tag"))
heatmap.plot <- ggplot(data = df.long, aes(x = variable, y = id)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_text(size = 6))

# Preview the heatmap
print(heatmap.plot)

# order dendrogram
df.order <- order.dendrogram(df.dendro)
# Order the levels according to their position in the cluster
df.long$id<- factor(x = df.long$id,levels = df.scaled$id[df.order], ordered = TRUE)
# Heatmap
heatmap.plot <- ggplot(data = df.long, aes(x = variable, y = id)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_text(size = 6))

# Preview the heatmap
print(heatmap.plot)
library("grid")
grid.newpage()
print(heatmap.plot, vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(dendro.plot, vp = viewport(x = 0.90, y = 0.445, width = 0.2, height = 1.0))

################### second heatmap
df <- rbind(res.up,res.nochange[1:30,],res.down[1:17,]) 
df <- df %>% rownames_to_column("id")
df$tag <- c(rep("up",nrow(res.up)),rep("nochange",nrow(res.nochange[1:30,])),rep("down",nrow(res.down)))
df <- df %>% dplyr::select("id","tag")
df2 <- normalized_counts %>% rownames_to_column("id") %>% inner_join(., df, by="id")

# scale data
df.scaled <- df2
df.scaled[, -c(1,14)] <- scale(df.scaled[, -c(1,14)])

# Run clustering
df.matrix <- as.matrix(df.scaled[, -c(1,14)])
rownames(df.matrix) <- df.scaled$id
df.dendro <- as.dendrogram(hclust(d = dist(x = df.matrix)))

# Create dendro
library(ggdendro)
dendro.plot <- ggdendrogram(data = df.dendro, rotate = TRUE) + theme(axis.text.y = element_text(size = 6))
# Preview the plot
print(dendro.plot)

# melt data to long format
df.long <- reshape2::melt(df.scaled, id.vars = c("id", "tag"))
heatmap.plot <- ggplot(data = df.long, aes(x = variable, y = id)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_text(size = 6))

# Preview the heatmap
print(heatmap.plot)

# order dendrogram
df.order <- order.dendrogram(df.dendro)
# Order the levels according to their position in the cluster
df.long$id<- factor(x = df.long$id,levels = df.scaled$id[df.order], ordered = TRUE)
# Heatmap
heatmap.plot <- ggplot(data = df.long, aes(x = variable, y = id)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2() +
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(angle = 30,vjust = 1,hjust=1,size=6))

# Preview the heatmap
print(heatmap.plot)
library("grid")
grid.newpage()
print(heatmap.plot, vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(dendro.plot, vp = viewport(x = 0.90, y = 0.445, width = 0.2, height = 1.0))

######################################
# # Pranitha's old heatmap function
df3 <- df2[, -c(1,14)] %>% magrittr::set_rownames(df2$id)
heatmap.2(as.matrix(df3), trace="none", density="none", col=bluered(75),
          Colv = T, Rowv=T, 
          dendrogram="column", 
          scale="row",
          # distance: 1 minus Pearson correlation distance with average linkage
          distfun=function(x) as.dist((1-cor(t(x)))/2), # as.dist() computes distance matrix. 
          srtCol=45, cexCol=1, #labeling angle
          main="Melanocytes (IFNg stimulation)")
```
```{r pca_after_DESeq2norm}
#df <- raw_counts_wn
df <- normalized_counts

mat.pca.num <- t(df) %>% as.data.frame()
pca = prcomp(as.matrix(mat.pca.num),centered= T,scale = T )

# loadings in each pc
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x 
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# top 3 PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], PC4=pca$x[,4], PC5=pca$x[,5],PC6=pca$x[,6],name=rownames(pcs))
d=inner_join(d,lib_info,by="name") %>% magrittr::set_rownames(d$name)
colnames(d)


```
```{r beehive_per_gene}
# beehive plot of individual gene
g="SP100"
name.row <- colnames(normalized_counts)
df <- normalized_counts[g,] %>% t()
df <- cbind(name.row,df)
df <- as.data.frame(df)
colnames(df) <- c("name","expression")
df <- inner_join(df,lib_info2,by="name")
df$expression <- as.numeric(as.character(df$expression))
ggplot(df,aes(x=condition,y=expression,color=condition)) + 
  geom_point(size=4) + 
  scale_y_log10() +
  scale_color_manual(values=c("olivedrab","violetred")) +
  geom_line(aes(group = donor),color="grey") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  labs(title=g)

```
```{r DE_evaluation}
# sample-to-sample distances
vsd <- vst(dds, blind=FALSE)
sampleDists <- dist(t(assay(vsd)),method="manhattan")
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- vsd$name
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

lib_info$keep <- ifelse(lib_info$donor %in% c("F25","F51","F49","F27","F52","F55"),1,0)
lib_info <- lib_info %>% filter(condition=="IFNg") %>%
 arrange(desc(Unique.Reads.Aligned..STAR.)) 
lib_info$name <- ordered(lib_info$name,levels=lib_info$name)
ggplot(lib_info , aes(x=name,y=Unique.Reads.Aligned..STAR.,fill=keep)) + geom_bar(stat="identity") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text.x = element_text(angle = 30,vjust = 1,hjust=1,size=8))

lib_info$keep <- ifelse(lib_info$donor %in% c("F25","F51","F49","F27","F52","F55"),1,0)
lib_info <- lib_info %>% filter(condition=="PBS") %>%
 arrange(desc(Unique.Reads.Aligned..STAR.)) 
lib_info$name <- ordered(lib_info$name,levels=lib_info$name)
ggplot(lib_info , aes(x=name,y=Unique.Reads.Aligned..STAR.,fill=keep)) + geom_bar(stat="identity") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        axis.text.x = element_text(angle = 30,vjust = 1,hjust=1,size=8))

lib_info$condition <- ordered(lib_info$condition,levels=c("PBS","IFNg"))
ggplot(lib_info , aes(y=donor,x=Unique.Reads.Aligned..STAR.)) + geom_bar(stat="identity") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        axis.text.x = element_text(angle = 30,vjust = 1,hjust=1,size=8)) +
  facet_wrap(~condition)

```

```{r write_to_cluster, include=F}
# top 2500 genes (by variation across 28 samples)
n=rownames(mat.cpm.filtered.top)
write.table(n,file=paste0(Dir,"/R/celseq2_MEL_top2500_gene.txt"),col.names=NA)
write.table(mat.cpm.filtered.top,file=paste0(Dir,"/R/celseq2_MEL_top2500_cpm.txt"),quote=F,sep="\t",col.names=NA)

# loadings for all PCs
write.table(loadings,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_loadings.txt"),quote=F,sep="\t",col.names=NA)

# top genes in PC1:3
loadings.PC1=loadings %>% arrange(desc(PC1)) %>% dplyr::select(PC1)
write.table(loadings.PC1,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_loadings_PC1.txt"),quote=F,sep="\t",col.names=NA)
loadings.PC2=loadings %>% arrange(desc(PC2)) %>% dplyr::select(PC2)
write.table(loadings.PC2,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_loadings_PC2.txt"),quote=F,sep="\t",col.names=NA)
loadings.PC3=loadings %>% arrange(desc(PC3)) %>% dplyr::select(PC3)
write.table(loadings.PC3,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_loadings_PC3.txt"),quote=F,sep="\t",col.names=NA)

# metadata & top 4 PCs
write.table(d,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_metadata.txt"),quote=F,sep="\t",col.names=NA)
```
  
### heatmap with top 2500 variable genes
```{r heatmap}
heatmap.2(as.matrix(mat.cpm.filtered.log2.top), trace="none", density="none", col=bluered(75),
          Colv = T, Rowv=T, 
          dendrogram="column", 
          scale="row",
          # distance: 1 minus Pearson correlation distance with average linkage
          distfun=function(x) as.dist((1-cor(t(x)))/2), # as.dist() computes distance matrix. 
          srtCol=45, cexCol=0.40,  #labeling angle
          main="Melanocytes (IFNg stimulation)")
```


### generate edgeR normalized cpm matrix.  
```{r edgeR}
### edgeR normalized cpm
y_RNA <- DGEList(counts = mat.count, genes = rownames(mat.count))
y_RNA <- calcNormFactors(y_RNA)
y_RNA$samples
mat.edgeR <- cpm(y_RNA, log = F, prior.count = F) %>% as.data.frame()
mat.edgeR.filtered <- mat.edgeR %>% rowwise() %>% mutate(max=max(F25M_IFN:F63M_PBS)) %>% 
  as.data.frame() %>% magrittr::set_rownames(rownames(mat.edgeR)) %>%
  filter(max>=5) %>% dplyr::select(-max)
mat.edgeR.filtered <- mat.edgeR[rownames(mat.edgeR.filtered),]
mat.edgeR.PBS <- mat.edgeR.filtered %>% dplyr::select(contains(samples.PBS))
colnames(mat.cpm.PBS)<-sapply(colnames(mat.cpm.PBS),function(x) (as.matrix(unlist(strsplit(x,"_")))[1,])) %>% as.vector()
mat.edgeR.IFN <- mat.edgeR.filtered %>% dplyr::select(contains(samples.IFN))
colnames(mat.cpm.IFN)<-sapply(colnames(mat.cpm.IFN),function(x) (as.matrix(unlist(strsplit(x,"_")))[1,])) %>% as.vector()
mat.edgeR.log2FC <- log2(mat.edgeR.IFN+1) - log2(mat.edgeR.PBS+1)
colnames(mat.edgeR.log2FC) <- colnames(mat.edgeR.log2FC) %>% gsub("M.+","",.)
```

### PCA analysis
```{r PCA-log2fc-cpm}
# select top 2500 genes with most variable log2fc
mat.log2FC.lnsd <- apply(mat.cpm.log2FC,MARGIN=1,FUN=sd)
mat.log2FC.lncv <- sqrt( (exp(mat.log2FC.lnsd) )^2 -1) #coefficient of variance
# order by cv
mat.log2FC.ordered <- mat.cpm.log2FC[order(mat.log2FC.lncv, decreasing=T),]
# pick top 2500 genes, largest cv first
mat.log2FC.top = mat.log2FC.ordered[1:2500,]
mat.log2FC <- mat.log2FC.top

# PCA analysis
mat.pca.num <- t(mat.log2FC) %>% as.data.frame()
pca = prcomp(as.matrix(mat.pca.num),centered= T,scale = T )
loadings=as.data.frame(pca$rotation) %>% rownames_to_column("gene")

# percent variance in each pc
plot(pca)
pcs = pca$x 
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# top 2 PC loadings
loading.pc1 <- loadings[,c("gene","PC1")] %>% arrange(desc(PC1))
write.table(loading.pc1, file=paste0(Dir,"/16_donors/log2fc_pca_pc1_loadings.txt"), quote=F, sep="\t", row.names=F)
loading.pc2 <- loadings[,c("gene","PC2")] %>% arrange(desc(PC2))
write.table(loading.pc2, file=paste0(Dir,"/16_donors/log2fc_pca_pc2_loadings.txt"), quote=F, sep="\t", row.names=F)

# top 3 PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], donor=rownames(pcs))
d$donor <- d$donor %>% gsub("M","",.)
d <- inner_join(d,lib_info.log2fc,by="donor")

################# plots
# PC1 vs. PC2, not affected by PCR cycle
d %>%
  mutate(PCR.cycle=as.factor(PCR.cycle.number)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=PCR.cycle,label=donor)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'PCR.cycle.n') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))

# PC1 vs. PC2, by ancestry
d %>%
  mutate(genotype=as.factor(ancestry)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=genotype,label=donor)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'ancestry') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))

# PC2 vs. PC2, by lib prep date
d %>%
  mutate(genotype=as.factor(Lib.Prep.Date)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=genotype,label=donor)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'library prep date') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))
colnames(d)

```

```{r PCA-log2fc-edgeR}
# select top 2500 genes with most variable log2fc
mat.log2FC.lnsd <- apply(mat.edgeR.log2FC,MARGIN=1,FUN=sd)
mat.log2FC.lncv <- sqrt( (exp(mat.log2FC.lnsd) )^2 -1) #coefficient of variance
# order by cv
mat.log2FC.ordered <- mat.edgeR.log2FC[order(mat.log2FC.lncv, decreasing=T),]
# pick top 2500 genes, largest cv first
mat.log2FC.top = mat.log2FC.ordered[1:2500,]
mat.log2FC <- mat.log2FC.top

# PCA analysis
mat.pca.num <- t(mat.log2FC) %>% as.data.frame()
pca = prcomp(as.matrix(mat.pca.num),centered= T,scale = T )
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x 
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# top 3 PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], donor=rownames(pcs))
d$donor <- d$donor %>% gsub("M.+","",.)
d <- inner_join(d,lib_info.log2fc,by="donor")

################# plots
# PC1 vs. PC2, not affected by PCR cycle
d %>%
  mutate(PCR.cycle=as.factor(PCR.cycle.number)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=PCR.cycle,label=donor)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'PCR.cycle.n') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))

# PC1 vs. PC2, by ancestry
d %>%
  mutate(genotype=as.factor(ancestry)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=genotype,label=donor)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'ancestry') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))

# PC2 vs. PC2, by lib prep date
d %>%
  mutate(genotype=as.factor(Lib.Prep.Date)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=genotype,label=donor)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'library prep date') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))
colnames(d)
```

```{r cpm}
#################
this="PLA1A"
#log2FC.this<- mat.cpm.IFN %>% filter(rownames(mat.cpm.IFN)==this)
cpm.this <- mat.cpm.PBS %>% filter(rownames(mat.cpm.PBS)==this)
#genotype.this <- c("HomoRef","HomoAlt","HomoAlt","Het","Het","Het","HomoAlt","Het","HomoAlt","Het","HomoAlt","HomoAlt","Het","Het","Het","HomoAlt")
genotype.this <- colnames(log2FC.this) %>%
  gsub("F25M", "NorthernEuropean", .) %>%
  gsub("F27M", "NorthernEuropean", .) %>%
  gsub("F30M", "NorthernEuropean", .) %>%
  gsub("F31M", "Mediterranean", .) %>%
  gsub("F34M", "Mediterranean", .) %>%
  gsub("F41M", "Mediterranean", .) %>%
  gsub("F45M", "Mediterranean", .) %>%
  gsub("F46M", "Mediterranean", .) %>%
  gsub("F49M", "African", .) %>%
  gsub("F50M", "NorthernEuropean", .) %>%
  gsub("F51M", "Mediterranean", .) %>%
  gsub("F52M", "NorthernEuropean", .) %>%
  gsub("F55M", "African", .) %>%
  gsub("F56M", "Mediterranean", .) %>%
  gsub("F62M", "NorthernEuropean", .) %>%
  gsub("F63M", "African", .)

df.this <- rbind(colnames(cpm.this),cpm.this,genotype.this) %>% t() %>% as.data.frame() %>% magrittr::set_colnames(c("donor","cpm","genotype"))
df.this$genotype <- as.factor(df.this$genotype)
df.this$cpm <- as.numeric(as.character(df.this$cpm))

ggplot(df.this, aes(x=genotype,y=cpm,color=genotype)) + geom_point()
#Het=subset(df.this,df.this$genotype=="Het")
#HomoAlt=subset(df.this,df.this$genotype=="HomoAlt")
#wilcox.test(Het$cpm,HomoAlt$cpm)
African=subset(df.this,df.this$genotype=="African")
Mediterranean=subset(df.this,df.this$genotype=="Mediterranean")
NorthernEuropean=subset(df.this,df.this$genotype=="NorthernEuropean")
wilcox.test(African$cpm,NorthernEuropean$cpm)
wilcox.test(African$cpm,Mediterranean$cpm)


#################
# scatterplot
x="IRF4"
y="PLA1A"
#log2FC.this<- mat.cpm.IFN %>% filter(rownames(mat.cpm.IFN)==this)
cpm.this <- mat.cpm.PBS %>% filter(rownames(mat.cpm.PBS) %in% c(x,y))
#genotype.this <- c("HomoRef","HomoAlt","HomoAlt","Het","Het","Het","HomoAlt","Het","HomoAlt","Het","HomoAlt","HomoAlt","Het","Het","Het","HomoAlt")
genotype.this <- colnames(cpm.this) %>%
  gsub("F25M", "NorthernEuropean", .) %>%
  gsub("F27M", "NorthernEuropean", .) %>%
  gsub("F30M", "NorthernEuropean", .) %>%
  gsub("F31M", "Mixed", .) %>%
  gsub("F34M", "Mediterranean", .) %>%
  gsub("F41M", "Mixed", .) %>%
  gsub("F45M", "Mediterranean", .) %>%
  gsub("F46M", "Mixed", .) %>%
  gsub("F49M", "African", .) %>%
  gsub("F50M", "NorthernEuropean", .) %>%
  gsub("F51M", "Mixed", .) %>%
  gsub("F52M", "NorthernEuropean", .) %>%
  gsub("F55M", "African", .) %>%
  gsub("F56M", "Mediterranean", .) %>%
  gsub("F62M", "NorthernEuropean", .) %>%
  gsub("F63M", "African", .)


df.this <- rbind(colnames(cpm.this),cpm.this,genotype.this) %>% t() %>% as.data.frame() %>% magrittr::set_colnames(c("donor","cpm.x","cpm.y","genotype"))
df.this$genotype <- as.factor(df.this$genotype)
df.this$cpm.x <- as.numeric(as.character(df.this$cpm.x))
df.this$cpm.y <- as.numeric(as.character(df.this$cpm.y))

ggplot(df.this, aes(x=cpm.x,y=cpm.y,color=genotype)) + geom_point() + xlab(x) + ylab(y) + ggtitle("cpm, PBS samples")
cor(df.this$cpm.x,df.this$cpm.y,method="spearman")
#Het=subset(df.this,df.this$genotype=="Het")
#HomoAlt=subset(df.this,df.this$genotype=="HomoAlt")
#wilcox.test(Het$cpm,HomoAlt$cpm)
African=subset(df.this,df.this$genotype=="African")
Mediterranean=subset(df.this,df.this$genotype=="Mediterranean")
NorthernEuropean=subset(df.this,df.this$genotype=="NorthernEuropean")
wilcox.test(African$cpm,NorthernEuropean$cpm)
wilcox.test(African$cpm,Mediterranean$cpm)


###################
cSum <- apply(mat.count,2,sum)
mat.cpm <- 1000000*sweep(mat.count,MARGIN=2,FUN="/",STATS=cSum)

# filter out empty/low-expressed genes with < 5 cpm
mat.cpm$max = apply(mat.cpm, 1, function (x) max(x))
mat.cpm.filtered <- mat.cpm %>% rowwise() %>% mutate(max=max(F25M_IFNg:F55M_PBS)) %>% 
  as.data.frame() %>% magrittr::set_rownames(rownames(mat.cpm)) %>%
  filter(max>=5) %>% dplyr::select(-max)
mat.count.filtered <- mat.count[rownames(mat.cpm.filtered),]


write.table(mat.cpm, file=paste0(Dir,"/R/3donors.matrix.cpm.txt"), quote=F, sep="\t")
write.table(mat.cpm.filtered, file=paste0(Dir,"/R/3donors.matrix.cpm.filtered.txt"), quote=F, sep="\t")

```

### cpm vs. edgeR normalization
```{r cpm_vs_edgeR}
rm(list=setdiff(ls(), c("mat.cpm.log2FC","mat.edgeR.log2FC","lib_info.log2fc","Dir")))
common.genes <- intersect(row.names(mat.cpm.log2FC),row.names(mat.edgeR.log2FC))
mat.cpm.log2FC <- mat.cpm.log2FC[common.genes,]
mat.edgeR.log2FC <- mat.edgeR.log2FC[common.genes,]
n=16
donor=lib_info.log2fc$donor[n]
df=data.frame(gene=rownames(mat.cpm.log2FC),
              cpm=mat.cpm.log2FC[,donor],
              edgeR=mat.edgeR.log2FC[,donor],
              sig=abs(mat.cpm.log2FC[,donor])>1.5)
ggplot(df,aes(x=cpm,y=edgeR,color=sig)) + geom_point(alpha=0.4) +
  ggtitle(paste0("donor: ",donor," pcc: ",as.character(round(cor(df$cpm,df$edgeR,method="pearson"),3))))
```




## Genotype data.
```{r genotype}
library(vcfR)
library(cluster) # for gower similarity and pam
dir="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38"
vcf <- read.vcfR(paste0(dir,"/vcf/merged/bd841628-fcc2-487a-8460-f5428237f0c9.16donors.filtered2.pruned.vcf.gz"))
vcf_table <- as_tibble(cbind(vcf@fix,vcf@gt))

vcf_table3 <- data.frame(lapply(vcf_table, function(x) {gsub("0/0.+", "1", x)}))
vcf_table3 <- data.frame(lapply(vcf_table3, function(x) {gsub("1/1.+", "0", x)}))
vcf_table3 <- data.frame(lapply(vcf_table3, function(x) {gsub("1/0.+", "0.5", x)}))
vcf_table3 <- data.frame(lapply(vcf_table3, function(x) {gsub("0/1.+", "0.5", x)}))
vcf_table3 <- vcf_table3 %>% filter(!is.na(ID)) %>% unite(mergedID,c("ID","REF","ALT"),sep=":",remove=F)
vcf_table3 <- vcf_table3 %>% dplyr::select(-c("CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO","FORMAT")) %>% column_to_rownames("mergedID")
colnames(vcf_table3) <- colnames(vcf_table3) %>%
  gsub("\\.", "-", .) %>%
  gsub("X6d0b04d1-84e3-4979-8b8b-b93792da776f", "F25", .) %>%
  gsub("X149430e7-6d2d-4e50-baba-3e8f044ce2e3", "F27", .) %>%
  gsub("X5b67f1d5-29b4-448f-b9a3-d23a2d12cd93", "F30", .) %>%
  gsub("d8380c80-b025-4d9b-9d06-67ee06eac4ad",  "F31", .) %>%
  gsub("X5df7279b-bef9-44b3-a1c1-87b3318b79d4", "F34", .) %>%
  gsub("e7b548bc-bed5-4a30-b4d9-0151966e6b9d",  "F41", .) %>%
  gsub("f4a987ea-ca08-4b38-8668-48e5efe45076",  "F45", .) %>%
  gsub("f0b571af-c87e-4611-8288-bfcf4f9e9133",  "F46", .) %>%
  gsub("b0690c72-9cf6-465f-b5dc-e0b94fe8c8f3",  "F49", .) %>%
  gsub("edbb9163-65d7-4fc8-9d14-627d0987f087",  "F50", .) %>%
  gsub("X57948dd0-a324-4477-a917-c33f9c4e2daf", "F51", .) %>%
  gsub("X383fe24c-a10e-424e-abcf-29d1e2a61c5d", "F52", .) %>%
  gsub("X8f482258-bffd-40ca-be3a-487644f0c52f", "F55", .) %>%
  gsub("X29026b76-493d-4f87-9e3b-800e311c19ed", "F56", .) %>%
  gsub("X3e5e3f37-fe51-4fb7-a356-80d803c3db9b", "F62", .) %>%
  gsub("X413a4cbe-af1f-4cbb-a794-14c88b30da31", "F63", .) 
write.table(vcf_table3, file=paste0(dir,"/vcf/bd841628-fcc2-487a-8460-f5428237f0c9.16donors.pruned.phased_vcf_table3.txt"), row.names=T, quote=F)
vcf_table3 <- read.table(file=paste0(dir,"/vcf/bd841628-fcc2-487a-8460-f5428237f0c9.16donors.pruned.phased_vcf_table3.txt"), sep="\t")

# transpose vcf2_table_select
vcf_table3_transposed <- as_tibble(cbind(names=colnames(vcf_table3),t(vcf_table3)))
# turn values into numeric
vcf_table3_transposed[,2:ncol(vcf_table3_transposed)] <-
  apply(vcf_table3_transposed[,2:ncol(vcf_table3_transposed)], 2, function(x) as.numeric(as.character(x)))
manhattan_dist <- daisy(vcf_table3_transposed[, -1], metric = "manhattan")
aggl.clust.c.manhattan <- hclust(manhattan_dist, method = "complete")
aggl.clust.c.manhattan$labels=as.character(vcf_table3_transposed$names)

manhattan_cluster_cut3<- cutree(aggl.clust.c.manhattan, k=3)
donors_manhattan_cluster_cut3 <- tibble(vcf_table3_transposed$names,manhattan_cluster_cut3) %>% arrange(manhattan_cluster_cut3)
print(donors_manhattan_cluster_cut3, n=Inf)
write.table(donors_manhattan_cluster_cut3, file=paste0(dir,"/16donors.donors_manhattan_cluster_cut3.txt"), quote=F, sep="\t")

manhattan_cluster_cut3<- cutree(aggl.clust.c.manhattan, k=4)
donors_manhattan_cluster_cut3 <- tibble(vcf_table3_transposed$names,manhattan_cluster_cut3) %>% arrange(manhattan_cluster_cut3)
print(donors_manhattan_cluster_cut3, n=Inf)
write.table(donors_manhattan_cluster_cut3, file=paste0(dir,"/16donors.donors_manhattan_cluster_cut4.txt"), quote=F, sep="\t")

#pdf(paste0(dir,"/16donors.manhattan.clust.complete.pdf"))
plot(aggl.clust.c.manhattan, main = "Agglomerative, complete linkages")
#dev.off()
write("plotting done", stdout())
Sys.time()

```

```{r pca-variants}
# select top 2500 variants with most variable genotypes
vcf_table3.lnsd <- apply(vcf_table3,MARGIN=1,FUN=sd)
vcf_table3.lncv <- sqrt( (exp(vcf_table3.lnsd) )^2 -1) #coefficient of variance
# order by cv
vcf_table3.ordered <- vcf_table3[order(vcf_table3.lncv, decreasing=T),]
# pick top 2500 genes, largest cv first
vcf_table3.top = vcf_table3.ordered[1:2500,]

# PCA analysis
vcf_table3.t <- as.data.frame(t(vcf_table3.top))
vcf_table3.t <- mutate_all(vcf_table3.t, function(x) as.numeric(as.character(x)))
pca = prcomp(as.matrix(vcf_table3.t),centered= T,scale = T )

# loadings in each pc
loadings=as.data.frame(pca$rotation)
# percent variance in each pc
plot(pca)
pcs = pca$x 
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# top 3 PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], name=colnames(vcf_table3.top))
ancestry <- data.frame(name=c("F27","F30","F31","F34",
                              "F41","F45","F46","F49",
                              "F50","F51","F52","F55",
                              "F56","F25","F62","F63"),
                       ancestry=c("NorthernEuropean","NorthernEuropean","Mixed","Mediterranean",
                                  "Mixed","Mediterranean","Mixed","African",
                                  "NorthernEuropean","Mixed","NorthernEuropean","African",
                                  "Mediterranean","NorthernEuropean","NorthernEuropean","African"))
d=inner_join(d,ancestry,by="name") %>% magrittr::set_rownames(d$name)
cols <- c("NorthernEuropean" = "blue", "Mediterranean" = "orange", "Mixed" = "#990000", "African" = "#669900")

# PC1 vs. PC2, by donor vitiligo-GWAS genotype cluster
d %>%
  mutate(genotype=as.factor(ancestry)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=ancestry,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  scale_colour_manual(values = cols)
```

```{r qc-metric}
# allele frequency distribution, INDELs
df <- read.table(file="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38/vcf/merged/all.indels.AF.txt",header=F) %>% magrittr::set_colnames("AF")
df$AF <- df$AF * 100
ggplot(df,aes(x=AF)) + 
  geom_histogram(fill=I("blue"),color=I("white"),alpha=I(0.5),binwidth=5) + 
  xlab("Allele frequency (%)") + ylab("number of INDELs") +
  theme_minimal()

# allele frequency distribution, SNVs
df <- read.table(file="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38/vcf/merged/all.snps.AF.txt",header=F) %>% magrittr::set_colnames("AF")
df$AF <- df$AF * 100
df2 <- df %>% filter(AF>=3.44) %>% filter(AF<50) 
ggplot(df,aes(x=AF)) + 
  geom_histogram(fill=I("blue"),color=I("white"),alpha=I(0.5),binwidth=5) + 
  xlab("Allele frequency (%)") + ylab("number of SNVs") +
  theme_minimal()

# substitution patterns
df <- read.table(file="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38/vcf/merged/all.snps.genotypes.txt",header=F)
id_lookup <- read.table(file="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38/id_lookup.txt",header=F) %>% magrittr::set_colnames(c("donor","id"))
id_lst <- read.table(file="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38/vcf/merged/all.snps.vcf.colnames",header=F) %>% magrittr::set_colnames("id")
donor_lst <- id_lookup %>% dplyr::filter(id %in% id_lst$id) %>% dplyr::select(donor)
df.colnames <- c("REF","ALT",donor_lst$donor)
colnames(df) <- df.colnames
snv <- paste(df$REF,">",df$ALT) %>% as.data.frame() %>% magrittr::set_colnames("pattern")
snv$pattern <- as.factor(snv$pattern)
percentage <- table(snv$pattern) %>% as.data.frame() %>% magrittr::set_colnames(c("pattern","freq")) %>% mutate(pct=freq/nrow(snv))
ggplot(percentage,aes(x=pattern,y=pct)) + 
  geom_bar(stat="identity",fill=I("blue"),color=I("black"),alpha=I(0.5)) + 
  geom_text_repel(aes(label = round(pct,2)),size = 3.5,direction = "y",nudge_y=-1,segment.color = NA) +
  xlab("substitution types") + ylab("percentage") +
  theme_minimal()

# Ts/Tv and Het/HomoALT ratio by sample
dir="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38/vcf/organized"
files <- list.files(path=dir,pattern = "summary")
df <- data.frame()
for (f in files) {
  df1 <- read.table(file=paste0(dir,"/",f),header=F)
  df <- rbind(df,df1)
}
colnames(df) <- c("donor","Ts","Tv","HomoREF","Het","HomoALT")
df.rownames <- df$donor
df <- df %>% dplyr::select(-donor) %>% mutate(across(where(is.character), as.numeric)) %>% magrittr::set_rownames(df.rownames)
df$Ts_vs_Tv <- df$Ts / df$Tv
df$Het_vs_HomoALT <- df$Het / df$HomoALT
df$Het_vs_HomoREF <- df$Het / df$HomoREF

ggplot(df,aes(x=rownames(df))) +
  geom_point(aes(y=Ts_vs_Tv,color="a"),shape=7,size=3) +
  geom_point(aes(y=Het_vs_HomoALT,color="b"),shape=7,size=3) +
  geom_point(aes(y=Het_vs_HomoREF,color="c"),shape=7,size=3) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        legend.position="top",axis.text.x = element_text(angle = 30,vjust = 1,hjust=1,size=6)) +
  xlab("samples") + ylab("ratio") + 
  scale_color_manual(values=c("a"="orange","b"="purple","c"="green"),
                     labels=c("Ts/Tv ratio","Het/HomoALT ratio","Het/HomoREF ratio"),
                     name="")

# Counts by sample
dir="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38/vcf/organized"
files <- list.files(path=dir,pattern = "counts")
df <- data.frame()
for (f in files) {
  donor <- unlist(strsplit(f,"[.]"))[2]
  df1 <- read.table(file=paste0(dir,"/",f),header=F) %>% as.data.frame()
  n.all <- df1[6,4]
  n.indel <- df1[3,4]
  n.snv <- df1[2,4]
  n.lst <- c(donor,n.all,n.indel,n.snv)
  df <- rbind(df,n.lst)
}
colnames(df) <- c("donor","total_variants","INDELs","SNVs")
df.rownames=df$donor
df <- df %>% dplyr::select(-donor) %>% mutate(across(where(is.character), as.numeric)) %>% magrittr::set_rownames(df.rownames)
df <- df %>% arrange(desc(total_variants))
note1=paste0("total: min=",format(min(df$total_variants), big.mark=","),";  max=",format(max(df$total_variants), big.mark=","))
note2=paste0("SNVs: min=",format(min(df$SNVs), big.mark=","),";  max=",format(max(df$SNVs), big.mark=","))

ggplot(df,aes(x=reorder(rownames(df),total_variants))) +
  geom_point(aes(y=total_variants,color="a"),shape=7,size=3) +
  geom_point(aes(y=SNVs,color="b"),shape=7,size=3) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        legend.position="top",axis.text.x = element_text(angle = 30,vjust = 1,hjust=1,size=6)) +
  xlab("samples") + ylab("count") + 
  scale_color_manual(values=c("a"="orange","b"="green"),
                     labels=c("total variants","SNVs"),
                     name="") +
  labs(title = "variant counts per sample", caption = paste0(note1,"\n",note2))

```

### Data preparation.
```{r metadata}
# metadata
summary <- read.table(file=paste0(Dir,"/DolphinNext_052821/report4643/summary/","overall_summary.tsv"),
                 sep="\t",header=T,row.names=1) %>% rownames_to_column("name")
summary <- tidyr::separate(summary, col=name, into=c("bc","donor","condition"),sep="_",remove=F) %>%
 unite("name", donor:condition, sep = "_", remove = T) %>%
 dplyr::select(-bc)
lib_info <- read.table(file=paste0(Dir,"/R/celseq2_MEL_libinfo.txt"),header=F)
colnames(lib_info) <- c("name","barcode","barcode.factor","Pooled.Library.Fragment.Size","Pooled.Lib.Size.BioAnalyzer(bp)",
                           "Fragment.Size.Matched","Pooled.library.Total.Reads","Lib.Prep.Batch","Lib.Prep.Date",
                           "reads_ESAT","pct.reads","PCR.cycle.n","Lib Conc.Qubit(ng/μl)","Genotype.Cluster")
metadata <- data.frame(key=colnames(mat)[3:ncol(mat)]) %>%
  tidyr::separate(. , col=key, into=c("donor","condition"),sep="_",remove=F) %>%
  column_to_rownames(var="key")
library <- data.frame(donor=c("F27M","F34M","F30M","F31M","F41M","F45M","F46M","F47M","F49M","F50M","F51M","F52M","F55M","F56M"),
                    library=c("1","1","2","2","3","3","4","4","5","5","6","6","7","7"))
genotype.cluster <- data.frame(donor=c("F27M","F30M","F34M","F45M","F50M","F52M","F56M","F31M","F41M","F46M","F49M","F51M","F55M"),
                               cluster=c("1","1","1","1","1","1","1","2","3","4","5","6","7"))
metadata <- metadata %>% mutate(name=rownames(metadata))
metadata <- inner_join(metadata,library,by="donor")
metadata <- inner_join(metadata,genotype.cluster,by="donor")
metadata <- inner_join(metadata,summary,by="name")
metadata <- inner_join(metadata,lib_info,by="name") %>% magrittr::set_rownames(metadata$name)
metadata$condition <- factor(metadata$condition, levels=c("PBS","IFNg"))

# clean up
#rm(mat,mat.cpm)
rm(lib_info,summary,cSum)
```



  
### k-means clustering with top 2500 variable genes
```{r kmc, include=F}
k = 5;
source = t(scale(t(mat.cpm.filtered.top)));
kmeans = kmeans(source,centers = k,iter.max = 1000,algorithm="Lloyd");
clustered.e.eif4e = data.frame();
breaks = c();
for (i in 1:k) {
  print(sprintf("Doing cluster %i",i))
  cluster = source[kmeans$cluster==i,];
  rows = row.names(cluster);
  clust = heatmap.2(as.matrix(cluster), distfun=function(x) as.dist((1-cor(t(x)))/2),dendrogram = "none",srtCol=45,cexCol=0.4)
  cluster.ordered = cluster[clust$rowInd,];
  cluster.ordered.genes = rows[clust$rowInd];
  row.names(cluster.ordered) = cluster.ordered.genes;
  class = data.frame(row.names=cluster.ordered.genes);
  class$class = i 
  cluster.ordered = cbind(cluster.ordered, class);
  clustered.e.eif4e = rbind(clustered.e.eif4e, cluster.ordered);
  if(i > 1 & i < k) {
    breaks[i] = as.numeric(breaks[i-1]) + length(rows);
  } else if(i==1) {
    breaks[i] = length(rows);
  }
}

dims = dim(clustered.e.eif4e);

png(filename=paste0(Dir,"/R/plots/celseq2_MEL_top2500_kmc_heatmap.png"),width=2000,height=1500)
par(mar=c(10,4,10,4)+0.1)
heatmap.2(as.matrix(clustered.e.eif4e[1:100,1:dims[2]-1]), trace="none", density="none", 
          col=bluered(30), Colv = T,Rowv=F,dendrogram = "column",
          distfun=function(x) as.dist((1-cor(t(x)))/2),rowsep=breaks, sepcol="black",sepwidth=c(1,1),
          srtCol=40, cexCol=1, offsetCol=-0.4,
          cexRow=0.8,offsetRow=-0.3,
          main="Top 2500 genes Full clustering")
dev.off()
```
  
### heatmaps for Genes of interest (Identified by Ken) --> not done yet
```{r ken_genes, include=F}
ken_genes <- unlist(read.csv("pilot1_ken_genes.csv", header=FALSE), use.names=FALSE)
ken_genes2 <- ken_genes[ken_genes %in% rownames(mat.cpm.filtered.log2)]
#GPMNB (and other names for it) - not found in dataset

heatmap.2(as.matrix(mat.cpm.filtered.log2[ken_genes2,]), trace="none", density="none", col=bluered(75),
          Colv = T, Rowv=T, dendrogram="column", scale="row",
          distfun=function(x) as.dist((1-cor(t(x)))/2),
          srtCol=45, cexCol=0.4, main="Selected Genes, All samples (IFN stimulation)")

```

### DESeq2
```{r deseq2, include=F}
dds <- DESeqDataSetFromMatrix(countData=as.matrix(mat.count.filtered),
                               colData=metadata, 
                               design = ~ condition)
dr <- DESeq(dds);
res <- results(dr, contrast=c("condition","PBS","IFNg"));
res_tmp <- res[!is.na(res$padj) & !is.na(res$log2FoldChange),] %>% as.data.frame()
res_p01 <- res_tmp[(res_tmp$padj<0.01),] %>% as.data.frame()
View(as.data.frame(res_tmp))
```











## ANALYSIS PART TWO (DESeq2) 
This may not work because the DESeq2 calculates one log2FC value per gene, considering all 14 samples
### DESeq2 analysis

```{r DESeq2}
mat <- mat[,-c(1,2)]
dds <- DESeqDataSetFromMatrix(countData=mat,colData=metadata,design=~condition)
# pre-filtering
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
# re-factor
dds$condition <- relevel(dds$condition, ref="PBS")
# differential expression analysis
dds <- DESeq(dds)
res <- results(dds, name="condition_IFNg_vs_PBS")
mcols(res)$description
# log fold change shrinkage for visualization and ranking
resLFC <- lfcShrink(dds, coef="condition_IFNg_vs_PBS",type="apeglm")
resLFC
# MA plot & identify genes of interest
plotMA(resLFC)
idx <- identify(res$baseMean, res$log2FoldChange)
rownames(res)[idx]
# plot counts
g="CD74"
plotCounts(dds,gene=g,intgroup="condition")
# ^ another way of plotting:
d<-plotCounts(dds,gene=g,intgroup="condition",returnData=T)
ggplot(d, aes(x=condition, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) + 
  scale_y_log10(breaks=c(25,100,400))

saveRDS(dds, file=paste0(Dir,"/R/celseq2_",celltype,"_","DESeq2_dds.rds"))
saveRDS(res, file=paste0(Dir,"/R/celseq2_",celltype,"_","res.rds"))
saveRDS(resLFC, file=paste0(Dir,"/R/celseq2_",celltype,"_","resLFC.rds"))
```


## ANALYSIS PART THREE (PCA OF LOG2_FC_CPM)
### prep log2fc
```{r prep3}
mat.PBS <- mat.count.filtered %>% dplyr::select(ends_with("PBS"))
colnames(mat.PBS)<-sapply(colnames(mat.PBS),function(x) (as.matrix(unlist(strsplit(x,"_")))[1,])) %>% as.vector()
mat.IFN <- mat.count.filtered %>% dplyr::select(ends_with("IFNg"))
colnames(mat.IFN)<-sapply(colnames(mat.IFN),function(x) (as.matrix(unlist(strsplit(x,"_")))[1,])) %>% as.vector()

mat.log2FC <- log2(mat.IFN+1) - log2(mat.PBS+1)
write.table(mat.log2FC, file=paste0(Dir,"/R/3donors.matrix.log2FC.filtered.txt"), quote=F, sep="\t")

mat.PBS <- mat.count %>% dplyr::select(ends_with("PBS"))
colnames(mat.PBS)<-sapply(colnames(mat.PBS),function(x) (as.matrix(unlist(strsplit(x,"_")))[1,])) %>% as.vector()
mat.IFN <- mat.count %>% dplyr::select(ends_with("IFNg"))
colnames(mat.IFN)<-sapply(colnames(mat.IFN),function(x) (as.matrix(unlist(strsplit(x,"_")))[1,])) %>% as.vector()

mat.log2FC <- log2(mat.IFN+1) - log2(mat.PBS+1)
write.table(mat.log2FC, file=paste0(Dir,"/R/3donors.matrix.log2FC.txt"), quote=F, sep="\t")
#metadata.log2fc <- metadata %>% 
#  dplyr::select(c(donor,library,PCR.cycle.n,cluster,Genotype.Cluster)) %>% 
#  dplyr::distinct() %>% magrittr::set_rownames(.$donor)



mat.log2FC <- rownames_to_column(mat.log2FC,var="gene")
```

### PCA
```{r pca-plots}
# PC1 vs. PC2, by PCR cycle
d %>%
  mutate(PCR.cycle=as.factor(PCR.cycle.n)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=PCR.cycle,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'PCR.cycle.n') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  scale_color_manual(values = c("13" = "red",
                              "10"="blue")) 

# PC1 vs. PC2, by donor vitiligo-GWAS genotype cluster
d %>%
  mutate(genotype=as.factor(Genotype.Cluster)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=genotype,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'genotype') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))

# PC2 vs. PC2, by donor genotype cluster (of all variants with LD <= 0.8)
d %>%
  mutate(genotype=as.factor(cluster)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=genotype,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'genotype') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))
colnames(d)

```

```{r rs12203592}
g="RERE"
g1 <- mat.log2FC.ordered %>% rownames_to_column("gene") %>% filter(gene=="IRF4") %>% t() %>% as.data.frame()
g2 <- mat.log2FC.ordered %>% rownames_to_column("gene") %>% filter(gene=="TMEM140") %>% t() %>% as.data.frame()
g3 <- mat.log2FC.ordered %>% rownames_to_column("gene") %>% filter(gene=="PLA1A") %>% t() %>% as.data.frame()
g4 <- mat.log2FC.ordered %>% rownames_to_column("gene") %>% filter(gene=="NEO1") %>% t() %>% as.data.frame()
g <- cbind(g1,g2,g3,g4)
colnames(g) <- g[1,]
g <- g[2:14,]

g <- data.frame(name=c("F27","F30","F31","F34","F41","F45",	"F46",	"F49",	"F50",	"F51",	"F52",	"F55",	"F56",	"F25",	"F62",	"F63"),
                genotype=c("HomoAlt",	"HomoAlt",	"Het",	"Het",	"Het",	"HomoAlt",	"Het",	"HomoAlt",	"Het",	"HomoAlt",	"HomoAlt",	"Het",	"Het",	"HomoRef",	"Het",	"HomoAlt"))
g$IRF4 <- as.numeric(g$IRF4)
g$TMEM140 <- as.numeric(g$TMEM140)
g$PLA1A <- as.numeric(g$PLA1A)
g$NEO1 <- as.numeric(g$NEO1)

ggplot(g,aes(y=IRF4,x=genotype,col=genotype)) + geom_point()
ggplot(g,aes(y=TMEM140,x=genotype,col=genotype)) + geom_point()
ggplot(g,aes(y=PLA1A,x=genotype,col=genotype)) + geom_point()
ggplot(g,aes(y=NEO1,x=genotype,col=genotype)) + geom_point()


g.rand <-mat.PBS%>% rownames_to_column("gene") %>% filter(gene=="IRF4") %>% t() %>% as.data.frame()
colnames(g.rand) <- g.rand[1,]
g.rand <- g.rand[2:14,] %>% as.data.frame()
g.rand$genotype <- as.factor(c("NorthernEurope","NorthernEurope","Mixed","Mixed","African","MediterraneanEurope","African","Mixed","MediterraneanEurope","MediterraneanEurope","NorthernEurope","NorthernEurope","NorthernEurope"))
colnames(g.rand) <- c("IRF4","genotype")
g.rand$IRF4 <- as.numeric(g.rand$IRF4)
ggplot(g.rand,aes(y=IRF4,x=genotype,col=genotype)) + geom_point()
```

```{r clustering}
k = 5;
source = t(scale(t(mat.log2FC.ordered)));
kmeans = kmeans(source,centers = k,iter.max = 1000,algorithm="Lloyd");
clustered.e.eif4e = data.frame();
breaks = c();
for (i in 1:k) {
  print(sprintf("Doing cluster %i",i))
  cluster = source[kmeans$cluster==i,];
  rows = row.names(cluster);
  clust = heatmap.2(as.matrix(cluster), distfun=function(x) as.dist((1-cor(t(x)))/2),dendrogram = "none",srtCol=45,cexCol=0.4)
  cluster.ordered = cluster[clust$rowInd,];
  cluster.ordered.genes = rows[clust$rowInd];
  row.names(cluster.ordered) = cluster.ordered.genes;
  class = data.frame(row.names=cluster.ordered.genes);
  class$class = i 
  cluster.ordered = cbind(cluster.ordered, class);
  clustered.e.eif4e = rbind(clustered.e.eif4e, cluster.ordered);
  if(i > 1 & i < k) {
    breaks[i] = as.numeric(breaks[i-1]) + length(rows);
  } else if(i==1) {
    breaks[i] = length(rows);
  }
}

dims = dim(clustered.e.eif4e);

#png(filename=paste0(Dir,"/R/plots/celseq2_MEL_top2500_kmc_heatmap.png"),width=2000,height=1500)
par(mar=c(10,4,10,4)+0.1)
heatmap.2(as.matrix(clustered.e.eif4e[1:100,1:dims[2]-1]), trace="none", density="none", 
          col=bluered(30), Colv = T,Rowv=F,dendrogram = "column",
          distfun=function(x) as.dist((1-cor(t(x)))/2),rowsep=breaks, sepcol="black",sepwidth=c(1,1),
          srtCol=40, cexCol=1, offsetCol=-0.4,
          cexRow=0.8,offsetRow=-0.3,
          main="Top 2500 genes Full clustering")
#dev.off()
heatmap.2(as.matrix(mat.log2FC.ordered[1:2500,]), trace="none", density="none", col=bluered(75),
          Colv = T, Rowv=T, 
          dendrogram="column", 
          scale="row",
          # distance: 1 minus Pearson correlation distance with average linkage
          distfun=function(x) as.dist((1-cor(t(x)))/2), # as.dist() computes distance matrix. 
          srtCol=40, cexCol=1, offsetCol=-0.4,
          cexRow=0.8,offsetRow=-0.3,
          main="Top 2500 genes")

```

### glmnet model on log2fc -- Pranitha old script
```{r glmnet_Pranitha}
require("VGAM")
library(tibble, lib.loc = "/usr/local/R-3.6.1/lib64/R/library")
grDevices::pdf.options(useDingbats = FALSE)
require(data.table)
require(foreign)
require(ggplot2)
require(MASS)
require(Hmisc)
library("nnet")
require(reshape2)
require(dplyr,lib.loc = "/usr/local/R-3.6.1/lib64/R/library")
require(caret)
require(tidyr)
library(XML)
theme_set(theme_bw())
source("/project/umw_garberlab/vangala/bin/custom_functions_fromDB.R")
options(stringsAsFactors = F)
load("/nl/umw_manuel_garber/mouse/DC/SIP/merged_SIP_H3K4me3/perElement_interactions2-100/perLocus_5k/perLocus/geneReg_models/regression/SeqGL/SeqGL_sip_1h_FC_atacDyn_5k_heldout.rda")

######## training and testing #########
train=k[k$held=="train",]
rownames(train)=train$gene
test=k[k$held=="test",]
rownames(test)=test$gene
x.indep=grep("Group",colnames(k),value = T)
#x.indep=grep("not|early|down",x.indep,value = T,invert = F)
tuneGrid=expand.grid(
  .alpha=seq(0,10, by = 0.5),
  .lambda=seq(0,10, by = 0.5))
trainControl = trainControl(method="repeatedcv", number=10,allowParallel = T,repeats=100)
############################## 1hour Elastic Net Regression #########################
y.dep="log2fc_1h"
#ridgeGrid= data.frame(.lambda =seq(0,1, length =5))
cluster=parallel::makeCluster(29) # convention to leave 1 core for OS
doParallel::registerDoParallel(cluster)
modelFit_early <- train((log2fc_1h) ~., data = train[,c(y.dep,x.indep)],
                        method = "glmnet",
                        #preProc=c("center","scale"),
                        trControl = trainControl,
                        metric = "R2", # Optimize by F-measure
                        #alpha=0,
                        tuneGrid=tuneGrid,
                        #tuneLength=20#,
                        #MaxNWts=5000)
)
parallel::stopCluster(cluster)
foreach::registerDoSEQ()
modelFit_early$bestTune
test$pred = predict(modelFit_early,test)
# Model performance metrics
caret::postResample(test$pred,test$log2fc_1h)
#k$logpred=log2(k$pred)
plot(test$log2fc_1h,test$pred)
#cor(k$logpred,k$log2fc_1h,method = "spearman")
rho=cor(test$pred,test$log2fc_1h)
rho
r2=caret::postResample(test$pred,test$log2fc_1h)[2]
r2
#minl=min(c(k$log2fc_1h,k$pred))
#maxl=max(c(k$log2fc_1h,k$pred))
k$pred_early = predict(modelFit_early,k)
k=k[order(k$Clusters,decreasing = T),]
k$held="train"
k$held[rownames(k)%in%testRow]="test"
ggplot(k,aes(log2fc_1h,pred_early)) +
  geom_point(position=position_jitter(w=1),aes(color=Clusters))+ #scale_x_log10()+scale_y_log10() + 
  geom_smooth(method = "lm", se = FALSE,color="black",size=0.4)+
  ggtitle(label = paste("corr is", format(rho, digits=4),"R2 is", format(r2, digits=2)))+
  scale_color_brewer(palette="Set1")
ggplot(test,aes(log2fc_1h,pred)) +
  geom_point(position=position_jitter(w=1),aes(color=Clusters))+ #scale_x_log10()+scale_y_log10() + 
  geom_smooth(method = "lm", se = FALSE,color="black",size=0.4)+
  ggtitle(label = paste("corr is", format(rho, digits=4),"R2 is", format(r2, digits=2)))+
  scale_color_brewer(palette="Set1")

```

### glmnet model on log2fc -- my take
```{r glmnet_Crystal}



```














## Archived code
```{r cpm_variance_vs_mean}
df <- mat.cpm.filtered
### generate stats (mean, std, cv) of log2-transformed data
df.log2 <- log2(df+1)
df.mean <- log2(apply(df,MARGIN=1,FUN=mean)+1)
df.lnsd <- apply(df.log2,MARGIN=1,FUN=sd)
df.lncv <- sqrt((exp(df.lnsd) )^2 -1) # coefficient of variance

# order by cv
df.log2.ordered <- df.log2[order(df.lncv,decreasing=T),]
df.ordered <- df[order(df.lncv,decreasing=T),]

# pick top 2500 genes, largest cv first
df.log2.top = df.log2.ordered[1:2500,]
df.top = df.ordered[rownames(df.log2.top),]

# plot: mean vs. CV
df.plot <- data.frame(mean=df.mean,
                 cv=df.lncv)
ggplot(df.plot,aes(x=mean,y=cv)) + geom_point()
```
```