---
title: "skin-eQTL_celseq2_analysis"
author: "Shuo Shan"
date: "8/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages  
```{r package, warning=FALSE}
library(dplyr)
library(tidyverse)
library(DESeq2)
#library(DEXSeq)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
library(edgeR)
library(gridExtra)
library(ggpubr)
library(plotly)
library(RCurl)
library(devtools)
library(ComplexHeatmap)

Dir="~/Downloads/nl/human/skin/eQTLs/RNA-Seq"
setwd(Dir)

#save.image(file=paste0(Dir,'/myEnvironment_allcts.RData'))
load(paste0(Dir,'/myEnvironment_allcts.RData'))


#MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
#rm(list=setdiff(ls(), c("Dir","counts_KRT","counts_MEL","counts_FRB","metadata_KRT","metadata_MEL","metadata_FRB")))

```
  
### Overview
Celseq2 data for all 3 cell types. 
```{r countmatrix}
counts_FRB$gene=rownames(counts_FRB)
counts_MEL$gene=rownames(counts_MEL)
counts_KRT$gene=rownames(counts_KRT)
counts=left_join(left_join(counts_MEL,counts_KRT,by="gene"),counts_FRB,by="gene")
rownames(counts)=counts$gene
counts=counts %>% dplyr::filter(!gene %in% MTgenes$V1) %>% dplyr::select(-"gene") # remove mitochrondrial genes
```

### logCPM calculation
  
```{r logCPM, echo=FALSE, message=F}
col.sum=apply(counts,2,sum)
CPM=sweep(counts, 2, col.sum, `/`)*1000000


col.sum=apply(counts_MEL,2,sum)
CPM=sweep(counts_MEL, 2, col.sum, `/`)*1000000

df.skintone=data.table::fread(paste0(Dir,"/donor_skintone.txt"))
metadata=rbind(metadata_FRB,metadata_KRT,metadata_MEL) %>% distinct_at(.,"sample",.keep_all=TRUE) %>% left_join(. , df.skintone, by="donor")
metadata.pbs <- metadata %>% dplyr::filter(condition=="PBS")
metadata.ifn <- metadata %>% dplyr::filter(condition=="IFN")

# Getting normalized CPM matrix
CPM.pbs <- as.data.frame(CPM)[,as.character(metadata.pbs$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.pbs$biosample)) 
CPM.ifn <- as.data.frame(CPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$biosample)) 

# Getting log transformed CPM
logCPM <- log2(CPM+1) %>% as.data.frame() 

logCPM.pbs <- as.data.frame(logCPM)[,as.character(metadata.pbs$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.pbs$donor)) 

logCPM.ifn <- as.data.frame(logCPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) 

log2FC <- logCPM.ifn - logCPM.pbs

# inspect distribution of log-transformed CPM:
mean(apply(logCPM.pbs, 2, sd))
mean(apply(logCPM.pbs, 2, mean))
plot(density(logCPM.pbs[,1]))
```


### K-means-clustering & hierarchical clustering
```{r kmclusters}
df <- logCPM
n=20 # top number of genes with most abs(log2FC)

# # set of candidate genes for clustering
# candidate_genes <- res %>% as.data.frame %>%
#   dplyr::filter(padj < 0.05) %>%                   # restrict FDR
#   dplyr::filter(abs(log2FoldChange) >= 1.5) %>%    # restrict effect size
#   dplyr::top_n(. , n, abs(log2FoldChange)) %>%     # restrict top FC response
#   pull(gene) %>%
#   unique()

# another way of combining candidate genes
candidate_genes_up <- res %>% as.data.frame %>%
  dplyr::filter(padj < 0.05) %>%                   # restrict FDR
  dplyr::filter(log2FoldChange >= 1.5) %>%    # restrict effect size
  dplyr::top_n(. , n/2, abs(log2FoldChange)) %>%     # restrict top FC response
  pull(gene) %>%
  unique()
candidate_genes_down <- res %>% as.data.frame %>%
  dplyr::filter(padj < 0.05) %>%                   # restrict FDR
  dplyr::filter(log2FoldChange <= 1.5) %>%    # restrict effect size
  dplyr::top_n(. , n/2, abs(log2FoldChange)) %>%     # restrict top FC response
  pull(gene) %>%
  unique()
candidate_genes <- c(candidate_genes_up,candidate_genes_down)

# filter data matrix based on candidate genes
df <- df %>% dplyr::filter(rownames(.) %in% candidate_genes)
  
# perform k-means-clustering to determine k_row.
# wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
# for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
# plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# hierarchical clustering to check number of genes that fall in clusters.
gene_hclust <- hclust(dist(df,method="euclidean"), method = "complete")
cutree(gene_hclust, k = 2) %>% 
  enframe() %>% 
  rename(gene=name,cluster=value) %>%
  tidyr::separate(.,gene,into=c("gene","response"),sep="_") %>%
  group_by(response,cluster) %>%
  summarise(n=n())

# heatmap
heatmaply(df, 
          k_row = 2,
          scale = "row",
          fontsize_row = 5,
          fontsize_col = 5,
          col_side_colors = metadata[,c("condition","batch")],
          main = "row-scaled log2 TMM-normalized CPM",
          file = c(paste0(Dir,"/heatmaply_log2TMM-CPM_top20.html")))

```
```{r heatmap}
dge <- DGEList(counts)
dge <- calcNormFactors(dge)
CPM <- cpm(dge, log=FALSE)
logCPM <- log2(CPM+1) %>% as.data.frame() 

# heatmap of top 2000 most variable genes based on CoV
mat.logCPM.lnsd <- apply(logCPM,MARGIN=1,FUN=sd)
mat.logCPM.lncv <- sqrt( (exp(mat.logCPM.lnsd) )^2 -1) #coefficient of variance
logCPM.top1k <- logCPM[order(mat.logCPM.lncv,decreasing=T),][1:1000,] 

# filter data matrix based on candidate genes
df <- logCPM.top1k

# perform k-means-clustering to determine k_row.
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")

# hierarchical clustering to check number of genes that fall in clusters.
gene_hclust <- hclust(dist(df,method="euclidean")^2, method = "cen")
gene_clusters <- cutree(gene_hclust, k = 5) %>%
  enframe() %>%
  rename(gene=name,cluster=value)

# plot heatmap
library(ComplexHeatmap)
df.scaled=scale(df)
celltype_anno=sapply(strsplit(as.character(colnames(df.scaled)),split="_"),"[[",2)
condition_anno=sapply(strsplit(as.character(colnames(df.scaled)),split="_"),"[[",3)


gene_hclust <- hclust(dist(df,method="euclidean")^2, method = "cen")
sample_hclust <- hclust(dist(t(df),method="euclidean")^2,method="cen")
pdf(paste0(Dir,"/heatmap_RNAseq_allcts_top1kCOVgenes.pdf"))
Heatmap(df.scaled,
        name = "z-score",
        column_title ="3'RNAseq log2CPM of fibroblasts, keratinocytes, melanocytes",
        top_annotation=HeatmapAnnotation(celltype=celltype_anno,
                                         condition=condition_anno,
                                         col = list(celltype = c("FRB"="red","KRT"="green","MEL"="blue"), 
                                                    condition = c("PBS"="black","IFN"="magenta"))),
        cluster_rows = gene_hclust, 
        cluster_columns = sample_hclust,
        border = TRUE,
        column_names_gp = grid::gpar(fontsize = 2),
        show_column_names = TRUE,
        show_row_names = FALSE)
dev.off()

# look for any IFNg response or production genes in the top1kCOV genes
genes_ifng_related=read.table(file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/IFNg_signature/GO_GSEA_IFNg_production_or_response_related_genes.txt")
overlap=intersect(rownames(df.scaled),as.character(genes_ifng_related$V1))
ifng_genes_clusters = gene_clusters %>% dplyr::filter(gene %in% overlap)


pdf(paste0(Dir,"/heatmap_RNAseq_allcts_top1kCOVgenes_withIFNgGenes.pdf"))

rowha=rowAnnotation(link = anno_mark(at = which(rownames(df.scaled) %in% overlap), 
                                     labels = overlap,
                                     which="row",
                                     labels_gp = gpar(fontsize=5),
                                     padding=unit(0.5,"mm")))
Heatmap(df.scaled,
        name = "z-score",
        column_title ="3'RNAseq log2CPM of fibroblasts, keratinocytes, melanocytes",
        top_annotation=HeatmapAnnotation(celltype=celltype_anno,
                                         condition=condition_anno,
                                         col = list(celltype = c("FRB"="red","KRT"="green","MEL"="blue"), 
                                                    condition = c("PBS"="black","IFN"="magenta"))),
        row_km = 6,
        #cluster_rows = gene_hclust, 
        cluster_columns = sample_hclust,
        border = TRUE,
        show_column_names = FALSE,
        show_row_names = FALSE) + 
  rowha
dev.off()

# use DE genes in each cell type, and label IFNg genes and vitiligo DE genes
DEG=read.table(file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/ifng_responsive_genes_in_every_celltypes_from_pipeline08172022.txt")
df.scaled.focus = df.scaled[intersect(rownames(df.scaled),DEG$V1),]
genes_ifng_related=read.table(file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/IFNg_signature/GO_GSEA_IFNg_production_or_response_related_genes.txt")
overlap=intersect(rownames(df.scaled.focus),as.character(genes_ifng_related$V1))
  
ifng_genes_clusters = gene_clusters %>% dplyr::filter(gene %in% overlap)


pdf(paste0(Dir,"/heatmap_RNAseq_allcts_IFNgDEgenes_withGOtermIFNgGenes.pdf"))

df.scaled.focus = df.scaled.focus[,c(colnames.krt.pbs,colnames.krt.ifn,colnames.frb.pbs,colnames.frb.ifn,colnames.mel.pbs,colnames.mel.ifn)]
celltype_anno=sapply(strsplit(as.character(colnames(df.scaled.focus)),split="_"),"[[",2)
condition_anno=sapply(strsplit(as.character(colnames(df.scaled.focus)),split="_"),"[[",3)

rowha=rowAnnotation(link = anno_mark(at = which(rownames(df.scaled.focus) %in% overlap), 
                                     labels = overlap,
                                     which="row",
                                     labels_gp = gpar(fontsize=5),
                                     padding=unit(0.5,"mm")))
Heatmap(df.scaled.focus,
        name = "z-score",
        column_title ="3'RNAseq log2CPM of fibroblasts, keratinocytes, melanocytes",
        top_annotation=HeatmapAnnotation(celltype=celltype_anno,
                                         condition=condition_anno,
                                         col = list(celltype = c("FRB"="red","KRT"="green","MEL"="blue"), 
                                                    condition = c("PBS"="black","IFN"="magenta"))),
        row_km = 6,
        cluster_columns=FALSE,
        #cluster_rows = gene_hclust, 
        #cluster_columns = sample_hclust,
        border = TRUE,
        show_column_names = FALSE,
        show_row_names = FALSE) + 
  rowha
dev.off()


# look for differentially expressed genes in melanocytes from scRNAseq of vitiligo lesional vs. healthy control
genes_DE_in_vitiligo_MEL=data.table::fread(file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/IFNg_signature/SSC_MEL_Vitiligo-active_Control_Lesional.txt") %>% dplyr::filter(logFC > 0) %>% dplyr::filter(FDR < 0.01) %>% pull(V1) %>% as.character()

overlap=intersect(rownames(df.scaled),as.character(genes_DE_in_vitiligo_MEL))
vitiligo_up_genes_clusters = gene_clusters %>% dplyr::filter(gene %in% overlap)

pdf(paste0(Dir,"/heatmap_RNAseq_allcts_top1kCOVgenes_withVitiligoUpregulatedGenesInVitiligo.pdf"))

gene_hclust <- hclust(dist(df,method="euclidean")^2, method = "cen")
sample_hclust <- hclust(dist(t(df),method="euclidean")^2,method="single")

rowha=rowAnnotation(link = anno_mark(at = which(rownames(df.scaled) %in% overlap), 
                                     labels = overlap,
                                     which="row",
                                     labels_gp = gpar(fontsize=5),
                                     padding=unit(0.5,"mm")))
df.scaled = df.scaled[,c(colnames.krt.pbs,colnames.krt.ifn,colnames.frb.pbs,colnames.frb.ifn,colnames.mel.pbs,colnames.mel.ifn)]
celltype_anno=sapply(strsplit(as.character(colnames(df.scaled)),split="_"),"[[",2)
condition_anno=sapply(strsplit(as.character(colnames(df.scaled)),split="_"),"[[",3)

Heatmap(df.scaled[,c(colnames.krt.pbs,colnames.krt.ifn,colnames.frb.pbs,colnames.frb.ifn,colnames.mel.pbs,colnames.mel.ifn)],
        name = "z-score",
        column_title ="3'RNAseq log2CPM of fibroblasts, keratinocytes, melanocytes",
        top_annotation=HeatmapAnnotation(celltype=celltype_anno,
                                         condition=condition_anno,
                                         col = list(celltype = c("FRB"="red","KRT"="green","MEL"="blue"), 
                                                    condition = c("PBS"="black","IFN"="magenta"))),
        row_km = 6,
        #cluster_rows = gene_hclust, 
        #cluster_columns = sample_hclust,
        cluster_columns=FALSE,
        border = TRUE,
        show_column_names = FALSE,
        show_row_names = FALSE) + 
  rowha
dev.off()
```


### Write to file
```{r output}
# set global theme for reactable
options(reactable.theme = reactableTheme(borderColor = "#dfe2e5",stripedColor = "#f6f8fa",highlightColor = "#f0f5f9",cellPadding = "8px 12px",style = list(fontFamily = "Helvetica")))

# write raw ESAT counts
write.table(counts.keep, file=paste0(Dir,"/raw_counts.txt"), quote=F, sep="\t", row.names=T, col.names=T)
htmlwidgets::saveWidget(widget = reactable(counts.keep %>% rownames_to_column("gene"),
                                           filterable=T, searchable=T, striped=T, bordered=T), 
                        file = paste0(Dir,"/raw_counts.html"), selfcontained = TRUE)
# write TMM-normalized CPM 
write.table(CPM, file=paste0(Dir,"/TMM-normalized-CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(logCPM, file=paste0(Dir,"/log2-TMM-normalized-CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# write metadata
write.table(metadata, file=paste0(Dir,"/metadata.txt"),quote=F, sep="\t", row.names=T, col.names=T)


# write DESeq2 results
write.table(res,file=paste0(Dir,"/DESeq2_results_IFNvsPBS.txt"),quote=F,row.names=T,col.names=T)
write.table(res.up$gene,file=paste0(Dir,"/upregulated_genes.txt"),quote=F,row.names=F,col.names=F)
write.table(res.down$gene,file=paste0(Dir,"/downregulated_genes.txt"),quote=F,row.names=F,col.names=F)
htmlwidgets::saveWidget(widget = reactable(res,filterable=T, searchable=T, striped=T, bordered=T), 
                        file = paste0(Dir,"/DESeq2_results_IFNvsPBS.html"), selfcontained = TRUE)
write.table(res$gene, file=paste0(Dir,"/genes_used_in_DE.txt"), quote=F, sep="\t", row.names=F, col.names=F)


# save useful PCA plots

# write log2FoldChange of CPM data
log2FC=log2((CPM.ifn+0.0001)/(CPM.pbs+0.0001))
write.table(log2FC, file=paste0(Dir,"/log2FoldChange_by_IFNg.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# filter out genes that aren't highly expressed
gene_expressed_pbs = CPM.pbs %>% mutate(mean=apply(CPM.pbs, 1, mean)) %>% dplyr::filter(mean>10) %>% rownames(.) 
gene_expressed_ifn = CPM.ifn %>% mutate(mean=apply(CPM.ifn, 1, mean)) %>% dplyr::filter(mean>10) %>% rownames(.)
gene_expressed = intersect(gene_expressed_pbs,gene_expressed_ifn)
write.table(gene_expressed, file=paste0(Dir,"/highly_expressed_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write more genes (expressed AND IFNg responsive)
gene_expressed_and_ifng_responsive = intersect(gene_expressed, c(res.up$gene,res.down$gene))
write.table(gene_expressed_and_ifng_responsive, file=paste0(Dir,"/highly_expressed_and_ifng_responsive_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write down genes that are responsive to IFNg
write.table(c(res.up$gene,res.down$gene),file=paste0(Dir,"/ifng_responsive_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write down batch-corrected counts
write.table(counts.keep.corrected, file=paste0(Dir,"/batch-corrected-counts.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# write down batch-corrected TMM-normalized CPM
write.table(CPM.corrected, file=paste0(Dir,"/TMM-normalized-batch-corrected-CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)

```



















### DE analysis  
The goal of Differential gene expression analysis is to find 
genes or transcripts whose difference in expression, when accounting 
for the variance within condition, is higher than expected by chance. 

In this case a total sum of 10 counts separates well expressed 
from non-expressed genes. You can change this value and padj value and 
log2FoldChange cutoffs according to your data

```{r DESeq2}
### conduct DESeq2 analysis on skintone to look for DE genes
# melanocytes
temp.CPM <- CPM[which(rowSums(CPM[,])>=10),which(grepl("MEL",colnames(CPM)))]
df <- counts[rownames(temp.CPM),colnames(temp.CPM)]
df.metadata <- metadata[which(metadata$celltype=="MEL"),]
dds <- DESeqDataSetFromMatrix(countData = df, colData = df.metadata, design = ~ batch + condition)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("condition", "IFN", "PBS")))
res.mel <- results(dds,contrast=c("condition", "IFN", "PBS")) %>% as.data.frame %>% rownames_to_column("gene")

# keratinocytes
temp.CPM <- CPM[which(rowSums(CPM[,])>=10),which(grepl("KRT",colnames(CPM)))]
df <- counts[rownames(temp.CPM),colnames(temp.CPM)]
df.metadata <- metadata[which(metadata$celltype=="KRT"),]
dds <- DESeqDataSetFromMatrix(countData = df, colData = df.metadata, design = ~ batch + condition)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("condition", "IFN", "PBS")))
res.krt <- results(dds,contrast=c("condition", "IFN", "PBS")) %>% as.data.frame %>% rownames_to_column("gene")

# fibroblasts
temp.CPM <- CPM[which(rowSums(CPM[,])>=10),which(grepl("FRB",colnames(CPM)))]
df <- counts[rownames(temp.CPM),colnames(temp.CPM)]
df.metadata <- metadata[which(metadata$celltype=="FRB"),]
dds <- DESeqDataSetFromMatrix(countData = df, colData = df.metadata, design = ~ batch + condition)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("condition", "IFN", "PBS")))
res.frb <- results(dds,contrast=c("condition", "IFN", "PBS")) %>% as.data.frame %>% rownames_to_column("gene")

```
```{r DESeq2-skintone}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
library(gridExtra)
library(ggpubr)
library(plotly)
library(cowplot)

Dir="~/Downloads/nl/human/skin/eQTLs/RNA-Seq"
setwd(Dir)

# load data and metadata
#save.image(file=paste0("~/Downloads",'/myEnvironment_allcts_skintoneDE_forKen.RData'))
load(paste0("~/Downloads",'/myEnvironment_allcts_skintoneDE_forKen.RData'))
df.skintone=data.table::fread(paste0(Dir,"/donor_skintone.txt"))
metadata=rbind(metadata_FRB,metadata_KRT,metadata_MEL) %>% distinct_at(.,"sample",.keep_all=TRUE) %>% left_join(. , df.skintone, by="donor")
colnames(CPM) <- gsub("_S.*","",colnames(CPM))
CPM.pbs <- CPM[,which(grepl("_PBS",colnames(CPM)))]
colnames(CPM.pbs) <- gsub("_S.*","",colnames(CPM.pbs))
CPM.ifn <- CPM[,which(grepl("_IFN",colnames(CPM)))]
colnames(CPM.ifn) <- gsub("_S.*","",colnames(CPM.ifn))

### conduct DESeq2 analysis on skintone to look for DE genes
# melanocytes
temp.CPM <- CPM[which(rowSums(CPM[,])>=10),which(grepl("MEL_PBS",colnames(CPM)))]
df <- counts[rownames(temp.CPM),colnames(temp.CPM)]
df.metadata <- metadata[which(metadata$celltype=="MEL" & metadata$condition=="PBS"),]
dds <- DESeqDataSetFromMatrix(countData = df, colData = df.metadata, design = ~ batch + skintone)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("skintone", "medium_to_dark", "white")))
res.mel <- results(dds,contrast=c("skintone", "medium_to_dark", "white")) %>% as.data.frame %>% rownames_to_column("gene")
res.up <- res.mel %>% filter(log2FoldChange > 1.5) %>% filter(padj < 0.05)
res.down <- res.mel %>% filter(log2FoldChange < -1.5) %>% filter(padj < 0.05)
skintone.DEG.mel <- c(res.up$gene,res.down$gene)

# keratinocytes
temp.CPM <- CPM[which(rowSums(CPM[,])>=10),which(grepl("KRT_PBS",colnames(CPM)))]
df <- counts[rownames(temp.CPM),colnames(temp.CPM)]
df.metadata <- metadata[which(metadata$celltype=="KRT" & metadata$condition=="PBS"),]
dds <- DESeqDataSetFromMatrix(countData = df, colData = df.metadata, design = ~ batch + skintone)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("skintone", "medium_to_dark", "white")))
res.krt <- results(dds,contrast=c("skintone", "medium_to_dark", "white")) %>% as.data.frame %>% rownames_to_column("gene")
res.up <- res.krt %>% filter(log2FoldChange > 1.5) %>% filter(padj < 0.05)
res.down <- res.krt %>% filter(log2FoldChange < -1.5) %>% filter(padj < 0.05)
skintone.DEG.krt <- c(res.up$gene,res.down$gene)

# fibroblasts
temp.CPM <- CPM[which(rowSums(CPM[,])>=10),which(grepl("FRB_PBS",colnames(CPM)))]
df <- counts[rownames(temp.CPM),colnames(temp.CPM)]
df.metadata <- metadata[which(metadata$celltype=="FRB" & metadata$condition=="PBS"),]
dds <- DESeqDataSetFromMatrix(countData = df, colData = df.metadata, design = ~ batch + skintone)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("skintone", "medium_to_dark", "white")))
res.frb <- results(dds,contrast=c("skintone", "medium_to_dark", "white")) %>% as.data.frame %>% rownames_to_column("gene")
res.up <- res.frb %>% filter(log2FoldChange > 1.5) %>% filter(padj < 0.05)
res.down <- res.frb %>% filter(log2FoldChange < -1.5) %>% filter(padj < 0.05)
skintone.DEG.frb <- c(res.up$gene,res.down$gene)

########## plot CPM for all donors across all cell types for gene of interest #########
plot_list = list()
for (n in 1:nrow(skintone.DEG.CPM)) {
  this.CPM <- as.data.frame(skintone.DEG.CPM[n,]) %>% rownames_to_column("sample") %>% left_join( . , metadata, by="sample")
  g=skintone.DEG[n]

  # pick out TMM-normalized CPM paired data 
  df.paired.this.long <- data.frame(condition=c(rep("PBS",ncol(CPM.pbs)), rep("IFN",ncol(CPM.ifn))),
                               CPM=c(t(CPM.pbs[g,]),t(CPM.ifn[g,])),
                               sample_for_splitting=c(colnames(CPM.pbs),colnames(CPM.ifn))) %>%
    mutate(biosample=sample_for_splitting) %>%
    tidyr::separate(., col=sample_for_splitting, into=c("donor","celltype","condition"),sep="_") %>%
    left_join( . , metadata[,c("biosample","skintone")], by="biosample")
  
  df.paired.this.long$condition <- factor(df.paired.this.long$condition, levels=c("PBS","IFN"))
  df.paired.this.long$skintone <- factor(df.paired.this.long$skintone, levels=c("white","medium_to_dark"))

  # DESeq2 analysis result
  df.DEseq2.this <- rbind(res.mel[which(res.mel$gene==g),] %>% mutate(celltype="MEL"),
                          res.krt[which(res.krt$gene==g),] %>% mutate(celltype="KRT"),
                          res.frb[which(res.frb$gene==g),] %>% mutate(celltype="FRB"))
  
    ### Plot melanocyte data
  plt.MEL <- df.paired.this.long %>% 
    dplyr::filter(celltype=="MEL") %>%
    ggplot(.,aes(x=condition, y=CPM)) +
    geom_line(aes(group=donor)) +
    geom_point(aes(color=donor)) +
    labs(title=paste0(g,"\n",
                      "skintone DE analysis:","\n"," log2FC=", round(df.DEseq2.this[1,]$log2FoldChange,3),"\n"," p.adj=", 
                      format(df.DEseq2.this[1,]$padj,digits=3,scientific=T),"\n")) +
    theme(title =element_text(size=5, face='bold'), legend.position="none") +
    facet_wrap(~skintone)
  
  plt.KRT <- df.paired.this.long %>% 
    dplyr::filter(celltype=="KRT") %>%
    ggplot(.,aes(x=condition, y=CPM)) +
    geom_line(aes(group=donor)) +
    geom_point(aes(color=donor)) +
    labs(title=paste0(g,"\n",
                      "skintone DE analysis:","\n"," log2FC=", round(df.DEseq2.this[2,]$log2FoldChange,3),"\n"," p.adj=", 
                      format(df.DEseq2.this[2,]$padj,digits=3,scientific=T),"\n")) +
    theme(title =element_text(size=5, face='bold'), legend.position="none") +
    facet_wrap(~skintone)
  
  plt.FRB <- df.paired.this.long %>% 
    dplyr::filter(celltype=="FRB") %>%
    ggplot(.,aes(x=condition, y=CPM)) +
    geom_line(aes(group=donor)) +
    geom_point(aes(color=donor)) +
    labs(title=paste0(g,"\n",
                      "skintone DE analysis:","\n"," log2FC=", round(df.DEseq2.this[3,]$log2FoldChange,3),"\n"," p.adj=", 
                      format(df.DEseq2.this[3,]$padj,digits=3,scientific=T),"\n")) +
    theme(title =element_text(size=5, face='bold'), legend.position="none") +
    facet_wrap(~skintone)
  
  p <- plot_grid(plt.MEL, plt.KRT, plt.FRB, ncol = 3, nrow = 1)
  plot_list[[n]] <- p
}
pdf(paste0(Dir,"/skintone_DEG_paired_plots_3cts.pdf"),width=12,height=8)
for (n in 1:length(plot_list)) {
  print(plot_list[[n]])
}
dev.off()
```

```{r paired-plot}
g="TGFBR2"

# pick out TMM-normalized CPM paired data & run wilcox test
df.paired.this.long <- data.frame(condition=c(rep("PBS",ncol(CPM.pbs)), rep("IFN",ncol(CPM.ifn))),
                                  CPM=c(t(CPM.pbs[g,]),t(CPM.ifn[g,])),
                                  sample_for_splitting=c(colnames(CPM.pbs),colnames(CPM.ifn))) %>%
  mutate(biosample=sample_for_splitting) %>%
  tidyr::separate(., col=sample_for_splitting, into=c("donor","celltype","condition"),sep="_") 

df.paired.this.long$condition <- factor(df.paired.this.long$condition, levels=c("PBS","IFN"))

# DESeq2 analysis result
df.DEseq2.this <- rbind(res.mel[which(res.mel$gene==g),] %>% mutate(celltype="MEL"),
                        res.krt[which(res.krt$gene==g),] %>% mutate(celltype="KRT"),
                        res.frb[which(res.frb$gene==g),] %>% mutate(celltype="FRB"))
  
### Plot melanocyte data
plt.MEL <- df.paired.this.long %>% 
  dplyr::filter(celltype=="MEL") %>%
  ggplot(.,aes(x=condition, y=CPM)) +
  geom_line(aes(group=donor)) +
  geom_point(aes(color=donor)) +
  labs(title=paste0(g," CPM in melanocytes\n",
                    "IFNg DE analysis:","\n"," log2FC=", round(df.DEseq2.this[1,]$log2FoldChange,3),"\n"," p.adj=", 
                    format(df.DEseq2.this[1,]$padj,digits=3,scientific=T),"\n")) +
  theme(title =element_text(size=5, face='bold'), legend.position="none")

plt.KRT <- df.paired.this.long %>% 
  dplyr::filter(celltype=="KRT") %>%
  ggplot(.,aes(x=condition, y=CPM)) +
  geom_line(aes(group=donor)) +
  geom_point(aes(color=donor)) +
  labs(title=paste0(g," CPM in keratinocytes\n",
                    "IFNg DE analysis:","\n"," log2FC=", round(df.DEseq2.this[2,]$log2FoldChange,3),"\n"," p.adj=", 
                    format(df.DEseq2.this[2,]$padj,digits=3,scientific=T),"\n")) +
  theme(title =element_text(size=5, face='bold'), legend.position="none")

plt.FRB <- df.paired.this.long %>% 
  dplyr::filter(celltype=="FRB") %>%
  ggplot(.,aes(x=condition, y=CPM)) +
  geom_line(aes(group=donor)) +
  geom_point(aes(color=donor)) +
  labs(title=paste0(g," CPM in fibroblasts\n",
                    "IFNg DE analysis:","\n"," log2FC=", round(df.DEseq2.this[3,]$log2FoldChange,3),"\n"," p.adj=", 
                    format(df.DEseq2.this[3,]$padj,digits=3,scientific=T),"\n")) +
  theme(title =element_text(size=5, face='bold'), legend.position="none") 
  
p <- plot_grid(plt.MEL, plt.KRT, plt.FRB, ncol = 3, nrow = 1)
```
### Heatmap of top 2k cov DE genes across all samples
```{r heatmap-top2kcovDEG}
df <- apply(cov.merged[rowSums(CPM)>=10,],2,round,0) %>% as.data.frame()

rm(dds)
dds <- DESeqDataSetFromMatrix(countData = df,
                              colData = metadata,
                              design = ~ condition)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("condition", "IFN", "PBS")))

res <- results(dds,contrast=c("condition", "IFN", "PBS")) %>% as.data.frame %>% rownames_to_column("region")
res.up <- res %>% filter(log2FoldChange > 1.5) %>% filter(padj < 0.05) %>% filter(!is.na(padj))
res.down <- res %>% filter(log2FoldChange < -1.5) %>% filter(padj < 0.05) %>% filter(!is.na(padj))
DEG <- c(res.up$gene,res.down$gene)

# heatmap of top 2000 most variable genes based on CoV
logCPM.DEG = log2(CPM[DEG,]+1) %>% as.data.frame()%>% na.omit() 
mat.logCPM.lnsd.DEG <- apply(logCPM.DEG,MARGIN=1,FUN=sd)
mat.logCPM.lncv.DEG <- sqrt( (exp(mat.logCPM.lnsd.DEG) )^2 -1) #coefficient of variance
logCPM.top2k.DEG <- logCPM.DEG[order(mat.logCPM.lncv.DEG,decreasing=T),]

# filter data matrix based on candidate genes
df.DEG <- logCPM.top2k.DEG
df.scaled.DEG=scale(df.DEG)

# plot heatmap
library(ComplexHeatmap)
celltype_anno=sapply(strsplit(as.character(colnames(df.scaled.DEG)),split="_"),"[[",2)
condition_anno=sapply(strsplit(as.character(colnames(df.scaled)),split="_"),"[[",3)

#pdf(paste0(Dir,"/heatmap_RNAseq_allcts_top2kCOV_DEGenes.pdf"))
Heatmap(df.scaled.DEG,
        name = "z-score",
        column_title ="3'RNAseq log2CPM of fibroblasts, keratinocytes, melanocytes",
        top_annotation=HeatmapAnnotation(celltype=celltype_anno,
                                         condition=condition_anno,
                                         col = list(celltype = c("FRB"="red","KRT"="green","MEL"="blue"), 
                                                    condition = c("PBS"="black","IFN"="magenta"))),
        border = TRUE,
        column_names_gp = grid::gpar(fontsize = 8),
        show_column_names = TRUE,
        show_row_names = FALSE)
#dev.off()
#save.image(paste0(Dir,'/myEnvironment_allcts_heatmap_top2kcov_DEGenes.RData'))
```

### TF expression across 3 celltypes
```{r TF}
# load list of 1924 TFs curated according to notes in: "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/TF_list_curation_notes.txt
tf_tab <- data.table::fread(input="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/TF_list.txt")

# load dic that converts ENSG to gene symbol
library('biomaRt')
ensembl <- useDataset(dataset = "hsapiens_gene_ensembl", mart = ensembl)
attributes=listAttributes(ensembl)
dict <- getBM(attributes=c("ensembl_gene_id","external_gene_name"), mart=ensembl, useCache = FALSE)

# convert ENSG to gene name
tf <- left_join(tf_tab,dict,by=c("ENSG"="ensembl_gene_id")) %>% pull(external_gene_name)
rm(tf_tab,ensembl,attributes,dict)

tf <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/TF_list_Lambert2018.txt",header=FALSE)

# query TFs in all 3 cell types baseline CPM data
CPM.tf <- as.data.frame(CPM)[tf$V1,metadata.pbs$sample]

# what are the top 100 TFs expressed in each celltype?
top100tf.krt <- CPM.tf[,intersect(metadata.pbs$sample,metadata_KRT$sample)] %>% rowMeans(.) %>% sort(decreasing=TRUE) %>% head(100) %>% names()
top100tf.mel <- CPM.tf[,intersect(metadata.pbs$sample,metadata_MEL$sample)] %>% rowMeans(.) %>% sort(decreasing=TRUE) %>% head(100) %>% names()
top100tf.frb <- CPM.tf[,intersect(metadata.pbs$sample,metadata_FRB$sample)] %>% rowMeans(.) %>% sort(decreasing=TRUE) %>% head(100) %>% names()
toptf <- unique(c(top100tf.krt,top100tf.mel,top100tf.frb))

# check how many are shared in the celltypes
venn.list <- list(top100tf.krt,top100tf.mel,top100tf.frb)
names(venn.list) <- c("keratinocytes","melanocytes","fibroblasts")
pdf("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/venn_topexpressed_TFs.pdf")
venn.res <- venn(venn.list,names=c("keratinocytes","melanocytes","fibroblasts"))
dev.off()

# inspect results
str(venn.res)
# get names of cell type specific transcription factors as well as TFs that are expressed in all cell types
tf.cts.frb <- attr(venn.res,"intersection")[1] %>% unlist() %>% unname()
tf.cts.krt <- attr(venn.res,"intersection")[2] %>% unlist() %>% unname()
tf.cts.mel <- attr(venn.res,"intersection")[3] %>% unlist() %>% unname()
tf.cts <- c(tf.cts.frb,tf.cts.krt,tf.cts.mel)
tf.shared <- attr(venn.res,"intersection")[7] %>% unlist() %>% unname()
write.table(tf.cts.frb, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/top_expressed_TFs_celltypespecific_fibroblasts.txt", quote=F, sep="\t", row.names=F, col.names=F)
write.table(tf.cts.krt, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/top_expressed_TFs_celltypespecific_keratinocytes.txt", quote=F, sep="\t", row.names=F, col.names=F)
write.table(tf.cts.mel, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/top_expressed_TFs_celltypespecific_melanocytes.txt", quote=F, sep="\t", row.names=F, col.names=F)
write.table(tf.shared, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/top_expressed_TFs_shared_fib_krt_mel.txt", quote=F, sep="\t", row.names=F, col.names=F)


# plot heatmap of those transcription factors in baseline condition
df.scaled = scale(log2(CPM.tf+1))
df.scaled.focus = subset(df.scaled, rownames(df.scaled) %in% toptf) %>% as.data.frame()

pdf("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/heatmap_topexpressed_TFs_skin_PBS.pdf")

df.scaled.focus = df.scaled.focus[,c(colnames.krt.pbs,colnames.frb.pbs,colnames.mel.pbs)]
celltype_anno=sapply(strsplit(as.character(colnames(df.scaled.focus)),split="_"),"[[",2)
condition_anno=sapply(strsplit(as.character(colnames(df.scaled.focus)),split="_"),"[[",3)

rowha=rowAnnotation(link = anno_mark(at = which(rownames(df.scaled.focus) %in% tf.shared), 
                                     labels = tf.shared,
                                     which="row",
                                     labels_gp = gpar(fontsize=6),
                                     padding=unit(0.5,"mm")))

Heatmap(df.scaled.focus,
        name = "z-score",
        column_title ="top expressed TFs in 3'RNAseq log2CPM of fibroblasts, keratinocytes, melanocytes",
        column_title_gp = gpar(fontsize = 10, fontface = "bold"),
        top_annotation=HeatmapAnnotation(celltype=celltype_anno,
                                         condition=condition_anno,
                                         col = list(celltype = c("FRB"="red","KRT"="green","MEL"="blue"), 
                                                    condition = c("PBS"="black","IFN"="magenta"))),
        row_km = 6,
        cluster_columns=FALSE,
        border = TRUE,
        show_column_names = FALSE,
        show_row_names = FALSE) + 
  rowha
dev.off()

# plot heatmap of those transcription factors in baseline AND IFNg condition
# query TFs in all 3 cell types baseline CPM data
CPM.tf <- as.data.frame(CPM)[tf$V1,metadata$sample]
df.scaled = scale(log2(CPM.tf+1))
df.scaled.focus = subset(df.scaled, rownames(df.scaled) %in% toptf) %>% as.data.frame()

pdf("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/heatmap_topexpressed_TFs_skin_PBSandIFNg.pdf")

df.scaled.focus = df.scaled.focus[,c(colnames.krt.pbs,colnames.krt.ifn,colnames.frb.pbs,colnames.frb.ifn,colnames.mel.pbs,colnames.mel.ifn)]
celltype_anno=sapply(strsplit(as.character(colnames(df.scaled.focus)),split="_"),"[[",2)
condition_anno=sapply(strsplit(as.character(colnames(df.scaled.focus)),split="_"),"[[",3)

rowha=rowAnnotation(link = anno_mark(at = which(rownames(df.scaled.focus) %in% tf.shared), 
                                     labels = tf.shared,
                                     which="row",
                                     labels_gp = gpar(fontsize=6),
                                     padding=unit(0.5,"mm")))

Heatmap(df.scaled.focus,
        name = "z-score",
        column_title ="top expressed TFs in 3'RNAseq log2CPM of fibroblasts, keratinocytes, melanocytes",
        column_title_gp = gpar(fontsize = 10, fontface = "bold"),
        top_annotation=HeatmapAnnotation(celltype=celltype_anno,
                                         condition=condition_anno,
                                         col = list(celltype = c("FRB"="red","KRT"="green","MEL"="blue"), 
                                                    condition = c("PBS"="black","IFN"="magenta"))),
        row_km = 6,
        cluster_columns=FALSE,
        border = TRUE,
        show_column_names = FALSE,
        show_row_names = FALSE) + 
  rowha
dev.off()




# plot heatmap of those transcription factors that are cell type specific
df.scaled = scale(log2(CPM.tf+1))
df.scaled.focus = subset(df.scaled, rownames(df.scaled) %in% tf.cts) %>% as.data.frame()

pdf("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/TFs/heatmap_topexpressed_cts_TFs.pdf")

df.scaled.focus = df.scaled.focus[,c(colnames.krt.pbs,colnames.frb.pbs,colnames.mel.pbs)]
celltype_anno=sapply(strsplit(as.character(colnames(df.scaled.focus)),split="_"),"[[",2)
condition_anno=sapply(strsplit(as.character(colnames(df.scaled.focus)),split="_"),"[[",3)

Heatmap(df.scaled.focus,
        name = "z-score",
        column_title ="top expressed cts TFs in 3'RNAseq log2CPM of fibroblasts, keratinocytes, melanocytes",
        top_annotation=HeatmapAnnotation(celltype=celltype_anno,
                                         condition=condition_anno,
                                         col = list(celltype = c("FRB"="red","KRT"="green","MEL"="blue"), 
                                                    condition = c("PBS"="black","IFN"="magenta"))),
        row_km = 6,
        cluster_columns=FALSE,
        #cluster_rows = gene_hclust, 
        #cluster_columns = sample_hclust,
        border = TRUE,
        show_column_names = FALSE,
        show_row_names = FALSE) 
dev.off()

```


























### calculate PEER factors
```{r peer}


```




























## unused code block:
### Load metadata
```{r metadata, echo=FALSE, message=FALSE}
# parse kept samples into table
samples <- data.frame(sample=colnames(counts.keep))
# import metadata master table
metadata.master=read.table("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/metadata_08182022.txt", sep="\t", header=TRUE)
# import DolphinNext pipeline overall summary
pip.summary=read.table("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/melanocytes/pipeline_08172022/DolphinNext/report7114/summary/overall_summary.tsv", sep="\t", header=TRUE) %>% magrittr::set_colnames(c("sample","total_reads","Multimapped_Reads_Aligned_STAR","Unique_Reads_Aligned_STAR"))
pip.summary$sample <- pip.summary$sample %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)

# query imported tables with samples
rm(metadata)
metadata <- left_join(samples,metadata.master,by="sample")
metadata <- left_join(metadata,pip.summary,by="sample")

# compile 
metadata$seq_date <- metadata$seq_date %>% gsub("\\/",".",.)
metadata$lib_prep_date <- metadata$lib_prep_date %>% gsub("\\/",".",.)
metadata$batch <- paste(metadata$lib_prep_date,metadata$seq_date,sep="_") %>% as.factor()

# cleanup
rm(pip.summary, samples, samples.all)
```
  
  
### Overview of libraries  
Overview of library size distribution.
```{r lib-size, echo=F, message=F}
# library size distribution
df=data.frame(sample=metadata$sample,total_count=metadata$total_reads)
df=df[order(-df$total_count),]
plot_ly(y = df$sample,x = df$total_count,type = "bar")

# histogram for gene counts, all samples
h <- hist(log10(rowSums(counts.keep)), breaks = as.numeric(100), plot = FALSE) 
plot_ly(x = h$mids, y = h$counts, width = 500, height=300) %>% layout( title = "Histogram") %>% add_bars()

# which libraries have low reads? (<5M)
df$sample[which(df$total_count<5000000)] %>% gsub("_MEL.*","",.) %>% table()
```

### PCA
  
```{r pca-uncorrected}
# pick genes of top row variance
mat.logCPM.lnsd <- apply(logCPM,MARGIN=1,FUN=sd)
mat.logCPM.lncv <- sqrt( (exp(mat.logCPM.lnsd) )^2 -1) #coefficient of variance
logCPM.top <- logCPM[order(mat.logCPM.lncv,decreasing=T),][1:2000,] 

# PCA
pca = prcomp(t(logCPM.top),centered= T,scale = T)
loadings=as.data.frame(pca$rotation)
# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)
# PC1 is 43.6%, PC2 is 8%, PC3 is 6.7%. Top 2000 variable genes separate well by condition (PC1) in melanocytes, much more dramatic than fibroblasts

# plot
# top PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], PC4=pca$x[,4], PC5=pca$x[,5],PC6=pca$x[,6],sample=rownames(pcs))
d=right_join(d,metadata,by="sample")
rownames(d)<-d$sample
#d$group_tags <- cut(d$Unique_Reads_Aligned_STAR, breaks=6)
d$group_tags <- cut(d$total_reads, 
                    breaks=c(min(d$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d$total_reads)))

d %>%
  mutate(platform=d$seq_model) %>%
  ggplot(., aes(y=PC1,x=PC2,col=celltype,shape=condition)) + 
  geom_point(size=3) + 
  #geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  ggsave(filename=paste0(Dir,"/PCA_1.png"))
```

```{r pca-corrected}
# pick genes of top row variance
mat.logCPM.lnsd.corrected <- apply(logCPM.corrected,MARGIN=1,FUN=sd)
mat.logCPM.lncv.corrected <- sqrt( (exp(mat.logCPM.lnsd.corrected) )^2 -1) #coefficient of variance
logCPM.top.corrected <- logCPM[order(mat.logCPM.lncv.corrected,decreasing=T),][1:2000,] 

# PCA
pca.corrected = prcomp(t(logCPM.top.corrected),centered= T,scale = T)
loadings.corrected=as.data.frame(pca.corrected$rotation)
# percent variance in each pc
plot(pca.corrected)
pcs.corrected = pca.corrected$x
pca.pctVar.corrected = pca.corrected$sdev^2/sum(pca.corrected$sdev^2)
# PC1 is 47.7%, PC2 is 7%, PC3 is 5.2%. Top 2000 variable genes separate well by condition (PC1) in melanocytes, much more dramatic than fibroblasts

# plot
# top PC information
d.corrected=data.frame(PC1=pca.corrected$x[,1],PC2=pca.corrected$x[,2], PC3=pca.corrected$x[,3], PC4=pca.corrected$x[,4], PC5=pca.corrected$x[,5],PC6=pca.corrected$x[,6],sample=rownames(pcs.corrected))
d.corrected=inner_join(d.corrected,metadata,by="sample") %>% magrittr::set_rownames(d.corrected$sample)
colnames(metadata)
d.corrected$group_tags <- cut(d.corrected$total_reads, 
                    breaks=c(min(d.corrected$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d.corrected$total_reads)))

d.corrected %>%
  mutate(platform=d.corrected$seq_model) %>%
  ggplot(., aes(y=PC1,x=PC2,col=group_tags,shape=condition,label=sample)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar.corrected[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar.corrected[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  ggsave(filename=paste0(Dir,"/PCA.corrected_3.png"))
```
  

### Plot paired data in CPM

Harris lab Shany project: Olink proteomics data of vit patient blister fluids showed increase in IFNlambda1 protein. (not increased.) But scRNAseq data of the cells in blister fluid showed decrease in IFNlambda gene and receptor. the receptor decrease is almost unique to melanocytes. not as much of a change in other cell types.
Would IFNg stimulation increase IFN lambda receptor expression in melanocytes?
IFNlR1: IL-10R2 (or IL10Rbeta) form a deterodimer at the cell surface, recruits JAK2, JAK1 and TYK2. oerlap with IFNalpha receptor actiation. downstream, phosphorylate & actiates STAT1 and STAT2, NFkB, and enhancs interferon and antiiral protein production.
would IFNg stimulation increase expression of genes that belong to the interferon stimulated genes (ISG): e.g. IFIT1, IFIT2, MX1, OAS1.

```{r paired-beehive}
g="OAS1" # IFIT1, IFIT2, MX1, OAS1

# pick out TMM-normalized CPM paired data & run wilcox test
df.paired.this <- data.frame(PBS=as.data.frame(t(CPM.pbs[g,])),
                             IFN=as.data.frame(t(CPM.ifn[g,]))) %>% set_colnames(c("PBS","IFN"))
df.paired.this.test = wilcox.test(t(CPM.pbs[g,]),t(CPM.ifn[g,]))

# DESeq2 analysis result
df.DEseq2.this <- res %>% dplyr::filter(gene==g)

# Plot
ggpaired(df.paired.this, cond1="PBS", cond2="IFN", 
         color="condition",palette=c("grey","red"),
         title=paste0("TMM-normalized CPM of ",g,"\n",
                      "wilcox test: p=",format(df.paired.this.test$p.value, digits=3,scientific=T),"\n",
                      "DE analysis: log2FC=", round(df.DEseq2.this$log2FoldChange,3)," ; p.adj=", 
                      format(df.DEseq2.this$padj,digits=3,scientific=T),"\n",
                      "donors=",as.character(nrow(df.paired.this))),
         xlab="condition",ylab="TMM-normalized CPM")


# IFIT1: up regulated in MEL. log2FC=3.38, p.adj=1.1E-17
# IFIT2: up-regulated in MEL. log2FC=5.15, p.adj=1.1E-149
# MX1: up regulated in MEL. log2FC=3.82, p.adj=1.88E-41
# OAS1: up regulated in MEL. log2FC=3.58, p.adj=1.97E-38

rm(g,df.paired.this,df.paired.this.test,df,df.DEseq2.this,df.this)
```
  
  