---
title: "skin-eQTL_celseq2_analysis"
author: "Shuo Shan"
date: "1/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages  
```{r package, warning=FALSE}
library(dplyr)
library(tidyverse)
#library(DESeq2)
#library(DEXSeq)
#library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
#library(ggrepel)
#library(ggfortify)
#library(reactable)
#library(edgeR)
#library(gridExtra)
#library(ggpubr)
#library(plotly)
#library(RCurl)
#library(fgsea) # FastGeneSetEncirhmentAnalysis


Dir="/Users/crystal/Desktop/nl"
setwd(Dir)

#save.image(file=paste0(Dir,'/myEnvironment.RData'))
#load(paste0(Dir,'/myEnvironment.RData'))
```
  
### Overview

RNAseq data for mouse DC and MAC under LPS stimulation.

```{r functions}
create_table <- function(Dir, file_lst) {
  f=file_lst[1]
  sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
  df <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
    tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
    mutate(sum=rowSums(select_if(., is.numeric))) %>%
    dplyr::select(c("region",sum)) %>% 
    magrittr::set_colnames(c("region",sample))
  
  if (length(file_lst) > 1) {
    for (f in file_lst[2:length(file_lst)]) {
      sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
        tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
        mutate(sum=rowSums(select_if(., is.numeric))) %>%
        dplyr::select(c("region",sum)) %>% 
        magrittr::set_colnames(c("region",sample))
    
    df <- dplyr::left_join(df,df.this,by="region")
    }
  }
  df <- df %>% tibble::column_to_rownames( . , "region")
  return(df)
  rm(f,sample,df,df.this)
}

# check which variable correlates with each PC the most.
best_correlation_with_pc <- function(pcs, metadata, pc.n) {
  total.pcc=c()
  pca.this=pcs[,pc.n]
  for (x in colnames(metadata)) {
    metadata.this <- as.numeric(as.factor(metadata[,x]))
    total.pcc=c(total.pcc,cor(pca.this,metadata.this))
  }
  total.df=data.frame(feature=colnames(metadata),pcc=total.pcc,pcc.magnitude=abs(total.pcc))
  return(total.df)
}
```
  
### Load raw ESAT counts
```{r, echo=FALSE, message=FALSE}
f <- read.table(paste0(Dir,"/compare_DC_Mac.tsv"))
# 1. import ESAT counts into master table
flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
  tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
samples.all <- data.frame(sample=colnames(counts))

# 2. find list of samples to keep
# 2a. condition used from pilot (KRT, FRB: 100ng 24h, MEL: 10ng 48h)
samples.pilot.keep <- c("F25_FRB_PBS_0ng_S0", "F25_FRB_24h_100ng_S0",
                        "F25_MEL_PBS_0ng_S0", "F25_MEL_48h_10ng_S0",
                        "F25_KRT_PBS_0ng_S0", "F25_KRT_24h_100ng_S0")
# 2b. latest version of samples (merged from replicates if any)
samples <- samples.all %>%
  tidyr::separate( . , col=sample, into=c("donor","celltype","condition","replicate"), sep="_",remove=FALSE) %>%
  dplyr::filter(donor!="F25") %>% tidyr::unite(. , "biosample", c("donor","celltype","condition"),sep="_", remove=FALSE)
# go through the list, and for each donor, pick the row that contain the highest number of replicate
samples.latest <- c()
for (d in unique(samples$biosample)) {
  df.this <- samples %>% dplyr::filter(biosample==d) %>% arrange( . , desc(replicate))
  sample.pick <- df.this[1,1]
  samples.latest <- c(samples.latest, sample.pick)
  rm(df.this,sample.pick)
}
# 2c. only keratinocytes.
samples.keep <- c(samples.latest, samples.pilot.keep)
samples.keep <- samples.keep[grep("FRB",samples.keep)]
# 2d. remove samples with indicated keywords (after QC inspection. will update until more sequence reads come in)
#keyword1 <- "F22|F24|F28|F30|F31|F32|F33"
keyword1 <- "F22"
samples.keep <- samples.keep[-grep(keyword1,samples.keep)]
# 2e. specificy the exact biosample to keep
#samples.keep <- c(samples.keep[-grep("F34_KRT_PBS",samples.keep)],"F34_KRT_PBS_S2")

# 2 combine
counts.keep <- counts %>% dplyr::select(samples.keep)
colnames(counts.keep) <- colnames(counts.keep) %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)
```
  
### Load metadata
```{r metadata, echo=FALSE, message=FALSE}
# parse kept samples into table
samples <- data.frame(sample=colnames(counts.keep))
# import metadata master table
metadata.master=read.table("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/metadata_01192022.txt", sep="\t", header=TRUE)
# import DolphinNext pipeline overall summary
pip.summary=read.table("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_01152022/DolphinNext/report6124/summary/overall_summary.tsv", sep="\t", header=TRUE) %>% magrittr::set_colnames(c("sample","total_reads","Multimapped_Reads_Aligned_STAR","Unique_Reads_Aligned_STAR"))
pip.summary$sample <- pip.summary$sample %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)


# query imported tables with samples
rm(metadata)
metadata <- left_join(samples,metadata.master,by="sample")
metadata <- left_join(metadata,pip.summary,by="sample")

# compile 
metadata$seq_date <- metadata$seq_date %>% gsub("\\/",".",.)
metadata$lib_prep_date <- metadata$lib_prep_date %>% gsub("\\/",".",.)
metadata$batch <- paste(metadata$lib_prep_date,metadata$seq_date,sep="_") %>% as.factor()

# cleanup
rm(pip.summary, samples, samples.all)
```
  
  
### Overview of libraries  
Overview of library size distribution.
```{r lib-size, echo=F, message=F}
# library size distribution
df=data.frame(sample=metadata$sample,total_count=metadata$total_reads)
df=df[order(-df$total_count),]
plot_ly(y = df$sample,x = df$total_count,type = "bar")

# histogram for gene counts, all samples
h <- hist(log10(rowSums(data)), breaks = as.numeric(100), plot = FALSE) 
plot_ly(x = h$mids, y = h$counts, width = 500, height=300) %>% layout( title = "Histogram") %>% add_bars()

# which libraries have low reads? (<5M)
df$sample[which(df$total_count<5000000)] %>% gsub("_FRB.*","",.) %>% table()

```

### TMM normalization  
```{r tmm, echo=FALSE, message=F}
dge <- DGEList(counts.keep)
dge <- calcNormFactors(dge)
dge$samples # check normalizaiton factors
CPM <- cpm(dge, log=FALSE)

metadata.pbs <- metadata %>% dplyr::filter(condition=="PBS")
metadata.ifn <- metadata %>% dplyr::filter(condition=="IFN")
# Getting normalized CPM matrix
CPM.pbs <- as.data.frame(CPM)[,as.character(metadata.pbs$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.pbs$donor)) 
CPM.ifn <- as.data.frame(CPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) 

# Getting log transformed CPM
logCPM <- log2(CPM+1) %>% as.data.frame() 

logCPM.pbs <- as.data.frame(logCPM)[,as.character(metadata.pbs$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.pbs$donor)) 

logCPM.ifn <- as.data.frame(logCPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) 

log2FC <- logCPM.ifn - logCPM.pbs

# inspect distribution of log-transformed CPM:
mean(apply(logCPM.pbs, 2, sd))
mean(apply(logCPM.pbs, 2, mean))
plot(density(logCPM.pbs[,1]))
```

### PCA
  
```{r pca}
# pick genes of top row variance
mat.logCPM.lnsd <- apply(logCPM,MARGIN=1,FUN=sd)
mat.logCPM.lncv <- sqrt( (exp(mat.logCPM.lnsd) )^2 -1) #coefficient of variance
logCPM.top <- logCPM[order(mat.logCPM.lncv,decreasing=T),][1:2000,]

# PCA
pca = prcomp(t(logCPM.top),centered= T,scale = T)
loadings=as.data.frame(pca$rotation)
# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)
# check which variable correlates with each PC the most.
tot=best_correlation_with_pc(pcs,metadata,4)
# PC1 (condition,pcc=0.99), PC2,3 (batch-effect like prep&seq dates), PC4 (batch-effect like seq_model)

# plot
# top PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], PC4=pca$x[,4], PC5=pca$x[,5],PC6=pca$x[,6],sample=rownames(pcs))
d=inner_join(d,metadata,by="sample") %>% magrittr::set_rownames(d$sample)
colnames(metadata)
#d$group_tags <- cut(d$Unique_Reads_Aligned_STAR, breaks=6)
d$group_tags <- cut(d$total_reads, 
                    breaks=c(min(d$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d$total_reads)))
d %>%
  mutate(libprep_seq_dates=paste(d$lib_prep_date,d$seq_date)) %>%
  ggplot(., aes(y=PC1,x=PC2,col=libprep_seq_dates,shape=condition,label=biosample)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  ggsave(filename=paste0(Dir,"/PCA_1.png"))

```
  
  
### DE analysis  
The goal of Differential gene expression analysis is to find 
genes or transcripts whose difference in expression, when accounting 
for the variance within condition, is higher than expected by chance. 

In this case a total sum of 10 counts separates well expressed 
from non-expressed genes. You can change this value and padj value and 
log2FoldChange cutoffs according to your data

```{r DESeq2}
df <- apply(counts.keep[rowSums(counts.keep)>=10,],2,round,0) %>% as.data.frame()

dds <- DESeqDataSetFromMatrix(countData = df,
                              colData = metadata,
                              design = ~ batch + condition)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("condition", "IFN", "PBS")))

res <- results(dds,contrast=c("condition", "IFN", "PBS")) %>% as.data.frame %>% rownames_to_column("gene")
res.up <- res %>% filter(log2FoldChange > 1.5) %>% filter(padj < 0.05)
res.down <- res %>% filter(log2FoldChange < -1.5) %>% filter(padj < 0.05)
res.nochange <- res %>% filter(abs(log2FoldChange) <= 1.5) %>% filter(padj < 0.05) %>% filter(!is.na(padj))
# design = ~ batch + condition: 574 res.up, 151 res.down
```

### Plot paired data in CPM  

```{r paired-beehive}
g="IL15RA" # IL1A, IL1RN, GM-CSF, TNF

# pick out TMM-normalized CPM paired data & run wilcox test
df.paired.this <- data.frame(PBS=as.data.frame(t(CPM.pbs[g,])),
                             IFN=as.data.frame(t(CPM.ifn[g,]))) %>% set_colnames(c("PBS","IFN"))
df.paired.this.test = wilcox.test(t(CPM.pbs[g,]),t(CPM.ifn[g,]))

# DESeq2 analysis result
df.DEseq2.this <- res %>% dplyr::filter(gene==g)

# Plot
ggpaired(df.paired.this, cond1="PBS", cond2="IFN", 
         color="condition",palette=c("grey","red"),
         title=paste0("TMM-normalized CPM of ",g,"\n",
                      "wilcox test: p=",format(df.paired.this.test$p.value, digits=3,scientific=T),"\n",
                      "DE analysis: log2FC=", round(df.DEseq2.this$log2FoldChange,3)," ; p.adj=", 
                      format(df.DEseq2.this$padj,digits=3,scientific=T),"\n",
                      "donors=",as.character(nrow(df.paired.this))),
         xlab="condition",ylab="TMM-normalized CPM")

# IFIT1: mildly up-regulated in KRT. log2FC=1.2, p.adj=2.4E-05. Some samples had no change.
# IFIT2: extremely up-regulated in KRT. log2FC=5.3, p.adj=4.2E-214. very dramatic. PBS had level of 0, but IFN had wide range.
# MX1: up regulated in KRT. log2FC=2, p.adj=1.09E-27
# OAS1: up regulated in KRT. log2FC=2.26, p.adj=2.08E-61
# IL1A: mildly up-regulated in KRT. log2FC=0.55, p.adj=6.2E-07.
# IL1AN: very mildly up-regulated in KRT. log2FC=0.32, p.adj=2.48E-02.
# TNF: upregulated overall in KRT, but a few samples went down? log2FC=2.031, p.adj=3.14E-06.

rm(g,df.paired.this,df.paired.this.test,df,df.DEseq2.this,df.this)
```
  
  
### K-means-clustering & hierarchical clustering
```{r kmclusters}
df <- logCPM

# set of candidate genes for clustering
n=20 # top number of genes with most abs(log2FC)
# candidate_genes <- res %>% as.data.frame %>%
#   dplyr::filter(padj < 0.05) %>%                   # restrict FDR
#   dplyr::filter(abs(log2FoldChange) >= 1.5) %>%    # restrict effect size
#   dplyr::top_n(. , n, abs(log2FoldChange)) %>%     # restrict top FC response
#   pull(gene) %>%
#   unique()

# another way of combining candidate genes
candidate_genes_up <- res %>% as.data.frame %>%
  dplyr::filter(padj < 0.05) %>%                   # restrict FDR
  dplyr::filter(log2FoldChange >= 1.5) %>%    # restrict effect size
  dplyr::top_n(. , n/2, abs(log2FoldChange)) %>%     # restrict top FC response
  pull(gene) %>%
  unique()
candidate_genes_down <- res %>% as.data.frame %>%
  dplyr::filter(padj < 0.05) %>%                   # restrict FDR
  dplyr::filter(log2FoldChange <= 1.5) %>%    # restrict effect size
  dplyr::top_n(. , n/2, abs(log2FoldChange)) %>%     # restrict top FC response
  pull(gene) %>%
  unique()
candidate_genes <- c(candidate_genes_up,candidate_genes_down)

# filter data matrix based on candidate genes
df <- df %>% dplyr::filter(rownames(.) %in% candidate_genes)
  
# perform k-means-clustering to determine k_row.
# wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
# for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
# plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# hierarchical clustering to check number of genes that fall in clusters.
gene_hclust <- hclust(dist(df,method="euclidean"), method = "complete")
cutree(gene_hclust, k = 2) %>% 
  enframe() %>% 
  rename(gene=name,cluster=value) %>%
  tidyr::separate(.,gene,into=c("gene","response"),sep="_") %>%
  group_by(response,cluster) %>%
  summarise(n=n())

# heatmap
heatmaply(df, 
          k_row = 2,
          scale = "row",
          fontsize_row = 5,
          fontsize_col = 5,
          col_side_colors = metadata[,c("condition","batch")],
          main = "row-scaled log2 TMM-normalized CPM",
          file = c(paste0(Dir,"/heatmaply_log2TMM-CPM_top20.html")))

```


### Write to file
```{r output}
# set global theme for reactable
options(reactable.theme = reactableTheme(borderColor = "#dfe2e5",stripedColor = "#f6f8fa",highlightColor = "#f0f5f9",cellPadding = "8px 12px",style = list(fontFamily = "Helvetica")))

# write raw ESAT counts
write.table(counts.keep, file=paste0(Dir,"/raw_counts.txt"), quote=F, sep="\t", row.names=T, col.names=T)
htmlwidgets::saveWidget(widget = reactable(counts.keep %>% rownames_to_column("gene"),
                                           filterable=T, searchable=T, striped=T, bordered=T), 
                        file = paste0(Dir,"/raw_counts.html"), selfcontained = TRUE)
# write metadata
metadata$RPI_index[61:62]="NA"
metadata$CS2_index[61:62]="NA"
write.table(metadata, file=paste0(Dir,"/metadata.txt"),quote=F, sep="\t", row.names=T, col.names=T)

# write TMM-normalized CPM 
write.table(CPM, file=paste0(Dir,"/TMM-normalized-CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(logCPM, file=paste0(Dir,"/log2-TMM-normalized-CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(CPM.ifn, file=paste0(Dir,"/TMM-normalized-CPM_IFN.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(CPM.pbs, file=paste0(Dir,"/TMM-normalized-CPM_PBS.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# write DESeq2 results
write.table(res,file=paste0(Dir,"/DESeq2_results_IFNvsPBS.txt"),quote=F,row.names=T,col.names=T)
write.table(res.up$gene,file=paste0(Dir,"/upregulated_genes.txt"),quote=F,row.names=F,col.names=F)
write.table(res.down$gene,file=paste0(Dir,"/downregulated_genes.txt"),quote=F,row.names=F,col.names=F)
htmlwidgets::saveWidget(widget = reactable(res,filterable=T, searchable=T, striped=T, bordered=T), 
                        file = paste0(Dir,"/DESeq2_results_IFNvsPBS.html"), selfcontained = TRUE)
write.table(res$gene, file=paste0(Dir,"/genes_used_in_DE.txt"), quote=F, sep="\t", row.names=F, col.names=F)


# write log2FoldChange of CPM data
log2FC=log2((CPM.ifn+0.0001)/(CPM.pbs+0.0001))
write.table(log2FC, file=paste0(Dir,"/log2FoldChange_by_IFNg.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# filter out genes that aren't highly expressed
gene_expressed_pbs = CPM.pbs %>% mutate(mean=apply(CPM.pbs, 1, mean)) %>% dplyr::filter(mean>10) %>% rownames(.) 
gene_expressed_ifn = CPM.ifn %>% mutate(mean=apply(CPM.ifn, 1, mean)) %>% dplyr::filter(mean>10) %>% rownames(.)
gene_expressed = intersect(gene_expressed_pbs,gene_expressed_ifn)
write.table(gene_expressed, file=paste0(Dir,"/highly_expressed_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write more genes (expressed AND IFNg responsive)
gene_expressed_and_ifng_responsive = intersect(gene_expressed, c(res.up$gene,res.down$gene))
write.table(gene_expressed_and_ifng_responsive, file=paste0(Dir,"/highly_expressed_and_ifng_responsive_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write down genes that are responsive to IFNg
write.table(c(res.up$gene,res.down$gene),file=paste0(Dir,"/ifng_responsive_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)
```


















