```{r package, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(magrittr)
library(stringr)
library(gplots)
library(ggfortify)
library(gridExtra)
library(ggpubr)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
library(ggbeeswarm)
library(sva)
library(clusterProfiler)
library(plotly)
library(Mfuzz)
library(Biobase)
library(org.Hs.eg.db)  
library(enrichplot)


Dir="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/analysis_07142025/"
setwd(Dir)

#save.image(file=paste0(Dir,'/myEnvironment_main.RData'))
load(paste0(Dir,"/myEnvironment_main.RData")) # raw counts, CPM, log2(CPM+1), metadata, expressed genes by ct, dds object for each ct

# load meta genes to gene name dictionary
metagene_dict <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/metaidname.txt") %>% 
  dplyr::select(-id) %>% distinct()
metagene_dict_collapsed <- metagene_dict %>% group_by(meta) %>% 
  summarize(source=paste(name,collapse=','))
```

```{r functions}
# useful function
convert_meta_genes <- function(gene_vector) {
  # Identify meta genes in the input vector
  meta_genes <- gene_vector[grepl("meta", gene_vector)]
  non_meta_genes <- gene_vector[!grepl("meta", gene_vector)]
  
  # Convert meta genes to actual gene names using metagene_dict
  meta_to_gene <- metagene_dict$name[metagene_dict$meta %in% meta_genes]
  
  # Combine non-meta genes and converted meta genes, ensuring uniqueness
  converted_genes <- unique(c(non_meta_genes, meta_to_gene))
  
  return(converted_genes)
}
niceKmeans <-function (df, k, kmeans.algorithm = "Lloyd", iter.max = 1000, nstart=100) {
  
  result = data.frame();
  source = t(scale(t(df)));
  kmeans = kmeans(source, centers = k, iter.max = iter.max, algorithm=kmeans.algorithm, nstart=nstart);
  clustered = data.frame();
  breaks = c();
  for (i in 1:k) {
    print(sprintf("Doing cluster %i",i))
    cluster = source[kmeans$cluster==i,];
    rows = row.names(cluster);
    clust = heatmap.2(as.matrix(cluster), distfun=function(x) as.dist((1-cor(t(x)))/2))
    cluster.ordered = cluster[clust$rowInd,];
    cluster.ordered.genes = rows[clust$rowInd];
    row.names(cluster.ordered) = cluster.ordered.genes;
    class = data.frame(row.names=cluster.ordered.genes);
    class[,"class"] = i 
    cluster.ordered = cbind(cluster.ordered, class);
    clustered = rbind(clustered, cluster.ordered);
    if(i > 1 & i < k) {
      breaks[i] = as.numeric(breaks[i-1]) + length(rows);
    } else if(i==1) {
      breaks[i] = length(rows);
    }
  }
  
  result <- list();
  result$clustered <- clustered;
  result$breaks <- breaks;
  return(result);
}
create_table <- function(Dir, file_lst) {
  f=file_lst[1]
  sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
  df <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
    tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
    mutate(sum=rowSums(select_if(., is.numeric))) %>%
    dplyr::select(c("region",sum)) %>% 
    magrittr::set_colnames(c("region",sample))
  
  if (length(file_lst) > 1) {
    for (f in file_lst[2:length(file_lst)]) {
      sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
        tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
        mutate(sum=rowSums(select_if(., is.numeric))) %>%
        dplyr::select(c("region",sum)) %>% 
        magrittr::set_colnames(c("region",sample))
    
    df <- dplyr::left_join(df,df.this,by="region")
    }
  }
  df <- df %>% tibble::column_to_rownames( . , "region")
  return(df)
  rm(f,sample,df,df.this)
}
convert_to_entrez <- function(gene_list) {
  bitr(gene_list, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
}
convert_from_entrez <- function(id_list) {
  bitr(id_list, fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Hs.eg.db)
}

plot_paired_plot <- function(gene){
  g=gene
  df <- data.frame(CPM = as.numeric(t(CPM.clean[g,])), 
                   sample = colnames(CPM.clean)) %>%
    left_join(. , metadata[,c("sample","donor","celltype","condition")], by="sample")
  
  df.ifng <- df %>% dplyr::filter(condition%in%c("PBS","IFNG"))
  df.ifnb <- df %>% dplyr::filter(condition%in%c("PBS","IFNB"))
  df.tnf <- df %>% dplyr::filter(condition%in%c("PBS","TNF"))
  
  # make paired plots
  position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
  line.color = "grey"; linetype = "solid"; palette="bright"
  
  p.ifng <- ggplot(df.ifng, create_aes(list(x = "condition", y = "CPM"))) +
    geom_exec(geom_point, data = df.ifng, color = "condition", size = point.size,
              position = "jitter") +
    scale_color_manual(values=c("#999999","#E69F00"))+
    labs(title=g)+
    theme(legend.position="none") +
    facet_wrap(~ celltype)
  
  p.ifnb <- ggplot(df.ifnb, create_aes(list(x = "condition", y = "CPM"))) +
    geom_exec(geom_line, data = df.ifnb, group = "donor",
              color = line.color, size = line.size, linetype = linetype,
              position = position) +
    geom_exec(geom_point, data = df.ifnb, color = "condition", size = point.size,
              position = position) +
    scale_color_manual(values=c("#999999","darkgreen"))+
    labs(title=g)+
    theme(legend.position="none") +
    facet_wrap(~ celltype)
  
  p.tnf <- ggplot(df.tnf, create_aes(list(x = "condition", y = "CPM"))) +
    geom_exec(geom_line, data = df.tnf, group = "donor",
              color = line.color, size = line.size, linetype = linetype,
              position = position) +
    geom_exec(geom_point, data = df.tnf, color = "condition", size = point.size,
              position = position) +
    scale_color_manual(values=c("#999999","purple"))+
    labs(title=g)+
    theme(legend.position="none") +
    facet_wrap(~ celltype)
  
  final_plot <- plot_grid(
    p.ifng, p.ifnb, p.tnf,
    nrow = 1, ncol = 3
  )
  
  return(final_plot)
}

plot_violinbox_plot <- function(gene){
  g=gene
  df <- data.frame(CPM = as.numeric(t(CPM.clean[g,])), 
                   sample = colnames(CPM.clean)) %>%
    left_join(. , metadata[,c("sample","donor","celltype","condition")], by="sample")
  
  df.ifng <- df %>% dplyr::filter(condition%in%c("PBS","IFNG"))
  df.ifnb <- df %>% dplyr::filter(condition%in%c("PBS","IFNB"))
  df.tnf <- df %>% dplyr::filter(condition%in%c("PBS","TNF"))
  
  # make paired plots
  position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
  line.color = "grey"; linetype = "solid"; palette="bright"
  
  p.ifng <- ggplot(df.ifng, aes(x = condition, y = CPM, fill = condition, color = condition)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "IFNG" = "#E69F00")) +
    scale_color_manual(values = c("PBS" = "#999999", "IFNG" = "#E69F00")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~celltype)
  
  p.ifnb <- ggplot(df.ifnb, aes(x = condition, y = CPM, fill = condition, color = condition)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "IFNB" = "darkgreen")) +
    scale_color_manual(values = c("PBS" = "#999999", "IFNB" = "darkgreen")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~celltype)
  
  p.tnf <- ggplot(df.tnf, aes(x = condition, y = CPM, fill = condition, color = condition)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "TNF" = "purple")) +
    scale_color_manual(values = c("PBS" = "#999999", "TNF" = "purple")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~celltype)
  
  final_plot <- plot_grid(
    p.ifng, p.ifnb, p.tnf,
    nrow = 1, ncol = 3
  )
  
  return(final_plot)
}

plot_violinbox_plot_cytokine_faceted <- function(gene){
  g=gene
  df <- data.frame(CPM = as.numeric(t(CPM.clean[g,])), 
                   sample = colnames(CPM.clean)) %>%
    left_join(. , metadata[,c("sample","donor","celltype","condition","phase")], by="sample")
  
  df$condition <- ordered(df$condition, levels=c("PBS","IFNB","IFNG","TNF"))
  
  df.krt <- df %>% dplyr::filter(celltype=="KRT")
  df.mel <- df %>% dplyr::filter(celltype=="MEL")
  df.frb <- df %>% dplyr::filter(celltype=="FRB")
  
  # make paired plots
  position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
  line.color = "grey"; linetype = "solid"; palette="bright"
  
  p.krt <- ggplot(df.krt, aes(x = celltype, y = CPM, fill = celltype, color = celltype)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "KRT" = "#DAF7A6")) +
    scale_color_manual(values = c("PBS" = "#999999", "KRT" = "#DAF7A6")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~condition, nrow=1)
  
  p.mel <- ggplot(df.mel, aes(x = celltype, y = CPM, fill = celltype, color = celltype)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "MEL" = "#C68452")) +
    scale_color_manual(values = c("PBS" = "#999999", "MEL" = "#C68452")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~condition, nrow=1)
  
    p.frb <- ggplot(df.frb, aes(x = celltype, y = CPM, fill = celltype, color = celltype)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "FRB" = "#CCCCFF")) +
    scale_color_manual(values = c("PBS" = "#999999", "FRB" = "#CCCCFF")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~condition, nrow=1)
  
  final_plot <- plot_grid(
    p.krt, p.mel, p.frb,
    nrow = 1, ncol = 3
  )
  
  return(final_plot)
}

plot_violinbox_plot_cytokine_faceted_by_celltype_grouped_by_phase <- function(gene){
  g=gene
  df <- data.frame(CPM = as.numeric(t(CPM.clean[g,])), 
                   sample = colnames(CPM.clean)) %>%
    left_join(. , metadata[,c("sample","donor","celltype","condition","phase")], by="sample")
  
  df$condition <- ordered(df$condition, levels=c("PBS","IFNB","IFNG","TNF"))
  
  df.krt <- df %>% dplyr::filter(celltype=="KRT")
  df.mel <- df %>% dplyr::filter(celltype=="MEL")
  df.frb <- df %>% dplyr::filter(celltype=="FRB")
  
  # make paired plots
  position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
  line.color = "grey"; linetype = "solid"; palette="bright"
  
  p.krt <- ggplot(df.krt, aes(x = phase, y = CPM, fill = celltype, color = celltype, group = phase)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "KRT" = "#DAF7A6")) +
    scale_color_manual(values = c("PBS" = "#999999", "KRT" = "#DAF7A6")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~condition, nrow=1)
  
  p.mel <- ggplot(df.mel, aes(x = phase, y = CPM, fill = celltype, color = celltype)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "MEL" = "#C68452")) +
    scale_color_manual(values = c("PBS" = "#999999", "MEL" = "#C68452")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~condition, nrow=1)
  
    p.frb <- ggplot(df.frb, aes(x = phase, y = CPM, fill = celltype, color = celltype)) +
    geom_violin(width = 0.9, alpha = 0.35, trim = FALSE) +
    ggbeeswarm::geom_quasirandom(width = 0.18, size = 2, alpha = 0.9) +
    geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("PBS" = "#999999", "FRB" = "#CCCCFF")) +
    scale_color_manual(values = c("PBS" = "#999999", "FRB" = "#CCCCFF")) +
    labs(title = g, x = NULL, y = "CPM") +
    theme_classic() +
    theme(legend.position = "none") +
    facet_wrap(~condition, nrow=1)
  
  final_plot <- plot_grid(
    p.krt, p.mel, p.frb,
    nrow = 1, ncol = 3
  )
  
  return(final_plot)
}
```
## Overview
Celseq2 RNAseq data from skin-eQTL data. phase1 and phase2 fastq files were processed, aligned, and counted by Via Foundry's ESAT 3.0 pipeline with custom genome with meta genes.
## load ESAT count tables, compile metadata, and calculate CPM and log2(CPM+1), and filter for expressed genes
### phase1 samples (no merged samples, lengths vary from 36bp to 56bp)
```{r load-phase1-countmatrix}
Dir.esat.list=c("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/DolphinNext/phase1_rerun_customGenome/report10996/ESAT/",
                "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/DolphinNext/phase1_rerun_customGenome/F44_KRT_PBS/report11029/ESAT/")

counts.phase1.list <- list()
metadata.phase1.list <- list()

for (i in 1:length(Dir.esat.list)) {
  # fetch ESAT directory
  Dir.esat=Dir.esat.list[i]
  
  # 1. import ESAT counts into master table
  flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
  counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
    tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
  col.sum = data.frame(sample = colnames(counts), samplesum = apply(counts,2,sum))
  
  # remove mitochondria genes
  MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
  counts <- counts %>% mutate(gene=rownames(.)) %>% 
    dplyr::filter(!rownames(.) %in% MTgenes$V1) %>% dplyr::select(-"gene")
  dim(counts)
  
  # mapping summary
  info_alignment <- data.table::fread(paste0(Dir.esat,"/../summary/overall_summary.tsv")) %>%
    set_colnames(c("sample","total_reads","multimapped_reads_aligned","unique_reads_aligned")) %>%
    left_join( . , col.sum, by="sample") %>%
    dplyr::mutate(reads_aligned = multimapped_reads_aligned + unique_reads_aligned) %>%
    dplyr::mutate(all_aligned_reads_over_total = (multimapped_reads_aligned + unique_reads_aligned) / total_reads) %>%
    dplyr::mutate(unique_aligned_reads_over_total = unique_reads_aligned / total_reads)

  # metadata
  metadata <- info_alignment
  
  # populate large list
  counts.phase1.list[[i]] <- counts
  metadata.phase1.list[[i]] <- metadata
  rm(counts, metadata, info_alignment, col.sum, flst,Dir.esat)
}

# combine all data frames by column
counts.phase1 <- do.call(cbind, counts.phase1.list) %>%
  dplyr::select(-F42_KRT_PBS_S1)# this was accidentally included in the run, but I used NovaSeq data for this sample instead

# import metadata
temp <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/metadata_phase1_selected_metadata_07082025.txt") %>% arrange(sample)
temp2 <- do.call(rbind, metadata.phase1.list) %>%
  dplyr::filter(!sample=="F42_KRT_PBS_S1") %>%
  arrange(sample) %>%
  mutate(CS2=NA)
metadata.phase1 <- left_join(temp, temp2, by="sample")
metadata.phase1 <- metadata.phase1[,c("sample","read_length","donor","celltype","condition","CS2",
                                      "phase","technical_rep","sequencing_platform",
                                      "total_reads","multimapped_reads_aligned","unique_reads_aligned",
                                      "samplesum","reads_aligned","all_aligned_reads_over_total","unique_aligned_reads_over_total")]

rm(counts.phase1.list, metadata.phase1.list, temp, temp2)
```
### 56bp
```{r load-56bp}
Dir.esat.list=c("~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/trimmed_FRB/report11013/ESAT/",
                "~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/trimmed_MEL/report11012/ESAT/",
                "~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/trimmed_KRT/report11011/ESAT/")

counts.56bp.list <- list()
metadata.56bp.list <- list()

for (i in 1:length(Dir.esat.list)) {
  # fetch ESAT directory
  Dir.esat=Dir.esat.list[i]
    
  # 1. import ESAT counts into master table
  flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
  counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
    tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
  col.sum = data.frame(sample = colnames(counts), samplesum = apply(counts,2,sum))
  
  # remove mitochondria genes
  MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
  counts <- counts %>% mutate(gene=rownames(.)) %>% 
    dplyr::filter(!rownames(.) %in% MTgenes$V1) %>% dplyr::select(-"gene")
  print(Dir.esat)
  print(dim(counts))
  
  # mapping summary
  info_alignment <- data.table::fread(paste0(Dir.esat,"/../summary/overall_summary.tsv")) %>%
    set_colnames(c("sample","total_reads","multimapped_reads_aligned","unique_reads_aligned")) %>%
    left_join( . , col.sum, by="sample") %>%
    dplyr::mutate(reads_aligned = multimapped_reads_aligned + unique_reads_aligned) %>%
    dplyr::mutate(all_aligned_reads_over_total = (multimapped_reads_aligned + unique_reads_aligned) / total_reads) %>%
    dplyr::mutate(unique_aligned_reads_over_total = unique_reads_aligned / total_reads)
  
  # metadata
  metadata <- data.frame(sample = colnames(counts), tosplit = colnames(counts)) %>%
    tidyr::separate(., col=tosplit, into=c(NA,"CS2","donor","celltype","condition",NA,NA),sep="_") %>%
    dplyr::mutate(read_length="56bp") %>%
    dplyr::mutate(phase="phase2") %>%
    dplyr::mutate(technical_rep="S1") %>%
    dplyr::mutate(sequencing_platform="NovaSeq") %>%
    right_join( . , info_alignment)
  metadata$condition <- ordered(metadata$condition, levels=c("PBS","IFNG","IFNB","TNF"))
  metadata$celltype <- ordered(metadata$celltype, levels=c("FRB","KRT","MEL"))
  
  # populate large list
  counts.56bp.list[[i]] <- counts
  metadata.56bp.list[[i]] <- metadata
  rm(counts, metadata, info_alignment, col.sum, flst)
}
# combine all data frames by column
counts.56bp <- do.call(cbind, counts.56bp.list)
metadata.56bp <- do.call(rbind, metadata.56bp.list)
metadata.56bp <- metadata.56bp[,c("sample","read_length","donor","celltype","condition","CS2",
                                      "phase","technical_rep","sequencing_platform",
                                      "total_reads","multimapped_reads_aligned","unique_reads_aligned",
                                      "samplesum","reads_aligned","all_aligned_reads_over_total","unique_aligned_reads_over_total")]

rm(counts.56bp.list, metadata.56bp.list)
```

### combine count matrices and correct for mislabeled samples
```{r combine-counts-matrix}
tmp.counts <- cbind(counts.phase1, counts.56bp)
tmp.metadata <- rbind(metadata.phase1, metadata.56bp)
tmp.metadata <- tmp.metadata[match(colnames(tmp.counts), tmp.metadata$sample), ]
rm(counts.phase1, counts.56bp)
rm(metadata.phase1,metadata.56bp)
rm(MTgenes, Dir.esat.list, i)

# F72, F75, F76, F78, F80, F81's fibroblast RNA samples were mistakenly labeled as "keratinocyte" and placed in KRT's box. now we have duplicated FRB libraries for those 6 donors in all four conditions. Based on the aligned reads, I pick and choose which one of the duplicates to keep for the FRB samples, and discard the other copy.
to_keep <- tmp.metadata %>% 
  mutate(biosample = paste0(donor,"_",celltype,"_",condition)) %>%
  dplyr::filter(biosample %in% c("F72_FRB_PBS", "F72_FRB_IFNG", "F72_FRB_IFNB", "F72_FRB_TNF",
  "F75_FRB_PBS", "F75_FRB_IFNG", "F75_FRB_IFNB", "F75_FRB_TNF",
  "F76_KRT_PBS", "F76_KRT_IFNG", "F76_KRT_IFNB", "F76_FRB_TNF",
  "F78_KRT_PBS", "F78_KRT_IFNG", "F78_KRT_IFNB", "F78_FRB_TNF",
  "F80_KRT_PBS", "F80_KRT_IFNG", "F80_FRB_IFNB", "F80_FRB_TNF",
  "F81_KRT_PBS", "F81_FRB_IFNG", "F81_KRT_IFNB", "F81_KRT_TNF")) %>%
  pull(sample)

to_delete <- tmp.metadata %>% 
  mutate(biosample = paste0(donor,"_",celltype,"_",condition)) %>%
  dplyr::filter(biosample %in% c("F72_FRB_PBS", "F72_FRB_IFNG", "F72_FRB_IFNB", "F72_FRB_TNF",
  "F75_FRB_PBS", "F75_FRB_IFNG", "F75_FRB_IFNB", "F75_FRB_TNF",
  "F76_FRB_PBS", "F76_FRB_IFNG", "F76_FRB_IFNB", "F76_FRB_TNF",
  "F78_FRB_PBS", "F78_FRB_IFNG", "F78_FRB_IFNB", "F78_FRB_TNF",
  "F80_FRB_PBS", "F80_FRB_IFNG", "F80_FRB_IFNB", "F80_FRB_TNF",
  "F81_FRB_PBS", "F81_FRB_IFNG", "F81_FRB_IFNB", "F81_FRB_TNF",
  "F72_KRT_PBS", "F72_KRT_IFNG", "F72_KRT_IFNB", "F72_KRT_TNF",
  "F75_KRT_PBS", "F75_KRT_IFNG", "F75_KRT_IFNB", "F75_KRT_TNF",
  "F76_KRT_PBS", "F76_KRT_IFNG", "F76_KRT_IFNB", "F76_KRT_TNF",
  "F78_KRT_PBS", "F78_KRT_IFNG", "F78_KRT_IFNB", "F78_KRT_TNF",
  "F80_KRT_PBS", "F80_KRT_IFNG", "F80_KRT_IFNB", "F80_KRT_TNF",
  "F81_KRT_PBS", "F81_KRT_IFNG", "F81_KRT_IFNB", "F81_KRT_TNF")) %>%
  pull(sample)

counts.keep = tmp.counts[,to_keep]
metadata.keep = tmp.metadata[match(colnames(counts.keep), tmp.metadata$sample), ]
metadata.keep$sample = gsub("KRT", "FRB", metadata.keep$sample)
metadata.keep$celltype = gsub("KRT", "FRB", metadata.keep$celltype)
colnames(counts.keep) = gsub("KRT", "FRB", colnames(counts.keep))

counts.cleaned = tmp.counts %>% dplyr::select(-to_delete)
metadata.cleaned = tmp.metadata[match(colnames(counts.cleaned), tmp.metadata$sample), ]

metadata = rbind(metadata.cleaned, metadata.keep)
metadata <- metadata %>%
  mutate(donor_num = as.numeric(sub("F", "", metadata$donor))) %>%
  arrange(condition, celltype, donor_num) %>%
  mutate(biosample=paste0(donor,"_",celltype,"_",condition))
counts = cbind(counts.cleaned, counts.keep)
counts = counts[, metadata$sample]
rm(to_keep, to_delete, tmp.counts, tmp.metadata, counts.keep, metadata.keep, counts.cleaned, metadata.cleaned)
```
### calculate CPM
```{r CPM}
## Calculate CPM
col.sum=apply(counts,2,sum)
CPM=sweep(counts, 2, col.sum, `/`)*1000000
log2CPM=log2(CPM+1)
rm(col.sum)
```

### filter for expressed genes. An expressed gene shall have CPM >= 10 in at least 5 donors in a condition.
```{r filter}
conditions <- c("PBS", "IFNG", "IFNB", "TNF")
celltypes <-c("FRB","MEL","KRT")
expressed_genes_list <- list()

for (ct in celltypes) {
  gene_expression_status <- logical(nrow(CPM))
  
  # Get sample names for this cell type
  ct_samples <- colnames(CPM)[grepl(paste0("_", ct, "_"), colnames(CPM))]
  
  for (i in 1:nrow(CPM)) {
    is_expressed <- FALSE
    for (cond in conditions) {
      cond_samples <- ct_samples[grepl(paste0("_", cond), ct_samples)]
      # Extract CPM values for this gene under this cell type and condition
      cpm_values <- CPM[i, cond_samples]
      if (length(which(cpm_values>=10)) >= 5) {
        is_expressed <- TRUE
        break
      }
    }
    
    gene_expression_status[i] <- is_expressed
  }
  
  expressed_genes_list[[ct]] <- rownames(CPM)[gene_expression_status]
}

# Filter for expressed genes
expressed_genes <- unique(unlist(expressed_genes_list))
expressed_genes.mel <- unique(expressed_genes_list[["MEL"]])
rm(celltypes, cond, cond_samples, conditions, ct, ct_samples, gene_expression_status, is_expressed, donor_ids, exprs_donors)
```
## PCA
```{r pca}
# Calculate Coefficient of Variation (COV)
gene_means <- rowMeans(log2CPM.expressed)
gene_sds <- apply(log2CPM.expressed, 1, sd)
gene_cov <- gene_sds / gene_means

# Remove genes with mean = 0 to avoid NaNs
gene_cov[is.na(gene_cov)] <- 0

# Select top N variable genes
top_n <- 3000
top_genes <- names(sort(gene_cov, decreasing = TRUE))[1:top_n]

# Subset log2CPM
log2CPM_subset <- log2CPM[top_genes, ]

#calculate principal components for top 3000 COV genes' log2CPM
pca_obj <- prcomp(t(log2CPM_subset), scale. = TRUE)

pca_scores <- as.data.frame(pca_obj$x)  # this is samples × PCs
pca_scores$sample_id <- colnames(counts)     # optional
pca_scores$celltype <- metadata$celltype
pca_scores$condition <- metadata$condition
pca_scores$donor <- metadata$donor
pca_scores$read_length <- metadata$read_length
pca_scores$sequencing_platform <- metadata$sequencing_platform
pca_scores$unique_aligned_reads_over_total <- metadata$unique_aligned_reads_over_total
# Get variance explained
percent_var <- round(100 * (pca_obj$sdev^2 / sum(pca_obj$sdev^2)), 1)
# Now use:
xlab <- paste0("PC1 (", percent_var[1], "%)")
ylab <- paste0("PC2 (", percent_var[2], "%)")

p1 = ggplot(data=pca_scores, aes(
  x = PC1,
  y = PC2,
  color = condition,
  #shape = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "celltype", shape = "condition",
       x = xlab, y = ylab) 

# Then enable the hover text in plotly
p1_interactive <- ggplotly(p1, tooltip = "text")
p1_interactive


# Extract top PC1 and PC2 gene loadings
top_pc1_genes <- names(sort(pca_obj$rotation[, 1], decreasing = TRUE)[1:30])
top_pc2_genes <- names(sort(pca_obj$rotation[, 2], decreasing = TRUE)[1:30])

# Combine and deduplicate
top_genes <- unique(c(top_pc1_genes, top_pc2_genes))

# Plot mean expression
mean_CPM <- rowMeans(CPM[top_genes, ])
barplot(mean_CPM, las=2, main="Mean CPM of top PC1/PC2 genes", col="steelblue")


# Use directional projection of gene loadings onto the cell-type separation vector
# Estimate centroids of each cell type in PC1–PC2 space:
frb_center <- colMeans(pca_scores[pca_scores$celltype == "FRB", c("PC1", "PC2")])
krt_center <- colMeans(pca_scores[pca_scores$celltype == "KRT", c("PC1", "PC2")])
mel_center <- colMeans(pca_scores[pca_scores$celltype == "MEL", c("PC1", "PC2")])

# Define the separation vector:
# Vector from FRB to KRT
vec_frb_to_krt <- krt_center - frb_center
unit_frb_to_krt <- vec_frb_to_krt / sqrt(sum(vec_frb_to_krt^2))

# Vector from FRB to MEL
vec_frb_to_mel <- mel_center - frb_center
unit_frb_to_mel <- vec_frb_to_mel / sqrt(sum(vec_frb_to_mel^2))

# Vector from KRT to MEL
vec_krt_to_mel <- mel_center - krt_center
unit_krt_to_mel <- vec_krt_to_mel / sqrt(sum(vec_krt_to_mel^2))

# Project each gene's PC1–PC2 loading onto the separation vector:
# Get gene loading matrix for PC1 and PC2
gene_loadings <- pca_obj$rotation[, 1:2]

# Get top directional genes:
# top_mel_genes_from_krt: genes that push samples from KRT → MEL
proj_frb_to_krt <- gene_loadings %*% unit_frb_to_krt
proj_frb_to_mel <- gene_loadings %*% unit_frb_to_mel
proj_krt_to_mel <- gene_loadings %*% unit_krt_to_mel

top_frb_genes_from_krt <- names(sort(proj_frb_to_krt[,1], decreasing = TRUE)[1:20])
top_krt_genes_from_frb <- names(sort(proj_frb_to_krt[,1], decreasing = FALSE)[1:20])

top_mel_genes_from_frb <- names(sort(proj_frb_to_mel[,1], decreasing = TRUE)[1:20])
top_frb_genes_from_mel <- names(sort(proj_frb_to_mel[,1], decreasing = FALSE)[1:20])

top_mel_genes_from_krt <- names(sort(proj_krt_to_mel[,1], decreasing = TRUE)[1:20])
top_krt_genes_from_mel <- names(sort(proj_krt_to_mel[,1], decreasing = FALSE)[1:20])


#### PC2 vs PC3 ####
xlab <- paste0("PC2 (", percent_var[2], "%)")
ylab <- paste0("PC3 (", percent_var[3], "%)")

p2 = ggplot(data=pca_scores, aes(
  x = PC2,
  y = PC3,
  color = sequencing_platform,
  shape = celltype,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "sequencing_platform", shape = "celltype",
       x = xlab, y = ylab) 
p2_interactive <- ggplotly(p2, tooltip = "text")
p2_interactive

#### PC3 vs PC4 ####
xlab <- paste0("PC3 (", percent_var[3], "%)")
ylab <- paste0("PC4 (", percent_var[4], "%)")

p3 = ggplot(data=pca_scores, aes(
  x = PC3,
  y = PC4,
  color = condition,
  shape = celltype,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "condition", shape = "celltype",
       x = xlab, y = ylab) 
p3_interactive <- ggplotly(p3, tooltip = "text")
p3_interactive

# which PC captures condition effect? bar plot 
ggplot(pca_scores, aes(x = condition, y = PC4, fill = condition)) +
  geom_boxplot() +
  labs(title = "PC4 captures condition effect", y = paste0("PC4 (", percent_var[4], "%)"))

# Run ANOVA across PCs to see which PC corresponds to "CONDITION"
pca_scores$condition <- metadata$condition

# Loop through top 10 PCs
for (i in 1:10) {
  cat(paste0("PC", i, ": "))
  print(summary(aov(pca_scores[,i] ~ pca_scores$condition)))
}
```
```{r pca_MEL}
# Calculate Coefficient of Variation (COV)
this_log2CPM <- log2CPM.expressed[, metadata$celltype == "MEL"]
this_metadata <- metadata[metadata$celltype == "MEL",]
gene_means <- rowMeans(this_log2CPM)
gene_sds <- apply(this_log2CPM, 1, sd)
gene_cov <- gene_sds / gene_means

# Remove genes with mean = 0 to avoid NaNs
gene_cov[is.na(gene_cov)] <- 0

# Select top N variable genes
top_n <- 3000
top_genes <- names(sort(gene_cov, decreasing = TRUE))[1:top_n]

# Subset log2CPM
log2CPM_subset <- this_log2CPM[top_genes, ]

#calculate principal components for top 3000 COV genes' log2CPM
pca_obj <- prcomp(t(log2CPM_subset), scale. = TRUE)

pca_scores <- as.data.frame(pca_obj$x)  # this is samples × PCs
pca_scores$sample_id <- colnames(this_log2CPM)     # optional
pca_scores$celltype <- this_metadata$celltype
pca_scores$condition <- this_metadata$condition
pca_scores$donor <- this_metadata$donor
pca_scores$read_length <- this_metadata$read_length
pca_scores$sequencing_platform <- this_metadata$sequencing_platform
pca_scores$unique_aligned_reads_over_total <- this_metadata$unique_aligned_reads_over_total
pca_scores$total_reads <- this_metadata$total_reads
# Get variance explained
percent_var <- round(100 * (pca_obj$sdev^2 / sum(pca_obj$sdev^2)), 1)
# Now use:
xlab <- paste0("PC1 (", percent_var[1], "%)")
ylab <- paste0("PC2 (", percent_var[2], "%)")

p1 = ggplot(data=pca_scores, aes(
  x = PC1,
  y = PC2,
  color = condition,
  #shape = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total,
    "<br>Total reads:", total_reads
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 209 melanocytes samples\ntop 3000 COV expressed genes",
       color = "condition", shape = "condition",
       x = xlab, y = ylab) 

# Then enable the hover text in plotly
p1_interactive <- ggplotly(p1, tooltip = "text")
p1_interactive


# Extract top PC1 and PC2 gene loadings
top_pc1_genes <- names(sort(pca_obj$rotation[, 1], decreasing = TRUE)[1:30])
bottom_pc1_genes <- names(sort(pca_obj$rotation[, 1], decreasing = FALSE)[1:30])

top_pc2_genes <- names(sort(pca_obj$rotation[, 2], decreasing = TRUE)[1:30])
bottom_pc2_genes <- names(sort(pca_obj$rotation[, 2], decreasing = FALSE)[1:30])

#### PC2 vs PC3 ####
xlab <- paste0("PC2 (", percent_var[2], "%)")
ylab <- paste0("PC3 (", percent_var[3], "%)")

p2 = ggplot(data=pca_scores, aes(
  x = PC2,
  y = PC3,
  color = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "sequencing_platform", shape = "celltype",
       x = xlab, y = ylab) 
p2_interactive <- ggplotly(p2, tooltip = "text")
p2_interactive

#### PC3 vs PC4 ####
xlab <- paste0("PC3 (", percent_var[3], "%)")
ylab <- paste0("PC4 (", percent_var[4], "%)")

p3 = ggplot(data=pca_scores, aes(
  x = PC3,
  y = PC4,
  color = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "condition", shape = "celltype",
       x = xlab, y = ylab) 
p3_interactive <- ggplotly(p3, tooltip = "text")
p3_interactive

# which PC captures condition effect? bar plot 
ggplot(pca_scores, aes(x = condition, y = PC4, fill = condition)) +
  geom_boxplot() +
  labs(title = "PC4 captures condition effect", y = paste0("PC3 (", percent_var[3], "%)"))

# Run ANOVA across PCs to see which PC corresponds to "CONDITION"
pca_scores$condition <- metadata$condition

# Loop through top 10 PCs
for (i in 1:10) {
  cat(paste0("PC", i, ": "))
  print(summary(aov(pca_scores[,i] ~ pca_scores$condition)))
}
```
```{r pca-krt-and-frb}
this_samples = colnames(counts)[grepl("FRB|KRT", colnames(counts))]
this_counts = counts[, this_samples]
this_metadata = metadata[match(colnames(this_counts), metadata$sample), ]

#calculate principal components for the raw count matrix
pca_obj <- prcomp(t(this_counts))

pca_scores <- as.data.frame(pca_obj$x)  # this is samples × PCs
pca_scores$sample_id <- colnames(this_counts)     # optional
pca_scores$celltype <- this_metadata$celltype
pca_scores$condition <- this_metadata$condition
pca_scores$donor <- this_metadata$donor
pca_scores$read_length <- this_metadata$read_length
pca_scores$sequencing_platform <- this_metadata$sequencing_platform
pca_scores$unique_aligned_reads_over_total <- this_metadata$unique_aligned_reads_over_total

ggplot(pca_scores, aes(x = PC1, y = PC2, color = celltype, shape = condition)) +
  geom_point(size = 3) +
  labs(title = "PCA of Samples (566 KRT + FRB)", color = "Cell Type", shape = "Condition") +
  theme_minimal() + 
  stat_ellipse(aes(group = celltype), type = "norm", linetype = 2)


p1 = ggplot(data=pca_scores, aes(
  x = PC1,
  y = PC2,
  color = celltype,
  shape = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for 566 KRT and FRB samples",
       color = "celltype", shape = "condition") +
  stat_ellipse(aes(group = celltype), type = "norm", linetype = 2)

# Then enable the hover text in plotly
p1_interactive <- ggplotly(p1, tooltip = "text")
p1_interactive

#Use Combined Gene Contribution via Vector Magnitude
# Gene contributions to both PC1 and PC2
loadings <- pca_obj$rotation[, 1:2]  # genes × (PC1, PC2)
# Euclidean length (magnitude) of each gene's loading vector
loading_magnitude <- sqrt(loadings[,1]^2 + loadings[,2]^2)
# Use directional projection of gene loadings onto the cell-type separation vector
# Estimate centroids of each cell type in PC1–PC2 space:
frb_center <- colMeans(pca_scores[pca_scores$celltype == "FRB", c("PC1", "PC2")])
krt_center <- colMeans(pca_scores[pca_scores$celltype == "KRT", c("PC1", "PC2")])
# Define the separation vector:
separation_vector <- krt_center - frb_center  # Points from FRB → KRT
# Project each gene's PC1–PC2 loading onto the separation vector:
# Get gene loading matrix for PC1 and PC2
gene_loadings <- pca_obj$rotation[, 1:2]  # genes × (PC1, PC2)
# Normalize the separation vector
sep_unit <- separation_vector / sqrt(sum(separation_vector^2))
# Project gene loadings onto the separation direction
gene_projection_scores <- gene_loadings %*% sep_unit  # matrix multiplication
# Positive scores = genes that push samples toward KRT,
# Negative scores = genes that push samples toward FRB
top_krt_genes <- names(sort(gene_projection_scores[,1], decreasing = TRUE)[1:20])
top_frb_genes <- names(sort(gene_projection_scores[,1], decreasing = FALSE)[1:20])

# why are there KRT samples in the FRB cluster? (PC2 > 0)
temp = pca_scores[which(grepl("KRT",rownames(pca_scores))),] %>% 
  as.data.frame %>%
  dplyr::filter(PC2 > 0) %>%
  mutate(biosample=paste0(donor,"_",celltype,"_",condition)) %>%
  arrange(biosample)
temp$biosample
# F72, F75, F76, F78, F80, F81 all 4 conditions were in this list. 6donors x 4 conditions = 24 samples total. 
# I recall that during library prep, I noticed that some samples in the extracted RNA box of "KRT" were labeled as "FRB". and I had a concern that I mislabeled things. I even wrote a script "KRT_combination_with_FRB_investigation.R" in the RNA-Seq/keeratinocytes folder.
# I should check the freezer if that truly is the case. yes. totally. those frozen RNA samples are now labeled "KRT" because I relabeled them. They are physically next and in sequence to each other, 24 of them (which makes sense because I processed 24 samples at a time, so I must have mislabeled the whole batch). so I should discard these samples and also remove the rows from FS plates.

tmp = data.frame(F37_FRB_PBS=CPM[top_frb_genes,colnames(CPM)[grepl("F37_FRB_PBS", colnames(CPM))]], 
                 F75_FRB_PBS=CPM[top_frb_genes,colnames(CPM)[grepl("F75_FRB_PBS", colnames(CPM))]],
                 gene=top_frb_genes)

ggplot(tmp, aes(x=F37_FRB_PBS, y=F75_FRB_PBS, label = gene)) +
  geom_point() +
  geom_text(vjust = -0.5, hjust = 1, size = 3) + 
  labs(title="top FRB genes") +
  xlim(0, max(c(tmp$F37_FRB_PBS, tmp$F37_FRB_PBS))) +
  ylim(0, max(c(tmp$F75_KRT_PBS, tmp$F75_FRB_PBS)))
```
## DEseq2
```{r deseq2-mel}
this_counts <- counts[, metadata$celltype == "MEL"]
this_metadata <- metadata[metadata$celltype == "MEL",]
this_metadata$condition <- as.character(this_metadata$condition)

# Construct DESeqDataSet with all genes
dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata,
                              design = ~ donor + condition)
dds$condition <- relevel(factor(dds$condition), ref = "PBS")

# Run DESeq
dds.mel <- DESeq(dds)

# DESeq2 results for MEL comparisons
res.mel.ifng <- results(dds, contrast = c("condition", "IFNG", "PBS"))
res.mel.ifnb <- results(dds, contrast = c("condition", "IFNB", "PBS"))
res.mel.tnf  <- results(dds, contrast = c("condition", "TNF",  "PBS"))

# Filter significant DEGs (adj p < 0.05 and |log2FC| ≥ 1)
deg.mel.ifng <- as.data.frame(res.mel.ifng)[which(res.mel.ifng$padj < 0.05 & abs(res.mel.ifng$log2FoldChange) >= 1), ] %>% rownames_to_column("gene")  %>% na.omit() %>% arrange(padj)
deg.mel.ifnb <- as.data.frame(res.mel.ifnb)[which(res.mel.ifnb$padj < 0.05 & abs(res.mel.ifnb$log2FoldChange) >= 1), ] %>% rownames_to_column("gene")  %>% na.omit() %>% arrange(padj)
deg.mel.tnf  <- as.data.frame(res.mel.tnf)[which(res.mel.tnf$padj < 0.05 & abs(res.mel.tnf$log2FoldChange) >= 1), ] %>% rownames_to_column("gene")  %>% na.omit() %>% arrange(padj)

# MA plot
plotMA(res.mel.ifng, main = "IFNG vs PBS (MEL)", ylim = c(-5, 5))

# Volcano plot
res.mel.ifng.df <- as.data.frame(res.mel.ifng)
res.mel.ifng.df$significant <- with(res.mel.ifng.df, padj < 0.05 & abs(log2FoldChange) >= 1)

ggplot(res.mel.ifng.df, aes(x = log2FoldChange, y = -log10(padj), color = significant)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("grey", "red")) +
  theme_minimal() +
  labs(title = "IFNG vs PBS (MEL)", x = "log2FC", y = "-log10(adj. p)")


```
```{r deseq2-krt}
this_counts <- counts[, metadata$celltype == "KRT"]
this_metadata <- metadata[metadata$celltype == "KRT",]
this_metadata$condition <- as.character(this_metadata$condition)
colnames(this_counts)==this_metadata$sample

# Construct DESeqDataSet with all genes
dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata,
                              design = ~ donor + condition)
dds$condition <- relevel(factor(dds$condition), ref = "PBS")

# Run DESeq
dds.krt <- DESeq(dds)

# DESeq2 results for MEL comparisons
res.krt.ifng <- results(dds.krt, contrast = c("condition", "IFNG", "PBS"))
res.krt.ifnb <- results(dds.krt, contrast = c("condition", "IFNB", "PBS"))
res.krt.tnf  <- results(dds.krt, contrast = c("condition", "TNF",  "PBS"))

# Filter significant DEGs (adj p < 0.05 and |log2FC| ≥ 1)
deg.krt.ifng <- as.data.frame(res.krt.ifng)[which(res.krt.ifng$padj < 0.05 & abs(res.krt.ifng$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.krt.ifnb <- as.data.frame(res.krt.ifnb)[which(res.krt.ifnb$padj < 0.05 & abs(res.krt.ifnb$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.krt.tnf  <- as.data.frame(res.krt.tnf)[which(res.krt.tnf$padj < 0.05 & abs(res.krt.tnf$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
```
```{r deseq2-frb}
## ----------------- FRB
this_counts <- counts[, metadata$celltype == "FRB"]
this_metadata <- metadata[metadata$celltype == "FRB",]
this_metadata$condition <- as.character(this_metadata$condition)
colnames(this_counts)==this_metadata$sample

# Construct DESeqDataSet with all genes
dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata,
                              design = ~ donor + condition)
dds$condition <- relevel(factor(dds$condition), ref = "PBS")

# Run DESeq
dds.frb <- DESeq(dds)

# DESeq2 results for MEL comparisons
res.frb.ifng <- results(dds.frb, contrast = c("condition", "IFNG", "PBS"))
res.frb.ifnb <- results(dds.frb, contrast = c("condition", "IFNB", "PBS"))
res.frb.tnf  <- results(dds.frb, contrast = c("condition", "TNF",  "PBS"))

# Filter significant DEGs (adj p < 0.05 and |log2FC| ≥ 1)
deg.frb.ifng <- as.data.frame(res.frb.ifng)[which(res.frb.ifng$padj < 0.05 & abs(res.frb.ifng$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.frb.ifnb <- as.data.frame(res.frb.ifnb)[which(res.frb.ifnb$padj < 0.05 & abs(res.frb.ifnb$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.frb.tnf  <- as.data.frame(res.frb.tnf)[which(res.frb.tnf$padj < 0.05 & abs(res.frb.tnf$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)

save.image(file=paste0(Dir,'/myEnvironment_main.RData'))
```
```{r deseq2-all}
## ----------------- all cell types
this_counts <- counts
this_metadata <- metadata
this_metadata$condition <- as.character(this_metadata$condition)
this_metadata$celltype <- as.character(this_metadata$celltype)
colnames(this_counts)==this_metadata$sample

# Construct DESeqDataSet with all genes
dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata,
                              design = ~ donor + celltype + condition + celltype:condition)
dds$condition <- relevel(factor(dds$condition), ref = "PBS")

# Run DESeq
dds.3ct <- DESeq(dds)

# DESeq2 results for MEL comparisons
res.3ct.ifng <- results(dds.3ct, contrast = c("condition", "IFNG", "PBS"))
res.3ct.ifnb <- results(dds.3ct, contrast = c("condition", "IFNB", "PBS"))
res.3ct.tnf  <- results(dds.3ct, contrast = c("condition", "TNF",  "PBS"))

# Filter significant DEGs (adj p < 0.05 and |log2FC| ≥ 1)
deg.3ct.ifng <- as.data.frame(res.3ct.ifng)[which(res.3ct.ifng$padj < 0.05 & abs(res.3ct.ifng$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.3ct.ifnb <- as.data.frame(res.3ct.ifnb)[which(res.3ct.ifnb$padj < 0.05 & abs(res.3ct.ifnb$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.3ct.tnf  <- as.data.frame(res.3ct.tnf)[which(res.3ct.tnf$padj < 0.05 & abs(res.3ct.tnf$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)

vsd.3ct <- vst(dds.3ct, blind = FALSE) # this gives me globally transformed expression matrix
counts_norm <- counts(dds, normalized=TRUE)  # DESeq2-normalized counts (not library size scaled)
cpm_mat <- t(t(counts_norm) / colSums(counts_norm)) * 1e6

rm(this_counts, this_metadata, dds)
save.image(file=paste0(Dir,'/myEnvironment_main.RData'))
```
## Obtain VST from DESeq2 objects
```{r vst}
clean_gene_map <- metagene_dict %>% 
  select(metagene = meta, gene = name)

vst.mel <- assay(vst(dds.mel, blind = TRUE)) %>% 
  as.data.frame %>% 
  rownames_to_column("metagene") %>%
  dplyr::left_join(clean_gene_map, by = "metagene") %>%
  dplyr::mutate(gene = dplyr::coalesce(gene, metagene)) %>%
  dplyr::select(-metagene) %>%
  dplyr::filter(.$gene!="ARMCX5-GPRASP2") %>%
  column_to_rownames("gene") %>%
  rownames_to_column("gene")

vst.krt <- assay(vst(dds.krt, blind = TRUE)) %>% 
  as.data.frame %>% 
  rownames_to_column("metagene") %>%
  dplyr::left_join(clean_gene_map, by = "metagene") %>%
  dplyr::mutate(gene = dplyr::coalesce(gene, metagene)) %>%
  dplyr::select(-metagene) %>%
  dplyr::filter(.$gene!="ARMCX5-GPRASP2") %>%
  column_to_rownames("gene") %>%
  rownames_to_column("gene")

vst.frb <- assay(vst(dds.frb, blind = TRUE)) %>% 
  as.data.frame %>% 
  rownames_to_column("metagene") %>%
  dplyr::left_join(clean_gene_map, by = "metagene") %>%
  dplyr::mutate(gene = dplyr::coalesce(gene, metagene)) %>%
  dplyr::select(-metagene) %>%
  dplyr::filter(.$gene!="ARMCX5-GPRASP2") %>%
  column_to_rownames("gene") %>%
  rownames_to_column("gene")

data.table::fwrite(vst.mel, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/analysis_07142025/VST_MEL.txt",sep="\t",quote=F)
data.table::fwrite(vst.krt, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/analysis_07142025/VST_KRT.txt",sep="\t",quote=F)
data.table::fwrite(vst.frb, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/analysis_07142025/VST_FRB.txt",sep="\t",quote=F)
```
## Heatmap
### heatmap of raw log2(CPM+1)
```{r heatmap-log2(CPM+1)}
vars_deg <- ls(pattern = "^deg.")
gene_lists <- lapply(vars_deg, function(var) {
  get(var)$gene
})
expressed_DEG <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))

CPM_DEG <- CPM[expressed_DEG,]
log2CPM_DEG <- log2(CPM_DEG + 1)

# split rows by determined number of clusters
km_res <- niceKmeans(log2CPM_DEG, 14)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,km_res$breaks[5]-km_res$breaks[4]),
                        rep(6,km_res$breaks[6]-km_res$breaks[5]),
                        rep(7,km_res$breaks[7]-km_res$breaks[6]),
                        rep(8,km_res$breaks[8]-km_res$breaks[7]),
                        rep(9,km_res$breaks[9]-km_res$breaks[8]),
                        rep(10,km_res$breaks[10]-km_res$breaks[9]),
                        rep(11,km_res$breaks[11]-km_res$breaks[10]),
                        rep(12,km_res$breaks[12]-km_res$breaks[11]),
                        rep(13,km_res$breaks[13]-km_res$breaks[12]),
                        rep(14,nrow(km_res$clustered)-km_res$breaks[13]))
km_res.rna <- km_res
split_instructions.rna <- split_instructions
rm(km_res, split_instructions)

# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
colnames(zlogCPM.heatmap)==metadata$sample

column_ha = HeatmapAnnotation(
  condition = metadata$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = metadata$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "#999999", "IFNG" = "#D7263D", "IFNB" = "#1E88E5", "TNF" = "#FFC107"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNAseq_3ctk_allDEG_14kmm-hc_07172025_padj0.05_abslog2FC1_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, IFNg/IFNb/TNF stimulated, , known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(metadata$biosample[grepl("FRB_PBS",metadata$biosample)])), 
                                             rep(2,length(metadata$biosample[grepl("KRT_PBS",metadata$biosample)])), 
                                             rep(3,length(metadata$biosample[grepl("MEL_PBS",metadata$biosample)])), 
                                             rep(4,length(metadata$biosample[grepl("FRB_IFNG",metadata$biosample)])), 
                                             rep(5,length(metadata$biosample[grepl("KRT_IFNG",metadata$biosample)])),
                                             rep(6,length(metadata$biosample[grepl("MEL_IFNG",metadata$biosample)])),
                                             rep(7,length(metadata$biosample[grepl("FRB_IFNB",metadata$biosample)])), 
                                             rep(8,length(metadata$biosample[grepl("KRT_IFNB",metadata$biosample)])), 
                                             rep(9,length(metadata$biosample[grepl("MEL_IFNB",metadata$biosample)])), 
                                             rep(10,length(metadata$biosample[grepl("FRB_TNF",metadata$biosample)])), 
                                             rep(11,length(metadata$biosample[grepl("KRT_TNF",metadata$biosample)])),
                                             rep(12,length(metadata$biosample[grepl("MEL_TNF",metadata$biosample)]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()
cluster1_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==1)]

for (i in 1:14){
  this_cluster_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==i)]
  write(this_cluster_genes, file=paste0(Dir,"/inducedGenes_log2FC1.5_cluster",as.character(i),".txt"))
}

```
```{r heatmap-log2(CPM+1)-MEL}
vars_deg <- ls(pattern = "^deg.mel")
gene_lists <- lapply(vars_deg, function(var) {
  get(var)$gene
})
expressed_DEG <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))

CPM_DEG <- CPM[expressed_DEG,grepl("MEL",colnames(CPM))]
log2CPM_DEG <- log2(CPM_DEG + 1)

# split rows by determined number of clusters
km_res <- niceKmeans(log2CPM_DEG, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.rna <- km_res
split_instructions.rna <- split_instructions
rm(km_res, split_instructions)

# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
metadata.heatmap <- metadata[metadata$sample%in%colnames(zlogCPM.heatmap), ]
colnames(zlogCPM.heatmap)==metadata.heatmap$sample
metadata.heatmap$biosample <- paste0(metadata.heatmap$donor,"_",metadata.heatmap$celltype,"_",metadata.heatmap$condition)

column_ha = HeatmapAnnotation(
  condition = metadata.heatmap$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = metadata.heatmap$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "#999999", "IFNG" = "#D7263D", "IFNB" = "#1E88E5", "TNF" = "#FFC107"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNAseq_3ctk_MEL_MELDEG_5kmm-hc_07172025_padj0.05_abslog2FC1_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in MEL, IFNg/IFNb/TNF stimulated, , known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(metadata.heatmap$biosample[grepl("MEL_PBS",metadata.heatmap$biosample)])), 
                                             rep(2,length(metadata.heatmap$biosample[grepl("MEL_IFNG",metadata.heatmap$biosample)])),
                                             rep(3,length(metadata.heatmap$biosample[grepl("MEL_IFNB",metadata.heatmap$biosample)])), 
                                             rep(4,length(metadata.heatmap$biosample[grepl("MEL_TNF",metadata.heatmap$biosample)]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 5),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()
```
```{r heatmap-log2(CPM+1)-FRB}
vars_deg <- ls(pattern = "^deg.frb")
gene_lists <- lapply(vars_deg, function(var) {
  get(var)$gene
})
expressed_DEG <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))

CPM_DEG <- CPM[expressed_DEG,grepl("FRB",colnames(CPM))]
log2CPM_DEG <- log2(CPM_DEG + 1)

# split rows by determined number of clusters
km_res <- niceKmeans(log2CPM_DEG, 5)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
km_res.rna <- km_res
split_instructions.rna <- split_instructions
rm(km_res, split_instructions)

# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
metadata.heatmap <- metadata[metadata$sample%in%colnames(zlogCPM.heatmap), ]
colnames(zlogCPM.heatmap)==metadata.heatmap$sample
metadata.heatmap$biosample <- paste0(metadata.heatmap$donor,"_",metadata.heatmap$celltype,"_",metadata.heatmap$condition)

column_ha = HeatmapAnnotation(
  condition = metadata.heatmap$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = metadata.heatmap$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "#999999", "IFNG" = "#D7263D", "IFNB" = "#1E88E5", "TNF" = "#FFC107"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNAseq_3ctk_FRB_FRBDEG_5kmm-hc_07172025_padj0.05_abslog2FC1_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in FRB, IFNg/IFNb/TNF stimulated, , known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(metadata.heatmap$biosample[grepl("FRB_PBS",metadata.heatmap$biosample)])), 
                                             rep(2,length(metadata.heatmap$biosample[grepl("FRB_IFNG",metadata.heatmap$biosample)])),
                                             rep(3,length(metadata.heatmap$biosample[grepl("FRB_IFNB",metadata.heatmap$biosample)])), 
                                             rep(4,length(metadata.heatmap$biosample[grepl("FRB_TNF",metadata.heatmap$biosample)]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 5),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()
```
### heatmap of ComBat-seq donor corrected log2(CPM+1)
```{r combatseq}
# perform the batch correction
# transform the format of our groups and batches from names to numbers (e.g. 1, 2, etc.)
groups = sapply(as.character(metadata$condition), switch, "PBS" = 1, "IFNG" = 2, "IFNB" = 3, "TNF" = 4, USE.NAMES = F)
donor = as.numeric(as.factor(metadata$donor_num))

# perform batch correction
counts.corrected = ComBat_seq(counts = as.matrix(counts), batch = donor, group = groups)
col.sum=apply(counts.corrected,2,sum)
CPM.corrected=sweep(counts.corrected, 2, col.sum, `/`)*1000000
log2CPM.corrected=log2(CPM.corrected+1)
rm(groups, donor, col.sum)
```
```{r heatmap-corrected}
vars_deg <- ls(pattern = "^deg.")
gene_lists <- lapply(vars_deg, function(var) {
  get(var)$gene
})
expressed_DEG <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))


CPM_DEG <- CPM.corrected[expressed_DEG,]
log2CPM_DEG <- log2(CPM_DEG + 1)

# split rows by determined number of clusters
km_res <- niceKmeans(log2CPM_DEG, 14)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,km_res$breaks[5]-km_res$breaks[4]),
                        rep(6,km_res$breaks[6]-km_res$breaks[5]),
                        rep(7,km_res$breaks[7]-km_res$breaks[6]),
                        rep(8,km_res$breaks[8]-km_res$breaks[7]),
                        rep(9,km_res$breaks[9]-km_res$breaks[8]),
                        rep(10,km_res$breaks[10]-km_res$breaks[9]),
                        rep(11,km_res$breaks[11]-km_res$breaks[10]),
                        rep(12,km_res$breaks[12]-km_res$breaks[11]),
                        rep(13,km_res$breaks[13]-km_res$breaks[12]),
                        rep(14,nrow(km_res$clustered)-km_res$breaks[13]))
km_res.rna <- km_res
km_res.rna.corrected <- km_res
split_instructions.rna <- split_instructions
rm(km_res, split_instructions)


# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
colnames(zlogCPM.heatmap)==metadata$sample

column_ha = HeatmapAnnotation(
  condition = metadata$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = metadata$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "#999999", "IFNG" = "#D7263D", "IFNB" = "#1E88E5", "TNF" = "#FFC107"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNAseq_3ctk_allDEG_14kmm-hc_07182025_padj0.05_abslog2FC1_log2CPMcorrected.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM corrected",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, IFNg/IFNb/TNF stimulated, , known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(metadata$biosample[grepl("FRB_PBS",metadata$biosample)])), 
                                             rep(2,length(metadata$biosample[grepl("KRT_PBS",metadata$biosample)])), 
                                             rep(3,length(metadata$biosample[grepl("MEL_PBS",metadata$biosample)])), 
                                             rep(4,length(metadata$biosample[grepl("FRB_IFNG",metadata$biosample)])), 
                                             rep(5,length(metadata$biosample[grepl("KRT_IFNG",metadata$biosample)])),
                                             rep(6,length(metadata$biosample[grepl("MEL_IFNG",metadata$biosample)])),
                                             rep(7,length(metadata$biosample[grepl("FRB_IFNB",metadata$biosample)])), 
                                             rep(8,length(metadata$biosample[grepl("KRT_IFNB",metadata$biosample)])), 
                                             rep(9,length(metadata$biosample[grepl("MEL_IFNB",metadata$biosample)])), 
                                             rep(10,length(metadata$biosample[grepl("FRB_TNF",metadata$biosample)])), 
                                             rep(11,length(metadata$biosample[grepl("KRT_TNF",metadata$biosample)])),
                                             rep(12,length(metadata$biosample[grepl("MEL_TNF",metadata$biosample)]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()
```
### heatmap of DEseq2 variance-stabilized transformed data
## decide which samples to exclude
```{r exclude-sample-KRT}
### ---- use heatmap to decide outlier samples --- ####
outliers_list <- c("F26_KRT_PBS","F39_KRT_PBS","F42_KRT_PBS","F43_KRT_PBS","F60_KRT_PBS","F116_KRT_PBS","F120_KRT_PBS",
                   "F26_KRT_IFNG","F39_KFT_IFNG","F42_KFT_IFNG","F43_KRT_IFNG","F60_KRT_IFNG","F112_KRT_IFNG",
                   "F103_KRT_IFNB","F118_KRT_TNF")
pattern <- str_c(outliers_list, collapse = "|")
outlier_samples_metadata.krt <- metadata %>% dplyr::filter(str_detect(sample, pattern))
```
```{r exclude-sample-MEL}
### ---- use heatmap to decide outlier samples --- ####
outliers_list <- c("F108_MEL","F109_MEL","F110_MEL","F111_MEL","F51_MEL","F52_MEL","F61_MEL","F118_MEL","F133_MEL","F116_MEL_IFNG","F116_MEL_IFNG","F142_MEL_TNF","F42_MEL_PBS","F34_MEL_PBS","F30_MEL_IFNG")
lowcounts_list <- c("F116_MEL_IFNG","F142_MEL_TNF","F42_MEL_PBS","F34_MEL_PBS","F30_MEL_IFNG")
pattern <- str_c(outliers_list, collapse = "|")
outlier_samples_metadata.mel <- metadata %>% dplyr::filter(str_detect(sample, pattern))

```

### check excluded samples on PCA
```{r PCA-of-outliers}
outliers_list <- c("F26_KRT_PBS","F39_KRT_PBS","F42_KRT_PBS","F43_KRT_PBS","F60_KRT_PBS","F116_KRT_PBS","F120_KRT_PBS",
                   "F26_KRT_IFNG","F39_KFT_IFNG","F42_KFT_IFNG","F43_KRT_IFNG","F60_KRT_IFNG","F112_KRT_IFNG",
                   "F103_KRT_IFNB","F118_KRT_TNF")
pattern <- str_c(outliers_list, collapse = "|")
outlier_samples_metadata.krt <- metadata %>% dplyr::filter(str_detect(sample, pattern))

outliers_list <- c("F108_MEL","F109_MEL","F110_MEL","F111_MEL","F51_MEL","F52_MEL","F61_MEL","F118_MEL","F133_MEL","F116_MEL_IFNG","F116_MEL_IFNG","F142_MEL_TNF","F42_MEL_PBS","F34_MEL_PBS","F30_MEL_IFNG")
lowcounts_list <- c("F116_MEL_IFNG","F142_MEL_TNF","F42_MEL_PBS","F34_MEL_PBS","F30_MEL_IFNG")
pattern <- str_c(outliers_list, collapse = "|")
outlier_samples_metadata.mel <- metadata %>% dplyr::filter(str_detect(sample, pattern))

outlier.mel <- outlier_samples_metadata.mel$sample
outlier.krt <- outlier_samples_metadata.krt$sample

this_log2CPM <- log2CPM[expressed_genes,]

# Calculate Coefficient of Variation (COV)
gene_means <- rowMeans(this_log2CPM)
gene_sds <- apply(this_log2CPM, 1, sd)
gene_cov <- gene_sds / gene_means

# Remove genes with mean = 0 to avoid NaNs
gene_cov[is.na(gene_cov)] <- 0

# Select top N variable genes
top_n <- 3000
top_genes <- names(sort(gene_cov, decreasing = TRUE))[1:top_n]

# Subset log2CPM
log2CPM_subset <- log2CPM[top_genes, ]

#calculate principal components for top 3000 COV genes' log2CPM
pca_obj <- prcomp(t(log2CPM_subset), scale. = TRUE)

pca_scores <- as.data.frame(pca_obj$x)  # this is samples × PCs
pca_scores$sample_id <- colnames(counts)     # optional
pca_scores$celltype <- metadata$celltype
pca_scores$condition <- metadata$condition
pca_scores$donor <- metadata$donor
pca_scores$read_length <- metadata$read_length
pca_scores$sequencing_platform <- metadata$sequencing_platform
pca_scores$unique_aligned_reads_over_total <- metadata$unique_aligned_reads_over_total
pca_scores$is_outlier <- ifelse(rownames(pca_scores)%in%unique(c(outlier.mel,outlier.krt)),"yes","no")

# Get variance explained
percent_var <- round(100 * (pca_obj$sdev^2 / sum(pca_obj$sdev^2)), 1)
# Now use:
xlab <- paste0("PC1 (", percent_var[1], "%)")
ylab <- paste0("PC2 (", percent_var[2], "%)")

p1 = ggplot(data=pca_scores, aes(
  x = PC1,
  y = PC2,
  color = is_outlier,
  #shape = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total,
    "<br>outlier", is_outlier
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "outlier", shape = "condition",
       x = xlab, y = ylab) 

# Then enable the hover text in plotly
p1_interactive <- ggplotly(p1, tooltip = "text")
p1_interactive

### PC2 v PC3
xlab <- paste0("PC2 (", percent_var[2], "%)")
ylab <- paste0("PC3 (", percent_var[3], "%)")

p2 = ggplot(data=pca_scores, aes(
  x = PC2,
  y = PC3,
  color = is_outlier,
  #shape = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total,
    "<br>outlier", is_outlier
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "outlier", shape = "condition",
       x = xlab, y = ylab) 

# Then enable the hover text in plotly
p2_interactive <- ggplotly(p2, tooltip = "text")
p2_interactive
```
## after removing outliers, redo DEA, expressed gene identification, PCA
### compile count matrix, CPM, log2CPM after excluding outlier samples
```{r exclude-outliers}
outliers_list <- c("F26_KRT_PBS","F39_KRT_PBS","F42_KRT_PBS","F43_KRT_PBS","F60_KRT_PBS","F116_KRT_PBS","F120_KRT_PBS",
                   "F26_KRT_IFNG","F39_KFT_IFNG","F42_KFT_IFNG","F43_KRT_IFNG","F60_KRT_IFNG","F112_KRT_IFNG",
                   "F103_KRT_IFNB","F118_KRT_TNF")
pattern <- str_c(outliers_list, collapse = "|")
outlier_samples_metadata.krt <- metadata %>% dplyr::filter(str_detect(sample, pattern))
outliers_list <- c("F108_MEL","F109_MEL","F110_MEL","F111_MEL","F51_MEL","F52_MEL","F61_MEL","F118_MEL","F133_MEL","F116_MEL_IFNG","F116_MEL_IFNG","F142_MEL_TNF","F42_MEL_PBS","F34_MEL_PBS","F30_MEL_IFNG")
lowcounts_list <- c("F116_MEL_IFNG","F142_MEL_TNF","F42_MEL_PBS","F34_MEL_PBS","F30_MEL_IFNG")
pattern <- str_c(outliers_list, collapse = "|")
outlier_samples_metadata.mel <- metadata %>% dplyr::filter(str_detect(sample, pattern))

outlier.mel <- outlier_samples_metadata.mel$sample
outlier.krt <- outlier_samples_metadata.krt$sample

counts.clean <- counts %>% dplyr::select(!unique(c(outlier.mel, outlier.krt)))
metadata.clean <- metadata[metadata$sample%in%colnames(counts.clean),]
# metadata.clean$sample%in%colnames(counts.clean)
col.sum=apply(counts.clean,2,sum)
CPM.clean=sweep(counts.clean, 2, col.sum, `/`)*1000000
log2CPM.clean=log2(CPM.clean + 1)
rm(col.sum)

# convert meta gene to gene symbols
CPM.clean.metaconverted <- CPM.clean %>%
  rownames_to_column("gene") %>%
  left_join(metagene_dict, by = c("gene" = "meta")) %>%
  mutate(final_gene = if_else(is.na(name), gene, name)) %>%
  select(-gene, -name) %>%
  separate_rows(final_gene, sep = ",") %>%   # in case some name cells contain comma-separated genes
  distinct(final_gene, .keep_all = TRUE)

data.table::fwrite(CPM.clean.metaconverted, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/analysis_07142025/CPM.sampleFiltered.metaConverted.txt",
                   quote=F, sep="\t")

data.table::fwrite(metadata.clean,
                   file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/analysis_07142025/metadata.sampleFiltered.txt",
                   quote = F, sep = "\t")
```
### filter for expressed genes after outlier exclusion. An expressed gene shall have CPM >= 10 in at least 5 donors in a condition.
```{r filter}
conditions <- c("PBS", "IFNG", "IFNB", "TNF")
celltypes <-c("FRB","MEL","KRT")
expressed_genes_list <- list()

for (ct in celltypes) {
  gene_expression_status <- logical(nrow(CPM.clean))
  
  # Get sample names for this cell type
  ct_samples <- colnames(CPM.clean)[grepl(paste0("_", ct, "_"), colnames(CPM.clean))]
  
  for (i in 1:nrow(CPM.clean)) {
    is_expressed <- FALSE
    for (cond in conditions) {
      cond_samples <- ct_samples[grepl(paste0("_", cond), ct_samples)]
      # Extract CPM values for this gene under this cell type and condition
      cpm_values <- CPM.clean[i, cond_samples]
      if (length(which(cpm_values>=10)) >= 5) {
        is_expressed <- TRUE
        break
      }
    }
    
    gene_expression_status[i] <- is_expressed
  }
  
  expressed_genes_list[[ct]] <- rownames(CPM.clean)[gene_expression_status]
}

# Filter for expressed genes
expressed_genes <- unique(unlist(expressed_genes_list))
rm(celltypes, cond, cond_samples, conditions, ct, ct_samples, gene_expression_status, is_expressed)
```
### PCA after outlier exclusion
```{r pca}
this_matrix <- log2CPM.clean[expressed_genes, ]

# Calculate Coefficient of Variation (COV)
gene_means <- rowMeans(this_matrix)
gene_sds <- apply(this_matrix, 1, sd)
gene_cov <- gene_sds / gene_means

# Remove genes with mean = 0 to avoid NaNs
gene_cov[is.na(gene_cov)] <- 0

# Select top N variable genes
top_n <- 3000
top_genes <- names(sort(gene_cov, decreasing = TRUE))[1:top_n]

# Subset log2CPM
log2CPM_subset <- this_matrix[top_genes, ]

#calculate principal components for top 3000 COV genes' log2CPM
pca_obj <- prcomp(t(log2CPM_subset), scale. = TRUE)

pca_scores <- as.data.frame(pca_obj$x)  # this is samples × PCs
pca_scores$sample_id <- colnames(counts.clean)     # optional
pca_scores$celltype <- metadata.clean$celltype
pca_scores$condition <- metadata.clean$condition
pca_scores$donor <- metadata.clean$donor
pca_scores$read_length <- metadata.clean$read_length
pca_scores$sequencing_platform <- metadata.clean$sequencing_platform
pca_scores$unique_aligned_reads_over_total <- metadata.clean$unique_aligned_reads_over_total
pca_scores$total_reads <- metadata.clean$total_reads

# Get variance explained
percent_var <- round(100 * (pca_obj$sdev^2 / sum(pca_obj$sdev^2)), 1)
# Now use:
xlab <- paste0("PC1 (", percent_var[1], "%)")
ylab <- paste0("PC2 (", percent_var[2], "%)")

p1 = ggplot(data=pca_scores, aes(
  x = PC1,
  y = PC2,
  color = condition,
  #shape = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Tota_reads:", total_reads,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 696 samples\ntop 3000 COV expressed genes",
       color = "celltype", shape = "condition",
       x = xlab, y = ylab) 

# Then enable the hover text in plotly
p1_interactive <- ggplotly(p1, tooltip = "text")
p1_interactive


# Extract top PC1 and PC2 gene loadings
top_pc1_genes <- names(sort(pca_obj$rotation[, 1], decreasing = TRUE)[1:30])
top_pc2_genes <- names(sort(pca_obj$rotation[, 2], decreasing = TRUE)[1:30])

# Combine and deduplicate
top_genes <- unique(c(top_pc1_genes, top_pc2_genes))

# Plot mean expression
mean_CPM <- rowMeans(CPM[top_genes, ])
barplot(mean_CPM, las=2, main="Mean CPM of top PC1/PC2 genes", col="steelblue")


# Use directional projection of gene loadings onto the cell-type separation vector
# Estimate centroids of each cell type in PC1–PC2 space:
frb_center <- colMeans(pca_scores[pca_scores$celltype == "FRB", c("PC1", "PC2")])
krt_center <- colMeans(pca_scores[pca_scores$celltype == "KRT", c("PC1", "PC2")])
mel_center <- colMeans(pca_scores[pca_scores$celltype == "MEL", c("PC1", "PC2")])

# Define the separation vector:
# Vector from FRB to KRT
vec_frb_to_krt <- krt_center - frb_center
unit_frb_to_krt <- vec_frb_to_krt / sqrt(sum(vec_frb_to_krt^2))

# Vector from FRB to MEL
vec_frb_to_mel <- mel_center - frb_center
unit_frb_to_mel <- vec_frb_to_mel / sqrt(sum(vec_frb_to_mel^2))

# Vector from KRT to MEL
vec_krt_to_mel <- mel_center - krt_center
unit_krt_to_mel <- vec_krt_to_mel / sqrt(sum(vec_krt_to_mel^2))

# Project each gene's PC1–PC2 loading onto the separation vector:
# Get gene loading matrix for PC1 and PC2
gene_loadings <- pca_obj$rotation[, 1:2]

# Get top directional genes:
# top_mel_genes_from_krt: genes that push samples from KRT → MEL
proj_frb_to_krt <- gene_loadings %*% unit_frb_to_krt
proj_frb_to_mel <- gene_loadings %*% unit_frb_to_mel
proj_krt_to_mel <- gene_loadings %*% unit_krt_to_mel

top_frb_genes_from_krt <- names(sort(proj_frb_to_krt[,1], decreasing = TRUE)[1:20])
top_krt_genes_from_frb <- names(sort(proj_frb_to_krt[,1], decreasing = FALSE)[1:20])

top_mel_genes_from_frb <- names(sort(proj_frb_to_mel[,1], decreasing = TRUE)[1:20])
top_frb_genes_from_mel <- names(sort(proj_frb_to_mel[,1], decreasing = FALSE)[1:20])

top_mel_genes_from_krt <- names(sort(proj_krt_to_mel[,1], decreasing = TRUE)[1:20])
top_krt_genes_from_mel <- names(sort(proj_krt_to_mel[,1], decreasing = FALSE)[1:20])


#### PC2 vs PC3 ####
xlab <- paste0("PC2 (", percent_var[2], "%)")
ylab <- paste0("PC3 (", percent_var[3], "%)")

p2 = ggplot(data=pca_scores, aes(
  x = PC2,
  y = PC3,
  color = sequencing_platform,
  shape = celltype,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Tota_reads:", total_reads,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 696 samples\ntop 3000 COV expressed genes",
       color = "sequencing_platform", shape = "celltype",
       x = xlab, y = ylab) 
p2_interactive <- ggplotly(p2, tooltip = "text")
p2_interactive

#### PC3 vs PC4 ####
xlab <- paste0("PC3 (", percent_var[3], "%)")
ylab <- paste0("PC4 (", percent_var[4], "%)")

p3 = ggplot(data=pca_scores, aes(
  x = PC3,
  y = PC4,
  color = condition,
  shape = celltype,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Tota_reads:", total_reads,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 696 samples\ntop 3000 COV expressed genes",
       color = "condition", shape = "celltype",
       x = xlab, y = ylab) 
p3_interactive <- ggplotly(p3, tooltip = "text")
p3_interactive

#### PC4 vs PC5 ####
xlab <- paste0("PC4 (", percent_var[4], "%)")
ylab <- paste0("PC5 (", percent_var[5], "%)")

p4 = ggplot(data=pca_scores, aes(
  x = PC4,
  y = PC5,
  color = condition,
  shape = celltype,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Tota_reads:", total_reads,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 696 samples\ntop 3000 COV expressed genes",
       color = "condition", shape = "celltype",
       x = xlab, y = ylab) 
p4_interactive <- ggplotly(p4, tooltip = "text")
p4_interactive

# which PC captures condition effect? bar plot 
ggplot(pca_scores, aes(x = condition, y = PC6, fill = condition)) +
  geom_boxplot() +
  labs(title = "PC6 captures condition effect", y = paste0("PC6 (", percent_var[6], "%)"))

# Run ANOVA across PCs to see which PC corresponds to "CONDITION"
pca_scores$condition <- metadata.clean$condition

# Loop through top 10 PCs
for (i in 1:10) {
  cat(paste0("PC", i, ": "))
  print(summary(aov(pca_scores[,i] ~ pca_scores$condition)))
}
```
```{r pca_MEL}
this_matrix <- log2CPM.clean[expressed_genes, grepl("MEL",colnames(log2CPM.clean))]
this_metadata <- metadata.clean[metadata.clean$sample%in%colnames(this_matrix),]
# colnames(this_matrix)==this_metadata$sample

# Calculate Coefficient of Variation (COV)
gene_means <- rowMeans(this_matrix)
gene_sds <- apply(this_matrix, 1, sd)
gene_cov <- gene_sds / gene_means

# Remove genes with mean = 0 to avoid NaNs
gene_cov[is.na(gene_cov)] <- 0

# Select top N variable genes
top_n <- 3000
top_genes <- names(sort(gene_cov, decreasing = TRUE))[1:top_n]

# Subset log2CPM
log2CPM_subset <- this_matrix[top_genes, ]

#calculate principal components for top 3000 COV genes' log2CPM
pca_obj <- prcomp(t(log2CPM_subset), scale. = TRUE)

pca_scores <- as.data.frame(pca_obj$x)  # this is samples × PCs
pca_scores$sample_id <- colnames(this_matrix)     # optional
pca_scores$celltype <- this_metadata$celltype
pca_scores$condition <- this_metadata$condition
pca_scores$donor <- this_metadata$donor
pca_scores$read_length <- this_metadata$read_length
pca_scores$sequencing_platform <- this_metadata$sequencing_platform
pca_scores$unique_aligned_reads_over_total <- this_metadata$unique_aligned_reads_over_total
pca_scores$total_reads <- this_metadata$total_reads
# Get variance explained
percent_var <- round(100 * (pca_obj$sdev^2 / sum(pca_obj$sdev^2)), 1)
# Now use:
xlab <- paste0("PC1 (", percent_var[1], "%)")
ylab <- paste0("PC2 (", percent_var[2], "%)")

p1 = ggplot(data=pca_scores, aes(
  x = PC1,
  y = PC2,
  color = condition,
  shape = sequencing_platform,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total,
    "<br>Total reads:", total_reads
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 167 melanocytes samples\ntop 3000 COV expressed genes",
       color = "condition", shape = "sequencing_platform",
       x = xlab, y = ylab) 

# Then enable the hover text in plotly
p1_interactive <- ggplotly(p1, tooltip = "text")
p1_interactive


# Extract top PC1 and PC2 gene loadings
top_pc1_genes <- names(sort(pca_obj$rotation[, 1], decreasing = TRUE)[1:30])
bottom_pc1_genes <- names(sort(pca_obj$rotation[, 1], decreasing = FALSE)[1:30])

top_pc2_genes <- names(sort(pca_obj$rotation[, 2], decreasing = TRUE)[1:30])
bottom_pc2_genes <- names(sort(pca_obj$rotation[, 2], decreasing = FALSE)[1:30])

#### PC2 vs PC3 ####
xlab <- paste0("PC2 (", percent_var[2], "%)")
ylab <- paste0("PC3 (", percent_var[3], "%)")

p2 = ggplot(data=pca_scores, aes(
  x = PC2,
  y = PC3,
  color = sequencing_platform,
  shape = celltype,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 167 MEL samples\ntop 3000 COV expressed genes",
       color = "sequencing_platform", shape = "celltype",
       x = xlab, y = ylab) 
p2_interactive <- ggplotly(p2, tooltip = "text")
p2_interactive

#### PC3 vs PC4 ####
xlab <- paste0("PC3 (", percent_var[3], "%)")
ylab <- paste0("PC4 (", percent_var[4], "%)")

p3 = ggplot(data=pca_scores, aes(
  x = PC3,
  y = PC4,
  color = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "condition", shape = "celltype",
       x = xlab, y = ylab) 
p3_interactive <- ggplotly(p3, tooltip = "text")
p3_interactive

# which PC captures condition effect? bar plot 
ggplot(pca_scores, aes(x = condition, y = PC4, fill = condition)) +
  geom_boxplot() +
  labs(title = "PC4 captures condition effect", y = paste0("PC3 (", percent_var[3], "%)"))

# Run ANOVA across PCs to see which PC corresponds to "CONDITION"
pca_scores$condition <- metadata$condition

# Loop through top 10 PCs
for (i in 1:10) {
  cat(paste0("PC", i, ": "))
  print(summary(aov(pca_scores[,i] ~ pca_scores$condition)))
}
```
### DEseq2
```{r deseq2-mel}
this_counts <- counts.clean[, metadata.clean$celltype == "MEL"]
this_metadata <- metadata.clean[metadata.clean$celltype == "MEL",]
this_metadata$condition <- as.character(this_metadata$condition)

# Construct DESeqDataSet with all genes
dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata,
                              design = ~ sequencing_platform + condition)
dds$condition <- relevel(factor(dds$condition), ref = "PBS")

# Run DESeq
dds.mel <- DESeq(dds)

# DESeq2 results for MEL comparisons
res.mel.ifng <- results(dds.mel, contrast = c("condition", "IFNG", "PBS"))
res.mel.ifnb <- results(dds.mel, contrast = c("condition", "IFNB", "PBS"))
res.mel.tnf  <- results(dds.mel, contrast = c("condition", "TNF",  "PBS"))

# Filter significant DEGs (adj p < 0.05 and |log2FC| ≥ 1)
# before outlier exclusion: 4062 IFNG DEGs, 2333 IFNB DEGs, 1866 TNF DEGs
# after outlier exclusion: 5144 IFNG DEGs, 2572 IFNB DEGs, 1960 TNF DEGs
deg.mel.ifng <- as.data.frame(res.mel.ifng)[which(res.mel.ifng$padj < 0.05 & abs(res.mel.ifng$log2FoldChange) >= 1), ] %>% rownames_to_column("gene")  %>% na.omit() %>% arrange(padj)
deg.mel.ifnb <- as.data.frame(res.mel.ifnb)[which(res.mel.ifnb$padj < 0.05 & abs(res.mel.ifnb$log2FoldChange) >= 1), ] %>% rownames_to_column("gene")  %>% na.omit() %>% arrange(padj)
deg.mel.tnf  <- as.data.frame(res.mel.tnf)[which(res.mel.tnf$padj < 0.05 & abs(res.mel.tnf$log2FoldChange) >= 1), ] %>% rownames_to_column("gene")  %>% na.omit() %>% arrange(padj)

# MA plot
plotMA(res.mel.ifng, main = "IFNG vs PBS (MEL)", ylim = c(-5, 5))

# Volcano plot
res.mel.ifng.df <- as.data.frame(res.mel.ifng)
res.mel.ifng.df$significant <- with(res.mel.ifng.df, padj < 0.05 & abs(log2FoldChange) >= 1)

ggplot(res.mel.ifng.df, aes(x = log2FoldChange, y = -log10(padj), color = significant)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("grey", "red")) +
  theme_minimal() +
  labs(title = "IFNG vs PBS (MEL)", x = "log2FC", y = "-log10(adj. p)")

# make heatmap after removing batch effect
# VST transform first
vst.mel <- vst(dds.mel, blind = FALSE)
expr.mat <- assay(vst.mel)  # genes x samples

# Remove sequencing platform effect
platform <- colData(vst.mel)$sequencing_platform
expr.clean <- limma::removeBatchEffect(expr.mat, batch = platform)
expr.clean <- expr.mat
# plot heatmap
library(ComplexHeatmap)
library(circlize)

# Select top genes
top_genes <- unique(c(deg.mel.ifng$gene[1:50], deg.mel.ifnb$gene[1:50], deg.mel.tnf$gene[1:50]))
mat <- expr.clean[top_genes, ]
mat <- t(scale(t(mat))) # z-score by gene

# Annotate samples
ha = HeatmapAnnotation(
  condition = this_metadata$condition,
  col = list(condition = c("PBS" = "gray", "IFNG" = "red", "IFNB" = "blue", "TNF" = "gold"))
)

Heatmap(mat,
        name = "zscore vst (corrected)",
        top_annotation = ha,
        cluster_rows = TRUE,
        cluster_columns = FALSE,
        show_column_names = FALSE,
        show_row_names = TRUE)
```
```{r deseq2-krt}
this_counts <- counts.clean[, metadata.clean$celltype == "KRT"]
this_metadata <- metadata.clean[metadata.clean$celltype == "KRT",]
this_metadata$condition <- as.character(this_metadata$condition)
colnames(this_counts)==this_metadata$sample

# Construct DESeqDataSet with all genes
dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata,
                              design = ~ sequencing_platform + condition)
dds$condition <- relevel(factor(dds$condition), ref = "PBS")

# Run DESeq
dds.krt <- DESeq(dds)

# DESeq2 results for MEL comparisons
res.krt.ifng <- results(dds.krt, contrast = c("condition", "IFNG", "PBS"))
res.krt.ifnb <- results(dds.krt, contrast = c("condition", "IFNB", "PBS"))
res.krt.tnf  <- results(dds.krt, contrast = c("condition", "TNF",  "PBS"))

# Filter significant DEGs (adj p < 0.05 and |log2FC| ≥ 1)
# before outlier exclusion: 2931 IFNG DEGs, 759 IFNB DEGs, 1036 TNF DEGs
# after outlier exclusion: 3140 IFNG DEGs, 902 IFNB DEGs, 1107 TNF DEGs. gained some power.
deg.krt.ifng <- as.data.frame(res.krt.ifng)[which(res.krt.ifng$padj < 0.05 & abs(res.krt.ifng$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.krt.ifnb <- as.data.frame(res.krt.ifnb)[which(res.krt.ifnb$padj < 0.05 & abs(res.krt.ifnb$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.krt.tnf  <- as.data.frame(res.krt.tnf)[which(res.krt.tnf$padj < 0.05 & abs(res.krt.tnf$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
```
```{r deseq2-frb}
## ----------------- FRB
this_counts <- counts.clean[, metadata.clean$celltype == "FRB"]
this_metadata <- metadata.clean[metadata.clean$celltype == "FRB",]
this_metadata$condition <- as.character(this_metadata$condition)
colnames(this_counts)==this_metadata$sample

# Construct DESeqDataSet with all genes
dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata,
                              design = ~ sequencing_platform + condition)
dds$condition <- relevel(factor(dds$condition), ref = "PBS")

# Run DESeq
dds.frb <- DESeq(dds)

# DESeq2 results for MEL comparisons
res.frb.ifng <- results(dds.frb, contrast = c("condition", "IFNG", "PBS"))
res.frb.ifnb <- results(dds.frb, contrast = c("condition", "IFNB", "PBS"))
res.frb.tnf  <- results(dds.frb, contrast = c("condition", "TNF",  "PBS"))

# Filter significant DEGs (adj p < 0.05 and |log2FC| ≥ 1)
# before outlier exclusion: 1327 IFNG DEGs, 2157 IFNB DEGs, 3380 TNF DEGs
# after outlier exclusion: 1529 IFNG DEGs, 2439 IFNB DEGs, 3829 TNF DEGs. also gained power.
deg.frb.ifng <- as.data.frame(res.frb.ifng)[which(res.frb.ifng$padj < 0.05 & abs(res.frb.ifng$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.frb.ifnb <- as.data.frame(res.frb.ifnb)[which(res.frb.ifnb$padj < 0.05 & abs(res.frb.ifnb$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.frb.tnf  <- as.data.frame(res.frb.tnf)[which(res.frb.tnf$padj < 0.05 & abs(res.frb.tnf$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
```
```{r save}
save(deg.frb.ifnb, deg.frb.ifng, deg.frb.tnf, 
     deg.krt.ifnb, deg.krt.ifng, deg.krt.tnf, 
     deg.mel.ifnb, deg.mel.ifng, deg.mel.tnf, file = paste0(Dir,"/all_degs_abslog2FC1_padj0.05_post_outlier_exclusion.RData"))
```
# --- loading Rdata is up to here.
```{r deseq2-all}
## ----------------- all cell types
this_counts <- counts
this_metadata <- metadata
this_metadata$condition <- as.character(this_metadata$condition)
this_metadata$celltype <- as.character(this_metadata$celltype)
colnames(this_counts)==this_metadata$sample

# Construct DESeqDataSet with all genes
dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata,
                              design = ~ donor + celltype + condition + celltype:condition)
dds$condition <- relevel(factor(dds$condition), ref = "PBS")

# Run DESeq
dds.3ct <- DESeq(dds)

# DESeq2 results for MEL comparisons
res.3ct.ifng <- results(dds.3ct, contrast = c("condition", "IFNG", "PBS"))
res.3ct.ifnb <- results(dds.3ct, contrast = c("condition", "IFNB", "PBS"))
res.3ct.tnf  <- results(dds.3ct, contrast = c("condition", "TNF",  "PBS"))

# Filter significant DEGs (adj p < 0.05 and |log2FC| ≥ 1)
deg.3ct.ifng <- as.data.frame(res.3ct.ifng)[which(res.3ct.ifng$padj < 0.05 & abs(res.3ct.ifng$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.3ct.ifnb <- as.data.frame(res.3ct.ifnb)[which(res.3ct.ifnb$padj < 0.05 & abs(res.3ct.ifnb$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)
deg.3ct.tnf  <- as.data.frame(res.3ct.tnf)[which(res.3ct.tnf$padj < 0.05 & abs(res.3ct.tnf$log2FoldChange) >= 1), ] %>% rownames_to_column("gene") %>% na.omit() %>% arrange(padj)

vsd.3ct <- vst(dds.3ct, blind = FALSE) # this gives me globally transformed expression matrix
counts_norm <- counts(dds, normalized=TRUE)  # DESeq2-normalized counts (not library size scaled)
cpm_mat <- t(t(counts_norm) / colSums(counts_norm)) * 1e6

rm(this_counts, this_metadata, dds)
save.image(file=paste0(Dir,'/myEnvironment_main.RData'))
```
### Heatmap
Heatmap of all samples
```{r heatmap-log2(CPM+1)-outlier-excluded}
vars_deg <- ls(pattern = "^deg.")
gene_lists <- lapply(vars_deg, function(var) {
  get(var)$gene
})
expressed_DEG <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))

CPM_DEG <- CPM.clean[expressed_DEG,]
log2CPM_DEG <- log2(CPM_DEG + 1)

# split rows by determined number of clusters
km_res <- niceKmeans(log2CPM_DEG, 14)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,km_res$breaks[5]-km_res$breaks[4]),
                        rep(6,km_res$breaks[6]-km_res$breaks[5]),
                        rep(7,km_res$breaks[7]-km_res$breaks[6]),
                        rep(8,km_res$breaks[8]-km_res$breaks[7]),
                        rep(9,km_res$breaks[9]-km_res$breaks[8]),
                        rep(10,km_res$breaks[10]-km_res$breaks[9]),
                        rep(11,km_res$breaks[11]-km_res$breaks[10]),
                        rep(12,km_res$breaks[12]-km_res$breaks[11]),
                        rep(13,km_res$breaks[13]-km_res$breaks[12]),
                        rep(14,nrow(km_res$clustered)-km_res$breaks[13]))
km_res.rna <- km_res
split_instructions.rna <- split_instructions
rm(km_res, split_instructions)

# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
colnames(zlogCPM.heatmap)==metadata$sample

metadata.clean$biosample = paste0(metadata.clean$donor,"_",metadata.clean$celltype,"_",metadata.clean$condition)
column_ha = HeatmapAnnotation(
  condition = metadata.clean$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = metadata.clean$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "#999999", "IFNG" = "#D7263D", "IFNB" = "#1E88E5", "TNF" = "#FFC107"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNAseq_3ctk_outlier-excluded_allDEG_14kmm-hc_08052025_padj0.05_abslog2FC1_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, IFNg/IFNb/TNF stimulated, , known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(metadata.clean$biosample[grepl("FRB_PBS",metadata.clean$biosample)])), 
                                             rep(2,length(metadata.clean$biosample[grepl("KRT_PBS",metadata.clean$biosample)])), 
                                             rep(3,length(metadata.clean$biosample[grepl("MEL_PBS",metadata.clean$biosample)])), 
                                             rep(4,length(metadata.clean$biosample[grepl("FRB_IFNG",metadata.clean$biosample)])), 
                                             rep(5,length(metadata.clean$biosample[grepl("KRT_IFNG",metadata.clean$biosample)])),
                                             rep(6,length(metadata.clean$biosample[grepl("MEL_IFNG",metadata.clean$biosample)])),
                                             rep(7,length(metadata.clean$biosample[grepl("FRB_IFNB",metadata.clean$biosample)])), 
                                             rep(8,length(metadata.clean$biosample[grepl("KRT_IFNB",metadata.clean$biosample)])), 
                                             rep(9,length(metadata.clean$biosample[grepl("MEL_IFNB",metadata.clean$biosample)])), 
                                             rep(10,length(metadata.clean$biosample[grepl("FRB_TNF",metadata.clean$biosample)])), 
                                             rep(11,length(metadata.clean$biosample[grepl("KRT_TNF",metadata.clean$biosample)])),
                                             rep(12,length(metadata.clean$biosample[grepl("MEL_TNF",metadata.clean$biosample)]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()
cluster1_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==1)]

for (i in 1:14){
  this_cluster_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==i)]
  write(this_cluster_genes, file=paste0(Dir,"/inducedGenes_log2FC1.5_cluster",as.character(i),".txt"))
}

```
Heatmap per cytokine
```{r heatmap-PBS-IFNG}
# prep data table ####
vars_deg <- ls(pattern = "^deg.*ifng")
#gene_lists <- lapply(vars_deg, function(var) {
#  get(var)$gene
#})
gene_lists <- lapply(vars_deg, function(var) {
  df <- get(var)
  # keep only significant, upregulated genes
  df %>%
    filter(log2FoldChange > 1.5, padj < 0.01) %>%
    pull(gene)   # or `rownames(df)` if genes are rownames
})
expressed_DEG <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))

CPM_DEG <- CPM.clean[expressed_DEG,] %>% 
  dplyr::select(matches("PBS|IFNG"))
log2CPM_DEG <- log2(CPM_DEG + 1)
this_metadata <- metadata.clean %>% 
  dplyr::filter(sample%in%colnames(CPM_DEG)) %>%
  mutate(biosample = paste0(donor,"_",celltype,"_",condition))
this_metadata$sample==colnames(CPM_DEG)


# pheatmap
z_mat <- t(scale(t(log2CPM_DEG)))
my_cols  <- colorRampPalette(rev(brewer.pal(11, "RdBu")))(101)
my_breaks <- seq(-2.5, 2.5, length.out = 101)  # clip extremes for readability
pheatmap(
  z_mat,
  scale = "none",                         # already scaled
  color = my_cols, breaks = my_breaks,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  clustering_method = "ward.D2",
  show_rownames = FALSE, show_colnames = FALSE,
  fontsize = 10, border_color = NA
)

# cluster ####
# split rows by determined number of clusters
km_res <- readRDS(paste0(Dir,"/km_res_for_heatmap_RNAseq_3ctk_outlier-excluded_IFNGDEG_6kmm-hc_03102026_padj0.01_log2FC1.5_avgCPM10.rds"))
km_res <- niceKmeans(log2CPM_DEG, 6)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,km_res$breaks[5]-km_res$breaks[4]),
                        rep(6,nrow(km_res$clustered)-km_res$breaks[5]))
km_res.rna <- km_res
split_instructions.rna <- split_instructions
rm(km_res, split_instructions)
#saveRDS(km_res.rna, file = paste0(Dir,"/km_res_for_heatmap_RNAseq_3ctk_outlier-excluded_IFNGDEG_6kmm-hc_03102026_padj0.01_log2FC1.5_avgCPM10.rds"))


# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# plot heatmap ####
colnames(zlogCPM.heatmap)==this_metadata$sample

column_ha = HeatmapAnnotation(
  condition = this_metadata$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = this_metadata$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "#999999", "IFNG" = "#D7263D", "IFNB" = "#1E88E5", "TNF" = "#FFC107"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

# old heatmap: heatmap_RNAseq_3ctk_outlier-excluded_IFNGDEG_6kmm-hc_08082025_padj0.01_log2FC1.5_avgCPM10.pdf
# 
pdf(paste0(Dir,"/heatmap_RNAseq_3ctk_outlier-excluded_IFNGDEG_6kmm-hc_03102026_padj0.01_log2FC1.5_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "upregulated DE genes in 3 celltypes, Ctrl vs IFNg stimulated, , known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(this_metadata$biosample[grepl("FRB_PBS",this_metadata$biosample)])), 
                                             rep(2,length(this_metadata$biosample[grepl("KRT_PBS",this_metadata$biosample)])), 
                                             rep(3,length(this_metadata$biosample[grepl("MEL_PBS",this_metadata$biosample)])), 
                                             rep(4,length(this_metadata$biosample[grepl("FRB_IFNG",this_metadata$biosample)])), 
                                             rep(5,length(this_metadata$biosample[grepl("KRT_IFNG",this_metadata$biosample)])),
                                             rep(6,length(this_metadata$biosample[grepl("MEL_IFNG",this_metadata$biosample)])),
                                             rep(7,length(this_metadata$biosample[grepl("FRB_IFNB",this_metadata$biosample)])), 
                                             rep(8,length(this_metadata$biosample[grepl("KRT_IFNB",this_metadata$biosample)])), 
                                             rep(9,length(this_metadata$biosample[grepl("MEL_IFNB",this_metadata$biosample)])), 
                                             rep(10,length(this_metadata$biosample[grepl("FRB_TNF",this_metadata$biosample)])), 
                                             rep(11,length(this_metadata$biosample[grepl("KRT_TNF",this_metadata$biosample)])),
                                             rep(12,length(this_metadata$biosample[grepl("MEL_TNF",this_metadata$biosample)]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            show_column_names = FALSE,
                            row_title_rot = 0,
                            #column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            #column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()



# ---- Analysis for each cluster ---- ####
for (i in 1:6){
  this_cluster_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==i)]
  write(this_cluster_genes, file=paste0(Dir,"/IFNG_inducedGenes_log2FC1.5_cluster",as.character(i),".txt"))
}
#
vars_res <- ls(pattern = "^res.*ifng")
gene_lists <- lapply(vars_res, function(var) {
  get(var)$gene
})
expressed_genes.krt <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))
# cluster 3
i=c(5)
this_cluster_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna %in% i)]
this_metagenes_idx <- which(grepl("meta", this_cluster_genes))
temp1 <- data.frame(meta=this_cluster_genes[-this_metagenes_idx], 
                    name=this_cluster_genes[-this_metagenes_idx])
temp2 <- data.frame(meta=this_cluster_genes[this_metagenes_idx]) %>%
  left_join(., metagene_dict, join_by(meta))
temp3 <- rbind(temp1, temp2) %>% set_colnames(c("meta","gene"))
gene_ids <- convert_to_entrez(convert_meta_genes(temp3$gene))$ENTREZID

this_res <- as.data.frame(res.frb.ifng) %>%
  rownames_to_column("meta") %>%
  dplyr::filter(meta %in% temp3$meta) %>%
  left_join(temp3) %>%
  arrange(padj)

ego <- enrichGO(gene         = gene_ids,
                OrgDb        = org.Hs.eg.db,
                keyType = "ENTREZID",
                ont          = "BP",       # Biological Process
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.1,
                qvalueCutoff = 0.1,
                readable      = TRUE)

ego.sig <- ego@result[ego@result$p.adjust<0.05,]


```

Heatmap per celltype
```{r heatmap-KRT}
vars_deg <- ls(pattern = "^deg.krt")
#gene_lists <- lapply(vars_deg, function(var) {
#  get(var)$gene
#})
gene_lists <- lapply(vars_deg, function(var) {
  df <- get(var)
  # keep only significant, upregulated genes
  df %>%
    filter(log2FoldChange > 1.5, padj < 0.01) %>%
    pull(gene)   # or `rownames(df)` if genes are rownames
})
expressed_DEG <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))

CPM_DEG <- CPM.clean[expressed_DEG,] %>% 
  dplyr::select(matches("KRT"))
log2CPM_DEG <- log2(CPM_DEG + 1)
this_metadata <- metadata.clean %>% 
  dplyr::filter(sample%in%colnames(CPM_DEG)) %>%
  mutate(biosample = paste0(donor,"_",celltype,"_",condition))
this_metadata$sample==colnames(CPM_DEG)

# split rows by determined number of clusters
# km_res <- niceKmeans(log2CPM_DEG, 5)
# split_instructions <- c(rep(1,km_res$breaks[1]), 
#                         rep(2,km_res$breaks[2]-km_res$breaks[1]),
#                         rep(3,km_res$breaks[3]-km_res$breaks[2]),
#                         rep(4,km_res$breaks[4]-km_res$breaks[3]),
#                         rep(5,nrow(km_res$clustered)-km_res$breaks[4]))
#saveRDS(km_res, file=paste0(Dir,"km_res_for_heatmap_RNAseq_KRT_allcytokines_outlier-excluded_inducedGenes_5km_03162026_padj0.01_log2FC1.5_avgCPM10.rds"))
km_res.rna <- readRDS(paste0(Dir,"km_res_for_heatmap_RNAseq_KRT_allcytokines_outlier-excluded_inducedGenes_5km_03162026_padj0.01_log2FC1.5_avgCPM10.rds"))
# remap class based on my preference
km_res.rna$clustered <- km_res.rna$clustered %>%
  mutate(
    class = case_when(
      class == 5 ~ 1L,
      class == 1 ~ 2L,
      class == 2 ~ 3L,
      class == 3 ~ 4L,
      class == 4 ~ 5L
    )
  ) %>%
  arrange(class)

split_instructions.rna <- km_res.rna$clustered$class


# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
colnames(zlogCPM.heatmap)==this_metadata$sample

column_ha = HeatmapAnnotation(
  celltype = this_metadata$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  condition = this_metadata$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  
  col = list(celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF"),
             condition = c("PBS" = "#999999", "IFNG" = "#D7263D", "IFNB" = "#1E88E5", "TNF" = "#FFC107")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNAseq_3ctk_outlier-excluded_all-KRT-DEG_5kmm-hc_03132026_padj0.01_log2FC1.5_avgCPM10.pdf"),width=24,height=30)
# png(paste0(Dir,"/heatmap_RNAseq_3ctk_outlier-excluded_all-KRT-DEG_5kmm-hc_03132026_padj0.01_log2FC1.5_avgCPM10.png"),
#   width = 24, height = 30, units = "in", res = 300)

heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM",
                            top_annotation = column_ha,
                            #right_annotation = row_ha,
                            column_title = "upregulated DE genes in KRT, Ctrl vs IFNg/IFNb/TNFa stimulated",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(this_metadata$biosample[grepl("FRB_PBS",this_metadata$biosample)])), 
                                             rep(2,length(this_metadata$biosample[grepl("KRT_PBS",this_metadata$biosample)])), 
                                             rep(3,length(this_metadata$biosample[grepl("MEL_PBS",this_metadata$biosample)])), 
                                             rep(4,length(this_metadata$biosample[grepl("FRB_IFNG",this_metadata$biosample)])), 
                                             rep(5,length(this_metadata$biosample[grepl("KRT_IFNG",this_metadata$biosample)])),
                                             rep(6,length(this_metadata$biosample[grepl("MEL_IFNG",this_metadata$biosample)])),
                                             rep(7,length(this_metadata$biosample[grepl("FRB_IFNB",this_metadata$biosample)])), 
                                             rep(8,length(this_metadata$biosample[grepl("KRT_IFNB",this_metadata$biosample)])), 
                                             rep(9,length(this_metadata$biosample[grepl("MEL_IFNB",this_metadata$biosample)])), 
                                             rep(10,length(this_metadata$biosample[grepl("FRB_TNF",this_metadata$biosample)])), 
                                             rep(11,length(this_metadata$biosample[grepl("KRT_TNF",this_metadata$biosample)])),
                                             rep(12,length(this_metadata$biosample[grepl("MEL_TNF",this_metadata$biosample)]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            show_column_names = FALSE,
                            row_title_rot = 0,
                            #column_names_rot = 45,
                            row_gap = unit(3, "mm"),
                            column_gap = unit(3, "mm"),
                            border = TRUE,
                            #column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()

for (i in 1:5){
  this_cluster_genes = convert_meta_genes(rownames(km_res.rna$clustered)[which(split_instructions.rna==i)])
  write(this_cluster_genes, file=paste0(Dir,"/inducedGenes_log2FC1.5_cluster",as.character(i),".txt"))
}

this_cluster_genes
gene="MUC1"
#plot_violinbox_plot_cytokine_faceted(gene)
p = plot_violinbox_plot_cytokine_faceted_by_celltype_grouped_by_phase(gene) 

# plot all the genes
pdf(paste0(Dir, "KRT_allcytokines_cluster6_gene_violinplots_03162026.pdf"), width = 12, height = 6)
for (gene in this_cluster_genes) {
  p <- plot_violinbox_plot(gene) +
       ggtitle(gene)
  print(p)
  message(gene)
}
dev.off()

```

```{r pca for each cluster}
# ----------------------------
# 1. Subset matrix to cluster genes
# ----------------------------
cluster_genes_use <- intersect(this_cluster_genes, rownames(log2CPM.clean))

if (length(cluster_genes_use) < 2) {
  stop("Fewer than 2 genes from this_cluster_genes found in log2CPM.clean")
}

this_matrix <- log2CPM.clean[cluster_genes_use, , drop = FALSE]

# Optional: remove zero-variance genes
gene_sds <- apply(this_matrix, 1, sd, na.rm = TRUE)
this_matrix <- this_matrix[gene_sds > 0, , drop = FALSE]

if (nrow(this_matrix) < 2) {
  stop("Fewer than 2 non-zero-variance genes remain for PCA")
}

# ----------------------------
# 2. PCA on samples using cluster genes only
# ----------------------------
pca_obj <- prcomp(t(this_matrix), scale. = TRUE)

pca_scores <- as.data.frame(pca_obj$x)
pca_scores$sample_id <- colnames(this_matrix)
pca_scores$celltype <- metadata.clean$celltype
pca_scores$condition <- metadata.clean$condition
pca_scores$donor <- metadata.clean$donor
pca_scores$read_length <- metadata.clean$read_length
pca_scores$sequencing_platform <- metadata.clean$sequencing_platform
pca_scores$unique_aligned_reads_over_total <- metadata.clean$unique_aligned_reads_over_total
pca_scores$total_reads <- metadata.clean$total_reads

# Variance explained
percent_var <- round(100 * (pca_obj$sdev^2 / sum(pca_obj$sdev^2)), 1)
xlab <- paste0("PC1 (", percent_var[1], "%)")
ylab <- paste0("PC2 (", percent_var[2], "%)")

# ----------------------------
# 3. PCA plot colored by sequencing platform
# ----------------------------
library(ggplot2)
library(plotly)

p_platform <- ggplot(
  data = pca_scores,
  aes(
    x = PC1,
    y = PC2,
    color = sequencing_platform,
    shape = condition,
    text = paste(
      "Sample:", sample_id,
      "<br>Donor:", donor,
      "<br>Condition:", condition,
      "<br>Celltype:", celltype,
      "<br>Sequencing platform:", sequencing_platform,
      "<br>Read length:", read_length,
      "<br>Total reads:", total_reads,
      "<br>Aligned:", unique_aligned_reads_over_total
    )
  )
) +
  geom_point(size = 3, alpha = 0.9) +
  labs(
    title = paste0(
      "PCA using cluster genes only (n = ", nrow(this_matrix), ")"
    ),
    x = xlab,
    y = ylab,
    color = "Sequencing platform",
    shape = "Condition"
  ) +
  theme_bw()

p_platform_interactive <- ggplotly(p_platform, tooltip = "text")
p_platform_interactive
```

## Skip: Combat-seq correction necessary? Probably not. main heatmap still shows a cluster with batch separation. 
```{r heatmap-Combat-seq-corrected-outlier-excluded}
# perform the batch correction
# transform the format of our groups and batches from names to numbers (e.g. 1, 2, etc.)
groups = sapply(as.character(metadata.clean$condition), switch, "PBS" = 1, "IFNG" = 2, "IFNB" = 3, "TNF" = 4, USE.NAMES = F)
seq_platform = as.numeric(as.factor(metadata.clean$sequencing_platform))

# perform batch correction
counts.corrected = ComBat_seq(counts = as.matrix(counts.clean), batch = seq_platform, group = groups)
col.sum=apply(counts.corrected,2,sum)
CPM.clean.corrected=sweep(counts.corrected, 2, col.sum, `/`)*1000000
log2CPM.clean.corrected=log2(CPM.clean.corrected+1)
rm(groups, donor, col.sum)

# make heatmap after removing batch effect using Combat-seq
vars_deg <- ls(pattern = "^deg.")
gene_lists <- lapply(vars_deg, function(var) {
  get(var)$gene
})
expressed_DEG <- intersect(unique(unlist(expressed_genes_list)), 
                               unique(unlist(gene_lists)))

CPM_DEG <- CPM.clean.corrected[expressed_DEG,]
log2CPM_DEG <- log2(CPM_DEG + 1)

# split rows by determined number of clusters
km_res <- niceKmeans(log2CPM_DEG, 14)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]),
                        rep(3,km_res$breaks[3]-km_res$breaks[2]),
                        rep(4,km_res$breaks[4]-km_res$breaks[3]),
                        rep(5,km_res$breaks[5]-km_res$breaks[4]),
                        rep(6,km_res$breaks[6]-km_res$breaks[5]),
                        rep(7,km_res$breaks[7]-km_res$breaks[6]),
                        rep(8,km_res$breaks[8]-km_res$breaks[7]),
                        rep(9,km_res$breaks[9]-km_res$breaks[8]),
                        rep(10,km_res$breaks[10]-km_res$breaks[9]),
                        rep(11,km_res$breaks[11]-km_res$breaks[10]),
                        rep(12,km_res$breaks[12]-km_res$breaks[11]),
                        rep(13,km_res$breaks[13]-km_res$breaks[12]),
                        rep(14,nrow(km_res$clustered)-km_res$breaks[13]))
km_res.rna <- km_res
split_instructions.rna <- split_instructions
rm(km_res, split_instructions)

# 4. load genes of interest
g <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/KEGG/GSEA_hallmark-genes-upregulated-in-response-to-IFNG.txt",header=F)$V1
zlogCPM.heatmap <- km_res.rna$clustered[,1:ncol(CPM_DEG)]
zlogCPM.heatmap_with_gene <- zlogCPM.heatmap %>% dplyr::filter(rownames(.) %in% g)
row_anno_index <- which(rownames(zlogCPM.heatmap) %in% rownames(zlogCPM.heatmap_with_gene))
row_anno_label <- rownames(zlogCPM.heatmap)[row_anno_index]

# 5. plot heatmap
colnames(zlogCPM.heatmap)==metadata.clean$sample
metadata.clean$biosample = paste0(metadata.clean$donor,"_",metadata.clean$celltype,"_",metadata.clean$condition)

column_ha = HeatmapAnnotation(
  condition = metadata.clean$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
  celltype = metadata.clean$biosample %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[2]) %>% unlist(),
  col = list(condition = c("PBS" = "#999999", "IFNG" = "#D7263D", "IFNB" = "#1E88E5", "TNF" = "#FFC107"),
             celltype = c("KRT" = "#DAF7A6", "MEL" = "#964B00", "FRB" = "#CCCCFF")),
  show_legend = c(TRUE,TRUE))

row_ha = rowAnnotation(
  foo = anno_mark(at = row_anno_index, 
                  labels = row_anno_label,
                  which="row",
                  labels_gp = gpar(fontsize=10),
                  padding=unit(0.5,"mm")))

pdf(paste0(Dir,"/heatmap_RNAseq_3ctk_outlier-excluded_CombatCorrected_allDEG_14kmm-hc_08052025_padj0.05_abslog2FC1_avgCPM10.pdf"),width=24,height=30)
heatmap.rna <- draw(Heatmap(zlogCPM.heatmap, 
                            name = "z-score log2CPM Seq_platform corrected",
                            top_annotation = column_ha,
                            right_annotation = row_ha,
                            column_title = "DE genes in 3 celltypes, IFNg/IFNb/TNF stimulated, , known IFNg inducible genes labeled",
                            cluster_columns = FALSE,
                            cluster_rows = FALSE,
                            column_split = c(rep(1,length(metadata.clean$biosample[grepl("FRB_PBS",metadata.clean$biosample)])), 
                                             rep(2,length(metadata.clean$biosample[grepl("KRT_PBS",metadata.clean$biosample)])), 
                                             rep(3,length(metadata.clean$biosample[grepl("MEL_PBS",metadata.clean$biosample)])), 
                                             rep(4,length(metadata.clean$biosample[grepl("FRB_IFNG",metadata.clean$biosample)])), 
                                             rep(5,length(metadata.clean$biosample[grepl("KRT_IFNG",metadata.clean$biosample)])),
                                             rep(6,length(metadata.clean$biosample[grepl("MEL_IFNG",metadata.clean$biosample)])),
                                             rep(7,length(metadata.clean$biosample[grepl("FRB_IFNB",metadata.clean$biosample)])), 
                                             rep(8,length(metadata.clean$biosample[grepl("KRT_IFNB",metadata.clean$biosample)])), 
                                             rep(9,length(metadata.clean$biosample[grepl("MEL_IFNB",metadata.clean$biosample)])), 
                                             rep(10,length(metadata.clean$biosample[grepl("FRB_TNF",metadata.clean$biosample)])), 
                                             rep(11,length(metadata.clean$biosample[grepl("KRT_TNF",metadata.clean$biosample)])),
                                             rep(12,length(metadata.clean$biosample[grepl("MEL_TNF",metadata.clean$biosample)]))),
                            row_split = split_instructions.rna, 
                            show_row_names = FALSE,
                            row_title_rot = 0,
                            column_names_rot = 45,
                            row_gap = unit(0, "mm"),
                            border = TRUE,
                            column_names_gp = gpar(fontsize = 10),
                            column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                            row_title_gp = gpar(fontsize = 10, fontface = "bold")))
dev.off()
cluster1_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==1)]

for (i in 1:14){
  this_cluster_genes = rownames(km_res.rna$clustered)[which(split_instructions.rna==i)]
  write(this_cluster_genes, file=paste0(Dir,"/inducedGenes_log2FC1.5_cluster",as.character(i),".txt"))
}

```
```{r PCA-of-outliers}
#log2CPM.clean.corrected=log2(CPM.clean.corrected+1)
this_log2CPM <- log2CPM.clean.corrected[expressed_genes,]

# Calculate Coefficient of Variation (COV)
gene_means <- rowMeans(this_log2CPM)
gene_sds <- apply(this_log2CPM, 1, sd)
gene_cov <- gene_sds / gene_means

# Remove genes with mean = 0 to avoid NaNs
gene_cov[is.na(gene_cov)] <- 0

# Select top N variable genes
top_n <- 3000
top_genes <- names(sort(gene_cov, decreasing = TRUE))[1:top_n]

# Subset log2CPM
log2CPM_subset <- this_log2CPM[top_genes, ]

#calculate principal components for top 3000 COV genes' log2CPM
pca_obj <- prcomp(t(log2CPM_subset), scale. = TRUE)

pca_scores <- as.data.frame(pca_obj$x)  # this is samples × PCs
pca_scores$sample_id <- colnames(counts.clean)     # optional
pca_scores$celltype <- metadata.clean$celltype
pca_scores$condition <- metadata.clean$condition
pca_scores$donor <- metadata.clean$donor
pca_scores$read_length <- metadata.clean$read_length
pca_scores$sequencing_platform <- metadata.clean$sequencing_platform
pca_scores$unique_aligned_reads_over_total <- metadata.clean$unique_aligned_reads_over_total
pca_scores$is_outlier <- ifelse(rownames(pca_scores)%in%unique(c(outlier.mel,outlier.krt)),"yes","no")

# Get variance explained
percent_var <- round(100 * (pca_obj$sdev^2 / sum(pca_obj$sdev^2)), 1)
# Now use:
xlab <- paste0("PC1 (", percent_var[1], "%)")
ylab <- paste0("PC2 (", percent_var[2], "%)")

p1 = ggplot(data=pca_scores, aes(
  x = PC1,
  y = PC2,
  color = celltype,
  #shape = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total,
    "<br>outlier", is_outlier
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes\nSeq_platform corrected",
       color = "outlier", shape = "condition",
       x = xlab, y = ylab) 

# Then enable the hover text in plotly
p1_interactive <- ggplotly(p1, tooltip = "text")
p1_interactive

### PC2 v PC3
xlab <- paste0("PC2 (", percent_var[2], "%)")
ylab <- paste0("PC3 (", percent_var[3], "%)")

p2 = ggplot(data=pca_scores, aes(
  x = PC2,
  y = PC3,
  color = celltype,
  shape = condition,
  text = paste(
    "Sample:", rownames(pca_scores),
    "<br>Donor:", donor,
    "<br>Condition:", condition,
    "<br>Celltype:", celltype,
    "<br>read_length:", read_length,
    "<br>Platform:", sequencing_platform,
    "<br>Aligned:", unique_aligned_reads_over_total,
    "<br>outlier", is_outlier
  )
)) +
  geom_point(size = 3) +
  labs(title = "PCA, RNA-seq counts for all 751 samples\ntop 3000 COV expressed genes",
       color = "celltype", shape = "condition",
       x = xlab, y = ylab) 

# Then enable the hover text in plotly
p2_interactive <- ggplotly(p2, tooltip = "text")
p2_interactive
```