---
title: "IFNg_signature"
author: "Shuo Shan"
date: "6/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(DEXSeq)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
library(edgeR)
library(gridExtra)
library(ggpubr)
library(plotly)
library(RCurl)
library(fgsea) # FastGeneSetEncirhmentAnalysis

Dir.mel="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/melanocytes/pipeline_01122022/analysis"
Dir.krt="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/keratinocytes/pipeline_01142022/analysis"
Dir.frb="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_01152022/analysis"

Dir="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/IFNg_signature"
setwd(Dir)

#save.image(file=paste0(Dir,'/myEnvironment.RData'))
#load(paste0(Dir,'/myEnvironment.RData'))
```

### Load useful functions
```{r functions}
gsea_heatmap <- function(fgseaResTidy, gene.in.pathway, f, Dir, gene_of_interest, MSigDBtype) {
  print("compiling data...")
  # pathways with sig enrichment score
  sig.path <- fgseaResTidy$pathway[fgseaResTidy$padj <= 0.05]
  sig.gen <- unique(na.omit(gene.in.pathway$SYMBOL[gene.in.pathway$pathway %in% sig.path]))
  
  # 
  h.dat <- reshape2::dcast(gene.in.pathway[, c(1,2)], SYMBOL~pathway)
  rownames(h.dat) <- h.dat$SYMBOL
  h.dat <- h.dat[, -1]
  h.dat <- h.dat[rownames(h.dat) %in% sig.gen, ]
  h.dat <- h.dat[, colnames(h.dat) %in% sig.path]
  h.dat[!is.na(h.dat)] = 1
  h.dat[is.na(h.dat)] = 0
  h.dat.rownames <- rownames(h.dat)
  h.dat <- h.dat %>% mutate_if(is.character,as.numeric) %>% magrittr::set_rownames(h.dat.rownames)
  
  # keep those genes with 3 or more occurences in significant pathways
  table(data.frame(rowSums(h.dat)))
  h.dat <- h.dat[data.frame(rowSums(h.dat)) >= 3, ]
  
  # get DESeq2 result of the top genes 
  topTable <- f[f$SYMBOL %in% rownames(h.dat), ]
  rownames(topTable) <- topTable$SYMBOL
  # match the order of rownames in toptable with that of h.dat
  topTableAligned <- topTable[which(rownames(topTable) %in% rownames(h.dat)),]
  topTableAligned <- topTableAligned[match(rownames(h.dat), rownames(topTableAligned)),]
  all(rownames(topTableAligned) == rownames(h.dat))
  print("...done, preparing to plot")
  
  # color bar for -log10(adjusted p-value) for sig.genes
  dfMinusLog10FDRGenes <- data.frame(-log10(
    topTableAligned[which(rownames(topTableAligned) %in% rownames(h.dat)), 'padj']))
  dfMinusLog10FDRGenes[dfMinusLog10FDRGenes == 'Inf'] <- 0
  
  # color bar for fold changes for sigGenes
  dfFoldChangeGenes <- data.frame(
    topTableAligned[which(rownames(topTableAligned) %in% rownames(h.dat)), 'log2FoldChange'])
  
  # merge both
  dfGeneAnno <- data.frame(dfMinusLog10FDRGenes, dfFoldChangeGenes)
  colnames(dfGeneAnno) <- c('Gene score', 'Log2FC')
  dfGeneAnno[,2] <- ifelse(dfGeneAnno$Log2FC > 0, 'Up-regulated',
                           ifelse(dfGeneAnno$Log2FC < 0, 'Down-regulated', 'Unchanged'))
  colours <- list(
    'Log2FC' = c('Up-regulated' = 'red', 'Down-regulated' = 'blue'))
  haGenes <- ComplexHeatmap::rowAnnotation(
    df = dfGeneAnno,
    col = colours,
    width = unit(1,'cm'),
    annotation_name_side = 'top')
  
  # Now a separate color bar for the GSEA enrichment padj. This will 
  # also contain the enriched term names via annot_text()
  
  # colour bar for enrichment score from fgsea results
  dfEnrichment <- fgseaRes[, c("pathway", "NES")]
  dfEnrichment <- dfEnrichment[dfEnrichment$pathway %in% colnames(h.dat)]
  dd <- dfEnrichment$pathway
  dfEnrichment <- dfEnrichment[, -1]
  rownames(dfEnrichment) <- dd
  colnames(dfEnrichment) <- 'Normalized\n Enrichment\n Score'
  haTerms <- ComplexHeatmap::HeatmapAnnotation(
    df = dfEnrichment,
    Term = anno_text(
      colnames(h.dat),
      rot = 45,
      just = 'right',
      gp = gpar(fontsize = 12)),
    annotation_height = unit.c(unit(1, 'cm'), unit(8, 'cm')),
    annotation_name_side = 'left')
  
  # now generate the heatmap
  hmapGSEA <- Heatmap(h.dat,
                      name = 'GSEA hallmark pathways enrichment',
                      split = dfGeneAnno[,2],
                      col = c('0' = 'white', '1' = 'forestgreen'),
                      rect_gp = gpar(col = 'grey85'),
                      cluster_rows = TRUE,
                      show_row_dend = TRUE,
                      row_title = 'Top Genes',
                      row_title_side = 'left',
                      row_title_gp = gpar(fontsize = 11, fontface = 'bold'),
                      row_title_rot = 90,
                      show_row_names = TRUE,
                      row_names_gp = gpar(fontsize = 11, fontface = 'bold'),
                      row_names_side = 'left',
                      row_dend_width = unit(35, 'mm'),
                      cluster_columns = TRUE,
                      show_column_dend = TRUE,
                      column_title = 'Enriched terms',
                      column_title_side = 'top',
                      column_title_gp = gpar(fontsize = 12, fontface = 'bold'),
                      column_title_rot = 0,
                      show_column_names = FALSE,
                      show_heatmap_legend = FALSE,
                      clustering_distance_columns = 'euclidean',
                      clustering_method_columns = 'ward.D2',
                      clustering_distance_rows = 'euclidean',
                      clustering_method_rows = 'ward.D2',
                      bottom_annotation = haTerms)
  
  # gene score: -log10(adjusted p-value) for significant genes identified by DESeq2. larger score = smaller p.adj.
  dev.off()
  png(paste0("GSEA_enrichment_heatmap_",gene_of_interest,"_against_",MSigDBtype,"-pathways.png"), units="in", width=20, height=0.1*nrow(h.dat), res=500)
  draw(hmapGSEA + haGenes,
       heatmap_legend_side = 'right',
       annotation_legend_side = 'right')
  dev.off()
  print("...done! Have a great day :)")
}

# bar plot: Normalized Enrichment Scores
gsea_bar_plot <- function(fgseaResTidy, filtered="no", Dir, gene_of_interest, MSigDBtype) {
  # bar plot: Normalized Enrichment Scores
  fgseaResTidy$adjPvalue <- ifelse(fgseaResTidy$padj <= 0.05, "significant", "non-significant")
  cols <- c("non-significant"="grey","significant"="red")
  
  if (filtered=="no") {df <- fgseaResTidy} else {df <- fgseaResTidy %>% dplyr::filter(padj<=0.05)}
  
  p <- df %>% 
    ggplot(., aes(reorder(pathway, NES), NES, fill=adjPvalue)) +
    geom_col(width=0.8) +
    scale_fill_manual(values=cols) +
    coord_flip() +
    labs(x="Pathway", y="Normalized Enrichment Score",
         title=paste0("GSEA pathways enrichment score for\n",gene_of_interest," against\n ",MSigDBtype," pathways")) + 
    theme(plot.title = element_text(size=8),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1, size=10),
          legend.position = "none")
  
  p.interactive <- ggplotly(p, width=1200, height=15*nrow(df))
  htmlwidgets::saveWidget(p.interactive, paste0(Dir,"/GSEA_",gene_of_interest,"_against_",MSigDBtype,"-pathways.html"))
}

```

### Load data
```{r load}
up.krt=read.table(paste0(Dir,"/upregulated_genes_KRT.txt"), header=FALSE) %>% set_colnames("gene")
up.mel=read.table(paste0(Dir,"/upregulated_genes_MEL.txt"), header=FALSE) %>% set_colnames("gene")
up.frb=read.table(paste0(Dir,"/upregulated_genes_FRB.txt"), header=FALSE) %>% set_colnames("gene")

down.krt=read.table(paste0(Dir,"/downregulated_genes_KRT.txt"), header=FALSE) %>% set_colnames("gene")
down.mel=read.table(paste0(Dir,"/downregulated_genes_MEL.txt"), header=FALSE) %>% set_colnames("gene")
down.frb=read.table(paste0(Dir,"/downregulated_genes_FRB.txt"), header=FALSE) %>% set_colnames("gene")

all.gene.krt=read.table(paste0(Dir,"/genes_used_in_DE_KRT.txt"), header=FALSE) %>% set_colnames("gene")
all.gene.mel=read.table(paste0(Dir,"/genes_used_in_DE_MEL.txt"), header=FALSE) %>% set_colnames("gene")
all.gene.frb=read.table(paste0(Dir,"/genes_used_in_DE_FRB.txt"), header=FALSE) %>% set_colnames("gene")

res.krt=read.table(paste0(Dir.krt,"/DESeq2_results_IFNvsPBS.txt"), header=TRUE) %>% mutate(celltype=rep("krt",nrow(.)))
res.mel=read.table(paste0(Dir.mel,"/DESeq2_results_IFNvsPBS.txt"), header=TRUE) %>% mutate(celltype=rep("mel",nrow(.)))
res.frb=read.table(paste0(Dir.frb,"/DESeq2_results_IFNvsPBS.txt"), header=TRUE) %>% mutate(celltype=rep("frb",nrow(.)))

common.genes = Reduce(intersect, list(res.frb$gene,res.mel$gene,res.frb$gene))
res.merged = rbind(res.krt,res.mel,res.frb)
```

### Check to see which genes are shared across all 3 cell types
 234 common_upregulated_genes.txt
   8 common_downregulated_genes.txt
```{r common_IFNg_response}
# common up-regulated genes
common.up = Reduce(intersect, list(up.krt$gene,up.mel$gene,up.frb$gene))
write.table(common.up[order(common.up)], file=paste0(Dir,"/common_upregulated_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# common down-regulated genes
common.down = Reduce(intersect, list(down.krt$gene,down.mel$gene,down.frb$gene))
write.table(common.down[order(common.down)], file=paste0(Dir,"/common_downregulated_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)
```

### Which genes have cell-type-specific IFNg response? Also differentiate the genes that are cell-type-specific due to a change in response from not being expressed.
------------------------------------------  
  187 cts_upregulated_genes_FRB.txt
    - 64 are not expressed/lowly expressed in KRT
    - 41 are not expressed/lowly expressed in MEL
    - 159 are expressed over cutoff threshold in all cell types, but only significantly upregulated in FRB.
  105 cts_downregulated_genes_FRB.txt
    - 36 are not/lowly expressed in KRT
    - 29 are not/lowly expressed in MEL
    - 88 are expressed over cutoff threshold in all cell types, but only significantly downregulated in FRB.
------------------------------------------  
  349 cts_upregulated_genes_KRT.txt
    - 284 are expressed over cutoff threshold in all cell types, but only significantly upregulated in KRT.
  339 cts_downregulated_genes_KRT.txt
    - 282 are expressed over cutoff in all cell types.
------------------------------------------  
  871 cts_upregulated_genes_MEL.txt
    - 770 are expressed over cutoff threshold in all cell types, but only significantly upregulated in MEL.
  648 cts_downregulated_genes_MEL.txt
    - 558 are expressed over cutoff threshold in all cell types, but only significantly downregulated in MEL.
```{r cts_IFNg_response}
# cts IFNg response in fibroblasts
temp1<-unique(c(up.krt$gene,up.mel$gene))
cts.up.frb<-up.frb$gene[which(!up.frb$gene %in% temp1)]

write.table(cts.up.frb[order(cts.up.frb)], file=paste0(Dir,"/cts_upregulated_genes_FRB.txt"), quote=F, sep="\t", row.names=F, col.names=F)

temp2<-unique(c(down.krt$gene,down.mel$gene))
cts.down.frb<-down.frb$gene[which(!down.frb$gene %in% temp2)]
write.table(cts.down.frb[order(cts.down.frb)], file=paste0(Dir,"/cts_downregulated_genes_FRB.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# cts IFNg response in melanocytes
temp3<-unique(c(up.frb$gene,up.krt$gene))
cts.up.mel<-up.mel$gene[which(!up.mel$gene %in% temp3)]
write.table(cts.up.mel[order(cts.up.mel)], file=paste0(Dir,"/cts_upregulated_genes_MEL.txt"), quote=F, sep="\t", row.names=F, col.names=F)

temp4<-unique(c(down.frb$gene,down.krt$gene))
cts.down.mel<-down.mel$gene[which(!down.mel$gene %in% temp4)]
write.table(cts.down.mel[order(cts.down.mel)], file=paste0(Dir,"/cts_downregulated_genes_MEL.txt"), quote=F, sep="\t", row.names=F, col.names=F)


# cts IFNg response in keratinocytes
temp5<-unique(c(up.frb$gene,up.mel$gene))
cts.up.krt<-up.krt$gene[which(!up.krt$gene %in% temp5)]
write.table(cts.up.krt[order(cts.up.krt)], file=paste0(Dir,"/cts_upregulated_genes_KRT.txt"), quote=F, sep="\t", row.names=F, col.names=F)

temp6<-unique(c(down.frb$gene,down.mel$gene))
cts.down.krt<-down.krt$gene[which(!down.krt$gene %in% temp6)]
write.table(cts.down.krt[order(cts.down.krt)], file=paste0(Dir,"/cts_downregulated_genes_KRT.txt"), quote=F, sep="\t", row.names=F, col.names=F)

rm(temp1,temp2,temp3,temp4,temp5,temp6)
```


### Illustrate the results in Venn Diagram
```{r venn-up}
library(VennDiagram)
library(RColorBrewer)
# Generate 3 sets of 200 words
set1 <- up.frb$gene
set2 <- up.krt$gene
set3 <- up.mel$gene
 
# Prepare a palette of 3 colors with R colorbrewer:

myCol <- brewer.pal(3, "Pastel2")

# Chart
venn.diagram(
        x = list(set1, set2, set3),
        category.names = c("up.frb" , "up.krt" , "up.mel"),
        filename = 'venn_diagramm_upregulated_genes.png',
        output=TRUE,
        
        # Output features
        imagetype="png" ,
        height = 480 , 
        width = 480 , 
        resolution = 300,
        compression = "lzw",
        
        # Circles
        lwd = 2,
        lty = 'blank',
        fill = myCol,
        
        # Numbers
        cex = .6,
        fontface = "bold",
        fontfamily = "sans",
        
        # Set names
        cat.cex = 0.6,
        cat.fontface = "bold",
        cat.default.pos = "outer",
        cat.pos = c(-27, 27, 135),
        cat.dist = c(0.055, 0.055, 0.085),
        cat.fontfamily = "sans",
        rotation = 1
)
```
```{r venn-down}
library(VennDiagram)
library(RColorBrewer)
# Generate 3 sets of 200 words
rm(set1,set2,set3)
set1 <- down.frb$gene
set2 <- down.krt$gene
set3 <- down.mel$gene
 
# Prepare a palette of 3 colors with R colorbrewer:

myCol <- brewer.pal(3, "Pastel2")

# Chart
venn.diagram(
        x = list(set1, set2, set3),
        category.names = c("down.frb" , "down.krt" , "down.mel"),
        filename = 'venn_diagramm_downregulated_genes.png',
        output=TRUE,
        
        # Output features
        imagetype="png" ,
        height = 480 , 
        width = 480 , 
        resolution = 300,
        compression = "lzw",
        
        # Circles
        lwd = 2,
        lty = 'blank',
        fill = myCol,
        
        # Numbers
        cex = .6,
        fontface = "bold",
        fontfamily = "sans",
        
        # Set names
        cat.cex = 0.6,
        cat.fontface = "bold",
        cat.default.pos = "outer",
        cat.pos = c(-27, 27, 135),
        cat.dist = c(0.055, 0.055, 0.085),
        cat.fontfamily = "sans",
        rotation = 1
)
```

### Cytokines or Cytokine receptors
```{r cytokine}
library(SignallingSingleCell)
# https://rdrr.io/github/kgellatl/SignallingSingleCell/man/id_rl.html

#common.up

#common.down
#cts.up.krt
#cts.down.krt
#cts.up.mel
#cts.down.mel
#cts.up.frb
#cts.down.frb
```

### Fold Change Distribution Violin Plot
```{r fc-plot-common}
#violin of common up genes
this.genes=common.up
this.df=res.merged %>% dplyr::filter(gene %in% this.genes) 
this.title="common upregulated genes"
this.fname="common.up.png"
png(this.fname, width=800,height=600)
this.df %>%
  ggplot( . , aes(x=celltype,y=log2FoldChange,fill=celltype)) +
  geom_violin(alpha=0.4) +
  geom_boxplot(width=0.1,alpha=0.4) +
  theme(text = element_text(size = 20,face="bold")) + 
  labs(title=this.title, subtitle=paste0(as.character(length(this.genes))," genes")) 
dev.off()

#violin of common down genes
this.genes=common.down
this.df=res.merged %>% dplyr::filter(gene %in% this.genes) 
this.title="common downregulated genes"
this.fname="common.down.png"
png(this.fname, width=800,height=600)
this.df %>%
  ggplot( . , aes(x=celltype,y=log2FoldChange,fill=celltype)) +
  geom_violin(alpha=0.4) +
  geom_boxplot(width=0.1,alpha=0.4) +
  theme(text = element_text(size = 20,face="bold")) + 
  labs(title=this.title, subtitle=paste0(as.character(length(this.genes))," genes")) 
dev.off()

```
```{r fc-plot-cts-krt}
#violin of cts.up.krt genes
this.genes=cts.up.krt
this.df=res.merged %>% dplyr::filter(gene %in% this.genes) 
this.title="keratinocytes-speicfic upregulated genes"
this.fname="cts.up.krt.png"
png(this.fname, width=800,height=600)
this.df %>%
  ggplot( . , aes(x=celltype,y=log2FoldChange,fill=celltype)) +
  geom_violin(alpha=0.4) +
  geom_boxplot(width=0.1,alpha=0.4) +
  theme(text = element_text(size = 20,face="bold")) + 
  labs(title=this.title, subtitle=paste0(as.character(length(this.genes))," genes")) 
dev.off()

#violin of cts.down.krt genes
this.genes=cts.down.krt
this.df=res.merged %>% dplyr::filter(gene %in% this.genes) 
this.title="keratinocytes-speicfic downregulated genes"
this.fname="cts.down.krt.png"
png(this.fname, width=800,height=600)
this.df %>%
  ggplot( . , aes(x=celltype,y=log2FoldChange,fill=celltype)) +
  geom_violin(alpha=0.4) +
  geom_boxplot(width=0.1,alpha=0.4) +
  theme(text = element_text(size = 20,face="bold")) + 
  labs(title=this.title, subtitle=paste0(as.character(length(this.genes))," genes")) 
dev.off()


```
```{r fc-plot-cts-mel}
#violin of cts.up.mel genes
this.genes=cts.up.mel
this.df=res.merged %>% dplyr::filter(gene %in% this.genes) 
this.title="melanocyte-speicfic upregulated genes"
this.fname="cts.up.mel.png"
png(this.fname, width=800,height=600)
this.df %>%
  ggplot( . , aes(x=celltype,y=log2FoldChange,fill=celltype)) +
  geom_violin(alpha=0.4) +
  geom_boxplot(width=0.1,alpha=0.4) +
  theme(text = element_text(size = 20,face="bold")) + 
  labs(title=this.title, subtitle=paste0(as.character(length(this.genes))," genes")) 
dev.off()

#violin of cts.down.mel genes
this.genes=cts.down.mel
this.df=res.merged %>% dplyr::filter(gene %in% this.genes) 
this.title="melanocyte-speicfic downregulated genes"
this.fname="cts.down.mel.png"
png(this.fname, width=800,height=600)
this.df %>%
  ggplot( . , aes(x=celltype,y=log2FoldChange,fill=celltype)) +
  geom_violin(alpha=0.4) +
  geom_boxplot(width=0.1,alpha=0.4) +
  theme(text = element_text(size = 20,face="bold")) + 
  labs(title=this.title, subtitle=paste0(as.character(length(this.genes))," genes")) 
dev.off()


```
```{r fc-plot-cts-frb}
#violin of cts.up.frb genes
this.genes=cts.up.frb
this.df=res.merged %>% dplyr::filter(gene %in% this.genes) 
this.title="fibroblast-speicfic upregulated genes"
this.fname="cts.up.frb.png"
png(this.fname, width=800,height=600)
this.df %>%
  ggplot( . , aes(x=celltype,y=log2FoldChange,fill=celltype)) +
  geom_violin(alpha=0.4) +
  geom_boxplot(width=0.1,alpha=0.4) +
  theme(text = element_text(size = 20,face="bold")) + 
  labs(title=this.title, subtitle=paste0(as.character(length(this.genes))," genes")) 
dev.off()

#violin of cts.down.frb genes
this.genes=cts.down.frb
this.df=res.merged %>% dplyr::filter(gene %in% this.genes) 
this.title="fibroblast-speicfic downregulated genes"
this.fname="cts.down.frb.png"
png(this.fname, width=800,height=600)
this.df %>%
  ggplot( . , aes(x=celltype,y=log2FoldChange,fill=celltype)) +
  geom_violin(alpha=0.4) +
  geom_boxplot(width=0.1,alpha=0.4) +
  theme(text = element_text(size = 20,face="bold")) + 
  labs(title=this.title, subtitle=paste0(as.character(length(this.genes))," genes")) 
dev.off()


```


### Heatmap
```{r heatmap-data}
# merge metadata
metadata.mel=read.table(paste0(Dir.mel,"/metadata.txt"), header=TRUE) %>% dplyr::mutate(celltype=rep("MEL",nrow(.)))
metadata.krt=read.table(paste0(Dir.krt,"/metadata.txt"), header=TRUE) %>% dplyr::mutate(celltype=rep("KRT",nrow(.)))
metadata.frb=read.table(paste0(Dir.frb,"/metadata.txt"), header=TRUE) %>% dplyr::mutate(celltype=rep("FRB",nrow(.)))
metadata=rbind(metadata.mel,metadata.krt,metadata.frb) %>% as.data.frame()
rownames(metadata)=metadata$sample

# merge logCPM data
logCPM.mel=read.table(paste0(Dir.mel,"/log2-TMM-normalized-CPM.txt"), header=TRUE) %>% rownames_to_column("gene")
logCPM.krt=read.table(paste0(Dir.krt,"/log2-TMM-normalized-CPM.txt"), header=TRUE) %>% rownames_to_column("gene")
logCPM.frb=read.table(paste0(Dir.frb,"/log2-TMM-normalized-CPM.txt"), header=TRUE) %>% rownames_to_column("gene")
logCPM=left_join(logCPM.frb, left_join(logCPM.mel,logCPM.krt,by="gene"), by="gene") %>% column_to_rownames("gene")

# pick genes of top row variance
mat.logCPM.lnsd <- apply(logCPM,MARGIN=1,FUN=sd)
mat.logCPM.lncv <- sqrt( (exp(mat.logCPM.lnsd) )^2 -1) #coefficient of variance
```

```{r heatmap-top10000}
logCPM.top <- logCPM[order(mat.logCPM.lncv,decreasing=T),][1:10000,]

# subset logCPM to include top 2000 cov genes and selected samples.
df=logCPM[rownames(logCPM.top),metadata$sample]

# perform k-means-clustering to determine k_row.
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
png("elbow_plot_top10000cov_log2CPM.png", width=800,height=600)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 
dev.off()

# heatmap
library(pheatmap)
df.meta=metadata[,c("condition","celltype")]
png("heatmap_top10000cov_log2CPM_unscaled.png", width=800,height=600)
pheatmap(df, 
         annotation_col = df.meta,
         show_rownames = FALSE,
         show_colnames = FALSE,
         cutree_cols = 3,
         cutree_rows = 5)
dev.off()
```

```{r heatmap-IFNg-responsive-genes}
IFNg.response.genes=res.merged %>% 
  dplyr::filter(padj<0.05) %>%
  dplyr::filter(abs(log2FoldChange) > 1.5) %>%
  pull(gene)
logCPM.top <- logCPM[IFNg.response.genes,]

# subset logCPM to include top 2000 cov genes and selected samples.
df=logCPM[rownames(logCPM.top),metadata$sample]
df<- na.omit(df)

# perform k-means-clustering to determine k_row.
wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
png("elbow_plot_IFNg-responsive-genes_log2CPM.png", width=800,height=600)
plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 
dev.off()

# heatmap
library(pheatmap)
df.meta=metadata[,c("condition","celltype")]
png("heatmap_IFNg-responsive-genes_log2CPM_unscaled.png", width=800,height=600)
pheatmap(df, 
         annotation_col = df.meta,
         show_rownames = FALSE,
         show_colnames = FALSE,
         cutree_cols = 3,
         cutree_rows = 5)
dev.off()

png("heatmap_IFNg-responsive-genes_log2CPM_scaled.png", width=800,height=600)
pheatmap(df, 
         scale = "row",
         annotation_col = df.meta,
         show_rownames = FALSE,
         show_colnames = FALSE,
         cutree_cols = 3,
         cutree_rows = 5)
dev.off()
```


### GeneSetEnrichmentAnalysis
```{r MSigDB-gene-sets}
# load the MSigDB gene sets into a named list
pathways.kegg <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c2.cp.kegg.v7.5.1.symbols.gmt")
# pathways.hallmark <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/h.all.v7.5.1.symbols.gmt")
#pathways.allcurated <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c2.all.v7.5.1.symbols.gmt")
# pathways.canonical <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c2.cp.v7.5.1.symbols.gmt")
# pathways.chemical_and_genetic_perturbations <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c2.cgp.v7.5.1.symbols.gmt")
# pathways.regulatory_target <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c3.all.v7.5.1.symbols.gmt")
pathways.ontology <- 	gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c5.all.v7.5.1.symbols.gmt")
pathways.GO_BP <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c5.go.bp.v7.5.1.symbols.gmt")
# pathways.GO_MF <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c5.go.mf.v7.5.1.symbols.gmt")
# pathways.GO_CC <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c5.go.cc.v7.5.1.symbols.gmt")
#pathways.immunologic_signature <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c7.all.v7.5.1.symbols.gmt")
# pathways.celltype_signature <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/c8.all.v7.5.1.symbols.gmt")
#pathways.all <- gmtPathways("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/GSEA/MSigDB/msigdb.v7.5.1.symbols.gmt")
pathways <- pathways.ontology
MSigDBtype="pathways.ontology" # type of MSigDB gene set
```  

```{r gsea}
gene.lst=data.frame(SYMBOL=common.up)
gene_of_interest = "common up-regulated genes"

# create a named vector [ranked genes]
ranks <- rank(gene.lst$SYMBOL)
names(ranks) <- gene.lst$SYMBOL
#barplot(sort(ranks, decreasing = T))

# run fgsea algorithm:
fgseaRes <- fgseaMultilevel(pathways=pathways, stats=ranks)
head(fgseaRes[order(padj, -abs(NES)), ], n=10)

# tidy the results:
fgseaResTidy <- fgseaRes %>% as_tibble %>%
  arrange(desc(NES)) # order by normalized enrichement score

# to see what genes are in each of these pathways:
gene.in.pathway <- pathways %>%
  enframe("pathway", "SYMBOL") %>%
  unnest(cols=c(SYMBOL)) %>%
  inner_join(gene.lst, by="SYMBOL")
# PLOT HEATMAP
gsea_heatmap(fgseaResTidy, gene.in.pathway, f, Dir, gene_of_interest, MSigDBtype)
# PLOT BARPLOT
gsea_bar_plot(fgseaResTidy, filtered="yes", Dir, gene_of_interest, MSigDBtype)

```













