---
title: "celseq2_analysis_MEL"
author: "Shuo Shan"
date: "6/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Intro
  
Celseq2 data for melanotytes (MEL) from 14 donors X 2 culture conditions. Libraries were pooled and prepared by celseq2 protocol, pair-end sequenced by BGI, trimmed to (read 1 3bp, read 2 36bp), barcode in read 1 name was used to separate out pooled samples within a library, and read 2 was used for downstream processing and treated as single-end sequencing. Read 2 was aligned by star to hg38, and gene expression count matrix generated by ESAT.
  
### Load packages.    
```{r package, warning=FALSE}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(DEXSeq)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
library(vcfR)
library(cluster) # for gower similarity and pam
library(ggplot2) # for visualization

Dir="/nl/umw_manuel_garber/human/skin/eQTLs/dnbseq"
#setwd(paste0(Dir,"/R"))
celltype="MEL"
mat <- read.table(file=paste0(Dir,"/ESAT/output/gene/",celltype,".gene.txt"),
                  header=T,row.names=1)
mat <- mat %>% dplyr::select(!starts_with("F47"))
summary <- read.table(file=paste0(Dir,"/DolphinNext_052821/report4643/summary/","overall_summary.tsv"), 
                  sep="\t",header=T,row.names=1) %>% rownames_to_column("name")
summary <- tidyr::separate(summary, col=name, into=c("bc","donor","condition"),sep="_",remove=F) %>% 
  unite("name", donor:condition, sep = "_", remove = T) %>%
  dplyr::select(-bc)
lib_info <- read.table(file=paste0(Dir,"/R/celseq2_MEL_libinfo.txt"),header=F)
colnames(lib_info) <- c("name","barcode","barcode.factor","Pooled.Library.Fragment.Size","Pooled.Lib.Size.BioAnalyzer(bp)",
                            "Fragment.Size.Matched","Pooled.library.Total.Reads","Lib.Prep.Batch","Lib.Prep.Date",
                            "reads_ESAT","pct.reads","PCR.cycle.n","Lib Conc.Qubit(ng/μl)","Genotype.Cluster")

```

### Data preparation.
```{r prep}
# count matrix and cpm matrix. (cpm=counts per million)
mat.count <- mat %>% dplyr::select(-c(1,2))
cSum <- apply(mat.count,2,sum)
mat.cpm <- 1000000*sweep(mat.count,MARGIN=2,FUN="/",STATS=cSum)

# filter out empty/low-expressed genes with < 5 cpm
mat.cpm$max = apply(mat.cpm, 1, function (x) max(x))
mat.cpm.filtered <- mat.cpm %>% rowwise() %>% mutate(max=max(F27M_PBS:F56M_IFNg)) %>% 
  as.data.frame() %>% magrittr::set_rownames(rownames(mat.cpm)) %>%
  filter(max>=5) %>% dplyr::select(-max)
mat.count.filtered <- mat.count[rownames(mat.cpm.filtered),]

# metadata
metadata <- data.frame(key=colnames(mat)[3:ncol(mat)]) %>%
  tidyr::separate(. , col=key, into=c("donor","condition"),sep="_",remove=F) %>%
  column_to_rownames(var="key")
library <- data.frame(donor=c("F27M","F34M","F30M","F31M","F41M","F45M","F46M","F47M","F49M","F50M","F51M","F52M","F55M","F56M"),
                    library=c("1","1","2","2","3","3","4","4","5","5","6","6","7","7"))
genotype.cluster <- data.frame(donor=c("F27M","F30M","F34M","F45M","F50M","F52M","F56M","F31M","F41M","F46M","F49M","F51M","F55M"),
                               cluster=c("1","1","1","1","1","1","1","2","3","4","5","6","7"))
metadata <- metadata %>% mutate(name=rownames(metadata))
metadata <- inner_join(metadata,library,by="donor")
metadata <- inner_join(metadata,genotype.cluster,by="donor")
metadata <- inner_join(metadata,summary,by="name")
metadata <- inner_join(metadata,lib_info,by="name") %>% magrittr::set_rownames(metadata$name)
metadata$condition <- factor(metadata$condition, levels=c("PBS","IFNg"))

# clean up
#rm(mat,mat.cpm)
rm(lib_info,summary,cSum)
```

```{r genotype}
vcfdir="/nl/umw_manuel_garber/human/skin/eQTLs/donor_genotyping/hg38"
vcf <- read.vcfR(paste0(vcfdir,"/vcf/merged/bd841628-fcc2-487a-8460-f5428237f0c9.merged.filtered.2.16donors.vcf.gz"))
vcf_table <- as_tibble(cbind(vcf@fix,vcf@gt))
vcf_table <- vcf_table %>% filter(!is.na(ID)) %>% unite(mergedID,c("ID","REF","ALT"),sep=":",remove=F)
colnames(vcf_table3) <- colnames(vcf_table3) %>%
  gsub("\\.", "-", .) %>%
  gsub("6d0b04d1-84e3-4979-8b8b-b93792da776f",  "F25", .) %>%
  gsub("X149430e7-6d2d-4e50-baba-3e8f044ce2e3", "F27", .) %>%
  gsub("X5b67f1d5-29b4-448f-b9a3-d23a2d12cd93", "F30", .) %>%
  gsub("d8380c80-b025-4d9b-9d06-67ee06eac4ad",  "F31", .) %>%
  gsub("X5df7279b-bef9-44b3-a1c1-87b3318b79d4", "F34", .) %>%
  gsub("e7b548bc-bed5-4a30-b4d9-0151966e6b9d",  "F41", .) %>%
  gsub("f4a987ea-ca08-4b38-8668-48e5efe45076",  "F45", .) %>%
  gsub("f0b571af-c87e-4611-8288-bfcf4f9e9133",  "F46", .) %>%
  gsub("b0690c72-9cf6-465f-b5dc-e0b94fe8c8f3",  "F49", .) %>%
  gsub("edbb9163-65d7-4fc8-9d14-627d0987f087",  "F50", .) %>%
  gsub("X57948dd0-a324-4477-a917-c33f9c4e2daf", "F51", .) %>%
  gsub("X383fe24c-a10e-424e-abcf-29d1e2a61c5d", "F52", .) %>%
  gsub("X8f482258-bffd-40ca-be3a-487644f0c52f", "F55", .) %>%
  gsub("X29026b76-493d-4f87-9e3b-800e311c19ed", "F56", .) %>%
  gsub("3e5e3f37-fe51-4fb7-a356-80d803c3db9b",  "F62", .) %>%
  gsub("413a4cbe-af1f-4cbb-a794-14c88b30da31",  "F63", .)
```
## ANALYSIS PART ONE (PCA OF CPM DATA)
### get top 2500 genes with most variable cpm across all samples
```{r top2500, include=F}
### generate stats (mean, std, cv) of log2-transformed data
mat.cpm.filtered.log2 <- log2(mat.cpm.filtered+1)
mat.cpm.filtered.mean <- log2(apply(mat.cpm.filtered,MARGIN=1,FUN=mean)+1)
mat.cpm.filtered.lnsd <- apply(mat.cpm.filtered.log2,MARGIN=1,FUN=sd)
mat.cpm.filtered.lncv <- sqrt( (exp(mat.cpm.filtered.lnsd) )^2 -1) #coefficient of variance

# order by cv
mat.cpm.filtered.log2.ordered <- mat.cpm.filtered.log2[order(mat.cpm.filtered.lncv, decreasing=T),]
mat.cpm.filtered.ordered <- mat.cpm.filtered[order(mat.cpm.filtered.lncv, decreasing=T),]
mat.count.filtered.ordered <- mat.count.filtered[order(mat.cpm.filtered.lncv, decreasing=T),]

# pick top 2500 genes, largest cv first
mat.cpm.filtered.log2.top = mat.cpm.filtered.log2.ordered[1:2500,]
mat.cpm.filtered.top = mat.cpm.filtered.ordered[rownames(mat.cpm.filtered.log2.top),]

# plot: mean vs. CV
df <- data.frame(mean=mat.cpm.filtered.mean,
                 cv=mat.cpm.filtered.lncv)
ggplot(df,aes(x=mean,y=cv)) + geom_point()

# plot: density distribution
n=1
sample=mat.cpm.filtered[,colnames(mat.cpm.filtered)[n]] %>% as.data.frame() %>% magrittr::set_colnames("cpm")
ggplot(sample, aes(x=cpm)) + geom_density()

```
  
### PCA
```{r pca, echo=F}
mat.pca.num <- t(mat.cpm.filtered.log2.top) %>% as.data.frame()
pca = prcomp(as.matrix(mat.pca.num),centered= T,scale = T )

#mat.pca <- merge(metadata,mat.pca.num,by=0)
#autoplot(pca, data=mat.pca, colour='condition') ### this looks wrong??? why?

# loadings in each pc
loadings=as.data.frame(pca$rotation)
reactable(loadings)

# percent variance in each pc
plot(pca)
pcs = pca$x 
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# top 3 PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], PC4=pca$x[,4], name=rownames(pcs))
d=inner_join(d,metadata,by="name") %>% magrittr::set_rownames(d$name)
d$reads.ESAT=cSum
d$pct.reads=d$reads.ESAT/d$Unique.Reads.Aligned..STAR.
colnames(d)
#reactable(d)
```

```{r pca-plots}

# PC1 vs. PC2, by condition
#png(filename=paste0(Dir,"/R/plots/celseq2_MEL_top2500_PC1.vs.PC2_condition.png"))
ggplot(d, aes(x=PC1,y=PC2,col=condition,label=name)) + geom_point() + geom_text_repel(show.legend = F) + scale_color_discrete(name = 'condition') +xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))
#dev.off()

# PC1 vs. PC2, by batch
d %>%
  mutate(q=as.factor(Lib.Prep.Date)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=Lib.Prep.Date,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'Lib.Prep.Date') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  scale_color_manual(values = c("11/17/21" = "red",
                                "12/16/21"="blue")) 

# PC1 vs. PC2, by PCR cycle
d %>%
  mutate(PCR.cycle=as.factor(PCR.cycle.n)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=PCR.cycle,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'PCR.cycle.n') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  scale_color_manual(values = c("13" = "red",
                              "10"="blue")) 

# PC2 vs. PC3
d %>%
  mutate(Genotype.Cluster=as.factor(cluster)) %>%
  ggplot(., aes(x=PC1,y=PC3,col=cluster,label=name)) + 
  geom_point(size=3) + geom_text_repel(show.legend = F) + 
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC3 (%2.1f%% variance)",100*pca.pctVar[3])) +
  theme(legend.position = "right") +
  guides(color = guide_legend(ncol = 2, bycol = TRUE))

colnames(d)

```
 
```{r write_to_cluster, include=F}
# top 2500 genes (by variation across 28 samples)
n=rownames(mat.cpm.filtered.top)
write.table(n,file=paste0(Dir,"/R/celseq2_MEL_top2500_gene.txt"),col.names=NA)
write.table(mat.cpm.filtered.top,file=paste0(Dir,"/R/celseq2_MEL_top2500_cpm.txt"),quote=F,sep="\t",col.names=NA)

# loadings for all PCs
write.table(loadings,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_loadings.txt"),quote=F,sep="\t",col.names=NA)

# top genes in PC1:3
loadings.PC1=loadings %>% arrange(desc(PC1)) %>% dplyr::select(PC1)
write.table(loadings.PC1,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_loadings_PC1.txt"),quote=F,sep="\t",col.names=NA)
loadings.PC2=loadings %>% arrange(desc(PC2)) %>% dplyr::select(PC2)
write.table(loadings.PC2,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_loadings_PC2.txt"),quote=F,sep="\t",col.names=NA)
loadings.PC3=loadings %>% arrange(desc(PC3)) %>% dplyr::select(PC3)
write.table(loadings.PC3,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_loadings_PC3.txt"),quote=F,sep="\t",col.names=NA)

# metadata & top 4 PCs
write.table(d,file=paste0(Dir,"/R/celseq2_MEL_top2500_PCA_metadata.txt"),quote=F,sep="\t",col.names=NA)
```
  
### heatmap with top 2500 variable genes
```{r heatmap}
heatmap.2(as.matrix(mat.cpm.filtered.log2.top), trace="none", density="none", col=bluered(75),
          Colv = T, Rowv=T, 
          dendrogram="column", 
          scale="row",
          # distance: 1 minus Pearson correlation distance with average linkage
          distfun=function(x) as.dist((1-cor(t(x)))/2), # as.dist() computes distance matrix. 
          srtCol=45, cexCol=0.40,  #labeling angle
          main="Melanocytes (IFNg stimulation)")
```
  
### k-means clustering with top 2500 variable genes
```{r kmc, include=F}
k = 5;
source = t(scale(t(mat.cpm.filtered.top)));
kmeans = kmeans(source,centers = k,iter.max = 1000,algorithm="Lloyd");
clustered.e.eif4e = data.frame();
breaks = c();
for (i in 1:k) {
  print(sprintf("Doing cluster %i",i))
  cluster = source[kmeans$cluster==i,];
  rows = row.names(cluster);
  clust = heatmap.2(as.matrix(cluster), distfun=function(x) as.dist((1-cor(t(x)))/2),dendrogram = "none",srtCol=45,cexCol=0.4)
  cluster.ordered = cluster[clust$rowInd,];
  cluster.ordered.genes = rows[clust$rowInd];
  row.names(cluster.ordered) = cluster.ordered.genes;
  class = data.frame(row.names=cluster.ordered.genes);
  class$class = i 
  cluster.ordered = cbind(cluster.ordered, class);
  clustered.e.eif4e = rbind(clustered.e.eif4e, cluster.ordered);
  if(i > 1 & i < k) {
    breaks[i] = as.numeric(breaks[i-1]) + length(rows);
  } else if(i==1) {
    breaks[i] = length(rows);
  }
}

dims = dim(clustered.e.eif4e);

png(filename=paste0(Dir,"/R/plots/celseq2_MEL_top2500_kmc_heatmap.png"),width=2000,height=1500)
par(mar=c(10,4,10,4)+0.1)
heatmap.2(as.matrix(clustered.e.eif4e[1:100,1:dims[2]-1]), trace="none", density="none", 
          col=bluered(30), Colv = T,Rowv=F,dendrogram = "column",
          distfun=function(x) as.dist((1-cor(t(x)))/2),rowsep=breaks, sepcol="black",sepwidth=c(1,1),
          srtCol=40, cexCol=1, offsetCol=-0.4,
          cexRow=0.8,offsetRow=-0.3,
          main="Top 2500 genes Full clustering")
dev.off()
```
  
### heatmaps for Genes of interest (Identified by Ken) --> not done yet
```{r ken_genes, include=F}
ken_genes <- unlist(read.csv("pilot1_ken_genes.csv", header=FALSE), use.names=FALSE)
ken_genes2 <- ken_genes[ken_genes %in% rownames(mat.cpm.filtered.log2)]
#GPMNB (and other names for it) - not found in dataset

heatmap.2(as.matrix(mat.cpm.filtered.log2[ken_genes2,]), trace="none", density="none", col=bluered(75),
          Colv = T, Rowv=T, dendrogram="column", scale="row",
          distfun=function(x) as.dist((1-cor(t(x)))/2),
          srtCol=45, cexCol=0.4, main="Selected Genes, All samples (IFN stimulation)")

```

### DESeq2
```{r deseq2, include=F}
dds <- DESeqDataSetFromMatrix(countData=as.matrix(mat.count.filtered),
                               colData=metadata, 
                               design = ~ condition)
dr <- DESeq(dds);
res <- results(dr, contrast=c("condition","PBS","IFNg"));
res_tmp <- res[!is.na(res$padj) & !is.na(res$log2FoldChange),] %>% as.data.frame()
res_p01 <- res_tmp[(res_tmp$padj<0.01),] %>% as.data.frame()
View(as.data.frame(res_tmp))
```











## ANALYSIS PART TWO (DESeq2) 
This may not work because the DESeq2 calculates one log2FC value per gene, considering all 14 samples
### DESeq2 analysis

```{r DESeq2}
mat <- mat[,-c(1,2)]
dds <- DESeqDataSetFromMatrix(countData=mat,colData=metadata,design=~condition)
# pre-filtering
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
# re-factor
dds$condition <- relevel(dds$condition, ref="PBS")
# differential expression analysis
dds <- DESeq(dds)
res <- results(dds, name="condition_IFNg_vs_PBS")
mcols(res)$description
# log fold change shrinkage for visualization and ranking
resLFC <- lfcShrink(dds, coef="condition_IFNg_vs_PBS",type="apeglm")
resLFC
# MA plot & identify genes of interest
plotMA(resLFC)
idx <- identify(res$baseMean, res$log2FoldChange)
rownames(res)[idx]
# plot counts
g="CD74"
plotCounts(dds,gene=g,intgroup="condition")
# ^ another way of plotting:
d<-plotCounts(dds,gene=g,intgroup="condition",returnData=T)
ggplot(d, aes(x=condition, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) + 
  scale_y_log10(breaks=c(25,100,400))

saveRDS(dds, file=paste0(Dir,"/R/celseq2_",celltype,"_","DESeq2_dds.rds"))
saveRDS(res, file=paste0(Dir,"/R/celseq2_",celltype,"_","res.rds"))
saveRDS(resLFC, file=paste0(Dir,"/R/celseq2_",celltype,"_","resLFC.rds"))
```


## ANALYSIS PART THREE (PCA OF LOG2_FC_CPM)
### prep log2fc
```{r prep3}
mat.PBS <- mat.count.filtered %>% dplyr::select(ends_with("PBS"))
colnames(mat.PBS)<-sapply(colnames(mat.PBS),function(x) (as.matrix(unlist(strsplit(x,"_")))[1,])) %>% as.vector()
mat.IFN <- mat.count.filtered %>% dplyr::select(ends_with("IFNg"))
colnames(mat.IFN)<-sapply(colnames(mat.IFN),function(x) (as.matrix(unlist(strsplit(x,"_")))[1,])) %>% as.vector()

mat.log2FC <- log2(mat.IFN+1) - log2(mat.PBS+1)
metadata.log2fc <- metadata %>% 
  dplyr::select(c(donor,library,PCR.cycle.n,cluster,Genotype.Cluster)) %>% 
  dplyr::distinct() %>% magrittr::set_rownames(.$donor)
```

### PCA
```{r PCA-log2fc}
# select top 2500 genes with most variable log2fc
mat.log2FC.lnsd <- apply(mat.log2FC,MARGIN=1,FUN=sd)
mat.log2FC.lncv <- sqrt( (exp(mat.log2FC.lnsd) )^2 -1) #coefficient of variance
# order by cv
mat.log2FC.ordered <- mat.log2FC[order(mat.log2FC.lncv, decreasing=T),]
# pick top 2500 genes, largest cv first
mat.log2FC.top = mat.log2FC.ordered[1:2500,]
mat.log2FC <- mat.log2FC.top
#mat.log2FC <- mat.log2FC.ordered

# PCA analysis
mat.pca.num <- t(mat.log2FC) %>% as.data.frame()
pca = prcomp(as.matrix(mat.pca.num),centered= T,scale = T )

#mat.pca <- merge(metadata,mat.pca.num,by=0)
#autoplot(pca, data=mat.pca, colour='condition') ### this looks wrong??? why?

# loadings in each pc
loadings=as.data.frame(pca$rotation)
#reactable(loadings)

# percent variance in each pc
plot(pca)
pcs = pca$x 
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# top 3 PC information
d=data.frame(PC1=pca$x[,1],PC2=pca$x[,2], PC3=pca$x[,3], name=rownames(pcs))
metadata.log2fc$name<-rownames(metadata.log2fc)
d=inner_join(d,metadata.log2fc,by="name") %>% magrittr::set_rownames(d$name)
colnames(d)
#reactable(d)
```

```{r pca-plots}
# PC1 vs. PC2, by PCR cycle
d %>%
  mutate(PCR.cycle=as.factor(PCR.cycle.n)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=PCR.cycle,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'PCR.cycle.n') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  scale_color_manual(values = c("13" = "red",
                              "10"="blue")) 

# PC1 vs. PC2, by donor vitiligo-GWAS genotype cluster
d %>%
  mutate(genotype=as.factor(Genotype.Cluster)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=genotype,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'genotype') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))

# PC2 vs. PC2, by donor genotype cluster (of all variants with LD <= 0.8)
d %>%
  mutate(genotype=as.factor(cluster)) %>%
  ggplot(., aes(x=PC1,y=PC2,col=genotype,label=name)) + geom_point() + geom_text_repel(show.legend = F) + 
  scale_color_discrete(name = 'genotype') +
  xlab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  ylab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2]))
colnames(d)

```

```{r gene-ontology}
top.genes.log2FC <- rownames(mat.log2FC.top)
write.table(top.genes.log2FC,file=paste0(dir,"/13donors.MEL.Celseq2.top.genes.log2FC.txt"),sep="\n",quote=F,row.names=F,col.names=F)
```

```{r rs12203592}
g1 <- mat.log2FC.ordered %>% rownames_to_column("gene") %>% filter(gene=="IRF4") %>% t() %>% as.data.frame()
g2 <- mat.log2FC.ordered %>% rownames_to_column("gene") %>% filter(gene=="TMEM140") %>% t() %>% as.data.frame()
g3 <- mat.log2FC.ordered %>% rownames_to_column("gene") %>% filter(gene=="PLA1A") %>% t() %>% as.data.frame()
g4 <- mat.log2FC.ordered %>% rownames_to_column("gene") %>% filter(gene=="NEO1") %>% t() %>% as.data.frame()
g <- cbind(g1,g2,g3,g4)
colnames(g) <- g[1,]
g <- g[2:14,]
g$genotype <- as.factor(c("NorthernEurope","NorthernEurope","Mixed","Mixed","African","MediterraneanEurope","African","Mixed","MediterraneanEurope","MediterraneanEurope","NorthernEurope","NorthernEurope","NorthernEurope"))
g$IRF4 <- as.numeric(g$IRF4)
g$TMEM140 <- as.numeric(g$TMEM140)
g$PLA1A <- as.numeric(g$PLA1A)
g$NEO1 <- as.numeric(g$NEO1)

ggplot(g,aes(y=IRF4,x=genotype,col=genotype)) + geom_point()
ggplot(g,aes(y=TMEM140,x=genotype,col=genotype)) + geom_point()
ggplot(g,aes(y=PLA1A,x=genotype,col=genotype)) + geom_point()
ggplot(g,aes(y=NEO1,x=genotype,col=genotype)) + geom_point()

```

### glmnet model on log2fc -- Pranitha old script
```{r glmnet_Pranitha}
require("VGAM")
library(tibble, lib.loc = "/usr/local/R-3.6.1/lib64/R/library")
grDevices::pdf.options(useDingbats = FALSE)
require(data.table)
require(foreign)
require(ggplot2)
require(MASS)
require(Hmisc)
library("nnet")
require(reshape2)
require(dplyr,lib.loc = "/usr/local/R-3.6.1/lib64/R/library")
require(caret)
require(tidyr)
library(XML)
theme_set(theme_bw())
source("/project/umw_garberlab/vangala/bin/custom_functions_fromDB.R")
options(stringsAsFactors = F)
load("/nl/umw_manuel_garber/mouse/DC/SIP/merged_SIP_H3K4me3/perElement_interactions2-100/perLocus_5k/perLocus/geneReg_models/regression/SeqGL/SeqGL_sip_1h_FC_atacDyn_5k_heldout.rda")

######## training and testing #########
train=k[k$held=="train",]
rownames(train)=train$gene
test=k[k$held=="test",]
rownames(test)=test$gene
x.indep=grep("Group",colnames(k),value = T)
#x.indep=grep("not|early|down",x.indep,value = T,invert = F)
tuneGrid=expand.grid(
  .alpha=seq(0,10, by = 0.5),
  .lambda=seq(0,10, by = 0.5))
trainControl = trainControl(method="repeatedcv", number=10,allowParallel = T,repeats=100)
############################## 1hour Elastic Net Regression #########################
y.dep="log2fc_1h"
#ridgeGrid= data.frame(.lambda =seq(0,1, length =5))
cluster=parallel::makeCluster(29) # convention to leave 1 core for OS
doParallel::registerDoParallel(cluster)
modelFit_early <- train((log2fc_1h) ~., data = train[,c(y.dep,x.indep)],
                        method = "glmnet",
                        #preProc=c("center","scale"),
                        trControl = trainControl,
                        metric = "R2", # Optimize by F-measure
                        #alpha=0,
                        tuneGrid=tuneGrid,
                        #tuneLength=20#,
                        #MaxNWts=5000)
)
parallel::stopCluster(cluster)
foreach::registerDoSEQ()
modelFit_early$bestTune
test$pred = predict(modelFit_early,test)
# Model performance metrics
caret::postResample(test$pred,test$log2fc_1h)
#k$logpred=log2(k$pred)
plot(test$log2fc_1h,test$pred)
#cor(k$logpred,k$log2fc_1h,method = "spearman")
rho=cor(test$pred,test$log2fc_1h)
rho
r2=caret::postResample(test$pred,test$log2fc_1h)[2]
r2
#minl=min(c(k$log2fc_1h,k$pred))
#maxl=max(c(k$log2fc_1h,k$pred))
k$pred_early = predict(modelFit_early,k)
k=k[order(k$Clusters,decreasing = T),]
k$held="train"
k$held[rownames(k)%in%testRow]="test"
ggplot(k,aes(log2fc_1h,pred_early)) +
  geom_point(position=position_jitter(w=1),aes(color=Clusters))+ #scale_x_log10()+scale_y_log10() + 
  geom_smooth(method = "lm", se = FALSE,color="black",size=0.4)+
  ggtitle(label = paste("corr is", format(rho, digits=4),"R2 is", format(r2, digits=2)))+
  scale_color_brewer(palette="Set1")
ggplot(test,aes(log2fc_1h,pred)) +
  geom_point(position=position_jitter(w=1),aes(color=Clusters))+ #scale_x_log10()+scale_y_log10() + 
  geom_smooth(method = "lm", se = FALSE,color="black",size=0.4)+
  ggtitle(label = paste("corr is", format(rho, digits=4),"R2 is", format(r2, digits=2)))+
  scale_color_brewer(palette="Set1")

```

### glmnet model on log2fc -- my take
```{r glmnet_Crystal}



```













