---
title: "skineQTL_model2025_test1"
author: "Crystal Shan"
date: "2025-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

notes:
F138 doesn't have genotype data?
# [don't run] archive code for getting ERAP2 expression for IFNG eQTL and IFNG reQTL.
```{r rankNorm fun, warning=FALSE}
# INT stands for rank-based inverse normal transformation
# function
rankNorm <- function (y) {
  # input y: numeric vector of CPM across all genes per donor.
  k <- 0.375 # an offset to ensure the z-score is finite. from Blom transform.
  n <- length(y)
  # Ranks.
  r <- rank(y, ties.method = "average") # if same value, same rank
  # Apply transformation.
  r.prob <- (r - k) / (n - 2 * k + 1)
  y.rankNorm <- stats::qnorm(r.prob) # because qnorm(1) is infinite and qnorm(-1) is -Inf
  return(y.rankNorm)
}
```

metadata$biosample = paste0(metadata$donor,"_",metadata$celltype, "_", metadata$condition)
df = data.frame(biosample=metadata$biosample, 
                donor=metadata$donor_num,
                counts=as.numeric(counts["ERAP2",]), 
                CPM=as.numeric(CPM["ERAP2",]),
                libsize=colSums(counts))
df.MELIFNG = df[metadata$celltype=="MEL"&metadata$condition=="IFNG",]
df.MELIFNG$rankNormCPM=rankNorm(df.MELIFNG$CPM)
metadata.MELIFNG = metadata[metadata$celltype=="MEL"&metadata$condition=="IFNG",]

df.MEL = df[metadata$celltype=="MEL"&(metadata$condition=="PBS"|metadata$condition=="IFNG"),]
df.MEL$rankNormCPM=rankNorm(df.MEL$CPM)
metadata.MEL = metadata[metadata$celltype=="MEL"&(metadata$condition=="PBS"|metadata$condition=="IFNG"),]

outDir="/pi/manuel.garber-umw/human/skin/eQTLs/DREG/2025models/test1/"
saveRDS(df.MELIFNG, paste0(outDir,"/ERAP_expression_MEL_IFNG.rds"))
saveRDS(metadata.MELIFNG, paste0(outDir,"/metadata_MEL_IFNG.rds"))
saveRDS(df.MEL, paste0(outDir,"/ERAP_expression_MEL_PBSandIFNG.rds"))
saveRDS(metadata.MEL, paste0(outDir,"/metadata_MEL_PBSandIFNG.rds"))
```

# load data
```{r load}
library(dplyr)
library(ggplot2)
#library(DESeq2)
#library(BiocParallel)
library(magrittr)
library(stringr)
#library(gplots)
#library(ggrepel)
#library(ggfortify)
#library(reactable)
#library(gridExtra)

load(here::here("notebooks/QTL_model_testing1.RData"))
# Dir="~/Downloads/nl/human/skin/eQTLs/DREG/2025models/test1/"
# df.MEL.IFNG <- readRDS(paste0(Dir,"/ERAP_expression_MEL_IFNG.rds"))
# df.MEL.IFNG$donor <- paste0("F", df.MEL.IFNG$donor)
# df.MEL.IFNG$loglibsize <- log(df.MEL.IFNG$libsize)
# metadata.MEL.IFNG <- readRDS(paste0(Dir,"/metadata_MEL_IFNG.rds"))
# df.MEL.PBSIFNG <- readRDS(paste0(Dir,"/ERAP_expression_MEL_PBSandIFNG.rds"))
# df.MEL.PBSIFNG$donor <- paste0("F", df.MEL.PBSIFNG$donor)
# df.MEL.PBSIFNG$loglibsize <- log(df.MEL.PBSIFNG$libsize)
# df.MEL.PBSIFNG <- df.MEL.PBSIFNG[rownames(df.MEL.PBSIFNG)!="F34_MEL_PBS_S1", ]
# df.MEL.PBSIFNG$condition <- ordered(gsub("F.*_MEL_","", df.MEL.PBSIFNG$biosample),
#                                     levels = c("PBS","IFNG"))
# metadata.MEL.PBSIFNG <- readRDS(paste0(Dir,"/metadata_MEL_PBSandIFNG.rds"))
# metadata.MEL.PBSIFNG <- metadata.MEL.PBSIFNG[metadata.MEL.PBSIFNG$sample!="F34_MEL_PBS_S1", ]
# 
# snpinfo <- data.table::fread(paste0(Dir,"/snp.bed")) %>% t() %>% 
#   as.data.frame() %>% 
#   rownames_to_column("donor") %>%
#   set_colnames(c("donor","GT"))
# snpinfo$donor <- gsub("skineQTL-","",snpinfo$donor)
# REF <- "T"
# ALT <- "C"
# 
# genotype <- snpinfo[7:nrow(snpinfo), ]
# genotype <- genotype %>%
#   mutate(
#     genotype.num = case_when(
#       GT == "0|0" ~ 0,
#       GT %in% c("0|1", "1|0") ~ 1,
#       GT == "1|1" ~ 2,
#       TRUE ~ NA_real_
#     ),
#     genotype.nt = case_when(
#       GT == "0|0" ~ "TT",
#       GT == "0|1" ~ "TC|CT",
#       GT == "1|0" ~ "TC|CT",
#       GT == "1|1" ~ "CC",
#       TRUE ~ NA_character_
#     )
#   )
# genotype$donor <- sub("^F0", "F", genotype$donor)
```
# plotting
```{r plotting}
this.snp <- "rs2910686"
snp.ref <- "T"
snp.alt <- "C"
this.gene <- "ERAP2"
celltype <- "MEL"
this_reQTL_matrix <- left_join(df.MEL.PBSIFNG, genotype, by="donor") %>% na.omit()
df <- this_reQTL_matrix
df$condition <- gsub("F.*_","",df$biosample)
df$condition <- ordered(df$condition, levels=c("PBS","IFNG"))
df$genotype.nt <- ordered(df$genotype.nt, 
                       levels <- c(paste0(snp.ref,snp.ref),
                                   paste0(snp.ref,snp.alt,"|",snp.alt,snp.ref),
                                   paste0(snp.alt,snp.alt)))

donor_counts <- df %>%
  distinct(genotype.nt, donor) %>%
  group_by(genotype.nt) %>%
  summarise(n = n(), .groups = "drop")

facet_labels <- setNames(
  paste0(as.character(donor_counts$genotype.nt), " (n=", donor_counts$n, ")"),
  donor_counts$genotype.nt
)


# make paired plots
position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
line.color = "grey"; linetype = "solid"; palette="bright"
highlightdata1 <- df %>% filter(donor=="F25")
highlightdata2 <- df %>% filter(donor=="F49")
highlightdata3 <- df %>% filter(donor=="F55")

p <- ggplot(df, create_aes(list(x = "condition", y = "CPM"))) +
  geom_exec(geom_line, data = df, group = "donor",
            color = line.color, size = line.size, linetype = linetype,
            position = position) +
  geom_exec(geom_point, data = df, color = "condition", size = point.size,
            position = position) +
  scale_color_manual(values=c("#999999", "#E69F00")) +
  geom_exec(geom_line, data = highlightdata1, group = "donor",
            color = "blue", size = 0.7, linetype = linetype,
            position = position) +
  geom_exec(geom_line, data = highlightdata2, group = "donor",
            color = "brown", size = 0.7, linetype = linetype,
            position = position) +
  geom_exec(geom_line, data = highlightdata3, group = "donor",
            color = "black", size = 0.7, linetype = linetype,
            position = position) +
  ggtitle(paste0(celltype,"\n",this.snp,":",this.gene)) +
  theme(legend.position="none") +
  facet_wrap(~ genotype.nt, labeller = labeller(genotype.nt = facet_labels))


```
# model for IFNG eQTL
```{r model-eQTL}
library(MASS)
this_matrix <- left_join(df.MEL.IFNG, genotype, by="donor") %>% na.omit

# Model 1: Raw counts + Poisson + Fixed effect
model1 <- glm(counts ~ genotype.num, data = this_matrix, family = "poisson", offset = loglibsize)

# Model 2: Raw counts + Negative Binomial + Fixed effect
# Why use: More flexible than Poisson; handles overdispersion better (variance > mean).
# > overdispersion_ratio <- sum(residuals(model1, type="pearson")^2) / df.residual(model1)
# > print(overdispersion_ratio)
# [1] 35.30822 --> highly overdispersed. NB >> Poisson
model2 <- glm.nb(counts ~ genotype.num + offset(loglibsize), data = this_matrix)

# Model 3: Rank-normalized CPM + Linear + Fixed effect
# Why use: No need to model count distributions; robust to outliers. Good for visualization too.
model3 <- lm(rankNormCPM ~ genotype.num, data = this_matrix)

### ----- is it really okay to have number of levels of each grouping factor == number of observations? model4 and 5 could work but model6 threw an error. 
# Model 4: Raw counts + Poisson + Mixed effects (random intercept by donor)
# Why use: Controls for donor-specific baseline expression differences.
library(lme4)
model4 <- glmer(counts ~ genotype.num + (1 | donor), family = poisson, data = this_matrix, offset = loglibsize)

# Model 5: Raw counts + Negative Binomial + Mixed effects (random intercept)
# Why use: Like Model 4, but handles overdispersion.
library(glmmTMB)
model5 <- glmmTMB(counts ~ genotype.num + (1 | donor), 
                  family = nbinom2, data = this_matrix, offset = loglibsize)

### after this point, the models don't work anymore. 
# Model 6: Rank-normalized CPM + Linear + Mixed (random intercept)
# Why use: Allows for subject-specific baselines without worrying about raw count assumptions.
# If every donor appears only once (or the counts equal one another), lme4 can’t estimate a donor‑level variance component
this_matrix$donor <- as.factor(this_matrix$donor)
model6 <- lmer(rankNormCPM ~ genotype.num + (1 | donor), data = this_matrix)
summary(model6)

# Model 7: Raw counts + Poisson + Mixed (random intercept & slope for donor)
# Why use: Donor response to treatment may vary — captures this heterogeneity.
model7 <- glmer(counts ~ genotype.num + IFNg_treatment + genotype:IFNg_treatment + 
                  (1 + IFNg_treatment | donor), 
                family = poisson, data = df, offset = log_libsize)
summary(model7)

# Model 8: Raw counts + Negative Binomial + Mixed (RI + RS)
# Why use: Best model for real-life complexity: donor-specific intercepts and slopes + count overdispersion.
model8 <- glmmTMB(counts ~ genotype + IFNg_treatment + genotype:IFNg_treatment + 
                    (1 + IFNg_treatment | donor), 
                  family = nbinom2, data = df, offset = log_libsize)
summary(model8)

# Model 9: Rank-normalized CPM + Linear + Mixed (RI + RS)
# Why use: Simplified model structure, very fast, good for initial screening.
df9 <- data.frame(expr = rank_norm_expr, genotype = genotype_vector, 
                  donor = donor_vector, IFNg_treatment = treatment_vector)

model9 <- lmer(expr ~ genotype + IFNg_treatment + genotype:IFNg_treatment + 
                 (1 + IFNg_treatment | donor), data = df9)
summary(model9)

```
# model for reQTL
```{r model-reQTL}
this_matrix <- left_join(df.MEL.PBSIFNG, genotype, by="donor") %>% na.omit

# Model 1: beta comparison method: Raw counts + Poisson + Fixed effect
# model1.p.pnorm <- 0.3244923
library(MASS)
model1.pbs <- glm(counts ~ genotype.num, data = this_matrix[this_matrix$condition=="PBS",], family = "poisson", offset = loglibsize)
model1.ifng <- glm(counts ~ genotype.num, data = this_matrix[this_matrix$condition=="IFNG",], family = "poisson", offset = loglibsize)
model1.pbs.b <- summary(model1.pbs)$coefficients["genotype.num","Estimate"]
model1.pbs.se <- summary(model1.pbs)$coefficients["genotype.num", "Std. Error"]
model1.ifng.b <- summary(model1.ifng)$coefficients["genotype.num","Estimate"]
model1.ifng.se <- summary(model1.ifng)$coefficients["genotype.num", "Std. Error"]
beta_comparison <- (model1.pbs.b - model1.ifng.b) / (sqrt((model1.pbs.se^2) + (model1.ifng.se^2)))
model1.p.pnorm <- 2 * pmin(pnorm(beta_comparison), 1 - pnorm(beta_comparison))

# Model 2: Raw counts + Negative Binomial + Fixed effect
# Why use: More flexible than Poisson; handles overdispersion better (variance > mean).
# model2.p.pnorm = 0.6081773
model2 <- glm.nb(counts ~ genotype.num + offset(loglibsize), data = this_matrix)
model2.pbs <- glm.nb(counts ~ genotype.num + offset(loglibsize), data = this_matrix[this_matrix$condition=="PBS",])
model2.ifng <- glm.nb(counts ~ genotype.num + offset(loglibsize), data = this_matrix[this_matrix$condition=="IFNG",])
model2.pbs.b <- summary(model2.pbs)$coefficients["genotype.num","Estimate"]
model2.pbs.se <- summary(model2.pbs)$coefficients["genotype.num", "Std. Error"]
model2.ifng.b <- summary(model2.ifng)$coefficients["genotype.num","Estimate"]
model2.ifng.se <- summary(model2.ifng)$coefficients["genotype.num", "Std. Error"]
beta_comparison <- (model2.pbs.b - model2.ifng.b) / (sqrt((model2.pbs.se^2) + (model2.ifng.se^2)))
model2.p.pnorm <- 2 * pmin(pnorm(beta_comparison), 1 - pnorm(beta_comparison))

# Model 3: Rank-normalized CPM + Linear + Fixed effect
# Why use: No need to model count distributions; robust to outliers. Good for visualization too.
# model3.p.pnorm = 0.243769
model3 <- lm(rankNormCPM ~ genotype.num, data = this_matrix)
model3.pbs <- lm(rankNormCPM ~ genotype.num, data = this_matrix[this_matrix$condition=="PBS",])
model3.ifng <- lm(rankNormCPM ~ genotype.num, data = this_matrix[this_matrix$condition=="IFNG",])
model3.pbs.b <- summary(model3.pbs)$coefficients["genotype.num","Estimate"]
model3.pbs.se <- summary(model3.pbs)$coefficients["genotype.num", "Std. Error"]
model3.ifng.b <- summary(model3.ifng)$coefficients["genotype.num","Estimate"]
model3.ifng.se <- summary(model3.ifng)$coefficients["genotype.num", "Std. Error"]
beta_comparison <- (model3.pbs.b - model3.ifng.b) / (sqrt((model3.pbs.se^2) + (model3.ifng.se^2)))
model3.p.pnorm <- 2 * pmin(pnorm(beta_comparison), 1 - pnorm(beta_comparison))

# Model 4: Raw counts + Poisson + Mixed effects (random intercept by donor)
# Why use: Controls for donor-specific baseline expression differences.
library(lme4)
model4 <- glmer(counts ~ genotype.num + condition + genotype.num:condition + (1 | donor), 
                     family = poisson, data = this_matrix, offset = loglibsize)

# Model 5: Raw counts + Negative Binomial + Mixed effects (random intercept)
# Why use: Like Model 4, but handles overdispersion.
library(glmmTMB)
model5 <- glmmTMB(counts ~ genotype.num + condition + genotype.num:condition + (1 | donor), 
                  family = nbinom2, data = this_matrix, offset = loglibsize)

# Model 6: Rank-normalized CPM + Linear + Mixed (random intercept)
# Why use: Allows for subject-specific baselines without worrying about raw count assumptions.
# If every donor appears only once (or the counts equal one another), lme4 can’t estimate a donor‑level variance component
library(lmerTest)
this_matrix$donor <- as.factor(this_matrix$donor)
model6 <- lmer(rankNormCPM ~ genotype.num + condition + genotype.num:condition + (1 | donor), data = this_matrix)


# Model 7: Raw counts + Poisson + Mixed (random intercept & slope for donor)
# Why use: Donor response to treatment may vary — captures this heterogeneity.
model7 <- glmer(counts ~ genotype.num + condition + genotype.num:condition + 
                  (1 + condition | donor),
                family = poisson, data = this_matrix, offset = loglibsize)

# Model 8: Raw counts + Negative Binomial + Mixed (RI + RS)
# Why use: Best model for real-life complexity: donor-specific intercepts and slopes + count overdispersion.
model8 <- glmmTMB(counts ~ genotype.num + condition + genotype.num:condition + 
                  (1 + condition | donor), 
                  family = nbinom2, data = this_matrix, offset = loglibsize)

# Model 9: Rank-normalized CPM + Linear + Mixed (RI + RS)
# Why use: Simplified model structure, very fast, good for initial screening.
model9 <- lmer(rankNormCPM ~ genotype.num + condition + genotype.num:condition + 
                 (1 + condition | donor), data = this_matrix)
summary(model9)

# Model 10: Kruskal-Wallis rank sum test
# p-value < 2.2e-16
this_matrix <- this_matrix %>%
  group_by(donor) %>%
  mutate(delta_CPM = CPM[condition == "IFNG"] - CPM[condition == "PBS"]) %>%
  ungroup()

kruskal.test(delta_CPM ~ genotype.nt, data = this_matrix)

# Model 11: beta regression on scaled delta CPM
library(betareg)
# Wide format for delta computation
df_wide <- this_matrix %>%
  dplyr::select(donor, genotype.num, genotype.nt, condition, CPM) %>%
  tidyr::pivot_wider(names_from = condition, values_from = CPM) %>%
  mutate(
    delta = IFNG - PBS,
    rel_response = (IFNG - PBS) / (PBS + 1),  # % increase over PBS baseline
    rel_response_scaled = (rel_response - min(rel_response) + 0.001) /
                          (max(rel_response - min(rel_response)) + 0.002)  # map to (0,1)
  )
model_beta <- betareg(rel_response_scaled ~ genotype.num, data = df_wide)
summary(model_beta)
ggplot(df_wide, aes(x = as.factor(genotype.num), y = rel_response_scaled)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, height = 0, color = "orange", alpha = 0.6) +
  labs(x = "Genotype (0=TT, 1=TC/CT, 2=CC)", 
       y = "Scaled IFNG response / PBS", 
       title = "Beta regression target: Relative response to IFNG")

# Model 12: beta regression on raw delta CPM
summary(lm(delta ~ genotype.num, data = df_wide))
ggplot(df_wide, aes(x = as.factor(genotype.num), y = rel_response)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, height = 0, color = "orange", alpha = 0.6) +
  labs(x = "Genotype (0=TT, 1=TC/CT, 2=CC)", 
       y = "IFNG response / PBS", 
       title = "Beta regression target: Relative response to IFNG")
```