---
title: "skin-eQTL_celseq2_analysis"
author: "Shuo Shan"
date: "11/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## main data processing
### Load packages  
```{r package, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
library(gridExtra)
library(ggpubr)
library(plotly)
library(RCurl)
library(ComplexHeatmap)

Dir="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis"
setwd(Dir)

#save.image(file=paste0(Dir,'/myEnvironment_main.RData'))
load(paste0(Dir,'/myEnvironment_main.RData'))
```
### Overview

Celseq2 data for fibroblasts (FRB) from 37 donors X 2 culture conditions.(F47 is removed due to failed genotyping). Libraries were pooled and prepared by celseq2 protocol, pair-end sequenced by BGI DNBseq or Illumina NextSeq500/NextSeq550/NextSeq2000 in house. Barcode in read 1 name was used to separate out pooled samples within a library, and read 2 was used for STAR alignment to hg38 assembly.  ESAT was used to quantify raw read counts on the gene level with 1000bp extension from the 3'UTR.
  
The files generated from this markdown are:
1) raw ESAT counts of selected samples. 
  - raw_counts.txt
  - raw_counts.html
2) TMM-normalized CPM table. 
  - TMM-normalized-CPM.txt
3) log2 CPM table
  - log2-TMM-normalized-CPM.txt
3) DESeq2 results of 21,308 genes (filtered by rowsun>10), including log2FC and padj (FDR). 
  - upregulated_genes.txt (padj<0.05, log2FC>1.5)
  - downregulated_genes.txt (padj<0.05, log2FC<-1.5)
  - DESeq2_results_IFNvsPBS.txt
  - DESeq2_results_IFNvsPBS.html
4) interactive heatmap of top 200 genes with most dramatic FC
  - heatmaply_log2TMM-CPM_top200.html

```{r functions}
create_table <- function(Dir, file_lst) {
  f=file_lst[1]
  sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
  df <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
    tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
    mutate(sum=rowSums(select_if(., is.numeric))) %>%
    dplyr::select(c("region",sum)) %>% 
    magrittr::set_colnames(c("region",sample))
  
  if (length(file_lst) > 1) {
    for (f in file_lst[2:length(file_lst)]) {
      sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
        tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
        mutate(sum=rowSums(select_if(., is.numeric))) %>%
        dplyr::select(c("region",sum)) %>% 
        magrittr::set_colnames(c("region",sample))
    
    df <- dplyr::left_join(df,df.this,by="region")
    }
  }
  df <- df %>% tibble::column_to_rownames( . , "region")
  return(df)
  rm(f,sample,df,df.this)
}

# check which variable correlates with each PC the most.
best_correlation_with_pc <- function(pcs, metadata, pc.n) {
  total.pcc=c()
  pca.this=pcs[,pc.n]
  for (x in colnames(metadata)) {
    metadata.this <- as.numeric(as.factor(metadata[,x]))
    total.pcc=c(total.pcc,cor(pca.this,metadata.this))
  }
  total.df=data.frame(feature=colnames(metadata),pcc=total.pcc,pcc.magnitude=abs(total.pcc))
  return(total.df)
}
```
  
### Load raw ESAT counts
```{r, echo=FALSE, message=FALSE}
Dir.esat="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/DolphinNext/report7561/ESAT"

# 1. import ESAT counts into master table
flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
  tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
samples.all <- data.frame(sample=colnames(counts))

# 2. find list of samples to keep
# 2a. condition used from pilot (KRT, FRB: 100ng 24h, MEL: 10ng 48h)
samples.pilot.keep <- c("F25_FRB_PBS_0ng_S0", "F25_FRB_24h_100ng_S0",
                        "F25_MEL_PBS_0ng_S0", "F25_MEL_48h_10ng_S0",
                        "F25_KRT_PBS_0ng_S0", "F25_KRT_24h_100ng_S0")
# 2b. latest version of samples (merged from replicates if any)
samples <- samples.all %>%
  tidyr::separate( . , col=sample, into=c("donor","celltype","condition","replicate"), sep="_",remove=FALSE) %>%
  dplyr::filter(donor!="F25") %>% tidyr::unite(. , "biosample", c("donor","celltype","condition"),sep="_", remove=FALSE)
# go through the list, and for each donor, pick the row that contain the highest number of replicate
samples.latest <- c()
for (d in unique(samples$biosample)) {
  df.this <- samples %>% dplyr::filter(biosample==d) %>% arrange( . , desc(replicate))
  sample.pick <- df.this[1,1] %>% as.character()
  samples.latest <- c(samples.latest, sample.pick)
  rm(df.this,sample.pick)
}
# 2c. only keratinocytes.
samples.keep <- c(samples.latest, samples.pilot.keep)
samples.keep <- samples.keep[grep("FRB",samples.keep)]

# 2d. remove samples with indicated keywords (after QC inspection. will update until more sequence reads come in)
keyword1 <- "F47" # remove due to failed genotyping.
samples.keep <- samples.keep[-grep(keyword1,samples.keep)]

# 2c combine
counts.keep <- counts %>% dplyr::select(samples.keep)
colnames(counts.keep) <- colnames(counts.keep) %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)

# remove MT genes
MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
ribogenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/ribosome_genes_KEGG-hsa03010.txt")
counts.keep <- counts.keep %>% mutate(gene=rownames(.)) %>% dplyr::filter(!rownames(.) %in% MTgenes$V1) %>%
  dplyr::filter(!gene %in% ribogenes$V1) %>% dplyr::select(-"gene")
```

### Load metadata
```{r metadata, echo=FALSE, message=FALSE}
# parse kept samples into table
samples <- data.frame(sample=colnames(counts.keep))
# import metadata master table
metadata.master=read.table("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/metadata_08182022.txt", sep="\t", header=TRUE)
# import DolphinNext pipeline overall summary
pip.summary=read.table("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/DolphinNext/report7561/summary/overall_summary.tsv", sep="\t", header=TRUE) %>% magrittr::set_colnames(c("sample","total_reads","Multimapped_Reads_Aligned_STAR","Unique_Reads_Aligned_STAR"))
pip.summary$sample <- pip.summary$sample %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)
# import ancestry table
ancestry=read.table("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/ancestry.txt", sep="\t", header=FALSE) %>% set_colnames(c("donor","ancestry"))

# query imported tables with samples
rm(metadata)
metadata <- left_join(samples,metadata.master,by="sample")
metadata <- left_join(metadata,pip.summary,by="sample")
metadata <- left_join(metadata,ancestry,by="donor")

# compile 
metadata$seq_date <- metadata$seq_date %>% gsub("\\/",".",.)
metadata$lib_prep_date <- metadata$lib_prep_date %>% gsub("\\/",".",.)
metadata$batch <- paste(metadata$lib_prep_date,metadata$seq_date,sep="_") %>% as.factor()

# 
metadata.pbs <- metadata[which(metadata$condition=="PBS"),]
metadata.ifn <- metadata[which(metadata$condition=="IFN"),]
# cleanup
rm(pip.summary, samples, samples.all)
```
  
  
### Overview of libraries  
Overview of library size distribution.
```{r lib-size, echo=F, message=F}
# library size distribution
df=data.frame(sample=metadata$sample,total_count=metadata$total_reads)
df=df[order(-df$total_count),]
plot_ly(y = df$sample,x = df$total_count,type = "bar")

# total reads distribution
df=data.frame(sample=metadata$sample,total_count=metadata$total_reads)
df=df[order(-df$total_count),] %>% mutate(celltype="FRB")
write.table(df, file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/library_size_distribution_04202023.txt",append=TRUE,
            quote=FALSE,sep="\t",row.names=F, col.names=T)

# histogram for gene counts, all samples
h <- hist(log10(rowSums(counts.keep)), breaks = as.numeric(100), plot = FALSE) 
plot_ly(x = h$mids, y = h$counts, width = 500, height=300) %>% layout( title = "Histogram") %>% add_bars()

# which libraries have low reads? (<5M)
df$sample[which(df$total_count<5000000)] %>% gsub("_MEL.*","",.) %>% table()
```

### Batch-correction  
```{r batchcorrect}
library("sva")
#load in the uncorrected data as raw counts
uncorrected_data = counts.keep
#calculate principal components for the uncorrected data
pca_uncorrected_obj = prcomp(uncorrected_data)
#pull PCA values out of the PCA object
pca_uncorrected = as.data.frame(pca_uncorrected_obj[2]$rotation)
#assign labels to the data frame
pca_uncorrected[,"condition"] = metadata$condition
pca_uncorrected[,"batch"] = metadata$batch
pca_uncorrected[,"seq_model"] = metadata$seq_model

#plot the PCA
#create a classic 2-dimension PCA plot (first two principal components) with conditions and library methods indicated
p1 = ggplot(data=pca_uncorrected, aes(x=PC1, y=PC2, color=batch, shape=condition))
p1 = p1 + geom_point(size=3)
p1 = p1 + labs(title="PCA, RNA-seq counts for 37x2 samples (uncorrected data)", color="batch", shape="condition")


#perform the batch correction

#first we need to transform the format of our groups and batches from names (e.g. "UHR", "HBR", etc.) to numbers (e.g. 1, 2, etc.)
#in the command below "sapply" is used to apply the "switch" command to each element and convert names to numbers as we define
groups = sapply(as.character(metadata$condition), switch, "PBS" = 1, "IFN" = 2, USE.NAMES = F)
lib_prep_dates = as.numeric(as.factor(metadata$lib_prep_date))
seq_dates = as.numeric(as.factor(metadata$seq_date))
ancestry = as.numeric(as.factor(metadata$ancestry))

#now run ComBat_seq
corrected_data = ComBat_seq(counts = as.matrix(uncorrected_data), batch = lib_prep_dates + seq_dates, group = groups)

#calculate principal components for the uncorrected data
pca_corrected_obj = prcomp(corrected_data)

#pull PCA values out of the PCA object
pca_corrected = as.data.frame(pca_corrected_obj[2]$rotation)
#assign labels to the data frame
pca_corrected[,"condition"] = metadata$condition
pca_corrected[,"batch"] = metadata$batch
pca_corrected[,"seq_model"] = metadata$seq_model
pca_corrected[,"ancestry"] = metadata$ancestry

#plot the PCA
#create a classic 2-dimension PCA plot (first two principal components) with conditions and library methods indicated
p2 = ggplot(data=pca_corrected, aes(x=PC1, y=PC2, color=batch, shape=condition))
p2 = p2 + geom_point(size=3)
p2 = p2 + labs(title="PCA, RNA-seq counts for 37x2 samples (corrected data)", color="batch", shape="condition")

# save the plots
pdf(file="Uncorrected-vs-BatchCorrected-PCA.pdf")
grid.arrange(p1, p2, nrow = 2)
dev.off()

# rename the variable
counts.keep.corrected=corrected_data
rm(corrected_data)
```
### CPM calculation using batch corrected counts + keep expressed genes
```{r cpm}
col.sum=apply(counts.keep.corrected,2,sum)
CPM=sweep(counts.keep.corrected, 2, col.sum, `/`)*1000000

# Getting CPM matrix
CPM.pbs <- as.data.frame(CPM)[,as.character(metadata.pbs$sample)] %>%
  magrittr::set_colnames(as.character(metadata.pbs$donor)) %>%
  magrittr::set_rownames(rownames(CPM))
CPM.ifn <- as.data.frame(CPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) %>%
  magrittr::set_rownames(rownames(CPM))
```
```{r expressed_genes}
# keep a gene if it's expressed in either PBS or IFN condition. expressed = CPM >= 10 in at least 5 donors in each condition.
gene_expression_status <- c()
for (i in 1:nrow(CPM)) {
  df.pbs <- CPM.pbs[i,]
  df.ifn <- CPM.ifn[i,]
  gene_expression_status_pbs <- (length(df.pbs[which(df.pbs >= 10)]) >= 5)
  gene_expression_status_ifn <- (length(df.ifn[which(df.ifn >= 10)]) >= 5)
  gene_expression_status <- c(gene_expression_status, (gene_expression_status_pbs | gene_expression_status_ifn))
}
expressed_genes <- rownames(CPM)[gene_expression_status]
# filter by expressed genes only.
CPM.expressed = CPM[expressed_genes,]
CPM.pbs.expressed = CPM.pbs[expressed_genes,]
CPM.ifn.expressed = CPM.ifn[expressed_genes,]
```
```{r logCPM_and_log2FC}
# Getting log transformed CPM
logCPM <- log2(CPM+1) %>% as.data.frame() 
logCPM.pbs <- as.data.frame(logCPM)[,as.character(metadata.pbs$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.pbs$donor)) 
logCPM.ifn <- as.data.frame(logCPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) 
log2FC <- logCPM.ifn - logCPM.pbs

# inspect distribution of log-transformed CPM:
plot(density(as.numeric(CPM.pbs[1,])))
plot(density(as.numeric(logCPM.pbs[1,])))
```
### DE analysis  
The goal of Differential gene expression analysis is to find 
genes or transcripts whose difference in expression, when accounting 
for the variance within condition, is higher than expected by chance. 

In this case a total sum of 10 counts separates well expressed 
from non-expressed genes. You can change this value and padj value and 
log2FoldChange cutoffs according to your data

```{r DESeq2}
df <- apply(counts.keep.corrected[rowSums(counts.keep.corrected)>=10,],2,round,0) %>% as.data.frame()

dds <- DESeqDataSetFromMatrix(countData = df,
                              colData = metadata,
                              design = ~ condition)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("condition", "IFN", "PBS")))

res <- results(dds,contrast=c("condition", "IFN", "PBS")) %>% as.data.frame %>% rownames_to_column("gene")
res.up <- res %>% filter(log2FoldChange > 1) %>% filter(padj < 0.05)
res.down <- res %>% filter(log2FoldChange < -1) %>% filter(padj < 0.05)

# design = ~ condition: 958 res.up, 565 res.down
```

### Write to file
```{r output}
# write raw ESAT counts and batch-corrected counts
write.table(counts.keep, file=paste0(Dir,"/raw_counts.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(counts.keep.corrected, file=paste0(Dir,"/counts_batch_corrected.txt"), quote=F, sep="\t", row.names=T, col.names=T)


# write CPM 
write.table(CPM, file=paste0(Dir,"/CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(CPM.pbs, file=paste0(Dir,"/CPM_PBS.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(CPM.ifn, file=paste0(Dir,"/CPM_IFN.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(CPM.pbs.expressed, file=paste0(Dir,"/CPM_expressedGenes_PBS.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(CPM.ifn.expressed, file=paste0(Dir,"/CPM_expressedGenes_IFN.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# write metadata
write.table(metadata, file=paste0(Dir,"/metadata.txt"),quote=F, sep="\t", row.names=T, col.names=T)
write.table(metadata.pbs, file=paste0(Dir,"/metadata_PBS.txt"),quote=F, sep="\t", row.names=T, col.names=T)
write.table(metadata.ifn, file=paste0(Dir,"/metadata_IFN.txt"),quote=F, sep="\t", row.names=T, col.names=T)

# write DESeq2 results
write.table(res,file=paste0(Dir,"/DESeq2_results_IFNvsPBS.txt"),quote=F,row.names=T,col.names=T)
write.table(res.up$gene,file=paste0(Dir,"/upregulated_genes_p0.05_log2FC1.txt"),quote=F,row.names=F,col.names=F)
write.table(res.down$gene,file=paste0(Dir,"/downregulated_genes_p0.05_log2FC-1.txt"),quote=F,row.names=F,col.names=F)
write.table(res$gene, file=paste0(Dir,"/genes_used_in_DE.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write expressed genes
write.table(expressed_genes, file = paste0(Dir,"/expressedGenes.txt"), quote = FALSE, row.names = FALSE, col.names=FALSE)
```

















## stage one
### Load packages  
```{r package, warning=FALSE}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(sva)
#library(DEXSeq)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
#library(edgeR)
library(gridExtra)
library(ggpubr)
library(plotly)
library(RCurl)
library(ComplexHeatmap)

Dir="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis"
setwd(Dir)

#save.image(file=paste0(Dir,'/myEnvironment.RData'))
load(paste0(Dir,'/myEnvironment.RData'))
```
### Overview

Celseq2 data for fibroblasts (FRB) from 37 donors X 2 culture conditions.(F47 is removed due to failed genotyping). Libraries were pooled and prepared by celseq2 protocol, pair-end sequenced by BGI DNBseq or Illumina NextSeq500/NextSeq550/NextSeq2000 in house. Barcode in read 1 name was used to separate out pooled samples within a library, and read 2 was used for STAR alignment to hg38 assembly.  ESAT was used to quantify raw read counts on the gene level with 1000bp extension from the 3'UTR.
  
The files generated from this markdown are:
1) raw ESAT counts of selected samples. 
  - raw_counts.txt
  - raw_counts.html
2) TMM-normalized CPM table. 
  - TMM-normalized-CPM.txt
3) log2 CPM table
  - log2-TMM-normalized-CPM.txt
3) DESeq2 results of 21,308 genes (filtered by rowsun>10), including log2FC and padj (FDR). 
  - upregulated_genes.txt (padj<0.05, log2FC>1.5)
  - downregulated_genes.txt (padj<0.05, log2FC<-1.5)
  - DESeq2_results_IFNvsPBS.txt
  - DESeq2_results_IFNvsPBS.html
4) interactive heatmap of top 200 genes with most dramatic FC
  - heatmaply_log2TMM-CPM_top200.html

```{r functions}
create_table <- function(Dir, file_lst) {
  f=file_lst[1]
  sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
  df <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
    tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
    mutate(sum=rowSums(select_if(., is.numeric))) %>%
    dplyr::select(c("region",sum)) %>% 
    magrittr::set_colnames(c("region",sample))
  
  if (length(file_lst) > 1) {
    for (f in file_lst[2:length(file_lst)]) {
      sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
        tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
        mutate(sum=rowSums(select_if(., is.numeric))) %>%
        dplyr::select(c("region",sum)) %>% 
        magrittr::set_colnames(c("region",sample))
    
    df <- dplyr::left_join(df,df.this,by="region")
    }
  }
  df <- df %>% tibble::column_to_rownames( . , "region")
  return(df)
  rm(f,sample,df,df.this)
}

# check which variable correlates with each PC the most.
best_correlation_with_pc <- function(pcs, metadata, pc.n) {
  total.pcc=c()
  pca.this=pcs[,pc.n]
  for (x in colnames(metadata)) {
    metadata.this <- as.numeric(as.factor(metadata[,x]))
    total.pcc=c(total.pcc,cor(pca.this,metadata.this))
  }
  total.df=data.frame(feature=colnames(metadata),pcc=total.pcc,pcc.magnitude=abs(total.pcc))
  return(total.df)
}
```
  
### Load raw ESAT counts
```{r, echo=FALSE, message=FALSE}
Dir.esat="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/DolphinNext/report7561/ESAT"

# 1. import ESAT counts into master table
flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
  tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
samples.all <- data.frame(sample=colnames(counts))

# 2. find list of samples to keep
# 2a. condition used from pilot (KRT, FRB: 100ng 24h, MEL: 10ng 48h)
samples.pilot.keep <- c("F25_FRB_PBS_0ng_S0", "F25_FRB_24h_100ng_S0",
                        "F25_MEL_PBS_0ng_S0", "F25_MEL_48h_10ng_S0",
                        "F25_KRT_PBS_0ng_S0", "F25_KRT_24h_100ng_S0")
# 2b. latest version of samples (merged from replicates if any)
samples <- samples.all %>%
  tidyr::separate( . , col=sample, into=c("donor","celltype","condition","replicate"), sep="_",remove=FALSE) %>%
  dplyr::filter(donor!="F25") %>% tidyr::unite(. , "biosample", c("donor","celltype","condition"),sep="_", remove=FALSE)
# go through the list, and for each donor, pick the row that contain the highest number of replicate
samples.latest <- c()
for (d in unique(samples$biosample)) {
  df.this <- samples %>% dplyr::filter(biosample==d) %>% arrange( . , desc(replicate))
  sample.pick <- df.this[1,1] %>% as.character()
  samples.latest <- c(samples.latest, sample.pick)
  rm(df.this,sample.pick)
}
# 2c. only keratinocytes.
samples.keep <- c(samples.latest, samples.pilot.keep)
samples.keep <- samples.keep[grep("FRB",samples.keep)]

# 2d. remove samples with indicated keywords (after QC inspection. will update until more sequence reads come in)
keyword1 <- "F47" # remove due to failed genotyping.
samples.keep <- samples.keep[-grep(keyword1,samples.keep)]

# 2c combine
counts.keep <- counts %>% dplyr::select(samples.keep)
colnames(counts.keep) <- colnames(counts.keep) %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)

# remove MT genes
MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
ribogenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/ribosome_genes_KEGG-hsa03010.txt")
counts.keep <- counts.keep %>% mutate(gene=rownames(.)) %>% dplyr::filter(!rownames(.) %in% MTgenes$V1) %>%
  dplyr::filter(!gene %in% ribogenes$V1) %>% dplyr::select(-"gene")
```

### Load metadata
```{r metadata, echo=FALSE, message=FALSE}
# parse kept samples into table
samples <- data.frame(sample=colnames(counts.keep))
# import metadata master table
metadata.master=read.table("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/metadata_08182022.txt", sep="\t", header=TRUE)
# import DolphinNext pipeline overall summary
pip.summary=read.table("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/DolphinNext/report7561/summary/overall_summary.tsv", sep="\t", header=TRUE) %>% magrittr::set_colnames(c("sample","total_reads","Multimapped_Reads_Aligned_STAR","Unique_Reads_Aligned_STAR"))
pip.summary$sample <- pip.summary$sample %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)
# import ancestry table
ancestry=read.table("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/ancestry.txt", sep="\t", header=FALSE) %>% set_colnames(c("donor","ancestry"))

# query imported tables with samples
rm(metadata)
metadata <- left_join(samples,metadata.master,by="sample")
metadata <- left_join(metadata,pip.summary,by="sample")
metadata <- left_join(metadata,ancestry,by="donor")

# compile 
metadata$seq_date <- metadata$seq_date %>% gsub("\\/",".",.)
metadata$lib_prep_date <- metadata$lib_prep_date %>% gsub("\\/",".",.)
metadata$batch <- paste(metadata$lib_prep_date,metadata$seq_date,sep="_") %>% as.factor()

# cleanup
rm(pip.summary, samples, samples.all)
```
  
  
### Overview of libraries  
Overview of library size distribution.
```{r lib-size, echo=F, message=F}
# library size distribution
df=data.frame(sample=metadata$sample,total_count=metadata$total_reads)
df=df[order(-df$total_count),]
plot_ly(y = df$sample,x = df$total_count,type = "bar")

# histogram for gene counts, all samples
h <- hist(log10(rowSums(counts.keep)), breaks = as.numeric(100), plot = FALSE) 
plot_ly(x = h$mids, y = h$counts, width = 500, height=300) %>% layout( title = "Histogram") %>% add_bars()

# which libraries have low reads? (<5M)
df$sample[which(df$total_count<5000000)] %>% gsub("_MEL.*","",.) %>% table()
```

### Batch-correction  
```{r batchcorrect}
library("sva")
#load in the uncorrected data as raw counts
uncorrected_data = counts.keep
#calculate principal components for the uncorrected data
pca_uncorrected_obj = prcomp(uncorrected_data)
#pull PCA values out of the PCA object
pca_uncorrected = as.data.frame(pca_uncorrected_obj[2]$rotation)
#assign labels to the data frame
pca_uncorrected[,"condition"] = metadata$condition
pca_uncorrected[,"batch"] = metadata$batch
pca_uncorrected[,"seq_model"] = metadata$seq_model

#plot the PCA
#create a classic 2-dimension PCA plot (first two principal components) with conditions and library methods indicated
p1 = ggplot(data=pca_uncorrected, aes(x=PC1, y=PC2, color=batch, shape=condition))
p1 = p1 + geom_point(size=3)
p1 = p1 + labs(title="PCA, RNA-seq counts for 37x2 samples (uncorrected data)", color="batch", shape="condition")


#perform the batch correction

#first we need to transform the format of our groups and batches from names (e.g. "UHR", "HBR", etc.) to numbers (e.g. 1, 2, etc.)
#in the command below "sapply" is used to apply the "switch" command to each element and convert names to numbers as we define
groups = sapply(as.character(metadata$condition), switch, "PBS" = 1, "IFN" = 2, USE.NAMES = F)
lib_prep_dates = as.numeric(as.factor(metadata$lib_prep_date))
seq_dates = as.numeric(as.factor(metadata$seq_date))
ancestry = as.numeric(as.factor(metadata$ancestry))

#now run ComBat_seq
corrected_data = ComBat_seq(counts = as.matrix(uncorrected_data), batch = lib_prep_dates + seq_dates + ancestry, group = groups)

#calculate principal components for the uncorrected data
pca_corrected_obj = prcomp(corrected_data)

#pull PCA values out of the PCA object
pca_corrected = as.data.frame(pca_corrected_obj[2]$rotation)
#assign labels to the data frame
pca_corrected[,"condition"] = metadata$condition
pca_corrected[,"batch"] = metadata$batch
pca_corrected[,"seq_model"] = metadata$seq_model
pca_corrected[,"ancestry"] = metadata$ancestry

#plot the PCA
#create a classic 2-dimension PCA plot (first two principal components) with conditions and library methods indicated
p2 = ggplot(data=pca_corrected, aes(x=PC1, y=PC2, color=batch, shape=condition))
p2 = p2 + geom_point(size=3)
p2 = p2 + labs(title="PCA, RNA-seq counts for 37x2 samples (corrected data)", color="batch", shape="condition")

# save the plots
pdf(file="Uncorrected-vs-BatchCorrected-PCA.pdf")
grid.arrange(p1, p2, nrow = 2)
dev.off()

# rename the variable
counts.keep.corrected=corrected_data
rm(corrected_data)
```
  
### TMM normalization  
```{r tmm, echo=FALSE, message=F}
dge <- DGEList(counts.keep)
dge <- calcNormFactors(dge)
dge$samples # check normalizaiton factors
CPM <- cpm(dge, log=FALSE)

metadata.pbs <- metadata %>% dplyr::filter(condition=="PBS")
metadata.ifn <- metadata %>% dplyr::filter(condition=="IFN")
# Getting normalized CPM matrix
CPM.pbs <- as.data.frame(CPM)[,as.character(metadata.pbs$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.pbs$donor)) 
CPM.ifn <- as.data.frame(CPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) 

# Getting log transformed CPM
logCPM <- log2(CPM+1) %>% as.data.frame() 

logCPM.pbs <- as.data.frame(logCPM)[,as.character(metadata.pbs$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.pbs$donor)) 

logCPM.ifn <- as.data.frame(logCPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) 

log2FC <- logCPM.ifn - logCPM.pbs

# inspect distribution of log-transformed CPM:
mean(apply(logCPM.pbs, 2, sd))
mean(apply(logCPM.pbs, 2, mean))
plot(density(logCPM.pbs[,1]))


# CPM for batch-corrected reads:
dge.corrected <- DGEList(counts.keep.corrected)
dge.corrected <- calcNormFactors(dge.corrected)
CPM.corrected <- cpm(dge.corrected, log=FALSE)
logCPM.corrected <- log2(CPM.corrected+1) %>% as.data.frame() 
```


### PCA
  
```{r pca-uncorrected}
# pick genes of top row variance
mat.logCPM.lnsd <- apply(logCPM,MARGIN=1,FUN=sd)
mat.logCPM.lncv <- sqrt( (exp(mat.logCPM.lnsd) )^2 -1) #coefficient of variance
logCPM.top <- logCPM[order(mat.logCPM.lncv,decreasing=T),][1:2000,] 

# PCA
pca = prcomp(t(logCPM.top),centered= T,scale = T)
loadings=as.data.frame(pca$rotation)
# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)
# PC1 is 43.6%, PC2 is 8%, PC3 is 6.7%. Top 2000 variable genes separate well by condition (PC1) in melanocytes, much more dramatic than fibroblasts

# plot
# top PC information
d=data.frame(pca$x,sample=rownames(pcs))
d=inner_join(d,metadata,by="sample") %>% magrittr::set_rownames(d$sample)
colnames(metadata)
#d$group_tags <- cut(d$Unique_Reads_Aligned_STAR, breaks=6)
d$group_tags <- cut(d$total_reads, 
                    breaks=c(min(d$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d$total_reads)))

d %>%
  mutate(platform=d$seq_model) %>%
  ggplot(., aes(y=PC1,x=PC2,col=ancestry,shape=condition,label=sample)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) +
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  ggsave(filename=paste0(Dir,"/PCA_4.png"))
```

```{r pca-corrected}
# pick genes of top row variance
mat.logCPM.lnsd.corrected <- apply(logCPM.corrected,MARGIN=1,FUN=sd)
mat.logCPM.lncv.corrected <- sqrt( (exp(mat.logCPM.lnsd.corrected) )^2 -1) #coefficient of variance
logCPM.top.corrected <- logCPM[order(mat.logCPM.lncv.corrected,decreasing=T),][1:2000,] 

# PCA
pca.corrected = prcomp(t(logCPM.top.corrected),centered= T,scale = T)
loadings.corrected=as.data.frame(pca.corrected$rotation)
# percent variance in each pc
plot(pca.corrected)
pcs.corrected = pca.corrected$x
pca.pctVar.corrected = pca.corrected$sdev^2/sum(pca.corrected$sdev^2)
# PC1 is 47.7%, PC2 is 7%, PC3 is 5.2%. Top 2000 variable genes separate well by condition (PC1) in melanocytes, much more dramatic than fibroblasts

# plot
# top PC information
d.corrected=data.frame(PC1=pca.corrected$x[,1],PC2=pca.corrected$x[,2], PC3=pca.corrected$x[,3], PC4=pca.corrected$x[,4], PC5=pca.corrected$x[,5],PC6=pca.corrected$x[,6],sample=rownames(pcs.corrected))
d.corrected=inner_join(d.corrected,metadata,by="sample") %>% magrittr::set_rownames(d.corrected$sample)
colnames(metadata)
d.corrected$group_tags <- cut(d.corrected$total_reads, 
                    breaks=c(min(d.corrected$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d.corrected$total_reads)))

d.corrected %>%
  mutate(platform=d.corrected$seq_model) %>%
  ggplot(., aes(y=PC1,x=PC2,col=group_tags,shape=condition,label=sample)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar.corrected[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar.corrected[2])) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  ggsave(filename=paste0(Dir,"/PCA.corrected_3.png"))
```
  
### DE analysis  
The goal of Differential gene expression analysis is to find 
genes or transcripts whose difference in expression, when accounting 
for the variance within condition, is higher than expected by chance. 

In this case a total sum of 10 counts separates well expressed 
from non-expressed genes. You can change this value and padj value and 
log2FoldChange cutoffs according to your data

```{r DESeq2}
df <- apply(counts.keep[rowSums(counts.keep)>=10,],2,round,0) %>% as.data.frame()

dds <- DESeqDataSetFromMatrix(countData = df,
                              colData = metadata,
                              design = ~ batch + condition)
dds <- DESeq(dds)
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
summary(results(dds,contrast=c("condition", "IFN", "PBS")))

res <- results(dds,contrast=c("condition", "IFN", "PBS")) %>% as.data.frame %>% rownames_to_column("gene")
res.up <- res %>% filter(log2FoldChange > 1.5) %>% filter(padj < 0.05)
res.down <- res %>% filter(log2FoldChange < -1.5) %>% filter(padj < 0.05)
res.nochange <- res %>% filter(abs(log2FoldChange) <= 1.5) %>% filter(padj < 0.05) %>% filter(!is.na(padj))
# design = ~ batch + condition: 565 res.up, 151 res.down

# next step: interactive heatmap and interactive MA plot
```

### Plot paired data in CPM

Harris lab Shany project: Olink proteomics data of vit patient blister fluids showed increase in IFNlambda1 protein. (not increased.) But scRNAseq data of the cells in blister fluid showed decrease in IFNlambda gene and receptor. the receptor decrease is almost unique to melanocytes. not as much of a change in other cell types.
Would IFNg stimulation increase IFN lambda receptor expression in melanocytes?
IFNlR1: IL-10R2 (or IL10Rbeta) form a deterodimer at the cell surface, recruits JAK2, JAK1 and TYK2. oerlap with IFNalpha receptor actiation. downstream, phosphorylate & actiates STAT1 and STAT2, NFkB, and enhancs interferon and antiiral protein production.
would IFNg stimulation increase expression of genes that belong to the interferon stimulated genes (ISG): e.g. IFIT1, IFIT2, MX1, OAS1.

```{r paired-beehive}
g="OAS1" # IFIT1, IFIT2, MX1, OAS1

# pick out TMM-normalized CPM paired data & run wilcox test
df.paired.this <- data.frame(PBS=as.data.frame(t(CPM.pbs[g,])),
                             IFN=as.data.frame(t(CPM.ifn[g,]))) %>% set_colnames(c("PBS","IFN"))
df.paired.this.test = wilcox.test(t(CPM.pbs[g,]),t(CPM.ifn[g,]))

# DESeq2 analysis result
df.DEseq2.this <- res %>% dplyr::filter(gene==g)

# Plot
ggpaired(df.paired.this, cond1="PBS", cond2="IFN", 
         color="condition",palette=c("grey","red"),
         title=paste0("TMM-normalized CPM of ",g,"\n",
                      "wilcox test: p=",format(df.paired.this.test$p.value, digits=3,scientific=T),"\n",
                      "DE analysis: log2FC=", round(df.DEseq2.this$log2FoldChange,3)," ; p.adj=", 
                      format(df.DEseq2.this$padj,digits=3,scientific=T),"\n",
                      "donors=",as.character(nrow(df.paired.this))),
         xlab="condition",ylab="TMM-normalized CPM")


# IFIT1: up regulated in MEL. log2FC=3.38, p.adj=1.1E-17
# IFIT2: up-regulated in MEL. log2FC=5.15, p.adj=1.1E-149
# MX1: up regulated in MEL. log2FC=3.82, p.adj=1.88E-41
# OAS1: up regulated in MEL. log2FC=3.58, p.adj=1.97E-38

rm(g,df.paired.this,df.paired.this.test,df,df.DEseq2.this,df.this)
```
  
  
### K-means-clustering & hierarchical clustering
```{r kmclusters}
df <- logCPM.corrected
n=20 # top number of genes with most abs(log2FC)

# # set of candidate genes for clustering
# candidate_genes <- res %>% as.data.frame %>%
#   dplyr::filter(padj < 0.05) %>%                   # restrict FDR
#   dplyr::filter(abs(log2FoldChange) >= 1.5) %>%    # restrict effect size
#   dplyr::top_n(. , n, abs(log2FoldChange)) %>%     # restrict top FC response
#   pull(gene) %>%
#   unique()

# another way of combining candidate genes
candidate_genes_up <- res %>% as.data.frame %>%
  dplyr::filter(padj < 0.05) %>%                   # restrict FDR
  dplyr::filter(log2FoldChange >= 1.5) %>%    # restrict effect size
  dplyr::top_n(. , n/2, abs(log2FoldChange)) %>%     # restrict top FC response
  pull(gene) %>%
  unique()
candidate_genes_down <- res %>% as.data.frame %>%
  dplyr::filter(padj < 0.05) %>%                   # restrict FDR
  dplyr::filter(log2FoldChange <= 1.5) %>%    # restrict effect size
  dplyr::top_n(. , n/2, abs(log2FoldChange)) %>%     # restrict top FC response
  pull(gene) %>%
  unique()
candidate_genes <- c(candidate_genes_up,candidate_genes_down)

# filter data matrix based on candidate genes
df <- df %>% dplyr::filter(rownames(.) %in% candidate_genes)
  
# perform k-means-clustering to determine k_row.
# wss <- (nrow(scale(df))-1)*sum(apply(scale(df),2,var))
# for (i in 2:20) wss[i] <- sum(kmeans(scale(df),centers=i)$withinss)
# plot(1:20, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 

# hierarchical clustering to check number of genes that fall in clusters.
gene_hclust <- hclust(dist(df,method="euclidean"), method = "complete")
cutree(gene_hclust, k = 2) %>% 
  enframe() %>% 
  rename(gene=name,cluster=value) %>%
  tidyr::separate(.,gene,into=c("gene","response"),sep="_") %>%
  group_by(response,cluster) %>%
  summarise(n=n())

# heatmap
heatmaply(df, 
          k_row = 2,
          scale = "row",
          fontsize_row = 5,
          fontsize_col = 5,
          col_side_colors = metadata[,c("condition","batch")],
          main = "row-scaled log2 TMM-normalized CPM",
          file = c(paste0(Dir,"/heatmaply_log2TMM-CPM_top20.html")))

```
### Write to file
```{r output}
# set global theme for reactable
options(reactable.theme = reactableTheme(borderColor = "#dfe2e5",stripedColor = "#f6f8fa",highlightColor = "#f0f5f9",cellPadding = "8px 12px",style = list(fontFamily = "Helvetica")))

# write raw ESAT counts
write.table(counts.keep, file=paste0(Dir,"/raw_counts.txt"), quote=F, sep="\t", row.names=T, col.names=T)
#htmlwidgets::saveWidget(widget = reactable(counts.keep %>% rownames_to_column("gene"),
#                                           filterable=T, searchable=T, striped=T, bordered=T), 
#                        file = paste0(Dir,"/raw_counts.html"), selfcontained = TRUE)
# write TMM-normalized CPM 
write.table(CPM, file=paste0(Dir,"/TMM-normalized-CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)
write.table(logCPM, file=paste0(Dir,"/log2-TMM-normalized-CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# write metadata
write.table(metadata, file=paste0(Dir,"/metadata.txt"),quote=F, sep="\t", row.names=T, col.names=T)


# write DESeq2 results
write.table(res,file=paste0(Dir,"/DESeq2_results_IFNvsPBS.txt"),quote=F,row.names=T,col.names=T)
write.table(res.up$gene,file=paste0(Dir,"/upregulated_genes.txt"),quote=F,row.names=F,col.names=F)
write.table(res.down$gene,file=paste0(Dir,"/downregulated_genes.txt"),quote=F,row.names=F,col.names=F)
#htmlwidgets::saveWidget(widget = reactable(res,filterable=T, searchable=T, striped=T, bordered=T), 
#                        file = paste0(Dir,"/DESeq2_results_IFNvsPBS.html"), selfcontained = TRUE)
write.table(res$gene, file=paste0(Dir,"/genes_used_in_DE.txt"), quote=F, sep="\t", row.names=F, col.names=F)


# save useful PCA plots

# write log2FoldChange of CPM data
log2FC=log2((CPM.ifn+0.0001)/(CPM.pbs+0.0001))
write.table(log2FC, file=paste0(Dir,"/log2FoldChange_by_IFNg.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# filter out genes that aren't highly expressed
gene_expressed_pbs = CPM.pbs %>% mutate(mean=apply(CPM.pbs, 1, mean)) %>% dplyr::filter(mean>10) %>% rownames(.) 
gene_expressed_ifn = CPM.ifn %>% mutate(mean=apply(CPM.ifn, 1, mean)) %>% dplyr::filter(mean>10) %>% rownames(.)
gene_expressed = intersect(gene_expressed_pbs,gene_expressed_ifn)
write.table(gene_expressed, file=paste0(Dir,"/highly_expressed_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write more genes (expressed AND IFNg responsive)
gene_expressed_and_ifng_responsive = intersect(gene_expressed, c(res.up$gene,res.down$gene))
write.table(gene_expressed_and_ifng_responsive, file=paste0(Dir,"/highly_expressed_and_ifng_responsive_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write down genes that are responsive to IFNg
write.table(c(res.up$gene,res.down$gene),file=paste0(Dir,"/ifng_responsive_genes.txt"), quote=F, sep="\t", row.names=F, col.names=F)

# write down batch-corrected counts
write.table(counts.keep.corrected, file=paste0(Dir,"/batch-corrected-counts.txt"), quote=F, sep="\t", row.names=T, col.names=T)

# write down batch-corrected TMM-normalized CPM
write.table(CPM.corrected, file=paste0(Dir,"/TMM-normalized-batch-corrected-CPM.txt"), quote=F, sep="\t", row.names=T, col.names=T)

```

















## stage two. confounder adjustment and normalization before eQTL and reQTL detection.
### Overview

Celseq2 data for fibroblasts (FRB) from 37 donors X 2 culture conditions.(F47 is removed due to failed genotyping). Libraries were pooled and prepared by celseq2 protocol, pair-end sequenced by BGI DNBseq or Illumina NextSeq500/NextSeq550/NextSeq2000 in house. Barcode in read 1 name was used to separate out pooled samples within a library, and read 2 was used for STAR alignment to hg38 assembly.  ESAT was used to quantify raw read counts on the gene level with 1000bp extension from the 3'UTR.
  
The files generated from this markdown are:
1) raw ESAT counts of selected samples. 
  - raw_counts.txt
  - raw_counts.html
2) TMM-normalized CPM table. 
  - TMM-normalized-CPM.txt
3) log2 CPM table
  - log2-TMM-normalized-CPM.txt
3) DESeq2 results of 21,308 genes (filtered by rowsun>10), including log2FC and padj (FDR). 
  - upregulated_genes.txt (padj<0.05, log2FC>1.5)
  - downregulated_genes.txt (padj<0.05, log2FC<-1.5)
  - DESeq2_results_IFNvsPBS.txt
  - DESeq2_results_IFNvsPBS.html
4) interactive heatmap of top 200 genes with most dramatic FC
  - heatmaply_log2TMM-CPM_top200.html

```{r functions}
create_table <- function(Dir, file_lst) {
  f=file_lst[1]
  sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
  df <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
    tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
    mutate(sum=rowSums(select_if(., is.numeric))) %>%
    dplyr::select(c("region",sum)) %>% 
    magrittr::set_colnames(c("region",sample))
  
  if (length(file_lst) > 1) {
    for (f in file_lst[2:length(file_lst)]) {
      sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
        tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
        mutate(sum=rowSums(select_if(., is.numeric))) %>%
        dplyr::select(c("region",sum)) %>% 
        magrittr::set_colnames(c("region",sample))
    
    df <- dplyr::left_join(df,df.this,by="region")
    }
  }
  df <- df %>% tibble::column_to_rownames( . , "region")
  return(df)
  rm(f,sample,df,df.this)
}

# check which variable correlates with each PC the most.
best_correlation_with_pc <- function(pcs, metadata, pc.n) {
  total.pcc=c()
  pca.this=pcs[,pc.n]
  for (x in colnames(metadata)) {
    metadata.this <- as.numeric(as.factor(metadata[,x]))
    total.pcc=c(total.pcc,cor(pca.this,metadata.this))
  }
  total.df=data.frame(feature=colnames(metadata),pcc=total.pcc,pcc.magnitude=abs(total.pcc))
  return(total.df)
}
```
  
### Load raw ESAT counts
```{r, echo=FALSE, message=FALSE}
Dir.esat="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/DolphinNext/report7561/ESAT"

# 1. import ESAT counts into master table
flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
  tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
samples.all <- data.frame(sample=colnames(counts))

# 2. find list of samples to keep
# 2a. condition used from pilot (KRT, FRB: 100ng 24h, MEL: 10ng 48h)
samples.pilot.keep <- c("F25_FRB_PBS_0ng_S0", "F25_FRB_24h_100ng_S0",
                        "F25_MEL_PBS_0ng_S0", "F25_MEL_48h_10ng_S0",
                        "F25_KRT_PBS_0ng_S0", "F25_KRT_24h_100ng_S0")
# 2b. latest version of samples (merged from replicates if any)
samples <- samples.all %>%
  tidyr::separate( . , col=sample, into=c("donor","celltype","condition","replicate"), sep="_",remove=FALSE) %>%
  dplyr::filter(donor!="F25") %>% tidyr::unite(. , "biosample", c("donor","celltype","condition"),sep="_", remove=FALSE)
# go through the list, and for each donor, pick the row that contain the highest number of replicate
samples.latest <- c()
for (d in unique(samples$biosample)) {
  df.this <- samples %>% dplyr::filter(biosample==d) %>% arrange( . , desc(replicate))
  sample.pick <- df.this[1,1] %>% as.character()
  samples.latest <- c(samples.latest, sample.pick)
  rm(df.this,sample.pick)
}
# 2c. only keratinocytes.
samples.keep <- c(samples.latest, samples.pilot.keep)
samples.keep <- samples.keep[grep("FRB",samples.keep)]

# 2d. remove samples with indicated keywords (after QC inspection. will update until more sequence reads come in)
keyword1 <- "F47" # remove due to failed genotyping.
samples.keep <- samples.keep[-grep(keyword1,samples.keep)]

# 2c combine
counts.keep <- counts %>% dplyr::select(samples.keep)
colnames(counts.keep) <- colnames(counts.keep) %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)

# remove MT genes
MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
counts.keep <- counts.keep %>% mutate(gene=rownames(.)) %>% dplyr::filter(!rownames(.) %in% MTgenes$V1) %>% dplyr::select(-"gene")
```

### Load metadata
```{r metadata, echo=FALSE, message=FALSE}
# parse kept samples into table
samples <- data.frame(sample=colnames(counts.keep))
# import metadata master table
metadata.master=read.table("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/metadata_08182022.txt", sep="\t", header=TRUE)
# import DolphinNext pipeline overall summary
pip.summary=read.table("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/DolphinNext/report7561/summary/overall_summary.tsv", sep="\t", header=TRUE) %>% magrittr::set_colnames(c("sample","total_reads","Multimapped_Reads_Aligned_STAR","Unique_Reads_Aligned_STAR"))
pip.summary$sample <- pip.summary$sample %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)
# import ancestry table
ancestry=read.table("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/ancestry.txt", sep="\t", header=FALSE) %>% set_colnames(c("donor","ancestry"))

# query imported tables with samples
rm(metadata)
metadata <- left_join(samples,metadata.master,by="sample")
metadata <- left_join(metadata,pip.summary,by="sample")
metadata <- left_join(metadata,ancestry,by="donor")

# compile 
metadata$seq_date <- metadata$seq_date %>% gsub("\\/",".",.)
metadata$lib_prep_date <- metadata$lib_prep_date %>% gsub("\\/",".",.)
metadata$batch <- paste(metadata$lib_prep_date,metadata$seq_date,sep="_") %>% as.factor()
metadata.pbs <- metadata %>% dplyr::filter(condition=="PBS")
metadata.ifn <- metadata %>% dplyr::filter(condition=="IFN")

data.table::fwrite(metadata.pbs, file = paste0(Dir,"/metadata_PBS.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
data.table::fwrite(metadata.ifn, file = paste0(Dir,"/metadata_IFN.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)




# cleanup
rm(pip.summary, samples, samples.all)
```
  
  
### CPM calculation + keep expressed genes
```{r cpm}
col.sum=apply(counts.keep,2,sum)
CPM=sweep(counts.keep, 2, col.sum, `/`)*1000000


# Getting CPM matrix
CPM.pbs <- as.data.frame(CPM)[,as.character(metadata.pbs$sample)] %>%
  magrittr::set_colnames(as.character(metadata.pbs$donor)) %>%
  magrittr::set_rownames(rownames(CPM))
CPM.ifn <- as.data.frame(CPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) %>%
  magrittr::set_rownames(rownames(CPM))
```
```{r expressed_genes}
# keep a gene if it's expressed in either PBS or IFN condition. expressed = CPM >= 10 in at least 5 donors in each condition.
gene_expression_status <- c()
for (i in 1:nrow(CPM)) {
  df.pbs <- CPM.pbs[i,]
  df.ifn <- CPM.ifn[i,]
  gene_expression_status_pbs <- (length(df.pbs[which(df.pbs >= 10)]) >= 5)
  gene_expression_status_ifn <- (length(df.ifn[which(df.ifn >= 10)]) >= 5)
  gene_expression_status <- c(gene_expression_status, (gene_expression_status_pbs | gene_expression_status_ifn))
}
expressed_genes <- rownames(CPM)[gene_expression_status]
data.table::fwrite(as.data.frame(expressed_genes), file = paste0(Dir,"/expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)
# filter by expressed genes only.
CPM = CPM[expressed_genes,]
CPM.pbs = CPM.pbs[expressed_genes,]
CPM.ifn = CPM.ifn[expressed_genes,]
```
```{r logCPM_and_log2FC}
# Getting log transformed CPM
logCPM <- log2(CPM+1) %>% as.data.frame() 
logCPM.pbs <- as.data.frame(logCPM)[,as.character(metadata.pbs$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.pbs$donor)) 
logCPM.ifn <- as.data.frame(logCPM)[,as.character(metadata.ifn$sample)] %>% 
  magrittr::set_colnames(as.character(metadata.ifn$donor)) 
log2FC <- logCPM.ifn - logCPM.pbs

# inspect distribution of log-transformed CPM:
plot(density(as.numeric(CPM.pbs[1,])))
plot(density(as.numeric(logCPM.pbs[1,])))
```
### INT transformation
```{r int}
# INT stands for rank-based inverse normal transformation
# function
rankNorm <- function (y) {
  # input y: CPM across all genes per donor.
  k <- 0.375 # an offset to ensure the z-score is finite. from Blom transform.
  n <- length(y)
  # Ranks.
  r <- rank(y, ties.method = "average")
  # Apply transformation.
  y.rankNorm <- stats::qnorm((r - k) / (n - 2 * k + 1))
  return(y.rankNorm)
}

# compare before vs. after INT 
plot(density((CPM.pbs[,2])))
plot(density(rankNorm(CPM.pbs[,2])))
df = data.frame(cpm=CPM.pbs[,2], rankNorm=rankNorm(CPM.pbs[,2]), gene=rownames(CPM.pbs))
df$class <- cut(df$cpm, 
                breaks=c(min(df$cpm)-1,
                         50, 100, 150, 200, 400, 600, 800, 1000, 1200,
                         max(df$cpm)))
ggplot(df , aes(x=cpm, y=rankNorm, color=class)) +
  geom_point()

# apply rankNorm on all CPM
CPM.pbs.rankNorm <- apply(CPM.pbs,2,rankNorm) %>% as.data.frame()
CPM.ifn.rankNorm <- apply(CPM.ifn,2,rankNorm) %>% as.data.frame()

# write to file
data.table::fwrite(CPM.pbs.rankNorm, file = paste0(Dir,"/rankNorm_expressedGenes_PBS.txt"), quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)
data.table::fwrite(CPM.ifn.rankNorm, file = paste0(Dir,"/rankNorm_expressedGenes_IFN.txt"), quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)
data.table::fwrite(CPM.pbs, file = paste0(Dir,"/CPM_expressedGenes_PBS.txt"), quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)
data.table::fwrite(CPM.ifn, file = paste0(Dir,"/CPM_expressedGenes_IFN.txt"), quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)
```

### PCA on rank normalized data
```{r PCA_PBS}
# PCA of CPM.pbs
df=t(CPM.pbs[expressed_genes,])
pca = prcomp(df,centered= T,scale = T)
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# plot
# top PC information
d <- as.data.frame(pcs) %>% mutate(sample=rownames(.))
d <- inner_join(d,metadata.pbs,by=c("sample"="donor")) %>% magrittr::set_rownames(.$sample)

colnames(metadata)
#d$group_tags <- cut(d$Unique_Reads_Aligned_STAR, breaks=6)
d$group_tags <- cut(d$total_reads, 
                    breaks=c(min(d$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d$total_reads)))

# write to file
d2 <- as.data.frame(pcs)
colnames(d2) <- lapply(colnames(d2), function (x) paste0("e",x)) # to denote they are expressionPCs
d2 <- d2 %>% mutate(donor=rownames(.)) %>% dplyr::select(donor, everything())
data.table::fwrite(d2, file = paste0(Dir,"/PCs_CPM_PBS_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

d %>%
  ggplot(., aes(y=PC1,x=PC2,col=seq_date,shape=condition,label=sample)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) 
  ggsave(filename=paste0(Dir,"/PCA_3.png"))
```
```{r PCA_IFN}
# PCA of CPM.ifn
df=t(CPM.ifn[expressed_genes,])
pca = prcomp(df,centered= T,scale = T)
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# plot
# top PC information
d <- as.data.frame(pcs) %>% mutate(sample=rownames(.))
d <- inner_join(d,metadata.pbs,by=c("sample"="donor")) %>% magrittr::set_rownames(.$sample)

colnames(metadata)
#d$group_tags <- cut(d$Unique_Reads_Aligned_STAR, breaks=6)
d$group_tags <- cut(d$total_reads, 
                    breaks=c(min(d$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d$total_reads)))

# write to file
d2 <- as.data.frame(pcs)
colnames(d2) <- lapply(colnames(d2), function (x) paste0("e",x)) # to denote they are expressionPCs
d2 <- d2 %>% mutate(donor=rownames(.)) %>% dplyr::select(donor, everything())
data.table::fwrite(d2, file = paste0(Dir,"/PCs_CPM_IFN_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

d %>%
  ggplot(., aes(y=PC1,x=PC2,col=seq_date,shape=condition,label=sample)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) 
  ggsave(filename=paste0(Dir,"/PCA_3.png"))
```
```{r PCA_rankNormPBS}
# PCA of CPM.pbs
df=t(CPM.pbs.rankNorm[expressed_genes,])
pca = prcomp(df,centered= T,scale = T)
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# plot
# top PC information
d <- as.data.frame(pcs) %>% mutate(sample=rownames(.))
d <- inner_join(d,metadata.pbs,by=c("sample"="donor")) %>% magrittr::set_rownames(.$sample)

colnames(metadata)
#d$group_tags <- cut(d$Unique_Reads_Aligned_STAR, breaks=6)
d$group_tags <- cut(d$total_reads, 
                    breaks=c(min(d$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d$total_reads)))

# write to file
d2 <- as.data.frame(pcs)
colnames(d2) <- lapply(colnames(d2), function (x) paste0("e",x)) # to denote they are expressionPCs
d2 <- d2 %>% mutate(donor=rownames(.)) %>% dplyr::select(donor, everything())
data.table::fwrite(d2, file = paste0(Dir,"/PCs_rankNormCPM_PBS_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

d %>%
  ggplot(., aes(y=PC1,x=PC2,col=seq_date,shape=condition,label=sample)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) 
  ggsave(filename=paste0(Dir,"/PCA_3.png"))
```
```{r PCA_rankNormIFN}
# PCA of CPM.ifn
df=t(CPM.ifn.rankNorm[expressed_genes])
pca = prcomp(df,centered= T,scale = T)
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)

# plot
# top PC information
d <- as.data.frame(pcs) %>% mutate(sample=rownames(.))
d <- inner_join(d,metadata.pbs,by=c("sample"="donor")) %>% magrittr::set_rownames(.$sample)

colnames(metadata)
#d$group_tags <- cut(d$Unique_Reads_Aligned_STAR, breaks=6)
d$group_tags <- cut(d$total_reads, 
                    breaks=c(min(d$total_reads)-1,
                             10e+06, 15e+06, 20e+06,
                             max(d$total_reads)))

# write to file
d2 <- as.data.frame(pcs)
colnames(d2) <- lapply(colnames(d2), function (x) paste0("e",x)) # to denote they are expressionPCs
d2 <- d2 %>% mutate(donor=rownames(.)) %>% dplyr::select(donor, everything())
data.table::fwrite(d2, file = paste0(Dir,"/PCs_rankNormCPM_IFN_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

d %>%
  ggplot(., aes(y=PC1,x=PC2,col=seq_date,shape=condition,label=sample)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) 
  ggsave(filename=paste0(Dir,"/PCA_3.png"))
```
## stage three. percentage change after IFNg stimulation
```{r cpm-fold-change}
absolute.change = CPM.ifn - CPM.pbs

g="ITGA1"
par(mfrow = c(2, 2))
fold.change = (CPM.ifn + 0.1) / (CPM.pbs + 0.1)
plot(density(as.numeric(fold.change[g,])),main="FC:ITGA1(CPM_PBS~14)")
fold.change.dummy = (CPM.ifn + 1) / (CPM.pbs + 1)
plot(density(as.numeric(fold.change.dummy[g,])),main="fold change with dummy=1")
fold.change.dummy = (CPM.ifn + 10) / (CPM.pbs + 10)
plot(density(as.numeric(fold.change.dummy[g,])),main="fold change with dummy=10")
fold.change.dummy = (CPM.ifn + 100) / (CPM.pbs + 100)
plot(density(as.numeric(fold.change.dummy[g,])),main="fold change with dummy=100")

# log2FC+dummy vs absolute change
dummy = 100
log2fc.dummy = log2((CPM.ifn + dummy) / (CPM.pbs + dummy))
df = data.frame(absChange=absolute.change[,33], log2fc=log2fc.dummy[,33], CPM.pbs=CPM.pbs[,33], CPM.ifn=CPM.ifn[,33])
df$CPM_pbs_group <- cut(df$CPM.pbs, breaks=c(min(df$CPM.pbs)-1,1,10,100,1000,max(df$CPM.pbs))) %>% as.factor()
ggplot(df, aes(x=absChange, y=log2fc, color=CPM_pbs_group)) + geom_point(alpha=0.4) + labs(title=">15M seq depth sample F62 with dummy=100")

# distribution of log2FC
dummy=10
coln=33
log2fc.dummy = log2((CPM.ifn + dummy) / (CPM.pbs + dummy))
df = data.frame(absChange=absolute.change[,coln], log2fc=log2fc.dummy[,coln], CPM.pbs=CPM.pbs[,coln], CPM.ifn=CPM.ifn[,coln])
ggplot(df, aes(x=log2fc)) + geom_density() + labs(title="distribution of log2fc wth dummy=10 for donor F62")


```
```{r normalize-log2fc}
dummy=10
log2fc.dummy = log2((CPM.ifn + dummy) / (CPM.pbs + dummy))
data.table::fwrite(log2fc.dummy, file = paste0(Dir,"/Log2FCwithDummy10_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)
# normalize the log2FC scores
# INT stands for rank-based inverse normal transformation
# function
rankNorm <- function (y) {
  # input y: CPM across all genes per donor.
  k <- 0.375 # an offset to ensure the z-score is finite. from Blom transform.
  n <- length(y)
  # Ranks.
  r <- rank(y, ties.method = "average")
  # Apply transformation.
  y.rankNorm <- stats::qnorm((r - k) / (n - 2 * k + 1))
  return(y.rankNorm)
}
rankNorm.log2fc <- apply(log2fc.dummy,2,rankNorm) %>% as.data.frame()
data.table::fwrite(rankNorm.log2fc, file = paste0(Dir,"/rankNormLog2FCwithDummy10_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)


# compare before vs. after INT 
coln=2
df <- data.frame(log2fcWithDummy=log2fc.dummy[,coln], 
                 rankNormLog2fcWithDummy=rankNorm.log2fc[,coln], 
                 absChange=absolute.change[,coln],
                 cpm.pbs=CPM.pbs[,coln])
df$log2fc_group <- cut(round(df$log2fcWithDummy,2), breaks=summary(df$log2fcWithDummy) %>% as.numeric()) %>% as.factor()
df$cpm_group <- cut(round(df$cpm.pbs,1), breaks=c(-1,20,50,100,200) %>% as.numeric()) %>% as.factor()
ggplot(df, aes(x=absChange, y=rankNormLog2fcWithDummy, color=log2fc_group)) + geom_point(alpha=0.4) + labs(title="")

temp=df[which(round(df$rankNormLog2fcWithDummy,1)==2),]
ggplot(temp, aes(x=absChange, y=rankNormLog2fcWithDummy, color=cpm_group)) + geom_point()
```
```{r pca-rankNormLog2fcDummy}
# PCA of log2FCdummy
df=t(rankNorm.log2fc[expressed_genes,])
pca = prcomp(df,centered= T,scale = T)
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)
pca.pctVar.df = data.frame(PC=seq(1,length(pca.pctVar),1), percentVar=round(pca.pctVar*100,2))
ggplot(pca.pctVar.df, aes(x=PC, y=percentVar, label=percentVar)) + geom_point() + geom_text(nudge_x = 0.8,nudge_y = 0.8,size=3,angle = 30) + labs(title="PC of rankNorm log2FC+dummy for fibroblasts") + ylab("percentage of variance (%)")

# metadata for each donor
PBS_knownVar_file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis/metadata_PBS.txt"
IFN_knownVar_file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis/metadata_IFN.txt"
metadata.pbs = data.table::fread(PBS_knownVar_file)
metadata.ifn = data.table::fread(IFN_knownVar_file)
metadata.fc = data.frame(donor=metadata.pbs$donor, ancestry=metadata.pbs$ancestry, 
                         total_reads_pbs=metadata.pbs$total_reads, total_reads_ifn=metadata.ifn$total_reads,
                         lib_prep_date_pbs=metadata.pbs$lib_prep_date,
                         lib_prep_date_ifn=metadata.ifn$lib_prep_date,
                         seq_date_pbs=metadata.pbs$seq_date,
                         seq_date_ifn=metadata.ifn$seq_date,
                         seq_model_pbs=metadata.pbs$seq_model,
                         seq_model_ifn=metadata.ifn$seq_model)
metadata.fc$difference_total_reads <- metadata.fc$total_reads_ifn - metadata.fc$total_reads_pbs
summary(metadata.fc$difference_total_reads)
metadata.fc$difference_total_reads_groups <- cut(metadata.fc$difference_total_reads,
                                                 breaks=c(summary(metadata.fc$difference_total_reads)[1]-1, 
                                                          summary(metadata.fc$difference_total_reads)[2:3],
                                                          summary(metadata.fc$difference_total_reads)[5],
                                                          summary(metadata.fc$difference_total_reads)[6]+1)) %>% as.factor()

metadata.fc$lib_prep_date <- paste0(metadata.pbs$lib_prep_date, "_", metadata.ifn$lib_prep_date) %>% as.factor()
metadata.fc$seq_date <- paste0(metadata.pbs$seq_date, "_", metadata.ifn$seq_date) %>% as.factor()
metadata.fc$seq_model <- paste0(metadata.pbs$seq_model, "_", metadata.ifn$seq_model) %>% as.factor()
data.table::fwrite(metadata.fc, file = paste0(Dir,"/metadata_rankNormLog2FCwithDummy10_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)


# write to file
d3 <- as.data.frame(pcs)
colnames(d3) <- lapply(colnames(d3), function (x) paste0("e",x)) # to denote they are expressionPCs
d3 <- d3 %>% mutate(donor=rownames(.)) %>% dplyr::select(donor, everything())
data.table::fwrite(d3, file = paste0(Dir,"/PCs_rankNormLog2FCwithDummy10_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

# plot
d3.plot <- as.data.frame(pcs) %>% mutate(donor=rownames(.))
d3.plot <- inner_join(d3.plot,metadata.fc,by="donor") %>% magrittr::set_rownames(.$donor)

colnames(metadata.fc)
d3.plot %>%
  ggplot(., aes(y=PC1,x=PC2,col=difference_total_reads_groups,label=donor)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  ggsave(filename=paste0(Dir,"/rankNormlog2FC_PCA_PC1PC2_read-difference.png"))
```
```{r pca-Log2fcDummy}
# PCA of log2FCdummy
df=t(log2fc.dummy[expressed_genes,])
pca = prcomp(df,centered= T,scale = T)
loadings=as.data.frame(pca$rotation)

# percent variance in each pc
plot(pca)
pcs = pca$x
pca.pctVar = pca$sdev^2/sum(pca$sdev^2)
pca.pctVar.df = data.frame(PC=seq(1,length(pca.pctVar),1), percentVar=round(pca.pctVar*100,2))
ggplot(pca.pctVar.df, aes(x=PC, y=percentVar, label=percentVar)) + geom_point() + geom_text(nudge_x = 0.8,nudge_y = 0.8,size=3,angle = 30) + 
  labs(title="PC of log2FC+dummy for fibroblasts") + ylab("percentage of variance (%)")

# metadata for each donor
PBS_knownVar_file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis/metadata_PBS.txt"
IFN_knownVar_file="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_11192022/analysis/metadata_IFN.txt"
metadata.pbs = data.table::fread(PBS_knownVar_file)
metadata.ifn = data.table::fread(IFN_knownVar_file)
metadata.fc = data.frame(donor=metadata.pbs$donor, ancestry=metadata.pbs$ancestry, 
                         total_reads_pbs=metadata.pbs$total_reads, total_reads_ifn=metadata.ifn$total_reads,
                         lib_prep_date_pbs=metadata.pbs$lib_prep_date,
                         lib_prep_date_ifn=metadata.ifn$lib_prep_date,
                         seq_date_pbs=metadata.pbs$seq_date,
                         seq_date_ifn=metadata.ifn$seq_date,
                         seq_model_pbs=metadata.pbs$seq_model,
                         seq_model_ifn=metadata.ifn$seq_model)
metadata.fc$difference_total_reads <- metadata.fc$total_reads_ifn - metadata.fc$total_reads_pbs
summary(metadata.fc$difference_total_reads)
metadata.fc$difference_total_reads_groups <- cut(metadata.fc$difference_total_reads,
                                                 breaks=c(summary(metadata.fc$difference_total_reads)[1]-1, 
                                                          summary(metadata.fc$difference_total_reads)[2:3],
                                                          summary(metadata.fc$difference_total_reads)[5],
                                                          summary(metadata.fc$difference_total_reads)[6]+1)) %>% as.factor()

metadata.fc$lib_prep_date <- paste0(metadata.pbs$lib_prep_date, "_", metadata.ifn$lib_prep_date) %>% as.factor()
metadata.fc$seq_date <- paste0(metadata.pbs$seq_date, "_", metadata.ifn$seq_date) %>% as.factor()
metadata.fc$seq_model <- paste0(metadata.pbs$seq_model, "_", metadata.ifn$seq_model) %>% as.factor()

# write to file
d4 <- as.data.frame(pcs)
colnames(d4) <- lapply(colnames(d4), function (x) paste0("e",x)) # to denote they are expressionPCs
d4 <- d4 %>% mutate(donor=rownames(.)) %>% dplyr::select(donor, everything())
data.table::fwrite(d4, file = paste0(Dir,"/PCs_Log2FCwithDummy10_expressedGenes.txt"), quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

# plot
d4.plot <- as.data.frame(pcs) %>% mutate(donor=rownames(.))
d4.plot <- inner_join(d4.plot,metadata.fc,by="donor") %>% magrittr::set_rownames(.$donor)

colnames(metadata.fc)
d4.plot %>%
  ggplot(., aes(y=PC1,x=PC2,col=ancestry,label=donor)) + 
  geom_point(size=3) + 
  geom_text_repel(show.legend = F,size=2) + 
  ylab(sprintf("PC1 (%2.1f%% variance)",100*pca.pctVar[1])) + 
  xlab(sprintf("PC2 (%2.1f%% variance)",100*pca.pctVar[2])) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  ggsave(filename=paste0(Dir,"/log2FC_PCA_PC1PC2_ancestry.png"))
```






