---
title: "skineQTL_model2025_test1"
author: "Crystal Shan"
date: "2025-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
notes:
F78 and F138 don't have genotype data?
# [don't run] archive code for getting ERAP2 expression for IFNG eQTL and IFNG reQTL.
```{r archive, warning=FALSE}
Dir="/pi/manuel.garber-umw/human/skin/eQTLs/RNA-Seq/analysis_07142025/"
setwd(Dir)

#save.image(file=paste0(Dir,'/myEnvironment_main.RData'))
#load("/pi/manuel.garber-umw/human/skin/eQTLs/RNA-Seq/analysis_07142025/myEnvironment_main.RData") # raw counts, CPM, log2(CPM+1), metadata, expressed genes by ct

# INT stands for rank-based inverse normal transformation
# function
rankNorm <- function (y) {
  # input y: numeric vector of CPM across all genes per donor.
  k <- 0.375 # an offset to ensure the z-score is finite. from Blom transform.
  n <- length(y)
  # Ranks.
  r <- rank(y, ties.method = "average") # if same value, same rank
  # Apply transformation.
  r.prob <- (r - k) / (n - 2 * k + 1)
  y.rankNorm <- stats::qnorm(r.prob) # because qnorm(1) is infinite and qnorm(-1) is -Inf
  return(y.rankNorm)
}

metadata$biosample = paste0(metadata$donor,"_",metadata$celltype, "_", metadata$condition)
df = data.frame(biosample=metadata$biosample, 
                donor=metadata$donor_num,
                counts=as.numeric(counts["ERAP2",]), 
                CPM=as.numeric(CPM["ERAP2",]),
                libsize=colSums(counts))
df.MELIFNG = df[metadata$celltype=="MEL"&metadata$condition=="IFNG",]
df.MELIFNG$rankNormCPM=rankNorm(df.MELIFNG$CPM)
metadata.MELIFNG = metadata[metadata$celltype=="MEL"&metadata$condition=="IFNG",]

df.MEL = df[metadata$celltype=="MEL"&(metadata$condition=="PBS"|metadata$condition=="IFNG"),]
df.MEL$rankNormCPM=rankNorm(df.MEL$CPM)
metadata.MEL = metadata[metadata$celltype=="MEL"&(metadata$condition=="PBS"|metadata$condition=="IFNG"),]

outDir="/pi/manuel.garber-umw/human/skin/eQTLs/DREG/2025models/test1/"
saveRDS(df.MELIFNG, paste0(outDir,"/ERAP_expression_MEL_IFNG.rds"))
saveRDS(metadata.MELIFNG, paste0(outDir,"/metadata_MEL_IFNG.rds"))
saveRDS(df.MEL, paste0(outDir,"/ERAP_expression_MEL_PBSandIFNG.rds"))
saveRDS(metadata.MEL, paste0(outDir,"/metadata_MEL_PBSandIFNG.rds"))
```
# load data
```{r load}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggpubr)
library(ggrepel)
library(ggfortify)
library(reactable)
library(gridExtra)
library(lme4)
library(lmerTest)
library(tidyr)
library(tibble)
library(purrr)
library(kableExtra)
library(pander)
library(emmeans)
library(here)
library(broom)
library(broom.mixed)
library(MuMIn)

Dir="~/Downloads/nl/human/skin/eQTLs/DREG/2025models/test1/"
setwd(Dir)
load(paste0(Dir,"/QTL_model_testing1.RData"))

# df.MEL.IFNG <- readRDS(paste0(Dir,"/ERAP_expression_MEL_IFNG.rds"))
# df.MEL.IFNG$donor <- paste0("F", df.MEL.IFNG$donor)
# df.MEL.IFNG$loglibsize <- log(df.MEL.IFNG$libsize)
# metadata.MEL.IFNG <- readRDS(paste0(Dir,"/metadata_MEL_IFNG.rds"))
# df.MEL.PBSIFNG <- readRDS(paste0(Dir,"/ERAP_expression_MEL_PBSandIFNG.rds"))
# df.MEL.PBSIFNG$donor <- paste0("F", df.MEL.PBSIFNG$donor)
# df.MEL.PBSIFNG$loglibsize <- log(df.MEL.PBSIFNG$libsize)
# df.MEL.PBSIFNG <- df.MEL.PBSIFNG[rownames(df.MEL.PBSIFNG)!="F34_MEL_PBS_S1", ]
# df.MEL.PBSIFNG$condition <- ordered(gsub("F.*_MEL_","", df.MEL.PBSIFNG$biosample),
#                                     levels = c("PBS","IFNG"))
# metadata.MEL.PBSIFNG <- readRDS(paste0(Dir,"/metadata_MEL_PBSandIFNG.rds"))
# metadata.MEL.PBSIFNG <- metadata.MEL.PBSIFNG[metadata.MEL.PBSIFNG$sample!="F34_MEL_PBS_S1", ]
# 
# snpinfo <- data.table::fread(paste0(Dir,"/snp.bed")) %>% t() %>% 
#   as.data.frame() %>% 
#   rownames_to_column("donor") %>%
#   set_colnames(c("donor","GT"))
# snpinfo$donor <- gsub("skineQTL-","",snpinfo$donor)
# REF <- "T"
# ALT <- "C"
# 
# genotype <- snpinfo[7:nrow(snpinfo), ]
# genotype <- genotype %>%
#   mutate(
#     genotype.num = case_when(
#       GT == "0|0" ~ 0,
#       GT %in% c("0|1", "1|0") ~ 1,
#       GT == "1|1" ~ 2,
#       TRUE ~ NA_real_
#     ),
#     genotype.nt = case_when(
#       GT == "0|0" ~ "TT",
#       GT == "0|1" ~ "TC|CT",
#       GT == "1|0" ~ "TC|CT",
#       GT == "1|1" ~ "CC",
#       TRUE ~ NA_character_
#     )
#   )
# genotype$donor <- sub("^F0", "F", genotype$donor)

outliers_list <- c("F108_MEL","F109_MEL","F110_MEL","F111_MEL","F51_MEL","F52_MEL","F61_MEL","F118_MEL","F133_MEL","F116_MEL_IFNG","F116_MEL_IFNG","F142_MEL_TNF","F42_MEL_PBS","F34_MEL_PBS","F30_MEL_IFNG")
lowcounts_list <- c("F116_MEL_IFNG","F142_MEL_TNF","F42_MEL_PBS","F34_MEL_PBS","F30_MEL_IFNG")
pattern <- str_c(outliers_list, collapse = "|")
outlier_samples_metadata.mel <- metadata.MEL.PBSIFNG %>% dplyr::filter(str_detect(sample, pattern))

metadata.MEL.PBSIFNG <- metadata.MEL.PBSIFNG %>% dplyr::filter(!sample%in% outlier_samples_metadata.mel$sample)
metadata.MEL.IFNG <- metadata.MEL.IFNG %>% dplyr::filter(!sample%in% outlier_samples_metadata.mel$sample)

df.MEL.PBSIFNG <- df.MEL.PBSIFNG[metadata.MEL.PBSIFNG$sample, ]
df.MEL.IFNG <- df.MEL.IFNG[metadata.MEL.IFNG$sample, ]

rownames(df.MEL.PBSIFNG)==metadata.MEL.PBSIFNG$sample
rownames(df.MEL.IFNG)==metadata.MEL.IFNG$sample

real_dat <- left_join(df.MEL.PBSIFNG, metadata.MEL.PBSIFNG[, c("biosample","phase","sequencing_platform")], by="biosample") %>%
  left_join(. , genotype, by="donor")
#make sure the reference group is set to "PBS"
real_dat$condition <- relevel(factor(as.character(real_dat$condition)), ref = "PBS")
real_dat$genotype.nt <- relevel(factor(real_dat$genotype.nt), ref = "TT")
real_dat$log2CPM <- log2(real_dat$CPM + 1)
```
# plotting
```{r plotting-x-is-genotype}
this.snp <- "rs2910686"
snp.ref <- "T"
snp.alt <- "C"
this.gene <- "ERAP2"
celltype <- "MEL"
this_reQTL_matrix <- left_join(df.MEL.PBSIFNG, genotype, by="donor") %>% na.omit()
df <- this_reQTL_matrix
df$condition <- gsub("F.*_","",df$biosample)
df$condition <- ordered(df$condition, levels=c("PBS","IFNG"))
df$genotype.nt <- ordered(df$genotype.nt, 
                       levels <- c(paste0(snp.ref,snp.ref),
                                   paste0(snp.ref,snp.alt,"|",snp.alt,snp.ref),
                                   paste0(snp.alt,snp.alt)))

donor_counts <- df %>%
  distinct(genotype.nt, donor) %>%
  group_by(genotype.nt) %>%
  summarise(n = n(), .groups = "drop")

facet_labels <- setNames(
  paste0(as.character(donor_counts$genotype.nt), " (n=", donor_counts$n, ")"),
  donor_counts$genotype.nt
)


# make paired plots
position <- "identity"; width = 0.5; point.size = 2; line.size = 0.5
line.color = "grey"; linetype = "solid"; palette="bright"
highlightdata1 <- df %>% filter(donor=="F25")
highlightdata2 <- df %>% filter(donor=="F49")
highlightdata3 <- df %>% filter(donor=="F55")

p <- ggplot(df, create_aes(list(x = "condition", y = "CPM"))) +
  geom_exec(geom_line, data = df, group = "donor",
            color = line.color, size = line.size, linetype = linetype,
            position = position) +
  geom_exec(geom_point, data = df, color = "condition", size = point.size,
            position = position) +
  scale_color_manual(values=c("#999999", "#E69F00")) +
  geom_exec(geom_line, data = highlightdata1, group = "donor",
            color = "blue", size = 0.7, linetype = linetype,
            position = position) +
  geom_exec(geom_line, data = highlightdata2, group = "donor",
            color = "brown", size = 0.7, linetype = linetype,
            position = position) +
  geom_exec(geom_line, data = highlightdata3, group = "donor",
            color = "black", size = 0.7, linetype = linetype,
            position = position) +
  ggtitle(paste0(celltype,"\n",this.snp,":",this.gene)) +
  theme(legend.position="none") +
  facet_wrap(~ genotype.nt, labeller = labeller(genotype.nt = facet_labels))


p2 <- ggplot(df, create_aes(list(x = "condition", y = "CPM"))) +
  geom_exec(geom_line, data = df, group = "donor",
            color = line.color, size = line.size, linetype = linetype,
            position = position) +
  geom_exec(geom_point, data = df, color = "condition", size = point.size,
            position = position) +
  scale_color_manual(values=c("#999999", "#E69F00")) +
  ggtitle(paste0(celltype,"\n",this.snp,":",this.gene)) +
  theme(legend.position="none") +
  facet_wrap(~ genotype.nt, labeller = labeller(genotype.nt = facet_labels))
```
```{r plotting-x-is-condition}
# Reshape to long format for faceting
real_dat_plot <- real_dat %>%
  pivot_longer(cols = c(counts, CPM), names_to = "scale_type", values_to = "expression_value")

# Create the plot
ggplot(real_dat_plot, aes(x = factor(condition),
                          y = expression_value,
                          color = factor(genotype.num))) +
  geom_point(alpha = 0.6) +
  geom_line(aes(group = donor), alpha = 0.6) + 
  labs(
    title = "Interaction between Genotype and IFNg_treatment on Expression",
    x = "IFNg Treatment",
    y = "Expression Value",
    color = "Genotype"
  ) +
  facet_wrap(scale_type ~ phase, scales = "free_y") +
  theme_bw()
```
# model for IFNG eQTL
```{r model-eQTL}
library(MASS)
this_matrix <- left_join(df.MEL.IFNG, genotype, by="donor") %>% na.omit

# Model 1: Raw counts + Poisson + Fixed effect
model1 <- glm(counts ~ genotype.num, data = this_matrix, family = "poisson", offset = loglibsize)

# Model 2: Raw counts + Negative Binomial + Fixed effect
# Why use: More flexible than Poisson; handles overdispersion better (variance > mean).
# > overdispersion_ratio <- sum(residuals(model1, type="pearson")^2) / df.residual(model1)
# > print(overdispersion_ratio)
# [1] 35.30822 --> highly overdispersed. NB >> Poisson
model2 <- glm.nb(counts ~ genotype.num + offset(loglibsize), data = this_matrix)

# Model 3: Rank-normalized CPM + Linear + Fixed effect
# Why use: No need to model count distributions; robust to outliers. Good for visualization too.
model3 <- lm(rankNormCPM ~ genotype.num, data = this_matrix)

### ----- is it really okay to have number of levels of each grouping factor == number of observations? model4 and 5 could work but model6 threw an error. 
# Model 4: Raw counts + Poisson + Mixed effects (random intercept by donor)
# Why use: Controls for donor-specific baseline expression differences.
library(lme4)
model4 <- glmer(counts ~ genotype.num + (1 | donor), family = poisson, data = this_matrix, offset = loglibsize)

# Model 5: Raw counts + Negative Binomial + Mixed effects (random intercept)
# Why use: Like Model 4, but handles overdispersion.
library(glmmTMB)
model5 <- glmmTMB(counts ~ genotype.num + (1 | donor), 
                  family = nbinom2, data = this_matrix, offset = loglibsize)

### after this point, the models don't work anymore. 
# Model 6: Rank-normalized CPM + Linear + Mixed (random intercept)
# Why use: Allows for subject-specific baselines without worrying about raw count assumptions.
# If every donor appears only once (or the counts equal one another), lme4 can’t estimate a donor‑level variance component
this_matrix$donor <- as.factor(this_matrix$donor)
model6 <- lmer(rankNormCPM ~ genotype.num + (1 | donor), data = this_matrix)
summary(model6)

# Model 7: Raw counts + Poisson + Mixed (random intercept & slope for donor)
# Why use: Donor response to treatment may vary — captures this heterogeneity.
model7 <- glmer(counts ~ genotype.num + IFNg_treatment + genotype:IFNg_treatment + 
                  (1 + IFNg_treatment | donor), 
                family = poisson, data = df, offset = log_libsize)
summary(model7)

# Model 8: Raw counts + Negative Binomial + Mixed (RI + RS)
# Why use: Best model for real-life complexity: donor-specific intercepts and slopes + count overdispersion.
model8 <- glmmTMB(counts ~ genotype + IFNg_treatment + genotype:IFNg_treatment + 
                    (1 + IFNg_treatment | donor), 
                  family = nbinom2, data = df, offset = log_libsize)
summary(model8)

# Model 9: Rank-normalized CPM + Linear + Mixed (RI + RS)
# Why use: Simplified model structure, very fast, good for initial screening.
df9 <- data.frame(expr = rank_norm_expr, genotype = genotype_vector, 
                  donor = donor_vector, IFNg_treatment = treatment_vector)

model9 <- lmer(expr ~ genotype + IFNg_treatment + genotype:IFNg_treatment + 
                 (1 + IFNg_treatment | donor), data = df9)
summary(model9)

```
# model for reQTL
```{r model-reQTL}
this_matrix <- left_join(df.MEL.PBSIFNG, metadata.MEL.PBSIFNG[, c("biosample","phase","sequencing_platform")], by="biosample") %>%
  left_join(. , genotype, by="donor")

# Model 1: beta comparison method: Raw counts + Poisson + Fixed effect
# model1.p.pnorm <- 0.3244923
library(MASS)
model1.pbs <- glm(counts ~ genotype.num, data = this_matrix[this_matrix$condition=="PBS",], family = "poisson", offset = loglibsize)
model1.ifng <- glm(counts ~ genotype.num, data = this_matrix[this_matrix$condition=="IFNG",], family = "poisson", offset = loglibsize)
model1.pbs.b <- summary(model1.pbs)$coefficients["genotype.num","Estimate"]
model1.pbs.se <- summary(model1.pbs)$coefficients["genotype.num", "Std. Error"]
model1.ifng.b <- summary(model1.ifng)$coefficients["genotype.num","Estimate"]
model1.ifng.se <- summary(model1.ifng)$coefficients["genotype.num", "Std. Error"]
beta_comparison <- (model1.pbs.b - model1.ifng.b) / (sqrt((model1.pbs.se^2) + (model1.ifng.se^2)))
model1.p.pnorm <- 2 * pmin(pnorm(beta_comparison), 1 - pnorm(beta_comparison))

# Model 2: Raw counts + Negative Binomial + Fixed effect
# Why use: More flexible than Poisson; handles overdispersion better (variance > mean).
# model2.p.pnorm = 0.6081773
model2 <- glm.nb(counts ~ genotype.num + offset(loglibsize), data = this_matrix)
model2.pbs <- glm.nb(counts ~ genotype.num + offset(loglibsize), data = this_matrix[this_matrix$condition=="PBS",])
model2.ifng <- glm.nb(counts ~ genotype.num + offset(loglibsize), data = this_matrix[this_matrix$condition=="IFNG",])
model2.pbs.b <- summary(model2.pbs)$coefficients["genotype.num","Estimate"]
model2.pbs.se <- summary(model2.pbs)$coefficients["genotype.num", "Std. Error"]
model2.ifng.b <- summary(model2.ifng)$coefficients["genotype.num","Estimate"]
model2.ifng.se <- summary(model2.ifng)$coefficients["genotype.num", "Std. Error"]
beta_comparison <- (model2.pbs.b - model2.ifng.b) / (sqrt((model2.pbs.se^2) + (model2.ifng.se^2)))
model2.p.pnorm <- 2 * pmin(pnorm(beta_comparison), 1 - pnorm(beta_comparison))

# Model 3: Rank-normalized CPM + Linear + Fixed effect
# Why use: No need to model count distributions; robust to outliers. Good for visualization too.
# model3.p.pnorm = 0.243769
model3 <- lm(rankNormCPM ~ genotype.num, data = this_matrix)
model3.pbs <- lm(rankNormCPM ~ genotype.num, data = this_matrix[this_matrix$condition=="PBS",])
model3.ifng <- lm(rankNormCPM ~ genotype.num, data = this_matrix[this_matrix$condition=="IFNG",])
model3.pbs.b <- summary(model3.pbs)$coefficients["genotype.num","Estimate"]
model3.pbs.se <- summary(model3.pbs)$coefficients["genotype.num", "Std. Error"]
model3.ifng.b <- summary(model3.ifng)$coefficients["genotype.num","Estimate"]
model3.ifng.se <- summary(model3.ifng)$coefficients["genotype.num", "Std. Error"]
beta_comparison <- (model3.pbs.b - model3.ifng.b) / (sqrt((model3.pbs.se^2) + (model3.ifng.se^2)))
model3.p.pnorm <- 2 * pmin(pnorm(beta_comparison), 1 - pnorm(beta_comparison))

# Model 4: Raw counts + Poisson + Mixed effects (random intercept by donor)
# Why use: Controls for donor-specific baseline expression differences.
library(lme4)
model4 <- glmer(counts ~ genotype.num + condition + genotype.num:condition + (1 | donor), 
                     family = poisson, data = this_matrix, offset = loglibsize)

# Model 5: Raw counts + Negative Binomial + Mixed effects (random intercept)
# Why use: Like Model 4, but handles overdispersion.
library(glmmTMB)
model5 <- glmmTMB(counts ~ genotype.num + condition + genotype.num:condition + (1 | donor), 
                  family = nbinom2, data = this_matrix, offset = loglibsize)

# Model 6: Rank-normalized CPM + Linear + Mixed (random intercept)
# Why use: Allows for subject-specific baselines without worrying about raw count assumptions.
# If every donor appears only once (or the counts equal one another), lme4 can’t estimate a donor‑level variance component
library(lmerTest)
this_matrix$donor <- as.factor(this_matrix$donor)
model6 <- lmer(rankNormCPM ~ genotype.num + condition + genotype.num:condition + (1 | donor), data = this_matrix)


# Model 7: Raw counts + Poisson + Mixed (random intercept & slope for donor)
# Why use: Donor response to treatment may vary — captures this heterogeneity.
model7 <- glmer(counts ~ genotype.num + condition + genotype.num:condition + 
                  (1 + condition | donor),
                family = poisson, data = this_matrix, offset = loglibsize)

# Model 8: Raw counts + Negative Binomial + Mixed (RI + RS)
# Why use: Best model for real-life complexity: donor-specific intercepts and slopes + count overdispersion.
model8 <- glmmTMB(counts ~ genotype.num + condition + genotype.num:condition + 
                  (1 + condition | donor), 
                  family = nbinom2, data = this_matrix, offset = loglibsize)

# Model 9: Rank-normalized CPM + Linear + Mixed (RI + RS)
# Why use: Simplified model structure, very fast, good for initial screening.
model9 <- lmer(rankNormCPM ~ genotype.num + condition + genotype.num:condition + 
                 (1 + condition | donor), data = this_matrix)
summary(model9)

# Model 10: Kruskal-Wallis rank sum test
# p-value < 2.2e-16
this_matrix <- this_matrix %>%
  group_by(donor) %>%
  mutate(delta_CPM = CPM[condition == "IFNG"] - CPM[condition == "PBS"]) %>%
  ungroup()

kruskal.test(delta_CPM ~ genotype.nt, data = this_matrix)

# Model 11: beta regression on scaled delta CPM
library(betareg)
# Wide format for delta computation
df_wide <- this_matrix %>%
  dplyr::select(donor, genotype.num, genotype.nt, condition, CPM) %>%
  tidyr::pivot_wider(names_from = condition, values_from = CPM) %>%
  mutate(
    delta = IFNG - PBS,
    rel_response = (IFNG - PBS) / (PBS + 1),  # % increase over PBS baseline
    rel_response_scaled = (rel_response - min(rel_response) + 0.001) /
                          (max(rel_response - min(rel_response)) + 0.002)  # map to (0,1)
  )
model_beta <- betareg(rel_response_scaled ~ genotype.num, data = df_wide)
summary(model_beta)
ggplot(df_wide, aes(x = as.factor(genotype.num), y = rel_response_scaled)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, height = 0, color = "orange", alpha = 0.6) +
  labs(x = "Genotype (0=TT, 1=TC/CT, 2=CC)", 
       y = "Scaled IFNG response / PBS", 
       title = "Beta regression target: Relative response to IFNG")

# Model 12: beta regression on raw delta CPM
summary(lm(delta ~ genotype.num, data = df_wide))
ggplot(df_wide, aes(x = as.factor(genotype.num), y = rel_response)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, height = 0, color = "orange", alpha = 0.6) +
  labs(x = "Genotype (0=TT, 1=TC/CT, 2=CC)", 
       y = "IFNG response / PBS", 
       title = "Beta regression target: Relative response to IFNG")

# Model 13: Gen suggested: linear model on CPM with random intercept for donor
model13 <- lmerTest::lmer(CPM ~ condition * genotype.num  +
                  (1 | donor), #random intercept for donor
                data = this_matrix)
summary(model13)
tidy(model13)
pander(broom.mixed::tidy(model13))
# check residual normality
plot(model13, which = 1)  # residuals vs fitted
qqnorm(residuals(model13)); qqline(residuals(model13))  # normal Q-Q plot
hist(residuals(model13))  # optional

# Model 14: log2(CPM+1) linear model with random intercept for donor
this_matrix$log2CPM <- log2(this_matrix$CPM + 1)
model14 <- lmerTest::lmer(log2CPM ~ condition * genotype.num  +
                  (1 | donor), #random intercept for donor
                data = this_matrix)
broom.mixed::tidy(model14)
# check residual normality
plot(model14, which = 1)  # residuals vs fitted
qqnorm(residuals(model14)); qqline(residuals(model14))  # normal Q-Q plot
hist(residuals(model14))  # optional
```
```{r compare-reQTL-models}
### ------- AIC --------- ####
AIC(model6, model13, model14)
```
# QQplot pval model calibration
```{r qqplot-functions-noDonorPairRetention}
### functions
#function to permute outcome and record results for various model formulas
generate_permuted_null <- function(model_formula,
                                   data,
                                   interaction_term_regex = ":",
                                   permute_var = NULL,
                                   n_perm = 100,
                                   model_type = c("lmer", "glmer", "glmer.nb", "glm"),
                                   family = poisson) {
  
  
  model_type <- match.arg(model_type)
  model_formula_str <- paste(deparse(model_formula), collapse = " ")
  
  if (is.null(permute_var)) {
    outcome_var <- all.vars(model_formula)[1]
  } else {
    outcome_var <- permute_var
  }
  
  # Helper to fit a model and capture warnings
  fit_model_safely <- function(formula, data, model_type, family) {
    warnings <- character()
    
    result <- withCallingHandlers(
      tryCatch({
        fit <- switch(
          model_type,
          lmer     = lmerTest::lmer(formula, data = data),
          glmer    = glmer(formula, data = data, family = family),
          glmer.nb = glmer.nb(formula, data = data),
          glm      = glm(formula, data = data, family = family)
        )
        list(fit = fit, warnings = warnings)
      }, error = function(e) {
        list(fit = NULL, warnings = paste("Error:", conditionMessage(e)))
      }),
      warning = function(w) {
        warnings <<- c(warnings, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    )
    
    result$warnings <- paste(result$warnings, collapse = " | ")
    result
  }
  
  # Fit observed model
  obs_model_result <- fit_model_safely(model_formula, data, model_type, family)
  if (is.null(obs_model_result$fit)) return(tibble())  # skip if fail
  
  observed_results <- tidy(obs_model_result$fit) %>%
    filter(str_detect(term, interaction_term_regex)) %>%
    mutate(
      type = "observed",
      model_formula = model_formula_str,
      warning = obs_model_result$warnings
    )
  
  # Permuted results
  permuted_results <- map_dfr(1:n_perm, function(i) {
    permuted_data <- data %>%
      mutate(!!sym(outcome_var) := sample(!!sym(outcome_var)))
    
    perm_model_result <- fit_model_safely(model_formula, permuted_data, model_type, family)
    
    if (is.null(perm_model_result$fit)) return(tibble())
    
    tidy(perm_model_result$fit) %>%
      filter(str_detect(term, interaction_term_regex)) %>%
      mutate(
        type = "null",
        model_formula = model_formula_str,
        warning = perm_model_result$warnings
      )
  })
  
  bind_rows(observed_results, permuted_results)
}


plot_qq_pval_multi <- function(sim_df) {
  # Load ggrepel for smart label positioning
  library(ggrepel)
  
  # Ensure p-values are available and create model identifier
  sim_df <- sim_df %>%
    mutate(
      model_formula_family = str_c(model_formula, " ", paste0(family), " ", paste0(model_type))
    )
  
  # Calculate empirical p-values per model_formula
  empirical_pvals <- sim_df %>%
    group_by(model_formula_family) %>%
    summarise(
      observed_pval = p.value[type == "observed"],
      null_pvals = list(p.value[type == "null"]),
      empirical_pval = mean(null_pvals[[1]] <= observed_pval, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Prepare data for QQ plot (null points only)
  # Convert p-values to -log10 scale for better visualization
  qq_df <- sim_df %>%
    filter(type == "null") %>%
    group_by(model_formula_family) %>%
    arrange(p.value) %>%
    mutate(
      neg_log10_pval = -log10(p.value),
      expected_neg_log10 = -log10(ppoints(n(), a = 0))
    ) %>%
    ungroup()
  
  # Prepare observed points (one per model)
  observed_points <- empirical_pvals %>%
    mutate(
      observed_neg_log10 = -log10(observed_pval),
      # Expected value for the most extreme point
      expected_neg_log10 = -log10(0.5 / lengths(null_pvals)),
      label = paste0("p = ", signif(observed_pval, 3))
    )
  
  # Plot
  ggplot(qq_df, aes(x = expected_neg_log10, y = neg_log10_pval, color = model_formula_family)) +
    geom_point(alpha = 0.4) +
    geom_abline(slope = 1, intercept = 0, color = "gray") +
    geom_point(data = observed_points, aes(x = expected_neg_log10, y = observed_neg_log10), 
               shape = 4, size = 4, stroke = 1.5) + # "X"
    geom_text_repel(data = observed_points, 
                    aes(x = expected_neg_log10,
                        y = observed_neg_log10,
                        label = label,
                        color = model_formula_family), 
                    size = 3,
                    min.segment.length = 0,  # Always draw connecting lines
                    segment.color = "black",
                    segment.alpha = 0.6,
                    box.padding = 0.5,       # Space around labels
                    point.padding = 0.3,     # Space around points
                    force = 2,               # Repulsion strength
                    max.overlaps = Inf) +    # Allow all labels to be shown
    scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    labs(
      x = "Expected -log10(p-value)",
      y = "Observed -log10(p-value)",
      color = "Model Formula",
      title = "QQ Plot of Permuted vs Observed P-values (Multiple Models)"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      legend.direction = "vertical"
    )
}

```
```{r qqplot-functions-DonorPairRetention}
### functions
#function to permute outcome and record results for various model formulas
generate_permuted_null <- function(model_formula,
                                   data,
                                   interaction_term_regex = ":",
                                   permute_var = NULL,
                                   n_perm = 100,
                                   model_type = c("lmer", "glmer", "glmer.nb", "glm"),
                                   family = gaussian(),
                                   id_var = "donor",             # NEW (required for donor-aware perms)
                                   condition_var = "condition",  # NEW (for swap_within_id)
                                   genotype_var = "genotype.num",# NEW (for permute_id_labels)
                                   perm_scheme = c("swap_within_id","permute_id_labels","free"),
                                   seed = NULL) {

  if (!is.null(seed)) set.seed(seed)
  model_type  <- match.arg(model_type)
  perm_scheme <- match.arg(perm_scheme)
  model_formula_str <- paste(deparse(model_formula), collapse = " ")

  outcome_var <- if (is.null(permute_var)) all.vars(model_formula)[1] else permute_var

  # ---- helper to fit model and capture warnings ----
  fit_model_safely <- function(formula, data, model_type, family) {
    warnings <- character()
    result <- withCallingHandlers(
      tryCatch({
        fit <- switch(model_type,
          lmer     = lmerTest::lmer(formula, data = data),
          glmer    = lme4::glmer(formula, data = data, family = family),
          glmer.nb = lme4::glmer.nb(formula, data = data),
          glm      = stats::glm(formula, data = data, family = family)
        )
        list(fit = fit, warnings = warnings)
      }, error = function(e) list(fit = NULL, warnings = paste("Error:", conditionMessage(e)))),
      warning = function(w) { warnings <<- c(warnings, conditionMessage(w)); invokeRestart("muffleWarning") }
    )
    result$warnings <- paste(result$warnings, collapse = " | ")
    result
  }

  # ---- observed fit ----
  obs_model_result <- fit_model_safely(model_formula, data, model_type, family)
  if (is.null(obs_model_result$fit)) return(tibble::tibble())

  observed_results <- broom.mixed::tidy(obs_model_result$fit) %>%
    dplyr::filter(stringr::str_detect(term, interaction_term_regex)) %>%
    dplyr::mutate(type = "observed", model_formula = model_formula_str,
                  warning = obs_model_result$warnings)

  # ---- permutation generators ----
  permute_once <- function(df) {
    if (perm_scheme == "free") {
      # (your original approach — not recommended for paired design)
      df %>% dplyr::mutate(!!rlang::sym(outcome_var) := sample(!!rlang::sym(outcome_var)))
    } else if (perm_scheme == "swap_within_id") {
      # swap PBS <-> IFNG labels within donor with prob 0.5
      # (assumes two conditions per donor)
      levs <- levels(df[[condition_var]]) %||% unique(df[[condition_var]])
      stopifnot(length(levs) == 2)
      swap_levels <- function(x) ifelse(x == levs[1], levs[2], levs[1])

      df %>%
        dplyr::group_by(!!rlang::sym(id_var)) %>%
        dplyr::mutate(.swap = rbinom(1L, 1L, 0.5),
                      !!rlang::sym(condition_var) :=
                        ifelse(.swap == 1L, swap_levels(!!rlang::sym(condition_var)),
                               !!rlang::sym(condition_var))) %>%
        dplyr::ungroup() %>%
        dplyr::select(-.swap)
    } else { # permute_id_labels
      donors <- unique(df[[id_var]])
      perm_donors <- sample(donors, length(donors), replace = FALSE)
      map_tab <- dplyr::tibble(!!id_var := donors, .perm_id = perm_donors)
      geno_map <- df %>%
        dplyr::distinct(!!rlang::sym(id_var), !!rlang::sym(genotype_var)) %>%
        dplyr::rename(.perm_id = !!rlang::sym(id_var), .perm_geno = !!rlang::sym(genotype_var))
      df %>%
        dplyr::left_join(map_tab, by = setNames(id_var, id_var)) %>%
        dplyr::left_join(geno_map, by = ".perm_id") %>%
        dplyr::mutate(!!rlang::sym(genotype_var) := .perm_geno) %>%
        dplyr::select(-.perm_id, -.perm_geno)
    }
  }

  permuted_results <- purrr::map_dfr(seq_len(n_perm), function(i) {
    permuted_data <- permute_once(data)
    perm_model_result <- fit_model_safely(model_formula, permuted_data, model_type, family)
    if (is.null(perm_model_result$fit)) return(tibble::tibble())
    broom.mixed::tidy(perm_model_result$fit) %>%
      dplyr::filter(stringr::str_detect(term, interaction_term_regex)) %>%
      dplyr::mutate(type = "null", model_formula = model_formula_str,
                    warning = perm_model_result$warnings)
  })

  dplyr::bind_rows(observed_results, permuted_results)
}



plot_qq_pval_multi <- function(sim_df) {
  # Load ggrepel for smart label positioning
  library(ggrepel)
  
  # Ensure p-values are available and create model identifier
  sim_df <- sim_df %>%
    mutate(
      model_formula_family = str_c(model_formula, " ", paste0(family), " ", paste0(model_type))
    )
  
  # Calculate empirical p-values per model_formula
  empirical_pvals <- sim_df %>%
    group_by(model_formula_family) %>%
    summarise(
      observed_pval = p.value[type == "observed"],
      null_pvals = list(p.value[type == "null"]),
      empirical_pval = mean(null_pvals[[1]] <= observed_pval, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Prepare data for QQ plot (null points only)
  # Convert p-values to -log10 scale for better visualization
  qq_df <- sim_df %>%
    filter(type == "null") %>%
    group_by(model_formula_family) %>%
    arrange(p.value) %>%
    mutate(
      neg_log10_pval = -log10(p.value),
      expected_neg_log10 = -log10(ppoints(n(), a = 0))
    ) %>%
    ungroup()
  
  # Prepare observed points (one per model)
  observed_points <- empirical_pvals %>%
    mutate(
      observed_neg_log10 = -log10(observed_pval),
      # Expected value for the most extreme point
      expected_neg_log10 = -log10(0.5 / lengths(null_pvals)),
      label = paste0("p = ", signif(observed_pval, 3))
    )
  
  # Plot
  ggplot(qq_df, aes(x = expected_neg_log10, y = neg_log10_pval, color = model_formula_family)) +
    geom_point(alpha = 0.4) +
    geom_abline(slope = 1, intercept = 0, color = "gray") +
    geom_point(data = observed_points, aes(x = expected_neg_log10, y = observed_neg_log10), 
               shape = 4, size = 4, stroke = 1.5) + # "X"
    geom_text_repel(data = observed_points, 
                    aes(x = expected_neg_log10,
                        y = observed_neg_log10,
                        label = label,
                        color = model_formula_family), 
                    size = 3,
                    min.segment.length = 0,  # Always draw connecting lines
                    segment.color = "black",
                    segment.alpha = 0.6,
                    box.padding = 0.5,       # Space around labels
                    point.padding = 0.3,     # Space around points
                    force = 2,               # Repulsion strength
                    max.overlaps = Inf) +    # Allow all labels to be shown
    scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    labs(
      x = "Expected -log10(p-value)",
      y = "Observed -log10(p-value)",
      color = "Model Formula",
      title = "QQ Plot of Permuted vs Observed P-values (Multiple Models)"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      legend.direction = "vertical"
    )
}

```
### Create a grid of models and plot them together
```{r qqplot-model-grid}
# Define models we'd like to try
model_grid <- tibble(
  model_formula = list(
    as.formula("CPM ~ condition * genotype.num + (1 | donor)"),
    as.formula("CPM ~ condition * genotype.num"),
    as.formula("counts ~ condition * genotype.num"),
    as.formula("counts ~ condition * genotype.num + (1 | donor)"),
    as.formula("counts ~ condition * genotype.num + (1 | donor)"),
    as.formula("log2CPM ~ condition * genotype.num + (1 | donor)"),
    as.formula("rankNormCPM ~ condition * genotype.num + (1 | donor)")
  ),
  model_type = c("lmer", "glm", "glm", "glmer", "glmer.nb","lmer", "lmer"),
  family = list(NA, gaussian(), poisson(), poisson(), NA, NA, NA)
)
```
```{r qqplot, include=FALSE, message=FALSE, warning=FALSE}
# Run all models and return one flat dataframe
flat_qq_results <- list()  # Initialize list to collect results

for (i in seq_len(nrow(model_grid))) {
  cat("Running model", i, "...\n")

  model_formula <- model_grid$model_formula[[i]]
  model_type <- model_grid$model_type[[i]]
  family <- model_grid$family[[i]]

  # Show formula and model type for debugging
  print(model_formula)
  cat("Model type:", model_type, "\n")
  if (inherits(family, "family")) {
    cat("Family:", family$family, "\n")
  } else {
    cat("Family: NA\n")
  }

  # Run permutation test
  # this block runs the function without retaining donor genotype pairing
  # result <- generate_permuted_null(
  #   model_formula = model_formula,
  #   data = real_dat,
  #   model_type = model_type,
  #   family = family,
  #   n_perm = 500
  # )
  
  # this block runs the function that retains genotype for pair of samples from the same donor.
  result <- generate_permuted_null(
    model_formula = model_formula,
    data = real_dat,
    interaction_term_regex = ":",
    n_perm = 500,
    model_type = model_type,
    family = family,
    id_var = "donor",
    condition_var = "condition",
    genotype_var = "genotype.num",
    perm_scheme = "permute_id_labels",   # or "permute_id_labels"
    seed = 1
  )

  # Skip failed model
  if (nrow(result) == 0) {
    cat("Model", i, "returned no result. Skipping.\n")
    next
  }

  # Add model metadata
  model_metadata <- tibble(
    model_type = model_type,
    family = if (inherits(family, "family")) family$family else NA_character_
  )

  # Combine metadata + result
  final_df <- bind_cols(model_metadata, result) %>%
    select(!contains("..."))

  # Save in list
  flat_qq_results[[i]] <- final_df
}

# Combine all results into one data frame
flat_qq_results <- bind_rows(flat_qq_results)

```

First plot is all tested models
```{r}
#plot the results
plot1 <- plot_qq_pval_multi(flat_qq_results)
plot1
```
Second plot is only the non-poisson models to "zoom in"
```{r}
plot2 <- plot_qq_pval_multi(
  filter(flat_qq_results, is.na(family) | family != "poisson")
)
plot2
```
Third plot is only the CPM models to "zoom in" even more
```{r}
plot3 <- plot_qq_pval_multi(
  filter(flat_qq_results, str_detect(model_formula, "CPM"))
)
plot3
```
Fourth plot is the CPM models with different mixed effects
```{r}
plot4 <- plot_qq_pval_multi(
  flat_qq_results %>%
    filter(
      str_detect(model_formula, "CPM"),
      !str_detect(model_formula, "log2CPM|rankNormCPM")
    )
)
plot4
```
Fifth plot is only the CPM models in different transformations
```{r}
plot5 <- plot_qq_pval_multi(
  flat_qq_results %>%
    filter(
      str_detect(model_formula, "CPM"),
      str_detect(model_formula, "(1 | donor)"),
      !str_detect(model_formula, "phase")
    )
)
plot5
```
# full model for reQTL with covariates
```{r make-full-log2CPM-reQTL-data-table}
# peer factor
dir.peer <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/peer/peer_factors"
sc <- "MEL_PBS-IFNG"
peer_factors <- data.table::fread(paste0(dir.peer, "/peer_factors_",sc,".tsv"))
colnames(peer_factors)[1] <- "sample"

# genotype PC
genotype_pcs <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/genotype_PCs_for_modeling_07242025.txt") %>% 
  dplyr::select(-c("FID","donor_num"))

# Inputs
this_matrix <- left_join(df.MEL.PBSIFNG, metadata.MEL.PBSIFNG[, c("sample","biosample","phase","sequencing_platform")], by="biosample") %>%
  left_join(. , genotype, by="donor") %>%
  left_join( . , peer_factors, by="sample") %>%
  left_join( . , genotype_pcs, by="donor") %>%
  mutate(log2CPM = log2(CPM + 1)) %>%
  na.omit()
peer_names <- paste0("PEER", 1:20)
pc_names <- paste0("PC", 1:10)

this_matrix$condition <- relevel(factor(as.character(this_matrix$condition)), ref = "PBS")
this_matrix$genotype.nt <- relevel(factor(this_matrix$genotype.nt), ref = "TT")

# Output results container
results <- data.frame()

# Loop over combinations of PEERs and PCs
for (n_peer in 0:20) {
  for (n_pc in 0:10) {
    
    print(paste0(n_peer,"_",n_pc))
   # Safely select covariates
    peer_vars <- if (n_peer > 0) peer_names[1:n_peer] else character(0)
    pc_vars <- if (n_pc > 0) pc_names[1:n_pc] else character(0)
    covariates <- c(peer_vars, pc_vars)
    
    # Build formula
    # Build formula string
    covariate_formula <- if (length(covariates) > 0) paste("+", paste(covariates, collapse = " + ")) else ""
    full_formula <- as.formula(
      paste("CPM ~ condition * genotype.num", covariate_formula, "+ (1 | donor)")
    )
    
    # Fit model
    model <- tryCatch(
      lmerTest::lmer(full_formula, data = this_matrix, REML = FALSE),
      error = function(e) NULL
    )
    
    if (!is.null(model)) {
      # Extract diagnostics
      aic <- AIC(model)
      bic <- BIC(model)
      r2 <- r.squaredGLMM(model)[1, "R2m"]  # marginal R² (fixed effects) # You're evaluating models that differ only in their fixed-effect covariates (PEERs, PCs), while donor is always included as a random intercept. So: R2m tells you how much variance is explained by genotype, condition, and covariates
      
      # Extract interaction p-value
      coef_table <- summary(model)$coefficients
      interaction_row <- grep("condition.*:genotype.num", rownames(coef_table))
      interaction_p <- if (length(interaction_row) > 0) coef_table[interaction_row, "Pr(>|t|)"] else NA
      
      # Shapiro-Wilk test of residuals
      resids <- resid(model)
      shapiro_p <- if (length(resids) >= 3) shapiro.test(resids)$p.value else NA # want it to be not significant
      
      # Save residual plot (optional)
      plot_name <- paste0(Dir,"/model13_covariate_selection/residplot_peer", n_peer, "_pc", n_pc, ".png")
      png(plot_name, width = 600, height = 600)
      print(plot(model, which = 1,
                 xlab = "Fitted values",
                 ylab = "Residuals",
                 main = paste("Residuals vs Fitted (PEER:", n_peer, "PC:", n_pc, ")"),
                 col = "steelblue"))  # Set y-axis limits
      # this_matrix$fitted <- fitted(model)
      # this_matrix$resid <- resid(model)
      # this_matrix$genotype_condition <- interaction(this_matrix$genotype.num, this_matrix$condition, sep = "_")
      # 
      # p <- ggplot(this_matrix, aes(x = fitted, y = resid, color = genotype_condition)) +
      #   geom_point(alpha = 0.7) +
      #   geom_hline(yintercept = 0, linetype = "solid", color = "gray40") +
      #   labs(x = "Fitted values", y = "Residuals") +
      #   theme_minimal()
      # print(p)   
      dev.off()



      
      # Save results
      results <- rbind(results, data.frame(
        n_peer = n_peer,
        n_pc = n_pc,
        AIC = aic,
        BIC = bic,
        adjR2 = r2,
        p_interaction = interaction_p,
        shapiro_p = shapiro_p,
        formula = paste(deparse(full_formula), collapse = "")
      ))
    }
  }
}

library(grid)
library(png)
plot_folder <- paste0(Dir,"/model13_covariate_selection")
output_pdf <- paste0(Dir,"/model13_covariate_selection/residplot_by_covariates.pdf")

# List all PNGs
png_files <- list.files(plot_folder, pattern = "\\.png$", full.names = TRUE)

# Extract PEER and PC numbers from file names
file_info <- data.frame(
  file = png_files,
  n_peer = as.numeric(sub(".*peer(\\d+)_pc\\d+.*", "\\1", png_files)),
  n_pc   = as.numeric(sub(".*pc(\\d+).*", "\\1", png_files)),
  stringsAsFactors = FALSE
)

# Sort by n_peer first, then n_pc
file_info <- file_info[order(file_info$n_peer, file_info$n_pc), ]

# Create a multipage PDF
pdf(output_pdf, width = 8, height = 8)
for (f in file_info$file) {
  img <- readPNG(f)
  grid::grid.newpage()
  grid::grid.raster(img)
}
dev.off()
file.remove(png_files)

# Save results table
write.csv(results, paste0(Dir,"/model13_covariate_selection/reQTL_model_selection_results.csv"), row.names = FALSE)

# Quick summary plot
p_pval <- ggplot(results, aes(x = n_peer, y = n_pc, fill = -log10(p_interaction))) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "Interaction p-value across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "-log10(p)") +
  theme_minimal()

p_aic <- ggplot(results, aes(x = n_peer, y = n_pc, fill = AIC)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "AIC across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "AIC") +
  theme_minimal()

p_bic <- ggplot(results, aes(x = n_peer, y = n_pc, fill = BIC)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "BIC across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "BIC") +
  theme_minimal()

p_shapirop <- ggplot(results, aes(x = n_peer, y = n_pc, fill = shapiro_p)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "shapiro_p across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "shapiro_p") +
  theme_minimal()

p_adjr2 <- ggplot(results, aes(x = n_peer, y = n_pc, fill = adjR2)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "adjR2 across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "adjR2") +
  theme_minimal()
# I'm going to go with peer = 7, pc = 8
```
```{r final_reQTL_model}
n_peer=5
n_pc=2
# Build covariate list
peer_vars <- peer_names[1:n_peer]
pc_vars <- pc_names[1:n_pc]
covariates <- c(peer_vars, pc_vars)
covariate_formula <- paste(covariates, collapse = "+ ")

# fit model with interaction term
full_formula <- as.formula(
  paste("CPM ~ condition * genotype.num",
        if (!is.null(covariate_formula)) paste("+", covariate_formula),
        "+ (1 | donor)")
)
final_reQTL_model_MEL_IFNG <- lmerTest::lmer(full_formula, data = this_matrix, REML = FALSE)

# check residual normality
plot(final_reQTL_model_MEL_IFNG, which = 1)  # residuals vs fitted
qqnorm(residuals(final_reQTL_model_MEL_IFNG)); qqline(residuals(final_reQTL_model_MEL_IFNG))  # normal Q-Q plot
hist(residuals(final_reQTL_model_MEL_IFNG))  # optional

```
```{r loglik-interaction-term}
n_peer=5
n_pc=2
# Build covariate list
peer_vars <- peer_names[1:n_peer]
pc_vars <- pc_names[1:n_pc]
covariates <- c(peer_vars, pc_vars)
covariate_formula <- paste(covariates, collapse = "+ ")

# fit model with interaction term
full_formula <- as.formula(
  paste("CPM ~ condition * genotype.num",
        if (!is.null(covariate_formula)) paste("+", covariate_formula),
        "+ (1 | donor)")
)
model_with_covariates <- lmerTest::lmer(full_formula, data = this_matrix, REML = FALSE)

# Fit model without covariates
nested_formula <- "CPM ~ condition*genotype.num + (1|donor)"
model_without_covariates <- lmerTest::lmer(nested_formula, data = this_matrix, REML = FALSE)

# compare the two models
tidy(anova(model_with_covariates, model_without_covariates, test="LRT"))
```
```{r make-full-CPM-reQTL-data-table}
# testing with covariates to see whether the residuals of CPM linear model#13 can resolve the heteroskedacity issue in basic model without covariates.
# peer factor
dir.peer <- "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/peer/peer_factors"
sc <- "MEL_PBS-IFNG"
peer_factors <- data.table::fread(paste0(dir.peer, "/peer_factors_",sc,".tsv"))
colnames(peer_factors)[1] <- "sample"

# genotype PC
genotype_pcs <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/genotype_PCs_for_modeling_07242025.txt") %>% 
  dplyr::select(-c("FID","donor_num"))

# Inputs
this_matrix <- left_join(df.MEL.PBSIFNG, metadata.MEL.PBSIFNG[, c("sample","biosample","phase","sequencing_platform")], by="biosample") %>%
  left_join(. , genotype, by="donor") %>%
  left_join( . , peer_factors, by="sample") %>%
  left_join( . , genotype_pcs, by="donor") %>%
  mutate(log2CPM = log2(CPM + 1))
peer_names <- paste0("PEER", 1:20)
pc_names <- paste0("PC", 1:10)

this_matrix$condition <- relevel(factor(as.character(this_matrix$condition)), ref = "PBS")
this_matrix$genotype.nt <- relevel(factor(this_matrix$genotype.nt), ref = "TT")

# Output results container
results <- data.frame()

# Loop over combinations of PEERs and PCs
for (n_peer in 0:20) {
  for (n_pc in 0:10) {
    
    print(paste0(n_peer,"_",n_pc))
   # Safely select covariates
    peer_vars <- if (n_peer > 0) peer_names[1:n_peer] else character(0)
    pc_vars <- if (n_pc > 0) pc_names[1:n_pc] else character(0)
    covariates <- c(peer_vars, pc_vars)
    
    # Build formula
    # Build formula string
    covariate_formula <- if (length(covariates) > 0) paste("+", paste(covariates, collapse = " + ")) else ""
    full_formula <- as.formula(
      paste("CPM ~ condition * genotype.num", covariate_formula, "+ (1 | donor)")
    )
    
    # Fit model
    model <- tryCatch(
      lmerTest::lmer(full_formula, data = this_matrix, REML = FALSE),
      error = function(e) NULL
    )
    
    if (!is.null(model)) {
      # Extract diagnostics
      aic <- AIC(model)
      bic <- BIC(model)
      r2 <- r.squaredGLMM(model)[1, "R2m"]  # marginal R² (fixed effects) # You're evaluating models that differ only in their fixed-effect covariates (PEERs, PCs), while donor is always included as a random intercept. So: R2m tells you how much variance is explained by genotype, condition, and covariates
      
      # Extract interaction p-value
      coef_table <- summary(model)$coefficients
      interaction_row <- grep("condition.*:genotype.num", rownames(coef_table))
      interaction_p <- if (length(interaction_row) > 0) coef_table[interaction_row, "Pr(>|t|)"] else NA
      
      # Shapiro-Wilk test of residuals
      resids <- resid(model)
      shapiro_p <- if (length(resids) >= 3) shapiro.test(resids)$p.value else NA # want it to be not significant
      
      # Save residual plot (optional)
      plot_name <- paste0(Dir,"/model13_covariate_selection/residplot_peer", n_peer, "_pc", n_pc, ".png")
      png(plot_name, width = 600, height = 600)
      print(plot(model, which = 1,
           xlab = "Fitted values",
           ylab = "Residuals",
           main = paste("Residuals vs Fitted (PEER:", n_peer, "PC:", n_pc, ")"),
           col = "steelblue"))  # Set y-axis limits
      dev.off()
      
      # Save results
      results <- rbind(results, data.frame(
        n_peer = n_peer,
        n_pc = n_pc,
        AIC = aic,
        BIC = bic,
        adjR2 = r2,
        p_interaction = interaction_p,
        shapiro_p = shapiro_p,
        formula = paste(deparse(full_formula), collapse = "")
      ))
    }
  }
}

library(grid)
library(png)
plot_folder <- paste0(Dir,"/model13_covariate_selection")
output_pdf <- paste0(Dir,"/model13_covariate_selection/residplot_by_covariates.pdf")

# List all PNGs
png_files <- list.files(plot_folder, pattern = "\\.png$", full.names = TRUE)

# Extract PEER and PC numbers from file names
file_info <- data.frame(
  file = png_files,
  n_peer = as.numeric(sub(".*peer(\\d+)_pc\\d+.*", "\\1", png_files)),
  n_pc   = as.numeric(sub(".*pc(\\d+).*", "\\1", png_files)),
  stringsAsFactors = FALSE
)

# Sort by n_peer first, then n_pc
file_info <- file_info[order(file_info$n_peer, file_info$n_pc), ]

# Create a multipage PDF
pdf(output_pdf, width = 8, height = 8)
for (f in file_info$file) {
  img <- readPNG(f)
  grid::grid.newpage()
  grid::grid.raster(img)
}
dev.off()
file.remove(png_files)

# Save results table
write.csv(results, paste0(Dir,"/model13_covariate_selection/reQTL_model_selection_results.csv"), row.names = FALSE)

# Quick summary plot
p_pval <- ggplot(results, aes(x = n_peer, y = n_pc, fill = -log10(p_interaction))) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "Interaction p-value across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "-log10(p)") +
  theme_minimal()

p_aic <- ggplot(results, aes(x = n_peer, y = n_pc, fill = AIC)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "AIC across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "AIC") +
  theme_minimal()

p_bic <- ggplot(results, aes(x = n_peer, y = n_pc, fill = BIC)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "BIC across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "BIC") +
  theme_minimal()

p_shapirop <- ggplot(results, aes(x = n_peer, y = n_pc, fill = shapiro_p)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "shapiro_p across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "shapiro_p") +
  theme_minimal()

p_adjr2 <- ggplot(results, aes(x = n_peer, y = n_pc, fill = adjR2)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "adjR2 across PEER/PC settings\n lm() CPM ~ covariates + genotype * condition + (1|donor)", fill = "adjR2") +
  theme_minimal()
# I'm going to go with: 
```













# coloc
```{r get-gwas-summary}
# === 1) Pick the file (GRCh38 harmonised recommended) ===
url <- "https://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/GCST90446001-GCST90447000/GCST90446792/harmonised/GCST90446792.h.tsv.gz"

# === 2) Read and normalise columns ===
ss <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/erap2/GCST90446792.h.tsv.gz")

# Harmonised files typically expose both original and 'hm_*' fields. 
# Prefer harmonised fields when available.
has_hm <- "hm_variant_id" %in% names(ss)

# Column selectors with fallbacks:
col_snp  <- if (has_hm) "hm_variant_id" else "rsid"
col_chr  <- "chromosome"
col_pos  <- "base_pair_location"
col_p    <- if ("p-value" %in% names(ss)) "p-value" else if ("p_value" %in% names(ss)) "p_value" else "p"
col_eaf  <- if (has_hm && "hm_effect_allele_frequency" %in% names(ss)) "hm_effect_allele_frequency" else "effect_allele_frequency"

need <- c(col_snp, col_chr, col_pos, col_p, col_eaf)
stopifnot(all(need %in% names(ss)))

data.table::setnames(ss, old = need,
              new = c("snp","chr","pos","p","EAF"), skip_absent = TRUE)

# Coerce chr to "chrN" or bare integer consistently (coloc doesn’t care either way)
# Here we keep it numeric-like:
ss[, chr := as.character(chr)]

# Drop missing essentials
ss <- ss[!is.na(p) & !is.na(pos) & !is.na(chr)]

# === Define ERAP2 window ===
# If your reQTL is GRCh38, these are GRCh38 coords; adjust padding as you like.
library(biomaRt)
mart <- useEnsembl("genes", dataset = "hsapiens_gene_ensembl")  # or GRCh=37
g <- getBM(attributes = c("hgnc_symbol","chromosome_name","start_position","end_position"),
           filters = "hgnc_symbol", values = "ERAP2", mart = mart)
chr <- paste0("chr", g$chromosome_name[1])
pad         <- 500000
w_start <- g$start_position[1] - pad
w_end   <- g$end_position[1] + pad

gwas_sub <- ss[ss$chr == erap2_chr & ss$pos >= w_start & ss$pos <= w_end,]
write(unique(gwas_sub$snp), "~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/snps.txt")
# out of 8899 GWAS SNPs, there are 1037 that have data in our genotype data.

# Derive MAF from EAF conservatively
#ss[, MAF := pmin(ss$EAF, 1 - ss$EAF)]

# Calculate MAF using my donor's data
# in cluster use PLINK

# Bring MAFs into R and merge with my GWAS data
# 1) Read MAFs from PLINK
maf <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/erap2_maf.tsv",
             col.names = c("CHR","POS","ID","REF","ALT","AN","AC","MAF"))

gwas_sub <- inner_join(gwas_sub, maf, by = c("snp"="ID")) %>% 
  dplyr::select(c("snp","p","MAF")) %>%
  na.omit() %>%
  dplyr::filter(!is.na(MAF) & MAF>0 & MAF<1) # 8899 -> 2812

# To remove duplicates while keeping the first occurrence:
data.table::setDT(gwas_sub)
data.table::setorder(gwas_sub, -MAF)
gwas_sub <- gwas_sub[!duplicated(gwas_sub$snp)] # 2812 -> 2783

# === 4) Add study-level N and case proportion s ===
# GWAS Catalog harmonised files often don’t carry per-variant N; use study totals.
# Fill in from the paper/GWAS Catalog study page (cases + controls).
N_cases    <- 20873   # <- SET THIS
N_controls <- 346719   # <- SET THIS
N_total    <- N_cases + N_controls
s_prop     <- if (N_total > 0) N_cases / N_total else NA_real_

# Build coloc input list using the p-value interface
gwas <- list(
  type    = "cc",
  snp     = gwas_sub$snp,
  pvalues = gwas_sub$p,
  MAF     = gwas_sub$MAF,
  N       = N_total,
  s       = s_prop
)
```
```{r reQTL-summary-for-coloc}
# simulate reQTL sumstats matched to SNP list
snps <- gwas_sub$hm_rsid
maf_dt <- gwas_sub$MAF
N_reqtl <- 124
# 4) Which SNPs you want to truly associate in the sim (can be empty)
causal_snps <- c("rs2910686")  # or character(0) for pure null
# 5) Effect size scale for causal SNP(s) (on standardized phenotype)
effect_sd <- 0.10  # larger -> stronger signals
# --- Build table and merge MAF ---
reqtl <- data.frame(snp=gwas_sub$hm_rsid, 
                       MAF=gwas_sub$MAF)

# Clean up MAF edge cases (1037 --> 950)
reqtl <- reqtl[!is.na(reqtl$MAF) & reqtl$MAF > 0 & reqtl$MAF < 1,]

# --- Simulate betas/SE and p-values ---
# For a simple linear model with standardized Y:
# var(beta_hat) ≈ 1 / (2 * N * MAF * (1 - MAF))
library(data.table)
setDT(reqtl)
reqtl$varbeta <- 1 / (2 * N_reqtl * reqtl$MAF * (1 - reqtl$MAF))
reqtl$se <- sqrt(reqtl$varbeta)

# Draw effects: zero for non-causal, N(0, effect_sd^2) for causal SNPs
reqtl[, beta := 0]
reqtl[snp %in% causal_snps, beta := rnorm(.N, mean = 0, sd = effect_sd)]

# Get Z and p
reqtl[, z := beta / se]
reqtl[, p := 2 * pnorm(-abs(z))]

# If you have beta_hat and se_hat from your reQTL interaction term:
s <- summary(final_reQTL_model_MEL_IFNG)
coef_table <- s$coefficients
beta_hat <- coef_table["conditionIFNG:genotype.num", "Estimate"]
se_hat   <- coef_table["conditionIFNG:genotype.num", "Std. Error"]
pval     <- coef_table["conditionIFNG:genotype.num", "Pr(>|t|)"]

# Update the row and recompute p
reqtl[snp == "rs2910686", `:=`(beta = beta_hat, se = se_hat, z = beta_hat/se_hat,
                               p = pval)]

# OR if you only have a p-value (no beta/se):
# real_p <- 3.2e-8
# reqtl[snp == "rs2910686", `:=`(p = real_p)]

# make a coloc-ready list right away (p-value interface) ---
D2_sim <- list(
  type    = "quant",
  snp     = reqtl$snp,
  pvalues = reqtl$p,
  MAF     = reqtl$MAF,
  N       = N_reqtl
)
```
```{r reQTL-run-model-ERAP2-locus}
# /pi/manuel.garber-umw/human/skin/eQTLs/donor_genotyping/hg38/erap2_subset.vcf.gz

# genotype table
genotype_table <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/erap2/ERAP2_snps.bed")
genotype_table <- genotype_table %>% 
  dplyr::filter(ID %in% gwas[["snp"]]) %>% # 961 snps also have GWAS results
  distinct(ID, .keep_all=TRUE)
colnames(genotype_table) <- gsub("F0","F", gsub(":GT", "", gsub("skineQTL-","",colnames(genotype_table))))


# expression table
this_df = df.MEL.PBSIFNG
this_metadata = metadata.MEL.PBSIFNG
this_peerfactors = peer_factors
  
pval_vector = c()
term <- "conditionIFNG:genotype.num"   # the interaction you care about
for (i in 1:nrow(genotype_table)) {
  print(i)
  snpinfo <- genotype_table[i, ] %>% t() %>%
    as.data.frame() %>%
    rownames_to_column("donor") %>%
    set_colnames(c("donor","GT"))
  REF <- snpinfo$GT[5]
  ALT <- snpinfo$GT[6]
  snpid <- snpinfo$GT[4]
  
  genotype <- snpinfo[7:nrow(snpinfo), ]
  genotype <- genotype %>%
    mutate(
      genotype.num = case_when(
        GT == "0|0" ~ 0,
        GT %in% c("0|1", "1|0") ~ 1,
        GT == "1|1" ~ 2,
        TRUE ~ NA_real_
      ),
      genotype.nt = case_when(
        GT == "0|0" ~ paste0(REF,"|",REF),
        GT == "0|1" ~ paste0(REF,ALT,"|",ALT,REF),
        GT == "1|0" ~ paste0(REF,ALT,"|",ALT,REF),
        GT == "1|1" ~ paste0(ALT,"|",ALT),
        TRUE ~ NA_character_
      )
    )
  
  # compile data matrix
  this_matrix <- left_join(this_df, this_metadata[, c("sample","biosample","phase","sequencing_platform")], by="biosample") %>%
    left_join(. , genotype, by="donor") %>%
    left_join( . , peer_factors, by="sample") %>%
    left_join( . , genotype_pcs, by="donor") %>%
    mutate(log2CPM = log2(CPM + 1)) %>%
    na.omit()
  this_matrix$condition <- relevel(factor(as.character(this_matrix$condition)), ref = "PBS")
  
  # add covariates
  n_peer=5
  n_pc=2
  # Build covariate list
  peer_vars <- peer_names[1:n_peer]
  pc_vars <- pc_names[1:n_pc]
  covariates <- c(peer_vars, pc_vars)
  covariate_formula <- paste(covariates, collapse = "+ ")
  
  # fit model with interaction term
  full_formula <- as.formula(
    paste("CPM ~ condition * genotype.num",
          if (!is.null(covariate_formula)) paste("+", covariate_formula),
          "+ (1 | donor)")
  )

  fit <- lmerTest::lmer(full_formula, data = this_matrix, REML = FALSE)
  tab <- summary(fit)$coefficients
  term <- "conditionIFNG:genotype.num"
  if (term %in% rownames(tab)) {
    pval <- tab[term, "Pr(>|t|)"]
  } else {
    pval <- NA_real_
  }
  pval_vector <- c(pval_vector, pval)
}

# compile output table
reqtl_res <- data.frame(snp=genotype_table$ID, pval=pval_vector) %>%
  na.omit() %>%
  left_join( . , gwas_sub[,c("snp","MAF")], by="snp")

# make a coloc-ready list right away (p-value interface) ---
N_reqtl <- nrow(this_matrix)
eqtl <- list(
  type    = "quant",
  snp     = reqtl_res$snp,
  pvalues = reqtl_res$pval,
  MAF     = reqtl_res$MAF,
  N       = N_reqtl
  )
```
```{r coloc}
library(coloc)
# https://chr1swallace.github.io/coloc/

##### harmonize gwas and eQTL data #####
common_snps <- intersect(gwas$snp, eqtl$snp)
idx_gwas <- gwas$snp %in% common_snps
idx_eqtl <- eqtl$snp %in% common_snps
# 3) Subset using those indices
gwas$snp     <- gwas$snp[idx_gwas]
gwas$pvalues <- gwas$pvalues[idx_gwas]
gwas$MAF     <- gwas$MAF[idx_gwas]

eqtl$snp       <- eqtl$snp[idx_eqtl]
eqtl$pvalues   <- eqtl$pvalues[idx_eqtl]
eqtl$MAF       <- eqtl$MAF[idx_eqtl]

# 4) Put both in the SAME order
ord_gwas <- match(common_snps, gwas$snp)
ord_eqtl <- match(common_snps, eqtl$snp)

gwas$snp     <- gwas$snp[ord_gwas]
gwas$pvalues <- gwas$pvalues[ord_gwas]
gwas$MAF     <- gwas$MAF[ord_gwas]

eqtl$snp       <- eqtl$snp[ord_eqtl]
eqtl$pvalues   <- eqtl$pvalues[ord_eqtl]
eqtl$MAF       <- eqtl$MAF[ord_eqtl]

# 5) Final sanity checks
stopifnot(identical(gwas$snp, eqtl$snp))
stopifnot(length(gwas$snp) == length(gwas$pvalues),
          length(gwas$snp) == length(gwas$MAF),
          length(eqtl$snp) == length(eqtl$pvalues),
          length(eqtl$snp) == length(eqtl$MAF))

# (optional) drop any NA / out-of-range MAFs or pvals
keep <- is.finite(gwas$MAF) & gwas$MAF>0 & gwas$MAF<1 &
        is.finite(eqtl$MAF) & eqtl$MAF>0 & eqtl$MAF<1 &
        is.finite(gwas$pvalues) & is.finite(eqtl$pvalues)
gwas$snp     <- gwas$snp[keep]
gwas$pvalues <- gwas$pvalues[keep]
gwas$MAF     <- gwas$MAF[keep]
eqtl$snp       <- eqtl$snp[keep]
eqtl$pvalues   <- eqtl$pvalues[keep]
eqtl$MAF       <- eqtl$MAF[keep]

##### choose priors #####
#Defaults are often fine, but for cis molecular traits a common choice is a bit more liberal on colocalization:
#p1 = prior a SNP is associated with trait 1 (GWAS), e.g. 1e-4
#p2 = prior a SNP is associated with trait 2 (eQTL/reQTL), e.g. 1e-4
#p12 = prior it’s associated with both, e.g. 1e-5 (some use 1e-6—you can sensitivity-check)
#res <- coloc.abf(gwas, D2_sim, p1=1e-4, p2=1e-4, p12=1e-5) # this uses simulated data with just rs2910686 as true signal
res <- coloc.abf(gwas, eqtl, p1=1e-4, p2=1e-4, p12=1e-5)
res$summary
#Interpretation (the big five):
#PP.H0: no association with either
#PP.H1: GWAS only
#PP.H2: reQTL only
#PP.H3: both, different causal variants
#PP.H4: both, shared causal variant (this is what you want high)
#Rules of thumb: PP.H4 > 0.8 is strong; 0.5–0.8 suggestive; also look at PP.H3.

##### Inspect the candidate SNPs #####
hits <- as.data.table(res$results)[order(-SNP.PP.H4)]
head(hits[, .(snp, SNP.PP.H4, pvalues.df1 = p.gwas, pvalues.df2 = p.eqtl)])

##### plot locus-compare #####
m <- merge(
  data.table(snp = gwas$snp, p_gwas = gwas$pvalues),
  data.table(snp = eqtl$snp, p_reqtl = eqtl$pvalues),
  by = "snp"
)

ggplot(m, aes(-log10(p_gwas), -log10(p_reqtl))) +
  geom_point(alpha = 0.6) +
  labs(x = "-log10 p (GWAS)", y = "-log10 p (reQTL)") +
  theme_minimal()

##### plot manhattan plot #####
# Example data frame for GWAS
df_gwas <- data.frame(
  snp = gwas$snp,
  pos = genotype_table[genotype_table$ID%in%gwas$snp,]$START,  # if IDs are chr:pos, extract pos
  pval = gwas$pvalues
)

# Example data frame for eQTL
df_eqtl <- data.frame(
  snp = eqtl$snp,
  pos = genotype_table[genotype_table$ID%in%eqtl$snp,]$START,
  pval = eqtl$pvalues
)

# GWAS Manhattan (top)
p1 <- ggplot(df_gwas, aes(x = pos, y = -log10(pval))) +
  geom_point(size = 1) +
  theme_bw() +
  labs(x = NULL, y = "-log10 p (GWAS)") +
  theme(axis.text.x = element_blank())

# eQTL Manhattan (bottom)
p2 <- ggplot(df_eqtl, aes(x = pos, y = -log10(pval))) +
  geom_point(size = 1, color = "firebrick") +
  theme_bw() +
  labs(x = "Genomic position", y = "-log10 p (reQTL)")

# Combine vertically
library(cowplot)
combined <- plot_grid(p1, p2, ncol = 1, align = "v", axis = "lr")
title <- ggdraw() + 
  draw_label("GWAS vs reQTL Manhattan Plots for ERAP2 locus", fontface = 'bold', x = 0.5, hjust = 0.5) 
final_plot <- plot_grid(title, combined, ncol = 1, rel_heights = c(0.06, 1))
final_plot

##### plot coloc with Manhattan Plot with LD colors #####
lead_snp_gwas <- gwas$snp[which.min(gwas$pvalues)]
lead_snp_eqtl <- eqtl$snp[which.min(eqtl$pvalues)]
# plink --vcf erap2_subset.vcf.gz \
#       --r2 gz \
#       --ld-snp rs2910686 \
#       --ld-window-kb 2000 --ld-window 999999 --ld-window-r2 0 \
#       --out ld_rs2910686
ld <- data.table::fread(paste0("~/Downloads/nl/human/skin/eQTLs/donor_genotyping/hg38/erap2/ld_",lead_snp_eqtl,".ld.gz"))[, .(snp = SNP_B, R2)]

# gwas/eqtl should already be intersected and aligned; add numeric pos
df_gwas <- data.frame(
  snp = gwas$snp,
  pos = genotype_table[genotype_table$ID%in%gwas$snp,]$START,  # if IDs are chr:pos, extract pos
  pval = gwas$pvalues
)

# Example data frame for eQTL
df_eqtl <- data.frame(
  snp = eqtl$snp,
  pos = genotype_table[genotype_table$ID%in%eqtl$snp,]$START,
  pval = eqtl$pvalues
)
# merge LD
df_gwas <- df_gwas %>% left_join(ld, by="snp")
df_eqtl <- df_eqtl %>% left_join(ld, by="snp")

# set lead to R2=1; NA -> 0
df_gwas$R2[is.na(df_gwas$R2) & df_gwas$snp==lead_snp_gwas] <- 1
df_eqtl$R2[is.na(df_eqtl$R2) & df_eqtl$snp==lead_snp_eqtl] <- 1
df_gwas$R2[is.na(df_gwas$R2)] <- 0
df_eqtl$R2[is.na(df_eqtl$R2)] <- 0

# LD bins (shared legend)
breaks <- c(0, .2, .4, .6, .8, 1)
labels <- c("0–0.2","0.2–0.4","0.4–0.6","0.6–0.8","0.8–1")
df_gwas$ld_bin <- cut(df_gwas$R2, breaks, labels, include.lowest=TRUE)
df_eqtl$ld_bin <- cut(df_eqtl$R2, breaks, labels, include.lowest=TRUE)

# x-range lock (so panels align)
xr <- range(c(df_gwas$pos, df_eqtl$pos), na.rm=TRUE)
#xr <- range(c(w_start, w_end), na.rm=TRUE)
ld_cols <- c("0–0.2"="#2166ac","0.2–0.4"="#4393c3","0.4–0.6"="#92c5de",
             "0.6–0.8"="#f4a582","0.8–1"="#d6604d")

p1 <- ggplot(df_gwas, aes(pos, -log10(pval), color=ld_bin)) +
  geom_point(size=1.5) +
  scale_color_manual(values=ld_cols, na.value="grey70", name=expression(r^2)) +
  coord_cartesian(xlim=xr) +
  labs(y="-log10 p (GWAS)", x=NULL) +
  theme_bw() +
  theme(legend.position="top", axis.text.x=element_blank())

p1 <- p1 +
  geom_point(data=subset(df_gwas, snp==lead_snp_gwas),
             size=2.5, shape=21, stroke=0.8, color="black", fill="gold") +
  geom_label_repel(data=subset(df_gwas, snp==lead_snp_gwas),
                   aes(label=snp), nudge_y=0.8, fill="gold", size=3,
                   color="black", label.size=0.2, seed=1)

p2 <- ggplot(df_eqtl, aes(pos, -log10(pval), color=ld_bin)) +
  geom_point(size=1.5) +
  scale_color_manual(values=ld_cols, na.value="grey70", name=expression(r^2)) +
  coord_cartesian(xlim=xr) +
  labs(y="-log10 p (reQTL)", x="Genomic position") +
  theme_bw() +
  theme(legend.position="none")

p2 <- p2 +
  geom_point(data=subset(df_eqtl, snp==lead_snp_eqtl),
             size=2.5, shape=21, stroke=0.8, color="black", fill="gold") +
  geom_label_repel(data=subset(df_eqtl, snp==lead_snp_eqtl),
                   aes(label=snp), nudge_y=0.8, fill="gold", size=3,
                   color="black", label.size=0.2, seed=2)

# generate track for the region
mart <- useEnsembl("genes", dataset = "hsapiens_gene_ensembl")  # or GRCh=37
g <- getBM(attributes = c("hgnc_symbol","chromosome_name","start_position","end_position"),
           filters = "hgnc_symbol", values = "ERAP2", mart = mart)
chr <- g$chromosome_name[1]
start <- min(c(df_gwas$pos, df_eqtl$pos))
end   <- max(c(df_gwas$pos, df_eqtl$pos))
genes <- getBM(attributes=c("hgnc_symbol","chromosome_name","start_position","end_position","strand"),
               filters=c("chromosome_name","start","end"),
               values=list(chr, start, end),
               mart=mart)
setDT(genes) 
genes <- genes[genes$hgnc_symbol!="", , drop=FALSE]
genes[, row := (seq_len(.N) - 1L) %% 3L]
y0 <- 0
step <- 0.25
genes[, `:=`(y = y0 + row*step, mid = (start + end)/2)]




combined <- plot_grid(p1, p2, ncol=1, align="v", axis="lr",
                      rel_heights=c(1,1,0.45))

title <- ggdraw() + draw_label(
  "Chron's Disease GWAS vs MEL IFNG reQTL at ERAP2 \nwith LD coloring r.s.t. lead reQTL",
  fontface='bold', x=.5, hjust=.5, size=8
)

final_plot <- plot_grid(title, combined, ncol=1, rel_heights=c(0.06,1))
final_plot

##### If multiple signals exist → use SuSiE colocalization #####
# If you suspect >1 causal variant in the window, switch to coloc.susie(). You’ll need approximate betas and variances. From p-values you can back-out Z and then beta/se if you have MAF & N:
# If you have actual betas/SEs for both studies, use those—much better than back-calculations.

# Approximate z from two-sided p and sign (if you have effect directions):
q2z <- function(p, sign=1) sign * qnorm(p/2, lower.tail=FALSE)

# For a quantitative trait with standardized phenotype:
# varbeta ≈ 1 / (2 * N * MAF * (1 - MAF)
varbeta <- function(N, maf) 1 / (2 * N * maf * (1 - maf))

# Build susie inputs:
D2_susie <- list(
  snp     = reqtl_sub$snp,
  beta    = q2z(reqtl_sub$p) * sqrt(varbeta(reqtl_sub$N[1], reqtl_sub$MAF)), # assumes sign=+
  varbeta = varbeta(reqtl_sub$N[1], reqtl_sub$MAF),
  type    = "quant",
  MAF     = reqtl_sub$MAF
)
# Do similar for GWAS (type="cc" is trickier; consider using published betas/SEs if available).


# QC checklist (don’t skip)
#Same build (hg19 vs hg38)
#Same rsIDs/positions
#Remove/handle palindromic A/T or C/G with high MAF
#Reasonable MAF source (study or matched ancestry reference)
#N and (for GWAS) s are correct
#Sensitivity-check priors (p12 especially)
#Consider trying ±250 kb vs ±500 kb windows; see if results are stable
```






















