---
title: "barcode_swap_investigation"
author: "Crystal Shan"
date: "2025-06-09"
output: html_document
---


```{r package, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(gridExtra)
library(ggpubr)
library(ComplexHeatmap)
library(cowplot)
library(sva)
library(clusterProfiler)
library(plotly)
library(Mfuzz)
library(Biobase)


Dir="~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1"
setwd(Dir)
```

```{r functions}
# useful function
convert_meta_genes <- function(gene_vector) {
  # Identify meta genes in the input vector
  meta_genes <- gene_vector[grepl("meta", gene_vector)]
  non_meta_genes <- gene_vector[!grepl("meta", gene_vector)]
  
  # Convert meta genes to actual gene names using metagene_dict
  meta_to_gene <- metagene_dict$name[metagene_dict$meta %in% meta_genes]
  
  # Combine non-meta genes and converted meta genes, ensuring uniqueness
  converted_genes <- unique(c(non_meta_genes, meta_to_gene))
  
  return(converted_genes)
}
niceKmeans <-function (df, k, kmeans.algorithm = "Lloyd", iter.max = 1000, nstart=100) {
  
  result = data.frame();
  source = t(scale(t(df)));
  kmeans = kmeans(source, centers = k, iter.max = iter.max, algorithm=kmeans.algorithm, nstart=nstart);
  clustered = data.frame();
  breaks = c();
  for (i in 1:k) {
    print(sprintf("Doing cluster %i",i))
    cluster = source[kmeans$cluster==i,];
    rows = row.names(cluster);
    clust = heatmap.2(as.matrix(cluster), distfun=function(x) as.dist((1-cor(t(x)))/2))
    cluster.ordered = cluster[clust$rowInd,];
    cluster.ordered.genes = rows[clust$rowInd];
    row.names(cluster.ordered) = cluster.ordered.genes;
    class = data.frame(row.names=cluster.ordered.genes);
    class[,"class"] = i 
    cluster.ordered = cbind(cluster.ordered, class);
    clustered = rbind(clustered, cluster.ordered);
    if(i > 1 & i < k) {
      breaks[i] = as.numeric(breaks[i-1]) + length(rows);
    } else if(i==1) {
      breaks[i] = length(rows);
    }
  }
  
  result <- list();
  result$clustered <- clustered;
  result$breaks <- breaks;
  return(result);
}
create_table <- function(Dir, file_lst) {
  f=file_lst[1]
  sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
  df <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
    tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
    mutate(sum=rowSums(select_if(., is.numeric))) %>%
    dplyr::select(c("region",sum)) %>% 
    magrittr::set_colnames(c("region",sample))
  
  if (length(file_lst) > 1) {
    for (f in file_lst[2:length(file_lst)]) {
      sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
        tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
        mutate(sum=rowSums(select_if(., is.numeric))) %>%
        dplyr::select(c("region",sum)) %>% 
        magrittr::set_colnames(c("region",sample))
    
    df <- dplyr::left_join(df,df.this,by="region")
    }
  }
  df <- df %>% tibble::column_to_rownames( . , "region")
  return(df)
  rm(f,sample,df,df.this)
}
```

### Overview
In Carol's notebook. When Yuming diluted the CS2 barcodes, she mislabeled 11s and 12s. Carol swapped them correctly. But she isn't sure if 11s is actually 11s anymore because her demultiplexing is giving her issues when she enters 11s as 11s, but could demultiplex when she enters 11s sequence as 12s.

my strategy: all my skin-eQTL pools have 11s and 12s. 03312025 prepped Celseq2 libraries have at least 3 pools where 11s and 12s correspond to a sample in PBS condition and another sample in IFNG condition. I will check key genes that are know to be only stimulated by IFNG, and also do correlation analyses to narrow down if 11s and 12s are mislabeled.
pool 12: 11s F122_FRB_PBS; 12s F133_FRB_IFNG
pool 18: 11s F77_KRT_PBS; 12s F77_KRT_IFNG
pool 19: 11s F92_KRT_IFNG; 12s F93_KRT_PBS

~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-12/report10956/ESAT
~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-18/report10963/ESAT
~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-19/report10964/ESAT

pool 12 seems like 11s and 12s are labeled correctly.
pool 18: donor F77 samples behaves very differently from other donors in heatmap. maybe F77 is bad.
pool 19: 11s and 12s are labeled correctly.

pool1 - 8 may have also had mislabeling because of pipetting error
just checked all the pools. pool1-8 original labeling is correct. 

## load ESAT count table for all samples
```{r load-skineQTL3ctk-pool12}
Dir.esat="~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-8_altBC/report10951/ESAT"

# 1. import ESAT counts into master table
flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
  tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
colnames(counts) <- gsub("h","",colnames(counts))
counts <- apply(counts[rowSums(counts)>=10,],2,round,0) %>% as.data.frame()
dim(counts)

# remove mitochondria genes
MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
counts <- counts %>% mutate(gene=rownames(.)) %>% 
  dplyr::filter(!rownames(.) %in% MTgenes$V1) %>% dplyr::select(-"gene")
dim(counts)

# load meta genes to gene name dictionary
metagene_dict <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/metaidname.txt") %>% dplyr::select(-id) %>% distinct()
metagene_dict_collapsed <- metagene_dict %>% group_by(meta) %>% 
  summarize(source=paste(name,collapse=','))

# metadata
sample_list <- colnames(counts) %>% gsub("^[ACGT]{6}_", "", . ) %>% gsub("_3ctk_S1mod","",.)
#metadata <- data.frame(sample = colnames(counts), tosplit = colnames(counts)) %>%
#  tidyr::separate(., col=tosplit, into=c("CS2","donor","celltype","condition",NA,"rep"),sep="_")
metadata <- data.frame(sample = colnames(counts), tosplit = colnames(counts)) %>%
  tidyr::separate(., col=tosplit, into=c("CS2",NA,"donor","celltype","condition",NA,"rep"),sep="_")
metadata$condition <- ordered(metadata$condition, levels=c("PBS","IFNG","IFNB","TNF"))
metadata$celltype <- ordered(metadata$celltype, levels=c("FRB","KRT","MEL"))

## Calculate CPM
col.sum=apply(counts,2,sum)
CPM=sweep(counts, 2, col.sum, `/`)*1000000
```

## check expressed immune response genes across samples
```{r immuneGenes}
library(clusterProfiler)
library(org.Hs.eg.db)
GO_table <- bitr(
  geneID = keys(org.Hs.eg.db, keytype = "GO"),
  fromType = "GO",
  toType = c("SYMBOL", "ENSEMBL"),
  OrgDb = org.Hs.eg.db
)

# Filter for GO:0006955: immune response. need to be our expressed gene table too.
immune_genes <- unique(GO_table[which(GO_table$GO == "GO:0006955"), "SYMBOL"])
df <- apply(counts[rowSums(counts)>=10,],2,round,0) %>% as.data.frame()
immune_genes <- intersect(immune_genes, rownames(df))
```
## heatmap (samples not clustered)
```{r heatmap-immuneGenes-fixedColumnOrder}
# make heatmap for the 202 immune genes. first not clustered
# order indices & log2CPM
ordered_indices <- order(metadata$celltype, metadata$condition, metadata$donor)
metadata.ordered <- metadata[ordered_indices, ]
mat_log2CPM <- log2(CPM[immune_genes,colnames(counts)] + 1)
log2CPM_ordered <- mat_log2CPM[,metadata.ordered$sample]

# perform k-means-clustering to determine k_row.
df <- log2CPM_ordered

# niceKmeans
km_res <- niceKmeans(df, 3)
split_instructions <- c(rep(1,km_res$breaks[1]), 
                        rep(2,km_res$breaks[2]-km_res$breaks[1]), 
                        rep(3,nrow(km_res$clustered)-km_res$breaks[2]))
zlogCPM.heatmap <- km_res$clustered %>% dplyr::select(-class)

# heatmap
# column_ha = HeatmapAnnotation(
#   celltype = colnames(zlogCPM.heatmap) %>% 
#     lapply( . , function(x) unlist(strsplit(x , "_", ))[3]) %>% unlist(),
#   condition = colnames(zlogCPM.heatmap) %>% 
#     lapply( . , function(x) unlist(strsplit(x , "_", ))[4]) %>% unlist(),
#   col = list(celltype = c("FRB" = "grey", "KRT" = "blue", "MEL" = "brown"),
#              condition = c("PBS" = "black", "IFNG" = "#dbfc03", 
#                            "IFNB" = "#fc036f", "TNF" = "#228B22")),
#   annotation_name_side = "left")

column_ha = HeatmapAnnotation(
  celltype = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[4]) %>% unlist(),
  condition = colnames(zlogCPM.heatmap) %>% 
    lapply( . , function(x) unlist(strsplit(x , "_", ))[5]) %>% unlist(),
  col = list(celltype = c("FRB" = "grey", "KRT" = "blue", "MEL" = "brown"),
             condition = c("PBS" = "black", "IFNG" = "#dbfc03", 
                           "IFNB" = "#fc036f", "TNF" = "#228B22")),
  annotation_name_side = "left")

pdf(paste0(Dir,"/analysis/heatmap_immuneGenes_log2CPM_skineQTL3ctk_pool8_altBC_06092025.pdf"))
p <- draw(Heatmap(as.matrix(zlogCPM.heatmap), 
        name = "z-score log2CPM",
        top_annotation = column_ha,
        column_title = "immune response genes in skin-eQTL-3ctk samples pool8_altBC\n GO: 0006955",
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        column_split = c(rep(1,length(which(grepl("PBS",colnames(counts))))),
                         rep(2,length(which(grepl("IFNG",colnames(counts))))),
                         rep(3,length(which(grepl("IFNB",colnames(counts))))),
                         rep(4,length(which(grepl("TNF",colnames(counts)))))),
        row_split = split_instructions, 
        show_row_names = TRUE,
        row_title_rot = 0,
        column_names_rot = 45,
        row_gap = unit(0, "mm"),
        border = TRUE,
        column_names_gp = gpar(fontsize = 5),
        row_names_gp = gpar(fontsize = 2),
        column_title_gp = gpar(fontsize = 5, fontface = "bold"),
        row_title_gp = gpar(fontsize = 5, fontface = "bold")))
dev.off()
```
### plot individual genes
GBP1, ISG15, MX1, IRF7, IRF9, IFIT1
```{r plot-per-gene}
gene = "DHX58"
# df <- CPM[gene,] %>% t() %>% as.data.frame() %>%
#   rownames_to_column("sample") %>%
#   set_colnames(c("sample","cpm")) %>%
#   tidyr::separate( . , col=sample, into=c("CS2","donor","celltype","condition",NA,NA), sep="_",remove=FALSE)
df <- CPM[gene,] %>% t() %>% as.data.frame() %>%
  rownames_to_column("sample") %>%
  set_colnames(c("sample","cpm")) %>%
  tidyr::separate( . , col=sample, into=c("CS2",NA,"donor","celltype","condition",NA,NA), sep="_",remove=FALSE)
df$condition = ordered(df$condition, levels=c("PBS","IFNG","IFNB","TNF"))

df$color <- ifelse(df$CS2 == "AGGATC", "red", 
                   ifelse(df$CS2 == "AGTGCA", "blue", "black"))

# Plot
p <- ggplot(df, aes(x = condition, y = cpm, color = color)) +
  geom_point() +
  geom_text(data = subset(df, CS2 %in% c("AGGATC", "AGTGCA")), 
            aes(label = sample), 
            vjust = -0.5, 
            size = 2, 
            show.legend = FALSE) +
  labs(title = gene,
       y = "Expression Level CPM") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~celltype, scales = "free", ncol = 7) +
  scale_color_identity()


pdf(paste0(Dir,"/analysis/",gene,"_pool8_altBC.pdf"))
p
dev.off()
```
