---
title: "compare_100bp_v_40bp_v_56bp"
author: "Crystal Shan"
date: "2025-06-16"
output: html_document
---

```{r package, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(gridExtra)
library(ggpubr)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
library(sva)
library(clusterProfiler)
library(plotly)
library(Mfuzz)
library(Biobase)


Dir="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fq_read_length_comparison"
setwd(Dir)

# load meta genes to gene name dictionary
metagene_dict <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/literature/metaidname.txt") %>% 
  dplyr::select(-id) %>% distinct()
metagene_dict_collapsed <- metagene_dict %>% group_by(meta) %>% 
  summarize(source=paste(name,collapse=','))

#save.image(file=paste0(Dir,'/myEnvironment_main.RData'))
load(paste0(Dir,'/myEnvironment_main.RData'))
```

```{r functions}
# useful function
convert_meta_genes <- function(gene_vector) {
  # Identify meta genes in the input vector
  meta_genes <- gene_vector[grepl("meta", gene_vector)]
  non_meta_genes <- gene_vector[!grepl("meta", gene_vector)]
  
  # Convert meta genes to actual gene names using metagene_dict
  meta_to_gene <- metagene_dict$name[metagene_dict$meta %in% meta_genes]
  
  # Combine non-meta genes and converted meta genes, ensuring uniqueness
  converted_genes <- unique(c(non_meta_genes, meta_to_gene))
  
  return(converted_genes)
}
niceKmeans <-function (df, k, kmeans.algorithm = "Lloyd", iter.max = 1000, nstart=100) {
  
  result = data.frame();
  source = t(scale(t(df)));
  kmeans = kmeans(source, centers = k, iter.max = iter.max, algorithm=kmeans.algorithm, nstart=nstart);
  clustered = data.frame();
  breaks = c();
  for (i in 1:k) {
    print(sprintf("Doing cluster %i",i))
    cluster = source[kmeans$cluster==i,];
    rows = row.names(cluster);
    clust = heatmap.2(as.matrix(cluster), distfun=function(x) as.dist((1-cor(t(x)))/2))
    cluster.ordered = cluster[clust$rowInd,];
    cluster.ordered.genes = rows[clust$rowInd];
    row.names(cluster.ordered) = cluster.ordered.genes;
    class = data.frame(row.names=cluster.ordered.genes);
    class[,"class"] = i 
    cluster.ordered = cbind(cluster.ordered, class);
    clustered = rbind(clustered, cluster.ordered);
    if(i > 1 & i < k) {
      breaks[i] = as.numeric(breaks[i-1]) + length(rows);
    } else if(i==1) {
      breaks[i] = length(rows);
    }
  }
  
  result <- list();
  result$clustered <- clustered;
  result$breaks <- breaks;
  return(result);
}
create_table <- function(Dir, file_lst) {
  f=file_lst[1]
  sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
  df <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
    tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
    mutate(sum=rowSums(select_if(., is.numeric))) %>%
    dplyr::select(c("region",sum)) %>% 
    magrittr::set_colnames(c("region",sample))
  
  if (length(file_lst) > 1) {
    for (f in file_lst[2:length(file_lst)]) {
      sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
        tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
        mutate(sum=rowSums(select_if(., is.numeric))) %>%
        dplyr::select(c("region",sum)) %>% 
        magrittr::set_colnames(c("region",sample))
    
    df <- dplyr::left_join(df,df.this,by="region")
    }
  }
  df <- df %>% tibble::column_to_rownames( . , "region")
  return(df)
  rm(f,sample,df,df.this)
}
```
## Overview
In my skin-eQTL project, there are multiple batches of fastq files. Some are 40bp, and the newest Azenta Novaseq sequenced ones are 100bp. I picked 3 donors and all their conditions and celltypes to investigate whether trimming the 100bp fastq reads to 40bp could result in a drastic bias in certain genes. This will inform me if I should keep the read length as a known factor, or trim the 100bp reads to 40bp before processing all fastqs on the Via Foundry ESAT pipeline. I also trimmed the 100bp reads to 56bp and processing it on ESAT pipeline.

## load ESAT count tables, compile metadata, and calculate CPM
### phase1 samples (no merged samples, lengths vary from 36bp to 56bp)
```{r load-phase1-countmatrix}
Dir.esat.list=c("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/DolphinNext/phase1_rerun_customGenome/report10996/ESAT/",
                "~/Downloads/nl/human/skin/eQTLs/RNA-Seq/DolphinNext/phase1_rerun_customGenome/F44_KRT_PBS/report11029/ESAT/")

counts.phase1.list <- list()
metadata.phase1.list <- list()

for (i in 1:length(Dir.esat.list)) {
  # fetch ESAT directory
  Dir.esat=Dir.esat.list[i]
  
  # 1. import ESAT counts into master table
  flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
  counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
    tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
  col.sum = data.frame(sample = colnames(counts), samplesum = apply(counts,2,sum))
  
  # remove mitochondria genes
  MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
  counts <- counts %>% mutate(gene=rownames(.)) %>% 
    dplyr::filter(!rownames(.) %in% MTgenes$V1) %>% dplyr::select(-"gene")
  dim(counts)
  
  # mapping summary
  info_alignment <- data.table::fread(paste0(Dir.esat,"/../summary/overall_summary.tsv")) %>%
    set_colnames(c("sample","total_reads","multimapped_reads_aligned","unique_reads_aligned")) %>%
    left_join( . , col.sum, by="sample") %>%
    dplyr::mutate(reads_aligned = multimapped_reads_aligned + unique_reads_aligned) %>%
    dplyr::mutate(all_aligned_reads_over_total = (multimapped_reads_aligned + unique_reads_aligned) / total_reads) %>%
    dplyr::mutate(unique_aligned_reads_over_total = unique_reads_aligned / total_reads)

  # metadata
  metadata <- info_alignment
  
  # populate large list
  counts.phase1.list[[i]] <- counts
  metadata.phase1.list[[i]] <- metadata
  rm(counts, metadata, info_alignment, col.sum, flst,Dir.esat)
}

# combine all data frames by column
counts.phase1 <- do.call(cbind, counts.phase1.list) %>%
  dplyr::select(-F42_KRT_PBS_S1)# this was accidentally included in the run, but I used NovaSeq data for this sample instead

# import metadata
temp <- data.table::fread("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/metadata_phase1_selected_metadata_07082025.txt") %>% arrange(sample)
temp2 <- do.call(rbind, metadata.phase1.list) %>%
  dplyr::filter(!sample=="F42_KRT_PBS_S1") %>%
  arrange(sample) %>%
  mutate(CS2=character())
metadata.phase1 <- left_join(temp, temp2, by="sample")
metadata.phase1 <- metadata.phase1[,c("sample","read_length","donor","celltype","condition","CS2",
                                      "phase","technical_rep","sequencing_platform",
                                      "total_reads","multimapped_reads_aligned","unique_reads_aligned",
                                      "samplesum","reads_aligned","all_aligned_reads_over_total","unique_aligned_reads_over_total")]

rm(counts.phase1.list, metadata.phase1.list, temp, temp2)
```
### 40bp
```{r load-40bp}
Dir.esat="~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fq_read_length_comparison/DolphinNext/40bp/report10990/ESAT"

# 1. import ESAT counts into master table
flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
  tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
col.sum = data.frame(sample = colnames(counts), samplesum = apply(counts,2,sum))

# remove mitochondria genes
MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
counts <- counts %>% mutate(gene=rownames(.)) %>% 
  dplyr::filter(!rownames(.) %in% MTgenes$V1) %>% dplyr::select(-"gene")
dim(counts)

# mapping summary
info_alignment <- data.table::fread(paste0(Dir.esat,"/../summary/overall_summary.tsv")) %>%
  set_colnames(c("sample","total_reads","multimapped_reads_aligned","unique_reads_aligned")) %>%
  left_join( . , col.sum, by="sample") %>%
  dplyr::mutate(reads_aligned = multimapped_reads_aligned + unique_reads_aligned) %>%
  dplyr::mutate(all_aligned_reads_over_total = (multimapped_reads_aligned + unique_reads_aligned) / total_reads) %>%
  dplyr::mutate(unique_aligned_reads_over_total = unique_reads_aligned / total_reads)

# metadata
metadata <- data.frame(sample = colnames(counts), tosplit = colnames(counts)) %>%
  tidyr::separate(., col=tosplit, into=c(NA,"CS2","donor","celltype","condition",NA,NA),sep="_") %>%
  dplyr::mutate(read_length="40bp") %>%
  dplyr::mutate(phase="phase2") %>%
  dplyr::mutate(technical_rep="S1") %>%
  dplyr::mutate(sequencing_platform="NovaSeq") %>%
  right_join( . , info_alignment) 
metadata$condition <- ordered(metadata$condition, levels=c("PBS","IFNG","IFNB","TNF"))
metadata$celltype <- ordered(metadata$celltype, levels=c("FRB","KRT","MEL"))

counts.40bp <- counts
metadata.40bp <- metadata
metadata.40bp <- metadata.40bp[,c("sample","read_length","donor","celltype","condition","CS2",
                                      "phase","technical_rep","sequencing_platform",
                                      "total_reads","multimapped_reads_aligned","unique_reads_aligned",
                                      "samplesum","reads_aligned","all_aligned_reads_over_total","unique_aligned_reads_over_total")]

rm(counts, metadata, info_alignment,col.sum)
```
### 56bp
```{r load-56bp}
Dir.esat.list=c("~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/trimmed_FRB/report11013/ESAT/",
                "~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/trimmed_MEL/report11012/ESAT/",
                "~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/trimmed_KRT/report11011/ESAT/")

counts.56bp.list <- list()
metadata.56bp.list <- list()

for (i in 1:length(Dir.esat.list)) {
  # fetch ESAT directory
  Dir.esat=Dir.esat.list[i]
    
  # 1. import ESAT counts into master table
  flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
  counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
    tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
  col.sum = data.frame(sample = colnames(counts), samplesum = apply(counts,2,sum))
  
  # remove mitochondria genes
  MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
  counts <- counts %>% mutate(gene=rownames(.)) %>% 
    dplyr::filter(!rownames(.) %in% MTgenes$V1) %>% dplyr::select(-"gene")
  print(Dir.esat)
  print(dim(counts))
  
  # mapping summary
  info_alignment <- data.table::fread(paste0(Dir.esat,"/../summary/overall_summary.tsv")) %>%
    set_colnames(c("sample","total_reads","multimapped_reads_aligned","unique_reads_aligned")) %>%
    left_join( . , col.sum, by="sample") %>%
    dplyr::mutate(reads_aligned = multimapped_reads_aligned + unique_reads_aligned) %>%
    dplyr::mutate(all_aligned_reads_over_total = (multimapped_reads_aligned + unique_reads_aligned) / total_reads) %>%
    dplyr::mutate(unique_aligned_reads_over_total = unique_reads_aligned / total_reads)
  
  # metadata
  metadata <- data.frame(sample = colnames(counts), tosplit = colnames(counts)) %>%
    tidyr::separate(., col=tosplit, into=c(NA,"CS2","donor","celltype","condition",NA,NA),sep="_") %>%
    dplyr::mutate(read_length="56bp") %>%
    dplyr::mutate(phase="phase2") %>%
    dplyr::mutate(technical_rep="S1") %>%
    dplyr::mutate(sequencing_platform="NovaSeq") %>%
    right_join( . , info_alignment)
  metadata$condition <- ordered(metadata$condition, levels=c("PBS","IFNG","IFNB","TNF"))
  metadata$celltype <- ordered(metadata$celltype, levels=c("FRB","KRT","MEL"))
  
  # populate large list
  counts.56bp.list[[i]] <- counts
  metadata.56bp.list[[i]] <- metadata
  rm(counts, metadata, info_alignment, col.sum, flst)
}
# combine all data frames by column
counts.56bp <- do.call(cbind, counts.56bp.list)
metadata.56bp <- do.call(rbind, metadata.56bp.list)
metadata.56bp <- metadata.56bp[,c("sample","read_length","donor","celltype","condition","CS2",
                                      "phase","technical_rep","sequencing_platform",
                                      "total_reads","multimapped_reads_aligned","unique_reads_aligned",
                                      "samplesum","reads_aligned","all_aligned_reads_over_total","unique_aligned_reads_over_total")]

rm(counts.56bp.list, metadata.56bp.list)
```
### 100bp
#### list all pools
```{r 100bp-pools}
Dir.esat.list=c("~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-1_origBC/report10931/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-2_origBC/report10940/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-3_origBC/report10942/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-4_origBC/report10944/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-5_origBC/report10946/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-6_origBC/report10948/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-7_origBC/report10950/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-8_origBC/report10952/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-9/report10953/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-10/report10954/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-11/report10955/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-12/report10956/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-13/report10957/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-14/report10958/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-15/report10959/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-16/report10960/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-17/report10961/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-18/report10963/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-19/report10964/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-20/report10965/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-21/report10966/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-22/report10967/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-23/report10968/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-24/report10969/ESAT",
"~/Downloads/nl/human/skin/eQTLs/Azenta_sequencing/03312025_3ctk_celseq2_atac_batch1/DolphinNext_celseq2/skineQTL-25/report10970/ESAT")
```
#### compile counts and metadata
```{r load-100bp}
counts.100bp.list <- list()
metadata.100bp.list <- list()

for (i in 1:length(Dir.esat.list)) {
  # fetch ESAT directory
  Dir.esat=Dir.esat.list[i]
  
  # 1. import ESAT counts into master table
  flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
  counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
    tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
  col.sum = data.frame(sample = colnames(counts), samplesum = apply(counts,2,sum))
  
  # remove mitochondria genes
  MTgenes=read.table("~/Downloads/nl/human/skin/eQTLs/literature/UCSC_tracks/Ensembl_GRCh38.105_mitochondrial_genes.txt")
  counts <- counts %>% mutate(gene=rownames(.)) %>% 
    dplyr::filter(!rownames(.) %in% MTgenes$V1) %>% dplyr::select(-"gene")
  print(Dir.esat)
  dim(counts)
  
  # mapping summary
  info_alignment <- data.table::fread(paste0(Dir.esat,"/../summary/overall_summary.tsv")) %>%
    set_colnames(c("sample","total_reads","multimapped_reads_aligned","unique_reads_aligned")) %>%
    left_join( . , col.sum, by="sample") %>%
    dplyr::mutate(reads_aligned = multimapped_reads_aligned + unique_reads_aligned) %>%
    dplyr::mutate(all_aligned_reads_over_total = (multimapped_reads_aligned + unique_reads_aligned) / total_reads) %>%
    dplyr::mutate(unique_aligned_reads_over_total = unique_reads_aligned / total_reads)
  
  # metadata
  metadata <- data.frame(sample = colnames(counts), tosplit = colnames(counts)) %>%
    tidyr::separate(., col=tosplit, into=c("CS2","donor","celltype","condition",NA,NA),sep="_") %>%
    dplyr::mutate(read_length="100bp") %>%
    dplyr::mutate(phase="phase2") %>%
    dplyr::mutate(technical_rep="S1") %>%
    dplyr::mutate(sequencing_platform="NovaSeq") %>%
    right_join( . , info_alignment)
  metadata$condition <- ordered(metadata$condition, levels=c("PBS","IFNG","IFNB","TNF"))
  metadata$celltype <- ordered(metadata$celltype, levels=c("FRB","KRT","MEL"))
  
  # populate large list
  counts.100bp.list[[i]] <- counts
  metadata.100bp.list[[i]] <- metadata
  rm(counts, metadata, info_alignment, col.sum, flst)
}

# combine all data frames by column
counts.100bp <- do.call(cbind, counts.100bp.list)
metadata.100bp <- do.call(rbind, metadata.100bp.list)
metadata.100bp <- metadata.100bp[,c("sample","read_length","donor","celltype","condition","CS2",
                                      "phase","technical_rep","sequencing_platform",
                                      "total_reads","multimapped_reads_aligned","unique_reads_aligned",
                                      "samplesum","reads_aligned","all_aligned_reads_over_total","unique_aligned_reads_over_total")]

rm(counts.100bp.list, metadata.100bp.list, i, Dir.esat, Dir.esat.list)
```

### combine count matrices
```{r combine-counts-matrix}
counts <- cbind(counts.phase1, counts.40bp, counts.56bp, counts.100bp)
metadata <- rbind(metadata.phase1, metadata.40bp, metadata.56bp, metadata.100bp)
rm(counts.phase1, counts.40bp, counts.56bp, counts.100bp)
rm(metadata.phase1, metadata.40bp, metadata.56bp, metadata.100bp)
rm(MTgenes)
```
### libray count matrix sum is not matching uniquely mapped reads?!
```{r wtf}
df.gene = data.table::fread("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fq_read_length_comparison/DolphinNext/40bp/report10990/ESAT/trimmed40bp_GTTGCA_F122_FRB_IFNB_3ctk_S1mod_sorted_esat.txt.gene.txt") %>% set_colnames(c("gene","chr","strand","count"))
df.window = data.table::fread("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fq_read_length_comparison/DolphinNext/40bp/report10990/ESAT/trimmed40bp_GTTGCA_F122_FRB_IFNB_3ctk_S1mod_sorted_esat.txt.window.txt") %>% set_colnames(c("gene","chr","stard","end","strand","count"))
df.summary = data.table::fread("~/Downloads/nl/human/skin/eQTLs/RNA-Seq/fq_read_length_comparison/DolphinNext/40bp/report10990/summary/overall_summary.tsv")

sum(df.gene$count) # 3,361,837
sum(df.window$count) # 2,412,922
df.summary[df.summary$Sample=="trimmed40bp_GTTGCA_F122_FRB_IFNB_3ctk_S1mod",]
```
```{r wtf2}
df.gene = data.table::fread("~/Downloads/nl/human/skin/RNA_seq/run_01_25_24/fastq2/Fibroblast/report8897/ESAT/F55_CXCL10_125_sorted_esat.txt.gene.txt") %>% set_colnames(c("gene","chr","strand","count"))

df.summary = data.table::fread("~/Downloads/nl/human/skin/RNA_seq/run_01_25_24/fastq2/Fibroblast/report8897/summary/overall_summary.tsv")

sum(df.gene$count) # 2,028,156
df.summary[df.summary$Sample=="F55_BSA_500",] # 13,394,453
```

### calculate CPM
```{r CPM}
## Calculate CPM
col.sum=apply(counts,2,sum)
CPM=sweep(counts, 2, col.sum, `/`)*1000000
rm(col.sum)
```



## correlation plots
### correlation plot for [aligned reads / total reads] %
```{r ggplot-aligned-reads-pct}
df <-  metadata[,c("donor","celltype","condition","total_reads","all_aligned_reads_over_total","length")] %>%
  mutate(biosample = paste(donor, celltype, condition, sep = "_")) %>%
  pivot_wider(
    names_from = length,
    values_from = c(all_aligned_reads_over_total, total_reads),
    names_sep = "_"
  ) %>%
  dplyr::mutate(total_reads_difference = abs(total_reads_100bp - total_reads_40bp))

p <- ggplot(df, aes(x = all_aligned_reads_over_total_40bp, y = all_aligned_reads_over_total_100bp, size = total_reads_difference, color = total_reads_difference)) +
  geom_point(alpha=0.8) +
  xlim(c(0.5, 1)) + ylim(c(0.5,1))
```
### correlation plot for [uniquely aligned reads / total reads] %
```{r ggplot-aligned-reads-pct}
df <-  metadata[,c("donor","celltype","condition","total_reads","unique_aligned_reads_over_total","length")] %>%
  mutate(biosample = paste(donor, celltype, condition, sep = "_")) %>%
  pivot_wider(
    names_from = length,
    values_from = c(unique_aligned_reads_over_total, total_reads),
    names_sep = "_"
  ) %>%
  dplyr::mutate(total_reads_difference = abs(total_reads_100bp - total_reads_40bp))

p <- ggplot(df, aes(x = unique_aligned_reads_over_total_40bp, y = unique_aligned_reads_over_total_100bp, size = total_reads_difference, 
                    color = celltype)) +
  geom_point(alpha=0.8) +
  xlim(c(0.5, 1)) + ylim(c(0.5,1))
```
### correlation plot for [multiple aligned reads / total reads] %
```{r ggplot-aligned-reads-pct}
df <-  metadata[,c("donor","celltype","condition","total_reads","multimapped_reads_aligned","length")] %>%
  mutate(multimapped_aligned_reads_over_total = multimapped_reads_aligned / total_reads) %>%
  dplyr::select(-multimapped_reads_aligned) %>%
  mutate(biosample = paste(donor, celltype, condition, sep = "_")) %>%
  pivot_wider(
    names_from = length,
    values_from = c(multimapped_aligned_reads_over_total, total_reads),
    names_sep = "_"
  ) %>%
  dplyr::mutate(total_reads_difference = abs(total_reads_100bp - total_reads_40bp))

p <- ggplot(df, aes(x = multimapped_aligned_reads_over_total_40bp, y = multimapped_aligned_reads_over_total_100bp, size = total_reads_difference, 
                    color = celltype)) +
  geom_point(alpha=0.8) +
  xlim(c(0, 0.25)) + ylim(c(0, 0.25))
```
### correlation plot for CPM
```{r cor-cpm}
biosamples = unique(paste(metadata$donor, metadata$celltype, metadata$condition,sep="_"))
for (i in 1:length(biosamples)) {
  sample_40bp <- metadata[grepl(biosamples[i], metadata$sample)&metadata$length=="40bp","sample"]
  sample_100bp <- metadata[grepl(biosamples[i], metadata$sample)&metadata$length=="100bp","sample"]
  this_CPM <- data.frame(CPM.40bp = CPM[,sample_40bp], CPM.100bp = CPM[,sample_100bp])
  
  ggplot(this_CPM, aes(x=CPM.40bp, y=CPM.100bp, color = CPM.40bp)) +
    geom_point(alpha=0.8) +
    xlim(c(0, max(c(this_CPM$CPM.40bp, this_CPM$CPM.100bp)))) +
    scale_x_continuous(trans = "log2") +
    scale_y_continuous(trans = "log2")
}
```
### DE analysis based on read length
### MEL PBS
```{r deseq2-based-on-read-length}
for (ct in c("MEL","KRT","FRB")) {
  for (stim in c("PBS","IFNG","IFNB","TNF")) {
    this_metadata = metadata %>% 
      dplyr::filter(celltype==ct) %>%
      dplyr::filter(condition==stim) 
    this_counts = counts[,this_metadata$sample]
    dds <- DESeqDataSetFromMatrix(countData = this_counts, colData = this_metadata, design = ~ length)
    dds <- DESeq(dds)
    res_mel_PBS <- results(dds,contrast=c("length", "100bp", "40bp")) %>% 
      as.data.frame %>% rownames_to_column("gene") %>%
      na.omit() %>%
      dplyr::filter(padj<0.05 & abs(log2FoldChange) >1) %>%
      mutate(celltype=ct) %>%
      mutate(condition=stim)
    data.table::fwrite(res_mel_PBS, file=paste0(Dir,"/DESeq2_res_",ct,"_",stim,".txt"), quote=F,sep="\t")
  }}

# combine all DESeq2 results
flst <- list.files(path = Dir, pattern = "DESeq2_res")
res_all <- data.frame()
DESeq2_DEG_count <- data.frame(file = flst, sample = gsub(".txt","", gsub("DESeq2_res_","", flst)), DEG_n = 0)
for (f in flst[1:length(flst)]) {
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) 
      res_all <- rbind(res_all, df.this)
      DESeq2_DEG_count[which(DESeq2_DEG_count$file==f), "DEG_n"] = nrow(df.this)
}

# somehow melanocytes are very vulnerable for trimming. trimming accounts for over 2000 DE genes!

# what genes are DE most often?
length(unique(res_all$gene)) # 5652 DE genes in total
df <- as.data.frame(table(res_all$gene))
table(df$Freq)
write(as.character(df[which(df$Freq>10),"Var1"]), paste0(Dir,"/DEGs_in_10+_scenarios.txt"))
```
```{r DEG-related-to-low-count-genes}
gene_means <- rowMeans(counts)
gene_categories <- cut(
  gene_means,
  breaks = c(-Inf, 0, 10, 100, 1000, Inf),
  labels = c("zero", "low (1-10)", "moderate (11-100)", "high (101-1000)", "very high (>1000)")
)

gene_category_df <- data.frame(
  gene = names(gene_means),
  expression_bin = as.character(gene_categories)
)
rownames(gene_category_df) <- gene_category_df$gene

# Attach to res_all
res_all$mean_count <- gene_means[res_all$gene]
res_all$expression_bin <- gene_category_df[res_all$gene, "expression_bin"]

# Add scenario
res_all <- res_all %>% mutate(scenario = paste0(celltype, "_", condition))
res_all$expression_bin <- ordered(res_all$expression_bin, levels = c("low (1-10)","moderate (11-100)","high (101-1000)","very high (>1000)"))

#Group and count
deg_summary_binlevel <- res_all %>%
  filter(padj < 0.05 & !is.na(expression_bin)) %>%
  group_by(scenario, expression_bin) %>%
  summarise(DEG_count = n(), .groups = "drop") %>%
  tidyr::complete(
    scenario,
    expression_bin = c("low (1-10)", "moderate (11-100)", "high (101-1000)", "very high (>1000)"),
    fill = list(DEG_count = 0)
  )

deg_summary_binlevel$expression_bin <- factor(
  deg_summary_binlevel$expression_bin,
  levels = c("low (1-10)", "moderate (11-100)", "high (101-1000)", "very high (>1000)")
)
# Define custom colors for the bins
custom_colors <- c(
  "low (1-10)" = "grey",
  "moderate (11-100)" = "yellow",
  "high (101-1000)" = "orange",
  "very high (>1000)" = "red"
)

# Plot
ggplot(deg_summary_binlevel, aes(x = scenario, y = DEG_count, fill = expression_bin)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Number of DEGs by Expression Bin per Scenario",
    x = "Scenario",
    y = "Number of DEGs",
    fill = "Expression Bin"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```
```{r MA-plot-MEL-PBS}
library(clusterProfiler)
library(org.Hs.eg.db)
GO_table <- bitr(
  geneID = keys(org.Hs.eg.db, keytype = "GO"),
  fromType = "GO",
  toType = c("SYMBOL", "ENSEMBL"),
  OrgDb = org.Hs.eg.db
)

# Filter for GO:0006955: immune response. need to be our expressed gene table too.
immune_genes <- unique(GO_table[which(GO_table$GO == "GO:0006955"), "SYMBOL"])
immune_genes <- intersect(immune_genes, rownames(counts))

# plot MA plot
this_scenario = "FRB_PBS"
this_res <- res_all %>% filter(scenario == this_scenario)

# Subset to immune genes that are in this result
immune_label_df <- this_res %>% 
  filter(gene %in% immune_genes)

# Plot with labels
ggplot(this_res, aes(x = log10(mean_count + 1), y = log2FoldChange)) +
  geom_point(color="grey", alpha = 0.5, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_text_repel(
    data = immune_label_df,
    aes(label = gene),
    size = 3,
    max.overlaps = 50,  # increase if too few are labeled
    box.padding = 0.3,
    point.padding = 0.2
  ) +
  labs(
    title = paste0("MA Plot of DEGs (40bp vs 100bp) in", this_scenario, "; nDEG = ",nrow(this_res)),
    x = "log10 Mean Count",
    y = "log2 Fold Change (40bp / 100bp)",
    color = "Significance"
  ) +
  theme_minimal()
```
```{r CPM_scatterplot_colored_by_DEG}
biosamples = unique(paste(metadata$donor, metadata$celltype, metadata$condition,sep="_"))
biosamples = biosamples[grep("MEL_PBS",biosamples)]

for (i in 1:length(biosamples)) {
  sample_40bp <- metadata[grepl(biosamples[i], metadata$sample)&metadata$length=="40bp","sample"]
  sample_100bp <- metadata[grepl(biosamples[i], metadata$sample)&metadata$length=="100bp","sample"]
  this_CPM <- data.frame(CPM.40bp = CPM[,sample_40bp], CPM.100bp = CPM[,sample_100bp])
  
  ggplot(this_CPM, aes(x=CPM.40bp, y=CPM.100bp, color = CPM.40bp)) +
    geom_point(alpha=0.8) +
    xlim(c(0, max(c(this_CPM$CPM.40bp, this_CPM$CPM.100bp)))) +
    scale_x_continuous(trans = "log2") +
    scale_y_continuous(trans = "log2")
}
```
### correlation matrix for CPM correlation between 100bp and 40bp
```{r cor-matrix}
scenarios = unique(paste(metadata$celltype, metadata$condition,sep="_"))

scenario_pcc_list <- list()

for (n in 1:length(scenarios)) {
  scenario <- scenarios[n]
  
  # loop through each scenario
  samples <- metadata[grepl(scenario, metadata$sample), "sample"]
  # subset cpm for these samples
  cpm_subset <- CPM[,samples, drop=F]
  # compute correlation matrix
  pcc_matrix <- cor(cpm_subset, method = "pearson")
  
  # store the matrix in the list
  scenario_pcc_list[[scenario]] <- pcc_matrix
  
  # Format annotation matrix
  anno_text <- matrix(sprintf("%.2f", pcc_matrix), nrow = nrow(pcc_matrix))
  rownames(anno_text) <- rownames(pcc_matrix)
  colnames(anno_text) <- colnames(pcc_matrix)
  
  # Create heatmap object
  ht <- Heatmap(
    pcc_matrix,
    name = "Pearson",
    col = colorRamp2(c(0, 0.5, 1), c("blue", "white", "red")),
    cell_fun = function(j, i, x, y, w, h, fill) {
      grid.text(sprintf("%.2f", pcc_matrix[i, j]), x, y, gp = gpar(fontsize = 8))
    },
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_names_side = "top",
    heatmap_legend_param = list(title = "Pearson"),
    row_names_gp = gpar(fontsize = 6),
    column_names_gp = gpar(fontsize = 6),
    column_names_rot = 45
  )
  
  # Save to PDF
  pdf(file = paste0(Dir,"/correlation_matrix_", scenario, ".pdf"), width = 7, height = 6)
  draw(ht)
  dev.off()
}
```

### ComBatseq correction
### combatseq
```{r combatseq}
library("sva")
uncorrected_data <- counts
#perform the batch correction
#first we need to transform the format of our groups and batches from names to numbers
batches = as.numeric(as.factor(metadata$length))
#now run ComBat_seq
counts.corrected = ComBat_seq(counts = as.matrix(uncorrected_data), batch = batches)
```
### CPM corrected
```{r corrected-cpm}
col.sum = apply(counts.corrected,2,sum)
CPM.corrected = sweep(counts.corrected, 2, col.sum, `/`)*1000000
rm(col.sum)
```
### correlation matrix of corrected matrix
```{r cor-matrix-corrected-CPM}
scenarios = unique(paste(metadata$celltype, metadata$condition,sep="_"))

scenario_pcc_list <- list()

for (n in 1:length(scenarios)) {
  scenario <- scenarios[n]
  
  # loop through each scenario
  samples <- metadata[grepl(scenario, metadata$sample), "sample"]
  # subset cpm for these samples
  cpm_subset <- CPM.corrected[,samples, drop=F]
  # compute correlation matrix
  pcc_matrix <- cor(cpm_subset, method = "pearson")
  
  # store the matrix in the list
  scenario_pcc_list[[scenario]] <- pcc_matrix
  
  # Format annotation matrix
  anno_text <- matrix(sprintf("%.2f", pcc_matrix), nrow = nrow(pcc_matrix))
  rownames(anno_text) <- rownames(pcc_matrix)
  colnames(anno_text) <- colnames(pcc_matrix)
  
  # Create heatmap object
  ht <- Heatmap(
    pcc_matrix,
    name = "Pearson",
    col = colorRamp2(c(0, 0.5, 1), c("blue", "white", "red")),
    cell_fun = function(j, i, x, y, w, h, fill) {
      grid.text(sprintf("%.2f", pcc_matrix[i, j]), x, y, gp = gpar(fontsize = 8))
    },
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_names_side = "top",
    heatmap_legend_param = list(title = "Pearson"),
    row_names_gp = gpar(fontsize = 6),
    column_names_gp = gpar(fontsize = 6),
    column_names_rot = 45
  )
  
  # Save to PDF
  pdf(file = paste0(Dir,"/correlation_matrix_combatseq_length_corrected_", scenario, ".pdf"), width = 7, height = 6)
  draw(ht)
  dev.off()
}
```
### Q: which genes have raw counts in 100bp but not in 40bp? and the other way around?
```{r genes-losing-raw-counts-in-40bp}
scenarios = unique(paste(metadata$celltype, metadata$condition,sep="_"))
scenario = "FRB_PBS"

# fetch count matrices
counts.40bp <- counts[,grepl("trimmed", colnames(counts))&grepl(scenario, colnames(counts))]
counts.100bp <- counts[,(!grepl("trimmed", colnames(counts)))&grepl(scenario, colnames(counts))]

# row sum raw counts for each gene
sum_40bp <- rowSums(counts.40bp)
sum_100bp <- rowSums(counts.100bp)

# Identify genes with nonzero in 100bp and zero in 40bp
genes_only_in_100bp <- names(which(sum_100bp > 0 & sum_40bp == 0))
length(genes_only_in_100bp)

# which genes have raw counts in 40bp but not in 100bp?
genes_only_in_40bp <- names(which(sum_40bp > 0 & sum_100bp == 0))
length(genes_only_in_40bp)

# inspect genes that are only in 100bp but not in 40bp
gene_totals <- rowSums(counts.100bp[genes_only_in_100bp,])
only_in_100bp_counts_sorted <- counts[genes_only_in_100bp,grepl(scenario, colnames(counts))][order(gene_totals, decreasing = TRUE), ]

# inspect genes that are only in 40bp but not in 100bp
gene_totals <- rowSums(counts.40bp[genes_only_in_40bp,])
only_in_40bp_counts_sorted <- counts[genes_only_in_40bp,grepl(scenario, colnames(counts))][order(gene_totals, decreasing = TRUE), ]
```
### Q: when I see differences in CPM, are they due to some genes simply dropping out, rather than true differences in normalized expression?
```{r cpm-dropout-test}
scenarios = unique(paste(metadata$celltype, metadata$condition,sep="_"))
scenario = "FRB_PBS"
# fetch count matrices
counts.40bp <- counts[,grepl("trimmed", colnames(counts))&grepl(scenario, colnames(counts))]
counts.100bp <- counts[,(!grepl("trimmed", colnames(counts)))&grepl(scenario, colnames(counts))]
CPM.40bp <- CPM[,colnames(counts.40bp)]
CPM.100bp <- CPM[,colnames(counts.100bp)]
# row sum raw counts for each gene
sum_40bp <- rowSums(counts.40bp)
sum_100bp <- rowSums(counts.100bp)
# Get dropout genes
gained_genes <- names(which(sum_40bp > 0 & sum_100bp == 0))
dropout_genes <- names(which(sum_100bp > 0 & sum_40bp == 0))
# Compare CPM differences for those genes vs. other genes
cpm_diff <- rowMeans(CPM.100bp) - rowMeans(CPM.40bp)
dropout_cpm_diff <- cpm_diff[names(cpm_diff) %in% dropout_genes]
gained_cpm_diff <- cpm_diff[names(cpm_diff) %in% gained_genes]
non_dropout_non_gained_cpm_diff <- cpm_diff[!names(cpm_diff) %in% c(dropout_genes, gained_genes)]

# Visualize and test 
boxplot(
  list(Gained = gained_cpm_diff, Dropout = dropout_cpm_diff, NonDropout = non_dropout_cpm_diff),
  main = "CPM difference (100bp - 40bp)",
  ylab = "Δ CPM",
  col = c("green","tomato", "skyblue")
)

# Or perform a statistical test
wilcox.test(dropout_cpm_diff, non_dropout_cpm_diff)
wilcox.test(gained_cpm_diff, non_dropout_cpm_diff)
```


## Q: for samples with multiple replicates, how do the read length affect their alignment performance & CPM correlation?
```{r find-samples-with-reps}
dups <- metadata %>%
  dplyr::mutate(biosample=paste0(donor,"_",celltype,"_",condition)) %>%
  group_by(biosample) %>%
  filter(n() > 2) %>%
  ungroup()
samples_with_dups <- unique(dups$biosample)

```