---
title: "skin-eQTL_celseq2_power_analysis"
author: "Shuo Shan"
date: "1/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages  
```{r package, warning=FALSE}
library(dplyr)
library(tidyverse)
library(DESeq2)
library(DEXSeq)
library(BiocParallel)
library(magrittr)
library(stringr)
library(gplots)
library(ggrepel)
library(ggfortify)
library(reactable)
library(edgeR)
library(gridExtra)
library(ggpubr)
library(debrowser)
library(plotly)
library(RCurl)


Dir="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/power_analysis"
setwd(Dir)

#save.image(file=paste0(Dir,'/myEnvironment.RData'))
#load(paste0(Dir,'/myEnvironment.RData'))
```
  
### Overview

Power analysis to determine how many donors do we need to identify 1.5 fold change differentially expressed genes under IFNg stimulation, and whether they allow us to investigate vitiligo SNPs or skin-disease SNPs with a certain level of minor allele frequency (MAF).

Plan:
1. Pick a cell type
2. Get their ESAT raw count data, and metadata, and randomly pick donors
3. 随机取5个patients，跟healthy control做对比find DE genes. 取很多次，看每次找到的DE genes重合的有多少。然后取6 个patients，7个，8个etc。然后发现7个时结果就稳定了.
4. Then ask what is the minor allele frequency corresponding to the donor number. Assuming donors are homozygous for minor allele.
5. Pick minimum and maximum of the 89 vitiligo SNPs. Can we study them with the number of donors we have?

```{r functions}
create_table <- function(Dir, file_lst) {
  f=file_lst[1]
  sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
  df <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
    tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
    mutate(sum=rowSums(select_if(., is.numeric))) %>%
    dplyr::select(c("region",sum)) %>% 
    magrittr::set_colnames(c("region",sample))
  
  if (length(file_lst) > 1) {
    for (f in file_lst[2:length(file_lst)]) {
      sample=gsub("_sorted_esat\\.txt\\.gene\\.txt","",f)
      df.this <- read.table(paste0(Dir,"/",f), sep="\t", header=TRUE) %>% 
        tidyr::unite( .,"region",c("Symbol","chr","strand"),sep=";",remove=TRUE) %>% 
        mutate(sum=rowSums(select_if(., is.numeric))) %>%
        dplyr::select(c("region",sum)) %>% 
        magrittr::set_colnames(c("region",sample))
    
    df <- dplyr::left_join(df,df.this,by="region")
    }
  }
  df <- df %>% tibble::column_to_rownames( . , "region")
  return(df)
  rm(f,sample,df,df.this)
}

# check which variable correlates with each PC the most.
best_correlation_with_pc <- function(pcs, metadata, pc.n) {
  total.pcc=c()
  pca.this=pcs[,pc.n]
  for (x in colnames(metadata)) {
    metadata.this <- as.numeric(as.factor(metadata[,x]))
    total.pcc=c(total.pcc,cor(pca.this,metadata.this))
  }
  total.df=data.frame(feature=colnames(metadata),pcc=total.pcc,pcc.magnitude=abs(total.pcc))
  return(total.df)
}
```
  
### Load raw ESAT counts
```{r, echo=FALSE, message=FALSE}
Dir.esat="/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_01152022/DolphinNext/report6124/ESAT"

# 1. import ESAT counts into master table
flst <- list.files(path = Dir.esat, pattern = "esat.txt.gene.txt")
counts <- create_table(Dir.esat, flst) %>% rownames_to_column("key") %>%
  tidyr::separate("key",into=c("gene",NA,NA),sep=";",remove=TRUE) %>% column_to_rownames("gene")
samples.all <- data.frame(sample=colnames(counts))

# 2. find list of samples to keep
# 2a. condition used from pilot (KRT, FRB: 100ng 24h, MEL: 10ng 48h)
samples.pilot.keep <- c("F25_FRB_PBS_0ng_S0", "F25_FRB_24h_100ng_S0",
                        "F25_MEL_PBS_0ng_S0", "F25_MEL_48h_10ng_S0",
                        "F25_KRT_PBS_0ng_S0", "F25_KRT_24h_100ng_S0")
# 2b. latest version of samples (merged from replicates if any)
samples <- samples.all %>%
  tidyr::separate( . , col=sample, into=c("donor","celltype","condition","replicate"), sep="_",remove=FALSE) %>%
  dplyr::filter(donor!="F25") %>% tidyr::unite(. , "biosample", c("donor","celltype","condition"),sep="_", remove=FALSE)
# go through the list, and for each donor, pick the row that contain the highest number of replicate
samples.latest <- c()
for (d in unique(samples$biosample)) {
  df.this <- samples %>% dplyr::filter(biosample==d) %>% arrange( . , desc(replicate))
  sample.pick <- df.this[1,1]
  samples.latest <- c(samples.latest, sample.pick)
  rm(df.this,sample.pick)
}
# 2c. only keratinocytes.
samples.keep <- c(samples.latest, samples.pilot.keep)
samples.keep <- samples.keep[grep("FRB",samples.keep)]
# 2d. remove samples with indicated keywords (after QC inspection. will update until more sequence reads come in)
keyword1 <- "F22|F24|F28|F30|F31|F32|F33"
samples.keep <- samples.keep[-grep(keyword1,samples.keep)]
# 2e. specificy the exact biosample to keep
#samples.keep <- c(samples.keep[-grep("F34_KRT_PBS",samples.keep)],"F34_KRT_PBS_S2")

# 2 combine
counts.keep <- counts %>% dplyr::select(samples.keep)
colnames(counts.keep) <- colnames(counts.keep) %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)
```
  
### Load metadata
```{r metadata, echo=FALSE, message=FALSE}
# parse kept samples into table
samples <- data.frame(sample=colnames(counts.keep))
# import metadata master table
metadata.master=read.table("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/metadata_01192022.txt", sep="\t", header=TRUE)
# import DolphinNext pipeline overall summary
pip.summary=read.table("/nl/umw_manuel_garber/human/skin/eQTLs/RNA-Seq/fibroblasts/pipeline_01152022/DolphinNext/report6124/summary/overall_summary.tsv", sep="\t", header=TRUE) %>% magrittr::set_colnames(c("sample","total_reads","Multimapped_Reads_Aligned_STAR","Unique_Reads_Aligned_STAR"))
pip.summary$sample <- pip.summary$sample %>%
  gsub("F25_FRB_PBS_0ng_S0","F25_FRB_PBS_S0",.) %>%
  gsub("F25_FRB_24h_100ng_S0","F25_FRB_IFN_S0",.) %>%
  gsub("F25_MEL_PBS_0ng_S0","F25_MEL_PBS_S0",.) %>%
  gsub("F25_MEL_48h_10ng_S0","F25_MEL_IFN_S0",.) %>%
  gsub("F25_KRT_PBS_0ng_S0","F25_KRT_PBS_S0",.) %>%
  gsub("F25_KRT_24h_100ng_S0","F25_KRT_IFN_S0",.)


# query imported tables with samples
rm(metadata)
metadata <- left_join(samples,metadata.master,by="sample")
metadata <- left_join(metadata,pip.summary,by="sample")

# compile 
metadata$seq_date <- metadata$seq_date %>% gsub("\\/",".",.)
metadata$lib_prep_date <- metadata$lib_prep_date %>% gsub("\\/",".",.)
metadata$batch <- paste(metadata$lib_prep_date,metadata$seq_date,sep="_") %>% as.factor()

# cleanup
rm(pip.summary, samples, samples.all)
```
  
### DE analysis  
The goal of Differential gene expression analysis is to find 
genes or transcripts whose difference in expression, when accounting 
for the variance within condition, is higher than expected by chance. 

In this case a total sum of 10 counts separates well expressed 
from non-expressed genes. You can change this value and padj value and 
log2FoldChange cutoffs according to your data

```{r DESeq2, echo=FALSE, message=FALSE}
# function that returns a tibble: donors, and a list of DE genes (padj<=0.05, abs(log2FC)>=1.5)
perform_deseq2 <- function(count.this, metadata.this, design.formula) {
  df <- apply(count.this[rowSums(count.this)>=10,],2,round,0) %>% as.data.frame()
  donors.this <- unique(metadata.this$donor)
  n.donors <- length(donors.this)
  
  dds <- DESeqDataSetFromMatrix(countData = df,
                                colData = metadata.this,
                                design = design.formula)
  dds <- DESeq(dds)
  dds <- estimateSizeFactors(dds)
  normalized_counts <- counts(dds, normalized=TRUE) %>% as.data.frame()
  summary(results(dds,contrast=c("condition", "IFN", "PBS")))
  
  res.this <- results(dds,contrast=c("condition", "IFN", "PBS")) %>% as.data.frame %>% 
    rownames_to_column("gene") %>% 
    dplyr::filter(abs(log2FoldChange)>=1.5) %>% 
    dplyr::filter(padj <= 0.05)
  
  answer <- tibble(n=n.donors, donors=paste(donors.this,collapse="_"), DEG=list(res.this$gene))
  return(answer)
}

# compare how many DE genes are shared when we select a range bewteen 3 donors to 30 donors. 10 reps each donor size.
tibble.master <- tibble(n=numeric(),donors="",DEG=list())
for (n in 3:30) {
  j=1
  while (j <= 10) {
    print(paste0("...working on ",as.character(n)," donors...iteration=",as.character(j)))
    donor.keep <- sample(unique(metadata$donor), n, replace=F)
    count.this <- counts.keep %>% dplyr::select(starts_with(donor.keep))
    metadata.this <- metadata %>% dplyr::filter(metadata$donor %in% donor.keep)
    tibble.this <- perform_deseq2(count.this, metadata.this, ~ donor + condition)
    tibble.master <- rbind(tibble.master, tibble.this)
    j=j+1
  }
}
tibble.master$n_DEG <- sapply(tibble.master$DEG, function(x) length(x))


# number of DE genes for each run vs. donor size
tibble.master <- readRDS(paste0(Dir,"/data_number_of_DE_genes_vs_number_of_donors.rds"))
df <- tibble.master
p<- df %>% 
  ggplot(., aes(x=n, y=n_DEG, group=n, label=donors)) +
  geom_boxplot(alpha=0.4,color='lightblue') +
  geom_point(alpha=0.4,color='blue',fill='none') +
  labs(title="number of DE genes identified",
       subtitle="p.adj<0.05, abs(log2FC)>1.5",
       x="number of donors included",
       y="number of DE genes")

p.interactive <- ggplotly(p)
htmlwidgets::saveWidget(widget = p.interactive, 
                        file = paste0(Dir,"/number_of_DE_genes_vs_number_of_donors.html"), 
                        selfcontained = TRUE)

save.image(file=paste0(Dir,'/myEnvironment.RData'))
#saveRDS(df4, paste0(Dir,"/data_number_of_DE_genes_vs_number_of_donors.rds"))


# common genes vs. donor size
tibble.master <- readRDS(paste0(Dir,"/data_number_of_DE_genes_vs_number_of_donors.rds"))
df <- tibble.master %>% dplyr::filter(n_DEG>120)
df.plt <- data.frame(n=numeric(), shared_genes=numeric())
for (i in 3:30) {
  n.donor=i
  df.this <- df %>% dplyr::filter(n==n.donor)
  df.plt.this <- data.frame(n=n.donor, shared_genes=length(Reduce(intersect, df.this$DEG)))
  df.plt <- rbind(df.plt, df.plt.this)
}
ggplot(df.plt, aes(x=n, y=shared_genes)) +
  geom_point()

```
  
### MAF analysis
```{r MAF}
snp.maf <- read.table("/nl/umw_manuel_garber/human/skin/eQTLs/fastQTL/GWAS_SNPs/01.24.2022/filtered_skin_disease_GWAS_SNPs_MAF.txt", sep="\t", header=FALSE) %>% magrittr::set_colnames(c("chr","position","ID","MAF"))

p <- ggplot(snp.maf, aes(x=MAF)) +
  geom_histogram(color="black", fill="white",bins=50) +
  geom_vline(xintercept = 5/20, color = "#aaf2d1") +
  geom_vline(xintercept = 5/25, color = "#76a5af") +
  geom_vline(xintercept = 5/30, color = "#0a75ad") +
  geom_vline(xintercept = 5/35, color = "#153963") +
  labs(title="MAF of 56,904 skin-disease GWAS SNPs in 37 genotyped donors")


```